### 2020-08-06 manuscript_00.Rmd
plae app manuscript setup

---
title: 'PLAE enables quick and easy searching across one million ocular transcriptomes'
author:
  - Vinay Swamy:
      institute: bg
  - Zachary Batz:
      institute: nnrl
  - Temesgen Fufa:
      institute: mgog
  - Robert Hufnagel:
      institute: mgog
  - David McGaughey:
      institute:
        - bg
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
  - nnrl: Neurobiology, Neurodegeneration & Repair Laboratory, National Eye Institute, National Institutes of Health
  - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health

date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: massive_integrated_eye_scRNA.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors. METHODS: We wrote the Shiny web application PLAE (https://plae.nei.nih.gov) and tailored the structure of the underlying data (sc Eye in a Disk, scEiaD) to display scEiaD. RESULTS: The new portal provides quick visualization of vertebrate eye-related transcriptomes published to date by gene/transcript, XX cell types, XX ocular tissues, XX body tissues. As a test of the value of this unified pan-eye dataset, we show XX. CONCLUSION: The PLAE v1.0 web app serves the pan-ocular and body dataset, scEiaD. This offers the eye community a powerful and quick means to test hypotheses on human gene and transcript expression across 54 body and 19 eye tissues. "
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
---

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)

# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

```

# Introduction
## Many published ocular atlases

## Data not easily available for all

## Some have accompanying apps

## But apps are generally slow and have minimal exploration tooling

## Cross comparisons difficult
But important as seeing trends trends across different groups strongly demonstrates that 
the signal is biological, not technical.

## What we did

Built an awesome app.

## Will do???

Trajectory

Velocity

# Methods

## Use scEiaD resource
We built, published [].

## pan mouse as non-ocular reference
Tabula Muris

# Results

## Rich functionality

## Web accessible

##

# Conclusion

### 2021-10-22 manuscript_00.Rmd
text updates

@@
       institute: nnrl
-  - Temesgen Fufa:
-      institute: mgog
   - Robert Hufnagel:
@@
 METHODS: We wrote the Shiny web application PLAE (https://plae.nei.nih.gov) and tailored the structure of the underlying data (sc Eye in a Disk, scEiaD) to display scEiaD.
-RESULTS: The new portal provides quick visualization of vertebrate eye-related transcriptomes published to date by gene/transcript, XX cell types, XX ocular tissues, XX body tissues.
+RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes., XX cell types, XX ocular tissues, XX body tissues.
 As a test of the value of this unified pan-eye dataset, we show XX.
-CONCLUSION: The PLAE v1.0 web app serves the pan-ocular and body dataset, scEiaD.
-This offers the eye community a powerful and quick means to test hypotheses on human gene and transcript expression across 54 body and 19 eye tissues.
+CONCLUSION: The PLAE v0.80 web app serves the pan-ocular and body dataset, scEiaD.
+This offers the eye community a powerful and quick means to test hypotheses on gene and transcript expression across 54 body and 19 eye tissues.
 "
@@
 
+library(pool)
+library(RSQLite)
+meta_filter <- fst::read_fst("~/data/scEiaD_v2/2021_10_15_meta_filter.fst")
 ```
 
+# Distribution of Tissues and Cell Types in scEiaD
+```{r, fig.width=8, fig.height=5}
+# organ
+a <- meta_filter %>%
+  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
+                           TRUE ~ 'Body'),
+         Tissue = gsub('\\.|_', ' ', Tissue),
+         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
+         Tissue = gsub(" and Aorta","", Tissue)) %>% 
+  group_by(Organ, Tissue,organism, study_accession) %>%
+  summarise(Count = n()) %>%
+  ungroup() %>% 
+  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
+  ylab('Number of Cells') +
+  geom_bar(stat='identity', position = 'dodge') +
+  cowplot::theme_cowplot() +
+  facet_wrap(~Organ, scales = 'free') +
+  coord_flip()
+
+# retina cell types
+b <- meta_filter %>% filter(Organ == 'Eye') %>% 
+  group_by(CellType_predict, study_accession, organism) %>%
+  summarise(Count = n()) %>% filter(Count > 10) %>% ungroup() %>% 
+  ggplot(aes(x=CellType_predict,y=Count, fill = organism, group = organism)) +
+  ylab('Number of Cells') +
+  geom_bar(stat='identity', position = 'dodge') +
+  cowplot::theme_cowplot() +
+  coord_flip()
+
+plot_grid(a,b)
+```
+# Distribution of ML Retina Cell Types Between Mouse and Human in Pan-Retina Studies
+Drop this figure?
+```{r, fig.height=5, fig.width=3}
+library(tidyverse)
+meta_filter <- fst::read_fst('~/data/scEiaD/2021_08_22_meta_filter.fst') %>%
+  as_tibble() %>%
+  mutate(CellType_predict = case_when(!is.na(TabulaMurisCellType_predict) ~ 'Tabula Muris',
+                                      is.na(CellType_predict) ~ 'Unlabelled',
+                                      TRUE ~ CellType_predict))
+
+pan_retina_studies <- c('SRP255195','E-MTAB-7316','SRP151023','SRP222001','SRP222958','SRP194595','SRP223254','SRP050054','SRP269634','SRP269635','SRP158528', 'SRP257758','SRP212788')
+
+meta_filter %>% 
+  filter(organism %in% c('Mus musculus','Homo sapiens'), 
+         Stage == 'Mature', 
+         study_accession %in% pan_retina_studies, 
+         grepl('Amacr|Bipol|Cone|Epithel|Endoth|Horizon|Muller|Pericy|Ganglion|Rod', CellType_predict)) %>% 
+  group_by(study_accession, CellType_predict) %>% 
+  summarise(Count = n()) %>% mutate(Percentage = 100 * (Count / sum(Count)))  %>% 
+  left_join(meta_filter %>% select(study_accession, Platform, organism) %>% unique()) %>% 
+  ggplot(aes(x=organism, y=Percentage)) + 
+  geom_boxplot() + 
+  coord_flip() + 
+  ggbeeswarm::geom_quasirandom(aes(color=study_accession)) + 
+  facet_wrap(~CellType_predict, ncol = 1) + 
+  cowplot::theme_cowplot()
+```
 # Introduction
@@
 
-## Will do???
-
-Trajectory
-
-Velocity
-
 # Methods
@@
 
-## pan mouse as non-ocular reference
+## pan mouse and pan human as non-ocular reference
 Tabula Muris

### 2021-10-22 manuscript_00.Rmd
text update

@@
 
-# Distribution of Tissues and Cell Types in scEiaD
+# (sup?) Table 1
+Studies in scEiaD
+
+# Figure 1 Distribution of Tissues and Cell Types in scEiaD
 ```{r, fig.width=8, fig.height=5}
@@
          Tissue = gsub(" and Aorta","", Tissue)) %>% 
+  filter(Organ == 'Body') %>% 
+  group_by(Organ, Tissue,organism, study_accession) %>%
+  summarise(Count = n()) %>%
+  ungroup() %>% 
+  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
+  ylab('Number of Cells') +
+  geom_bar(stat='identity', position = 'dodge') +
+  cowplot::theme_cowplot() +
+  coord_flip() +
+  scale_y_continuous(breaks = c(2500, 5000, 7500, 10000, 12500, 84000))
+
+b <-  meta_filter %>%
+  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
+                           TRUE ~ 'Body'),
+         Tissue = gsub('\\.|_', ' ', Tissue),
+         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
+         Tissue = gsub(" and Aorta","", Tissue)) %>% 
+  filter(Organ == 'Eye') %>% 
   group_by(Organ, Tissue,organism, study_accession) %>%
@@
 # retina cell types
-b <- meta_filter %>% filter(Organ == 'Eye') %>% 
+c <- meta_filter %>% filter(Organ == 'Eye') %>% 
   group_by(CellType_predict, study_accession, organism) %>%
@@
 
-plot_grid(a,b)
+plot_grid(a +  ggbreak::scale_y_break(c(13000, 82000)),b, c)
 ```
-# Distribution of ML Retina Cell Types Between Mouse and Human in Pan-Retina Studies
-Drop this figure?
-```{r, fig.height=5, fig.width=3}
-library(tidyverse)
-meta_filter <- fst::read_fst('~/data/scEiaD/2021_08_22_meta_filter.fst') %>%
-  as_tibble() %>%
-  mutate(CellType_predict = case_when(!is.na(TabulaMurisCellType_predict) ~ 'Tabula Muris',
-                                      is.na(CellType_predict) ~ 'Unlabelled',
-                                      TRUE ~ CellType_predict))
 
-pan_retina_studies <- c('SRP255195','E-MTAB-7316','SRP151023','SRP222001','SRP222958','SRP194595','SRP223254','SRP050054','SRP269634','SRP269635','SRP158528', 'SRP257758','SRP212788')
+# sup fig 1 
+Effect of decontx for Rho between scEiaD v2 and scEiaD v1
 
-meta_filter %>% 
-  filter(organism %in% c('Mus musculus','Homo sapiens'), 
-         Stage == 'Mature', 
-         study_accession %in% pan_retina_studies, 
-         grepl('Amacr|Bipol|Cone|Epithel|Endoth|Horizon|Muller|Pericy|Ganglion|Rod', CellType_predict)) %>% 
-  group_by(study_accession, CellType_predict) %>% 
-  summarise(Count = n()) %>% mutate(Percentage = 100 * (Count / sum(Count)))  %>% 
-  left_join(meta_filter %>% select(study_accession, Platform, organism) %>% unique()) %>% 
-  ggplot(aes(x=organism, y=Percentage)) + 
-  geom_boxplot() + 
-  coord_flip() + 
-  ggbeeswarm::geom_quasirandom(aes(color=study_accession)) + 
-  facet_wrap(~CellType_predict, ncol = 1) + 
-  cowplot::theme_cowplot()
-```
+<!-- ```{r, fig.height=5, fig.width=3} -->
+<!-- # Distribution of ML Retina Cell Types Between Mouse and Human in Pan-Retina Studies -->
+<!-- Drop this figure? -->
+<!-- Yes -->
+
+<!-- pan_retina_studies <- c('SRP255195','E-MTAB-7316','SRP151023','SRP222001','SRP222958','SRP194595','SRP223254','SRP050054','SRP269634','SRP269635','SRP158528', 'SRP257758','SRP212788') -->
+
+<!-- meta_filter %>%  -->
+<!--   filter(organism %in% c('Mus musculus','Homo sapiens'),  -->
+<!--          Stage == 'Mature',  -->
+<!--          study_accession %in% pan_retina_studies,  -->
+<!--          grepl('Amacr|Bipol|Cone|Epithel|Endoth|Horizon|Muller|Pericy|Ganglion|Rod', CellType_predict)) %>%  -->
+<!--   group_by(study_accession, CellType_predict) %>%  -->
+<!--   summarise(Count = n()) %>% mutate(Percentage = 100 * (Count / sum(Count)))  %>%  -->
+<!--   left_join(meta_filter %>% select(study_accession, Platform, organism) %>% unique()) %>%  -->
+<!--   ggplot(aes(x=organism, y=Percentage)) +  -->
+<!--   geom_boxplot() +  -->
+<!--   coord_flip() +  -->
+<!--   ggbeeswarm::geom_quasirandom(aes(color=study_accession)) +  -->
+<!--   facet_wrap(~CellType_predict, ncol = 1) +  -->
+<!--   cowplot::theme_cowplot() -->
+<!-- ``` -->
 # Introduction
@@
 
-## Data not easily available for all
-
-## Some have accompanying apps
-
-## But apps are generally slow and have minimal exploration tooling
+## Data difficult to access
 
@@
 
-Built an awesome app.
+Collated XX studies and processed
 
@@
 
-## Use scEiaD resource
-We built, published [].
+## Update scEiaD resource
+Our first version of scEiaD contained XX studies, 3 species, XX cells, and XX curated cell types.
+We updated the scEiaD resource for the PLAE web app in several important ways.
+First, we added two chicken (*Gallus gallus*) studies.
+Second, we added a human pan-body reference scRNA dataset.
+Third, we used an *in silico* method to remove background gene contamination.
+Fourth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained.
 
 ## pan mouse and pan human as non-ocular reference
-Tabula Muris
+Tabula Muris and
 

### 2021-10-23 manuscript_00.Rmd
text

@@
 library(RSQLite)
-meta_filter <- fst::read_fst("~/data/scEiaD_v2/2021_10_15_meta_filter.fst")
+meta_filter_v0 <- fst::read_fst("~/data/scEiaD/2021_03_17_meta_filter.fst")
+# remove eric bharti 2D/3D RPE
+meta_filter <- fst::read_fst("~/data/scEiaD_v2/2021_10_15_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D')
+
 ```
@@
 
+# eye studies
 b <-  meta_filter %>%
@@
   coord_flip()
+# eye studies samples
+c <-  meta_filter %>%
+  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
+                           TRUE ~ 'Body'),
+         Tissue = gsub('\\.|_', ' ', Tissue),
+         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
+         Tissue = gsub(" and Aorta","", Tissue)) %>% 
+  filter(Organ == 'Eye') %>% 
+  select(Organ, Tissue,organism, study_accession) %>%
+  unique() %>%
+  group_by(Organ, Tissue, organism) %>% 
+  summarise(Count = n()) %>%
+  ungroup() %>% 
+  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
+  ylab('Number of Cells') +
+  geom_bar(stat='identity', position = 'dodge') +
+  cowplot::theme_cowplot() +
+  facet_wrap(~Organ, scales = 'free') +
+  coord_flip()
 
 # retina cell types
-c <- meta_filter %>% filter(Organ == 'Eye') %>% 
+d <- meta_filter %>% filter(Organ == 'Eye') %>% 
   group_by(CellType_predict, study_accession, organism) %>%
@@
 
-## Update scEiaD resource
-Our first version of scEiaD contained XX studies, 3 species, XX cells, and XX curated cell types.
-We updated the scEiaD resource for the PLAE web app in several important ways.
-First, we added two chicken (*Gallus gallus*) studies.
-Second, we added a human pan-body reference scRNA dataset.
-Third, we used an *in silico* method to remove background gene contamination.
-Fourth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained.
-
 ## pan mouse and pan human as non-ocular reference
@@
 # Results
+
+## Updated the scEiaD backend
+Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types.
+We updated the scEiaD resource for the PLAE v0.80 web app in four important ways.
+First, we added two chicken retina (*Gallus gallus*) studies.
+Second we added an ocular outflow tract dataset.
+Third, we added a human pan-body reference scRNA dataset.
+Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX).
+Fourth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450.
+
+After the dataset and quality control updates the scEiaD v02 dataset contains `r meta_filter_$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).
+
+## PLAE contains 10 fold more cells and studies than the closest comparable web app
+
+## Extensive optimization makes the PLAE web app highly responsive despite containing huge amounts of data
 

### 2021-10-23 manuscript_00.Rmd
more text

@@
 # remove eric bharti 2D/3D RPE
-meta_filter <- fst::read_fst("~/data/scEiaD_v2/2021_10_15_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D')
+meta_filter <- fst::read_fst("~/data/scEiaD_v2/2021_10_22_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D')
 
@@
 # organ
+meta_filter <- meta_filter %>% mutate(organism = factor(organism, levels = meta_filter$organism %>% unique()))
 a <- meta_filter %>%
@@
   ungroup() %>% 
+  mutate(organism = factor(organism, levels = meta_filter$organism %>% unique())) %>% 
   ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
@@
   coord_flip() +
-  scale_y_continuous(breaks = c(2500, 5000, 7500, 10000, 12500, 84000))
+  scale_y_continuous(breaks = c(2500, 5000, 7500, 10000, 12500, 84000)) +
+  ggbreak::scale_y_break(c(13000, 82000)) +
+  scale_fill_discrete(drop=TRUE,
+                      limits = levels(meta_filter$organism)) +
+  theme(legend.position = "none")
 
@@
          Tissue = gsub("RPE$","RPE-Choroid", Tissue),
+         Tissue = case_when(Tissue == "Choroid" ~ 'RPE-Choroid',
+                            TRUE ~ Tissue),
          Tissue = gsub(" and Aorta","", Tissue)) %>% 
   filter(Organ == 'Eye') %>% 
-  group_by(Organ, Tissue,organism, study_accession) %>%
+  group_by(Organ, Tissue,organism) %>%
   summarise(Count = n()) %>%
@@
   cowplot::theme_cowplot() +
-  facet_wrap(~Organ, scales = 'free') +
-  coord_flip()
+  coord_flip() +
+  theme(legend.position = "none")
 # eye studies samples
@@
   ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
-  ylab('Number of Cells') +
+  ylab('Number of Studies') +
   geom_bar(stat='identity', position = 'dodge') +
   cowplot::theme_cowplot() +
-  facet_wrap(~Organ, scales = 'free') +
-  coord_flip()
+  coord_flip() +
+  theme(legend.position = "none")
 
@@
 
-plot_grid(a +  ggbreak::scale_y_break(c(13000, 82000)),b, c)
+plot_grid(a,b, c, d)
 ```

### 2021-10-25 manuscript_00.Rmd
text updates

@@
       - '--lua-filter=author-info-blocks.lua'
-bibliography: massive_integrated_eye_scRNA.bib
+bibliography: zotero_plaeApp.bib
 csl: investigative-ophthalmology-and-visual-science.csl
@@
 Studies in scEiaD
+```{r}
+meta_filter %>% group_by(Citation, PMID, study_accession) %>% summarise(Count = n()) %>% flextable()
+```
 
@@
   filter(Organ == 'Body') %>% 
+  filter(Tissue != 'Brain Choroid Plexus') %>% 
   group_by(Organ, Tissue,organism, study_accession) %>%
@@
   coord_flip() +
-  scale_y_continuous(breaks = c(2500, 5000, 7500, 10000, 12500, 84000)) +
-  ggbreak::scale_y_break(c(13000, 82000)) +
+  # scale_y_continuous(breaks = c(2500, 5000, 7500, 10000, 12500, 84000)) +
+  # ggbreak::scale_y_break(c(13000, 82000)) +
   scale_fill_discrete(drop=TRUE,
                       limits = levels(meta_filter$organism)) +
-  theme(legend.position = "none")
+  theme(legend.position = "none") +
+  scale_y_continuous(expand = c(0, 0)) + 
+  ylab('') + xlab('')
+
+aa <- meta_filter %>% 
+  filter(Tissue == 'Brain Choroid Plexus') %>% 
+  group_by(Organ, Tissue,organism, study_accession) %>%
+  summarise(Count = n()) %>%
+  ungroup() %>% 
+  mutate(organism = factor(organism, levels = meta_filter$organism %>% unique())) %>% 
+  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
+  ylab('Number of Cells') +
+  geom_bar(stat='identity', position = 'dodge') +
+  cowplot::theme_cowplot() +
+  coord_flip() +
+  scale_fill_discrete(drop=TRUE,
+                      limits = levels(meta_filter$organism)) +
+  theme(legend.position = "none") +
+  scale_y_continuous(expand = c(0, 0)) + 
+  ylab('') + xlab('')
 
@@
   coord_flip() +
-  theme(legend.position = "none")
+  theme(legend.position = "none") +
+  scale_y_continuous(expand = c(0, 0)) +
+  xlab('')
+
 # eye studies samples
@@
   coord_flip() +
-  theme(legend.position = "none")
+  theme(legend.position = "none") +
+  scale_y_continuous(expand = c(0, 0))  + xlab('')
 
@@
   cowplot::theme_cowplot() +
-  coord_flip()
+  coord_flip() + 
+  scale_y_continuous(expand = c(0, 0)) + xlab('')
 
-plot_grid(a,b, c, d)
+a_aa_b <- plot_grid(a,aa,b, ncol = 1, align = 'hv', rel_heights = c(10, 1.5,3.5), labels = 'a')
+c_d <- plot_grid(c,d, rel_heights = c(2,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))
+plot_grid(a_aa_b, c_d, ncol = 2, rel_widths = c(1,1.2))
 ```
@@
 
-## Data difficult to access
+## Pan Atlases contain limited retina cell types or lack integration
 
-## Cross comparisons difficult
+Tabula Sapiens, MOCA (?), BBrowser (\$\$\$), https://cells.ucsc.edu
+
+## Data difficult to access and cross comparisons impossible
+
 But important as seeing trends trends across different groups strongly demonstrates that 
@@
 
-## PLAE contains 10 fold more cells and studies than the closest comparable web app
+# Comparison of PLAE to "pan" web apps with retina data
+```{r}
+feature_table <- cbind(c('Spectacle', 'No' ,'No', ),
+                       c('UCSC Cell Browser', 'Yes', 'No', 'https://cells.ucsc.edu' ),
+                       c('Tabula Sapiens', 'No', 'N/A', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
+                       c('PLAE','Yes','Yes', 'plae.nei.nih.gov'),
+                       c('PanglaoDB', 'Yes', 'No'),
+                       c('Protein Atlas', 'Yes','No', 'https://www.proteinatlas.org/humanproteome/celltype'))
+colnames(feature_table) <- c('Resource Name', 'Independent Datasets', 'Harmonization', 'URL')
+```
+
+## PLAE contains over 10 fold more cells and studies than the closest comparable web app
+
+The Spectacle web app also provides a web app to access 87,305 single cell transcriptomes across four publications.
 

### 2021-10-25 manuscript_00.Rmd
text updates

--- 2021-10-23 manuscript_00.Rmd
@@
 Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types.
-We updated the scEiaD resource for the PLAE v0.80 web app in four important ways.
+We updated the scEiaD resource for the PLAE v0.80 web app in five important ways.
 First, we added two chicken retina (*Gallus gallus*) studies.
@@
 Fourth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450.
+Fifth, we carried and harmonnized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.
 
-After the dataset and quality control updates the scEiaD v02 dataset contains `r meta_filter_$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).
+After the dataset and quality control updates the scEiaD v02 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).
 
@@
 
-## Rich functionality
+## Rich web functionality
 
-## Web accessible
+## Differential testing identifies thousands of genes enriched in various ocular cell types
 

### 2021-11-13 manuscript_00.Rmd
text updates

@@
   - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health
+```{r}
+
+```
 
@@
 abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors.
-METHODS: We wrote the Shiny web application PLAE (https://plae.nei.nih.gov) and tailored the structure of the underlying data (sc Eye in a Disk, scEiaD) to display scEiaD.
+METHODS: We wrote the Shiny web application PLAE (https://plae.nei.nih.gov) and tailored the structure of the underlying data (single cell Eye in a Disk, scEiaD) to asdfasdf.
 RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes., XX cell types, XX ocular tissues, XX body tissues.
@@
 library(RSQLite)
-meta_filter_v0 <- fst::read_fst("~/data/scEiaD/2021_03_17_meta_filter.fst")
+meta_filter_v0 <- fst::read_fst("~/data/scEiaD/2021_03_17_meta_filter.fst") %>% 
+  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
+                                      TRUE ~ CellType_predict),
+         CellType= case_when(CellType == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
+                                      TRUE ~ CellType))
 # remove eric bharti 2D/3D RPE
-meta_filter <- fst::read_fst("~/data/scEiaD_v2/2021_10_22_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D')
+meta_filter <- fst::read_fst("/Volumes/McGaughey_S/scEiaD_v3/2021_11_09_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>% 
+  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
+                                      TRUE ~ CellType_predict),
+         CellType= case_when(CellType == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
+                                      TRUE ~ CellType))
 
@@
 
-# sup fig 1 
-Effect of decontx for Rho between scEiaD v2 and scEiaD v1
+# Sup fig 1 
+DecontX *in silico* ambient RNA contamination tool substantially removes Rhodopsin expression in non-rod cells while retaining high expression rods.
+First shown (a) is Rhodopsin expression across the scEiaD v0 dataset, without DecontX optimization.
+Note Rhodopsin expression is noticeable in many cell types beyond the rods.
+Next (b) is the scEiaD v1 dataset with DecontX optimization.
+Rhodopsin is nearly exclusively expressed in labelled rod cells.
+Despite the computational ambient RNA removal, Rhodopsin expression remains high in the Rods.
 
+```{r, fig.width=10, fig.height=12}
+source('src/make_exp_plot.R')
+scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  '~/data/scEiaD/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite')
+scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  '/Volumes/McGaughey_S/scEiaD_v3/MOARTABLES__anthology_limmaFALSE___6000-counts-universe-batch-scVIprojection-10-15-0.1-50-20.sqlite')
+
+input <- list()
+input$exp_plot_groups <- c('organism')
+input$exp_plot_facet <- c('CellType_predict')
+input$exp_plot_genes <- c('RHO (ENSG00000163914)')
+input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
+input$exp_plot_col_num <- 16
+input$exp_filter_min_cell_number <- 50
+
+v0_rho <- make_exp_plot(input, scEiaD_2020_v00, meta_filter_v0)
+v1_rho <-  make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+input$exp_filter_min_cell_number <- 1000
+v1.1_rho <- make_exp_plot(input, scEiaD_2020_v01, meta_filter %>% filter(CellType_predict %in% unique(meta_filter_v0$CellType_predict)))
+plot_grid(v0_rho, v1_rho, ncol = 1, labels = c('a','b'), rel_heights = c(1,1.5))
+```
 <!-- ```{r, fig.height=5, fig.width=3} -->
@@
 <!-- ``` -->
+
 # Introduction
+
 ## Many published ocular atlases
 
-## Pan Atlases contain limited retina cell types or lack integration
+## Several web atlases with ocular cells exist but have significant limitations
 
-Tabula Sapiens, MOCA (?), BBrowser (\$\$\$), https://cells.ucsc.edu
+There are several web app resources for querying gene expression across ocular cell types (table 1).
 
@@
 
-## Updated the scEiaD backend
-Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types.
+## Updates to the scEiaD backend
+Version zero of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types.
 We updated the scEiaD resource for the PLAE v0.80 web app in five important ways.
 First, we added two chicken retina (*Gallus gallus*) studies.
-Second we added an ocular outflow tract dataset.
+Second we added human ocular outflow tract and cornea datasets.
 Third, we added a human pan-body reference scRNA dataset.
 Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX).
-Fourth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450.
-Fifth, we carried and harmonnized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.
+Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450.
+Sixth, we carried and harmonized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.
 
@@
 
-# Comparison of PLAE to "pan" web apps with retina data
+# Comparison of PLAE to "pan" scRNA web apps containing ocular data
 ```{r}
-feature_table <- cbind(c('Spectacle', 'No' ,'No', ),
-                       c('UCSC Cell Browser', 'Yes', 'No', 'https://cells.ucsc.edu' ),
-                       c('Tabula Sapiens', 'No', 'N/A', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
-                       c('PLAE','Yes','Yes', 'plae.nei.nih.gov'),
-                       c('PanglaoDB', 'Yes', 'No'),
-                       c('Protein Atlas', 'Yes','No', 'https://www.proteinatlas.org/humanproteome/celltype'))
-colnames(feature_table) <- c('Resource Name', 'Independent Datasets', 'Harmonization', 'URL')
+feature_table <- rbind(
+  c('PLAE','Yes','Yes', 'Yes', '887', 'https://plae.nei.nih.gov'),
+  c('Spectacle', 'No' ,'No', 'No', '87', 'https://singlecell-eye.org'),
+  c('UCSC Cell Browser', 'Yes', 'No', 'Yes', '62', 'https://cells.ucsc.edu' ),
+  c('Tabula Sapiens', 'No', 'N/A', 'Yes', '10', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
+  c('PanglaoDB', 'Yes', 'No', 'Yes', '47', 'https://panglaodb.se'),
+  c('Protein Atlas', 'Yes*','No', 'Yes', '5',  'https://www.proteinatlas.org/humanproteome/celltype')) %>% 
+  as_tibble()
+colnames(feature_table) <- c('Resource Name', 'Independent Datasets', 'Harmonization', 'Ocular and non-Ocular Data', 'Ocular Cell Count\n(thousands)', 'URL')
+feature_table %>% flextable()
+# as of 2021 10 25
 ```

### 2022-03-12 manuscript_00.Rmd
text updates
text tweaks

--- 2021-10-25 manuscript_00.Rmd
@@
       institute: nnrl
-  - Robert Hufnagel:
-      institute: mgog
   - David McGaughey:
@@
 abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors.
-METHODS: We wrote the Shiny web application PLAE (https://plae.nei.nih.gov) and tailored the structure of the underlying data (sc Eye in a Disk, scEiaD) to display scEiaD.
+METHODS: We wrote the web application PLAE (https://plae.nei.nih.gov) and tailored the structure of the underlying data (sc Eye in a Disk, scEiaD) to display scEiaD.
 RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes., XX cell types, XX ocular tissues, XX body tissues.
@@
 
-## Pan Atlases contain limited retina cell types or lack integration
+## Pan Atlases contain limited ocular data or lack integration
 
@@
 Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types.
-We updated the scEiaD resource for the PLAE v0.80 web app in five important ways.
+We updated the scEiaD resource for the PLAE v0.90 web app in four important ways.
 First, we added two chicken retina (*Gallus gallus*) studies.
-Second we added an ocular outflow tract dataset.
+Second we added an ocular outflow tract dataset and two more cornea datasets.
 Third, we added a human pan-body reference scRNA dataset.
 Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX).
-Fourth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450.
-Fifth, we carried and harmonnized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.
+Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450.
 
-After the dataset and quality control updates the scEiaD v02 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).
+After the dataset and quality control updates the scEiaD v02 dataset contains `r meta_filter_$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).
 
@@
 ```{r}
-feature_table <- cbind(c('Spectacle', 'No' ,'No', ),
+feature_table <- rbind(c('Spectacle', 'No' ,'No', 'http://singlecell-eye.org'),
                        c('UCSC Cell Browser', 'Yes', 'No', 'https://cells.ucsc.edu' ),
@@
                        c('PLAE','Yes','Yes', 'plae.nei.nih.gov'),
-                       c('PanglaoDB', 'Yes', 'No'),
+                       c('PanglaoDB', 'Yes', 'No', 'https://panglaodb.se'),
                        c('Protein Atlas', 'Yes','No', 'https://www.proteinatlas.org/humanproteome/celltype'))
@@
 
-## PLAE contains over 10 fold more cells and studies than the closest comparable web app
+## PLAE differs from a complementary web app, Spectacle in several important ways
+
+While PLAE and Spectacle both contain large numbers of ocular-related scRNA datasets, they are structured differently.
+PLAE is built around the scEiaD database, which integrates all datasets together.
+This allows for queries to be constructed across studies and
 
@@
 
-## Rich web functionality
+## Rich functionality
 
-## Differential testing identifies thousands of genes enriched in various ocular cell types
+## Web accessible
 

### 2022-03-12 manuscript_00.Rmd

--- 2021-11-13 manuscript_00.Rmd
@@
   - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health
-```{r}
-
-```
 
@@
 abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors.
-METHODS: We wrote the Shiny web application PLAE (https://plae.nei.nih.gov) and tailored the structure of the underlying data (single cell Eye in a Disk, scEiaD) to asdfasdf.
-RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes., XX cell types, XX ocular tissues, XX body tissues.
+METHODS: We updated the content and structure of the underlying data (sc Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it.
+RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes, XX cell types, XX ocular tissues, XX body tissues.
 As a test of the value of this unified pan-eye dataset, we show XX.
-CONCLUSION: The PLAE v0.80 web app serves the pan-ocular and body dataset, scEiaD.
+CONCLUSION: The PLAE v0.90 web app serves the pan-ocular and body dataset, scEiaD.
 This offers the eye community a powerful and quick means to test hypotheses on gene and transcript expression across 54 body and 19 eye tissues.
 "
+
 keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
@@
 
-## Several web atlases with ocular cells exist but have significant limitations
+## Several web atlases with ocular cells exist but have limitations
 
@@
 
-## Updates to the scEiaD backend
-Version zero of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types.
-We updated the scEiaD resource for the PLAE v0.80 web app in five important ways.
+## Updated the scEiaD backend
+Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types.
+We updated the scEiaD resource for the PLAE v0.90 web app in four important ways.
 First, we added two chicken retina (*Gallus gallus*) studies.
-Second we added human ocular outflow tract and cornea datasets.
+Second we added an ocular outflow tract dataset and two more cornea datasets.
 Third, we added a human pan-body reference scRNA dataset.
@@
 ```{r}
+
 feature_table <- rbind(
@@
 
-## PLAE contains over 10 fold more cells and studies than the closest comparable web app
+## PLAE differs from a complementary web app, Spectacle in several important ways
 
-The Spectacle web app also provides a web app to access 87,305 single cell transcriptomes across four publications.
+While PLAE and Spectacle both contain large numbers of ocular-related scRNA datasets, they are structured differently.
+PLAE is built around the scEiaD database, which integrates all datasets together.
+This allows for queries to be constructed across studies and
+
+The Spectacle web app contains access 87,305 single cell transcriptomes across four publications.
 

### 2022-03-21 manuscript_00.Rmd
Merge branch 'master' of github.com:davemcg/plaeApp_manuscript
# Conflicts:
#	manuscript_00.Rmd

@@
 
+## New studies identified since 2019 data run
+
+## scEiaD re-created and pipeline updated
+
+The scEiaD pipeline is largely the same as in Swamy et al.
+[@swamy_building_2021].
+Briefly, we collect the raw fastq files and quantify with the kallisto with the bustools system [@melsted_barcode_2019; @melsted_modular_2019].
+The quantified counts are merged into one large matrix that is used for downstream processing with scvitools to batch correct the data and Seurat for quality control, cluster finding, UMAP finding.
+We use a xgboost-based machine learning approach to transfer cell type labels to unlabelled cells.
+
+The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300 Second, we use the DecontX algorithm [@yang_decontamination_2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination.
+Third, we altered the procedure for automatically aligning gene names across species (see below for details).
+Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in batch correction.
+
+## Cross species gene alignment
+
+As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify in a combined data analysis, we must align shared genes.
+We use a multi-step procedure.
+
+##
+
 ## pan mouse and pan human as non-ocular reference
@@
 First, we added two chicken retina (*Gallus gallus*) studies.
-Second we added an ocular outflow tract dataset and two more cornea datasets.
-Third, we added a human pan-body reference scRNA dataset.
+Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye.
+Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons.
 Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX).
-Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450.
-Sixth, we carried and harmonized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.
+Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450, which reduces the number of cells but improves the quality.
+Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.
 
@@
 feature_table <- rbind(
-  c('PLAE','Yes','Yes', 'Yes', '887', 'https://plae.nei.nih.gov'),
-  c('Spectacle', 'No' ,'No', 'No', '87', 'https://singlecell-eye.org'),
+  c('PLAE','Yes','Yes', 'Yes', '1004', 'https://plae.nei.nih.gov'),
+  c('Spectacle', 'Yes' ,'No', 'No', '1800', 'https://singlecell-eye.org'),
   c('UCSC Cell Browser', 'Yes', 'No', 'Yes', '62', 'https://cells.ucsc.edu' ),
@@
 While PLAE and Spectacle both contain large numbers of ocular-related scRNA datasets, they are structured differently.
-PLAE is built around the scEiaD database, which integrates all datasets together.
+PLAE is built around the scEiaD database, which runs an extensive multi-stage pipeline to build and identify a high performing scVI-based eye model that is used to integrate all datasets together.
 This allows for queries to be constructed across studies and

### 2022-03-22 manuscript_00.Rmd
many text adds

@@
 library(RSQLite)
-meta_filter_v0 <- fst::read_fst("~/data/scEiaD/2021_03_17_meta_filter.fst") %>% 
+meta_filter_v0 <- fst::read_fst("/Volumes/McGaughey_S/scEiaD/paper_data/2021_03_17_meta_filter.fst") %>% 
   mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
@@
 
+We added X studies from X species and X tissues. 
 ## scEiaD re-created and pipeline updated
@@
 [@swamy_building_2021].
-Briefly, we collect the raw fastq files and quantify with the kallisto with the bustools system [@melsted_barcode_2019; @melsted_modular_2019].
-The quantified counts are merged into one large matrix that is used for downstream processing with scvitools to batch correct the data and Seurat for quality control, cluster finding, UMAP finding.
-We use a xgboost-based machine learning approach to transfer cell type labels to unlabelled cells.
+Briefly, we collect the raw fastq files and quantify with the kallisto / bustools system [@melsted_barcode_2019; @melsted_modular_2019].
+The quantified counts are merged into one matrix that is used for downstream processing with scvitools to batch correct the data and Seurat for quality control, clustering, and 2D UMAP visualization.
+We use our previous xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamy_building_2021].
 
@@
 Third, we altered the procedure for automatically aligning gene names across species (see below for details).
-Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in batch correction.
+Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process.
+
+Two custom cutoffs for number of unique genes identified were used for the E12 chicken data and the cornea dataset (SRP362101) as an abnormally high number of cells were returned from our pipeline.
+We set the cutoffs for these studies to be 1000 and 800, respectively, as these values resulted in approximately the same number of cells being returned as the authors reported in their papers.
 
@@
 
-As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify in a combined data analysis, we must align shared genes.
-We use a multi-step procedure.
+As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify, we must align shared genes.
+We found homologs/orthologs by pulling from BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference.
+We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus Musculus*.
+This generates a table linked by Ensembl gene IDs and gene names.
+We then use a full join on non-duplicated Ensembl IDs between mouse and human.
+As we noticed a small number of gene names failed to be linked in this manner we ran another join on the remaining non-aligned genes, using gene name.
+This procedure gave us 17,769 shared human and mouse genes.
+To align the other species we again joined on Ensembl gene ID, removing genes where one chicken or macaque gene aligned to multiple human/mouse genes.
+In cases where multiple chicken or macaque genes aligned to one human/mouse gene, then we aggregated the counts by sum.
+In cases where there was no corresponding chicken or macaque gene, we filled in zeros.
+After these steps we had, as expected, 17,769 genes.
 
-##
+## Metadata curation
 
-## pan mouse and pan human as non-ocular reference
-Tabula Muris and
+Every study brought into scEiaD was hand curated to label biological sample, published paper ID (where available), organ, tissue (e.g.
+Cornea), source (iPSC based or tissue), single cell platform (e.g.
+10Xv1 or DropSeq).
+We also, where possible, annotated retina region (e.g.
+fovea), sex, and age or developmental stage.
+To bring across individual cell type labels (e.g.
+Rod, Muller Glia), we wrote custom code for each study that made this table available to link to the individual cells.
+We curated two types of cell type labels: "CellType" and "PubCellType" where the former is our normalized cell types taken from the published labels.
+In this field we normalized names, for example changin "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).
+We did retain the original published (except for the broader name normalization) labels under PubCellType.
+
+## Pan mouse and human non-ocular references
+
+The Tabula Muris and
+
+## scEiaD structure and web app optimization
+
+The web app at plae.nei.nih.gov can display million plus cell level gene expression in only a few seconds.
+This speed is only possible due to custom data structures to optimize for data querying.
+Like eyeIntegration's EiaD database, scEiaD is a sqlite database with tables for the core categories.
+This allows for app start up to complete in under 30 seconds on a modestly performant cloud server with memory usage under 8 GB.
+For the gene by cell expression matrix, the data was transformed into a "long" format where there are three columns: Gene, Cell Barcode, and Gene Count Value.
+This allows for cell - gene level queries to be completed, again with minimal memory usage, in under 0.5 seconds.
+To generate the aggregated information (in the Dot Plot, Expression Plot, and tables) we pre-calculated all queries by running a dplyr "group by" operation on all column fields.
+On user given query, the data is further aggregated to level request.
+This allows for complicated queries to complete in seconds rather than minutes.
+
+## Data reproducibility
+
+The codebase for scEiaD is available at github.com/davemcg/scEiaD.
+There are three relevent Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which takes the optimal parameters from SnakePOP to build the scEiaD sqlite database for the app.
 
@@
 
-## Updated the scEiaD backend
+## Updates to the scEiaD backend
 Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types.
-We updated the scEiaD resource for the PLAE v0.90 web app in four important ways.
+We updated the scEiaD resource for the PLAE v0.90 web app in six important ways.
 First, we added two chicken retina (*Gallus gallus*) studies.
@@
 Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX).
-Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450, which reduces the number of cells but improves the quality.
+Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 200 to 300, which reduces the number of cells that pass QC but improves the quality.
 Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.
 
-After the dataset and quality control updates the scEiaD v02 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).
+After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).
 
@@
 
-While PLAE and Spectacle both contain large numbers of ocular-related scRNA datasets, they are structured differently.
+While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured very differently.
 PLAE is built around the scEiaD database, which runs an extensive multi-stage pipeline to build and identify a high performing scVI-based eye model that is used to integrate all datasets together.
-This allows for queries to be constructed across studies and
+This allows for gene queries to be constructed and analyzed across studies.
+In contrast, Spectacle, as of March 2022, is a compilation of study-level datasets and thus you cannot run queries across studies in Spectacle.
 
-The Spectacle web app contains access 87,305 single cell transcriptomes across four publications.
+There are a few other differences in between PLAE and Spectacle.
+First, there are no non-ocular datasets in Spectacle.
+Second, the cell types labels have not been harmonized.
+Third, as this resource uses Seurat objects, the app takes many seconds to load an individual dataset and many more to display the data.
 
@@
 
-## Rich web functionality
+There are two interwoven reasons that PLAE is responsive, despite the huge amount of data present.
+First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables.
+A sqlite database allows for huge amounts of data to be stored for quick retrieval with minimal loading time.
+The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous.
+Second, the visualizations in PLAE (e.g.
+the UMAP view) pull from the custom database directly and use highly optimized plotting packages (scattermore) to reduce drawing time.
+Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study.
+In the same amount of time, PLAE can plot the cell type labels for over a million cells.
 
-## Differential testing identifies thousands of genes enriched in various ocular cell types
+## Rich web-available visualizations
 
-##
+PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values.
+So thus a user can, on the left, see Crx expression across over a million cells, and on the right, see cell type labels for those same million cells.
+The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells.
+Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area.
 
+For users interested in seeing how expression of a gene relates to study or other covariates, the "Exp Plot" view displays expression by cell type, predicted cell type, or cluster.
+The values can then be split and colored by multiple covariates like study or organism.
+This allows a user to see whether a gene pattern is consistent across studies or species.
+We also offer a "Dot Plot" view which can display a large number of genes in a space efficient manner.
+This allows a users to see whether a set of genes has related (or not) expression patterns across multiple cell types.
+
+While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina.
+
+## Differential testing identifies many genes consistently enriched in various ocular cell types
+
+## Ocular specific gene signatures
+
+## Fovea vs peripheral photoreceptor validation
+
+Sanes proposed X genes (https://www.nature.com/articles/s41598-020-66092-9 --> Figure 9)
+
+PRPH2 and RS1 for cones
+NR2E3, SLC24A1, PDE6H, GNAT1 for rods
+Also did RGC, Microglia, Astrocytes, Muller Glia
+
+## Design of scEiaD allows for projection of outside data
+
+scEiaD is built using scvitool's scVI / scANVI batch correction tool.
+As part of the process a computational model of the reference datasets are built.
+This model is only a few mb in size and can be used on a modest computer to project private scRNA datasets onto it.
+This allows an ocular researcher several functions.
+First, the ability to check of quality control of an existing dataset by looking at the distribution of their cells overlapped with our labelled cell types.
+Second, in perturbation experiments, to identify unusual distributions of cells.
+Third, to leverage our extensive pre-labelled cell information to transfer cell type labels onto private data.
 # Conclusion

### 2022-03-22 manuscript_00.Rmd
text / code tweaks

@@
 library(RSQLite)
-meta_filter_v0 <- fst::read_fst("/Volumes/McGaughey_S/scEiaD/paper_data/2021_03_17_meta_filter.fst") %>% 
+meta_filter_v0 <- fst::read_fst("~/data/scEiaD/paper_data/2021_03_17_meta_filter.fst") %>% 
   mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
@@
 # remove eric bharti 2D/3D RPE
-meta_filter <- fst::read_fst("/Volumes/McGaughey_S/scEiaD_v3/2021_11_09_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>% 
+meta_filter <- fst::read_fst("~/data/scEiaD_2022_02/2022_03_22_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>% 
   mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
@@
          CellType= case_when(CellType == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
-                                      TRUE ~ CellType))
+                             TRUE ~ CellType),  
+         Tissue = case_when(grepl('Iris', Tissue) ~ 'Iris',
+                            TRUE ~ Tissue))
 
@@
 # Figure 1 Distribution of Tissues and Cell Types in scEiaD
-```{r, fig.width=8, fig.height=5}
+```{r, fig.width=5.5, fig.height=5}
 # organ
@@
 c_d <- plot_grid(c,d, rel_heights = c(2,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))
-plot_grid(a_aa_b, c_d, ncol = 2, rel_widths = c(1,1.2))
+plot_grid(a_aa_b, c_d, ncol = 2, rel_widths = c(1,1.6))
 ```
@@
 scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  '~/data/scEiaD/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite')
-scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  '/Volumes/McGaughey_S/scEiaD_v3/MOARTABLES__anthology_limmaFALSE___6000-counts-universe-batch-scVIprojection-10-15-0.1-50-20.sqlite')
+scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  '~/data/scEiaD_2022_02/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20.sqlite')
 
@@
 ## Differential testing identifies many genes consistently enriched in various ocular cell types 
+```{r}
+scEiaD_2020_v01 %>% tbl('wilcox_diff_testing') %>% filter(Group == 'CellType (Predict)', FDR < 1e-5, mean_auc > 0.7 ) %>% group_by(Base) %>% summarise(Count = n()) %>% data.frame()
+```
 
-## Ocular specific gene signatures
-
-## Fovea vs peripheral photoreceptor validation
+## Fovea vs peripheral photoreceptor multi-study validation
 
@@
 
-## Design of scEiaD allows for projection of outside data
+## Projection of organoid data onto scEiaD demonstrates [SOMETHING]
 

### 2022-03-28 manuscript_00.Rmd
text updates

@@
 
-## New studies identified since 2019 data run
+## scEiaD re-built and pipeline updated
 
-We added X studies from X species and X tissues. 
-## scEiaD re-created and pipeline updated
-
-The scEiaD pipeline is largely the same as in Swamy et al.
+The scEiaD pipeline is largely the same as used in Swamy et al.
 [@swamy_building_2021].
@@
 
-The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300 Second, we use the DecontX algorithm [@yang_decontamination_2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination.
-Third, we altered the procedure for automatically aligning gene names across species (see below for details).
+The following alterations were made compared to our previous publication.
+First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300.
+Second, we use the DecontX algorithm [@yang_decontamination_2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination.
+This is run across each study.
+Third, we altered the procedure for aligning gene names across species (see below for details).
 Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process.
 
-Two custom cutoffs for number of unique genes identified were used for the E12 chicken data and the cornea dataset (SRP362101) as an abnormally high number of cells were returned from our pipeline.
+Two custom cutoffs for number of unique genes identified were used for E12 chicken data (SRP286543) and a cornea dataset (SRP362101) as an abnormally high number of cells were returned from our pipeline.
 We set the cutoffs for these studies to be 1000 and 800, respectively, as these values resulted in approximately the same number of cells being returned as the authors reported in their papers.
@@
 
-As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify, we must align shared genes.
 We found homologs/orthologs by pulling from BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference.
 We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus Musculus*.
-This generates a table linked by Ensembl gene IDs and gene names.
+This generates a table linked by Ensembl gene IDs (e.g.
+ENSG00000105392) and gene names (e.g.
+CRX).
 We then use a full join on non-duplicated Ensembl IDs between mouse and human.
@@
 Rod, Muller Glia), we wrote custom code for each study that made this table available to link to the individual cells.
-We curated two types of cell type labels: "CellType" and "PubCellType" where the former is our normalized cell types taken from the published labels.
-In this field we normalized names, for example changin "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).
-We did retain the original published (except for the broader name normalization) labels under PubCellType.
+We curated two types of cell type labels: "CellType" and "PubCellType".
+In the "CellType" field we normalized names, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).
+The original published (except for the broader name normalization) labels are retained in PubCellType.
 
-## Pan mouse and human non-ocular references
-
-The Tabula Muris and
+## Differential expression testing
 
@@
 
+Similar to our related bulk RNA-seq resource, eyeIntegration, PLAE is built with the R-based Shiny platform.
+It is hosted on a cloud server instance.
+We make all data files and code available (www.github.com/plaeApp) to enable local install of the resource, if desired.
+
 ## Data reproducibility
@@
 
-## Updates to the scEiaD backend
+## Updates to the scEiaD database
 Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types.
@@
 
-After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).
+After the dataset and quality control updates the scEiaD 2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).
 
@@
 Third, as this resource uses Seurat objects, the app takes many seconds to load an individual dataset and many more to display the data.
+Fourth, extensive manual curation was done to normalize cell type names, developmental stage, and other important experimental details (see Methods).
 
@@
 
-## Rich web-available visualizations
+## Rich interactive visualizations
 
-PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values.
-So thus a user can, on the left, see Crx expression across over a million cells, and on the right, see cell type labels for those same million cells.
+PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell value plots.
+For example, a user can see Crx expression across over a million cells on the left and cell type labels for those same million cells on the right.
 The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells.
@@
 
-For users interested in seeing how expression of a gene relates to study or other covariates, the "Exp Plot" view displays expression by cell type, predicted cell type, or cluster.
+For users interested in seeing how expression of a gene relates to study or other covariates, the "Exp Plot" (short for Expression) view displays expression by cell type, predicted cell type, or cluster.
 The values can then be split and colored by multiple covariates like study or organism.
 This allows a user to see whether a gene pattern is consistent across studies or species.
-We also offer a "Dot Plot" view which can display a large number of genes in a space efficient manner.
-This allows a users to see whether a set of genes has related (or not) expression patterns across multiple cell types.
+We also offer a "Dot Plot" view which can display expression as dots by gene and user-selected covariate (the dot color is level of expression and size is percentage).
+This allows a users to see whether a semi-large of genes has related (or not) expression patterns across multiple cell types.
 
-While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina.
+While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer.
+The major cells of the retina through to the RPE are displayed in a drawn cross-section.
+Each cell type is colored by the amount of expression of a user given gene.
 

### 2022-03-30 manuscript_00.Rmd
off sync text updates

--- 2022-03-22 manuscript_00.Rmd
@@
 Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX).
-Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 200 to 300, which reduces the number of cells that pass QC but improves the quality.
+Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300, which reduces the number of cells that pass QC but improves the quality.
 Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.
@@
 
-There are two interwoven reasons that PLAE is responsive, despite the huge amount of data present.
+There are two interwoven reasons why PLAE is responsive, despite the huge amount of data it contains.
 First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables.
-A sqlite database allows for huge amounts of data to be stored for quick retrieval with minimal loading time.
-The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous.
+A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage.
+The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous; the scEiaD v2022-03-21 sqlite database is abour 368 GB.
 Second, the visualizations in PLAE (e.g.
 the UMAP view) pull from the custom database directly and use highly optimized plotting packages (scattermore) to reduce drawing time.
-Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study.
-In the same amount of time, PLAE can plot the cell type labels for over a million cells.
+Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study (tested on 2022-03-23).
+In the same amount of time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies.
 
@@
 PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values.
-So thus a user can, on the left, see Crx expression across over a million cells, and on the right, see cell type labels for those same million cells.
+So thus a user can see Crx expression across over a million cells and separately see cell type labels for those same million cells.
 The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells.
@@
 While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina.
+The structure of the cartoon retina was based on the human retina.
+This unique visualization displays the major cell types of the retina and the RPE / choroid layer behind it.
+The cell types are colored by the relative amounts of expression in the user selected gene.
+Like all visualizations in PLAE, the user can filter to only display data from certain metadata characteristics, like species or publication.
 
-## Differential testing identifies many genes consistently enriched in various ocular cell types 
+## Differential testing identifies thousands genes frequently enriched across various ocular cell types 
 ```{r}
@@
 
-## Fovea vs peripheral photoreceptor multi-study validation
+## Pan-study fovea vs peripheral photoreceptor identifies N genes consistently differentially expressed
 

### 2022-03-30 manuscript_00.Rmd

@@
 
-## Pan mouse and human non-ocular references
+## Differential gene testing
 
-The Tabula Muris and
+Pseudobulk matrices were created for each organism with genes as the rows and study_accession - [test] as the columns; [test] is either cluster, CellType, or CellType_predict with the R package scater function XXX.
+Genes were removed which had zero expression across all cells.
+Columns were not retained if they were composed of fewer than 20 total cells.
+These pseudobulk matrices were used as the input for DESeq2 differential testing.
+The design of the test was "~study_accession + [test]" where, again, test is either cluster, CellType, or CellType_predict.
+This design sets the study_accession as the covariate, thus removing effects that are specific to a small number of studies.
+As the differential testing was done for each individual species, we aligned the separate result tables by the Ensembl gene ID to create a unified table.
+Values (e.g.
+pvalue) specific
 
@@
 The cell types are colored by the relative amounts of expression in the user selected gene.
-Like all visualizations in PLAE, the user can filter to only display data from certain metadata characteristics, like species or publication.
+Like all visualizations in PLAE, the user can use powerful filtering options to only display data from certain metadata characteristics, like species or publication.
 
@@
 
+To aid in in the analysis of ocular gene expression patterns we provide pre-computed differential tests run 
 ## Pan-study fovea vs peripheral photoreceptor identifies N genes consistently differentially expressed

### 2022-03-31 manuscript_00.Rmd
text updates

@@
 library(RSQLite)
-meta_filter_v0 <- fst::read_fst("~/data/scEiaD/paper_data/2021_03_17_meta_filter.fst") %>% 
+meta_filter_v0 <- fst::read_fst("/Volumes/McGaughey_S/scEiaD/paper_data/2021_03_17_meta_filter.fst") %>% 
   mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
@@
 # remove eric bharti 2D/3D RPE
-meta_filter <- fst::read_fst("~/data/scEiaD_2022_02/2022_03_22_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>% 
+meta_filter <- fst::read_fst("~/data/scEiaD_2022_02/meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>% 
   mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
@@
 ```{r}
-meta_filter %>% group_by(Citation, PMID, study_accession) %>% summarise(Count = n()) %>% flextable()
+meta_filter %>% group_by(Citation, PMID, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable()
 ```
@@
 # Figure 1 Distribution of Tissues and Cell Types in scEiaD
-```{r, fig.width=5.5, fig.height=5}
+```{r, fig.width=6, fig.height=5}
 # organ
@@
 d <- meta_filter %>% filter(Organ == 'Eye') %>% 
+  mutate(CellType_predict = case_when(is.na(CellType_predict) ~ 'Unlabelled', TRUE ~ CellType_predict)) %>% 
   group_by(CellType_predict, study_accession, organism) %>%
@@
 a_aa_b <- plot_grid(a,aa,b, ncol = 1, align = 'hv', rel_heights = c(10, 1.5,3.5), labels = 'a')
-c_d <- plot_grid(c,d, rel_heights = c(2,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))
+c_d <- plot_grid(c,d, rel_heights = c(3,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))
 plot_grid(a_aa_b, c_d, ncol = 2, rel_widths = c(1,1.6))
@@
 source('src/make_exp_plot.R')
-scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  '~/data/scEiaD/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite')
-scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  '~/data/scEiaD_2022_02/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20.sqlite')
+scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  '/Volumes/McGaughey_S/scEiaD/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite')
+scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  '/Volumes/McGaughey_S/scEiaD_2022/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20.sqlite')
 
@@
 
-## Several web atlases with ocular cells exist but have limitations
+## Web atlases containing ocular data have substantial limitations
 
@@
 
-## Data difficult to access and cross comparisons impossible
+Data difficult to access and cross comparisons impossible
 
@@
 
-## New studies identified since 2019 data run
-
-We added X studies from X species and X tissues. 
 ## scEiaD re-created and pipeline updated
 
-The scEiaD pipeline is largely the same as in Swamy et al.
+The scEiaD pipeline is structurally the same as in Swamy et al.
 [@swamy_building_2021].
-Briefly, we collect the raw fastq files and quantify with the kallisto / bustools system [@melsted_barcode_2019; @melsted_modular_2019].
+Very briefly, we collect the raw fastq files and quantify with the kallisto / bustools system [@melsted_barcode_2019; @melsted_modular_2019].
 The quantified counts are merged into one matrix that is used for downstream processing with scvitools to batch correct the data and Seurat for quality control, clustering, and 2D UMAP visualization.
@@
 
-## Updates to the scEiaD backend
-Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types.
+## New studies and improvements to the scEiaD database
+Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamy_building_2021].
 We updated the scEiaD resource for the PLAE v0.90 web app in six important ways.
@@
 feature_table <- rbind(
-  c('PLAE','Yes','Yes', 'Yes', '1004', 'https://plae.nei.nih.gov'),
+  c('PLAE','Yes','Yes', 'Yes', '1136 ', 'https://plae.nei.nih.gov'),
   c('Spectacle', 'Yes' ,'No', 'No', '1800', 'https://singlecell-eye.org'),

### 2022-03-31 manuscript_00.Rmd
alter data load

--- 2022-03-30 manuscript_00.Rmd
@@
 library(captioner)
+library(glue)
+data_dir_vGiga <- '~/data/scEiaD/'
+data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 
@@
 library(RSQLite)
-meta_filter_v0 <- fst::read_fst("~/data/scEiaD/paper_data/2021_03_17_meta_filter.fst") %>% 
+meta_filter_v0 <- fst::read_fst(glue("{data_dir_vGiga}/paper_data/2021_03_17_meta_filter.fst")) %>% 
   mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
@@
                              TRUE ~ CellType))
-# remove eric bharti 2D/3D RPE
-meta_filter <- fst::read_fst("~/data/scEiaD_2022_02/2022_03_22_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>% 
+
+meta_filter <- fst::read_fst(glue("{data_dir_vPLAE}/meta_filter.fst")) %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>% 
   mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
@@
          Tissue = case_when(grepl('Iris', Tissue) ~ 'Iris',
-                            TRUE ~ Tissue))
+                            TRUE ~ Tissue),
+
+         Citation = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'Swamy VS, Fufa TD, Hufnagel RB, McGaughey DM Building the Mega Single Cell Transcriptome Ocular Meta-Atlas Gigascience 2021 Oct 13;10(10)',
+                              TRUE ~ Citation),
+         PMID = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ '34651173',
+                          TRUE ~ PMID),
+         study_accession = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'SRP329495',
+                                     TRUE ~ study_accession))
 
@@
 Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX).
-Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300, which reduces the number of cells that pass QC but improves the quality.
+Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 200 to 300, which reduces the number of cells that pass QC but improves the quality.
 Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.
@@
 
-There are two interwoven reasons why PLAE is responsive, despite the huge amount of data it contains.
+There are two interwoven reasons that PLAE is responsive, despite the huge amount of data present.
 First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables.
-A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage.
-The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous; the scEiaD v2022-03-21 sqlite database is abour 368 GB.
+A sqlite database allows for huge amounts of data to be stored for quick retrieval with minimal loading time.
+The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous.
 Second, the visualizations in PLAE (e.g.
 the UMAP view) pull from the custom database directly and use highly optimized plotting packages (scattermore) to reduce drawing time.
-Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study (tested on 2022-03-23).
-In the same amount of time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies.
+Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study.
+In the same amount of time, PLAE can plot the cell type labels for over a million cells.
 
@@
 PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values.
-So thus a user can see Crx expression across over a million cells and separately see cell type labels for those same million cells.
+So thus a user can, on the left, see Crx expression across over a million cells, and on the right, see cell type labels for those same million cells.
 The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells.
@@
 While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina.
-The structure of the cartoon retina was based on the human retina.
-This unique visualization displays the major cell types of the retina and the RPE / choroid layer behind it.
-The cell types are colored by the relative amounts of expression in the user selected gene.
-Like all visualizations in PLAE, the user can filter to only display data from certain metadata characteristics, like species or publication.
 
-## Differential testing identifies thousands genes frequently enriched across various ocular cell types 
+## Differential testing identifies many genes consistently enriched in various ocular cell types 
 ```{r}
@@
 
-## Pan-study fovea vs peripheral photoreceptor identifies N genes consistently differentially expressed
+## Fovea vs peripheral photoreceptor multi-study validation
 

### 2022-03-31 manuscript_00.Rmd

--- 2022-03-31 manuscript_00.Rmd
@@
 library(RSQLite)
-meta_filter_v0 <- fst::read_fst("/Volumes/McGaughey_S/scEiaD/paper_data/2021_03_17_meta_filter.fst") %>% 
+meta_filter_v0 <- fst::read_fst(glue("{data_dir_vGiga}/paper_data/2021_03_17_meta_filter.fst")) %>%
+
   mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
@@
                              TRUE ~ CellType))
-# remove eric bharti 2D/3D RPE
-meta_filter <- fst::read_fst("~/data/scEiaD_2022_02/meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>% 
+
+meta_filter <- fst::read_fst(glue("{data_dir_vPLAE}/meta_filter.fst")) %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>%
+
   mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',

### 2022-03-31 manuscript_00.Rmd
updates

@@
 
-While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured very differently.
+While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently.
 PLAE is built around the scEiaD database, which runs an extensive multi-stage pipeline to build and identify a high performing scVI-based eye model that is used to integrate all datasets together.
@@
 In contrast, Spectacle, as of March 2022, is a compilation of study-level datasets and thus you cannot run queries across studies in Spectacle.
-
-There are a few other differences in between PLAE and Spectacle.
-First, there are no non-ocular datasets in Spectacle.
-Second, the cell types labels have not been harmonized.
-Third, as this resource uses Seurat objects, the app takes many seconds to load an individual dataset and many more to display the data.
+Also, Spectacle contains no non-ocular dataset and the cell types labels have not been harmonized between studies.
+Finally, as this Spectacle uses Seurat objects, the app takes many seconds to load an individual dataset and many more to display the data.
 
@@
 A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage.
-The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous; the scEiaD v2022-03-21 sqlite database is abour 368 GB.
+The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous; the scEiaD v2022-03-21 sqlite database is about 368 GB.
 Second, the visualizations in PLAE (e.g.
-the UMAP view) pull from the custom database directly and use highly optimized plotting packages (scattermore) to reduce drawing time.
+the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) along with customized R plotting code to reduce drawing time.
 Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study (tested on 2022-03-23).
@@
 ```{r}
-scEiaD_2020_v01 %>% tbl('wilcox_diff_testing') %>% filter(Group == 'CellType (Predict)', FDR < 1e-5, mean_auc > 0.7 ) %>% group_by(Base) %>% summarise(Count = n()) %>% data.frame()
+scEiaD_2020_v01 %>% tbl('wilcox_diff_testing') %>% filter(Group == 'CellType (Predict)', FDR < 1e-5, mean_auc > 0.7 ) %>% 
+  group_by(Base) %>% 
+  summarise(Count = n()) %>% 
+  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
+  as_tibble() %>% 
+  flextable()
 ```

### 2022-04-02 manuscript_00.Rmd
new text

@@
 library(glue)
-data_dir_vGiga <- '~/data/scEiaD/'
-data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+data_dir_vGiga <- '/Volumes/McGaughey_S/scEiaD/'
+data_dir_vPLAE <- '/Volumes/McGaughey_S/scEiaD_2022_02/'
 
@@
 source('src/make_exp_plot.R')
-scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  '/Volumes/McGaughey_S/scEiaD/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite')
-scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  '/Volumes/McGaughey_S/scEiaD_2022/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20.sqlite')
+scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vGiga}/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite'))
+scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vPLAE}/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20.sqlite'))
 
@@
 
-Collated XX studies and processed
+Collated `r meta_filter$study_accession %>% unique() %>% length()` studies across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues.
+We hand-curated `r meta_filter$CellType %>% unique() %>% length()` cell types, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the remaining cells.
+Extensive quality control metrics were tuned and applied to remove lower quality cells.
+In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values.
 
@@
 ## Differential gene testing
-
-Pseudobulk matrices were created for each organism with genes as the rows and study_accession - [test] as the columns; [test] is either cluster, CellType, or CellType_predict with the R package scater function XXX.
-Genes were removed which had zero expression across all cells.
-Columns were not retained if they were composed of fewer than 20 total cells.
-These pseudobulk matrices were used as the input for DESeq2 differential testing.
-The design of the test was "~study_accession + [test]" where, again, test is either cluster, CellType, or CellType_predict.
-This design sets the study_accession as the covariate, thus removing effects that are specific to a small number of studies.
-As the differential testing was done for each individual species, we aligned the separate result tables by the Ensembl gene ID to create a unified table.
-Values (e.g.
-pvalue) specific
 
@@
 To aid in in the analysis of ocular gene expression patterns we provide pre-computed differential tests run
+
 ## Pan-study fovea vs peripheral photoreceptor identifies N genes consistently differentially expressed

### 2022-04-13 manuscript_00.Rmd
new text

@@
                             TRUE ~ Tissue),
-
+         Tissue = case_when(grepl('RPE|Choroid', Tissue) ~ 'RPE-Choroid',
+                            TRUE ~ Tissue),
          Citation = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'Swamy VS, Fufa TD, Hufnagel RB, McGaughey DM Building the Mega Single Cell Transcriptome Ocular Meta-Atlas Gigascience 2021 Oct 13;10(10)',
@@
 
+scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vGiga}/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite'))
+scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vPLAE}/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20.sqlite'))
+
 ```
 
-# (sup?) Table 1
-Studies in scEiaD
-```{r}
-meta_filter %>% group_by(Citation, PMID, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable()
-```
+# Introduction
 
-# Figure 1 Distribution of Tissues and Cell Types in scEiaD
-```{r, fig.width=6, fig.height=5}
+## single cell RNA-seq allows for a higher resolution view of ocular transcriptomics
+
+Single cell RNA-sequencing dates back to . The first
+
+## Many published ocular single cell RNA-seq atlases
+
+## Web atlases containing ocular data have substantial limitations
+
+Data difficult to access and cross comparisons impossible
+
+But important as seeing trends trends across different groups strongly demonstrates that 
+the signal is biological, not technical.
+
+## What we did
+
+We published the ocular focused scEiaD v2021 resource in 2021, which was one of the first single cell meta-atlases.
+To broaden the usage of this powerful database, we restructured the data in a custom sqlite-based database and built a reactive web application around it.
+As many new studies were published since our last data build, we re-ran the scEiaD pipeline in March of 2022.
+In aggregate we have collated `r meta_filter$study_accession %>% unique() %>% length()` studies across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues.
+We hand-curated `r meta_filter$CellType %>% unique() %>% length()` cell types, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the remaining cells.
+Extensive quality control metrics were tuned and applied to remove lower quality cells.
+In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values.
+
+# Methods
+
+## scEiaD pipeline upgraded and the database re-built
+
+The scEiaD pipeline is structurally the same as in Swamy et al.
+[@swamy_building_2021].
+Very briefly, we collect the raw fastq files and quantify with the kallisto / bustools system [@melsted_barcode_2019; @melsted_modular_2019].
+The quantified counts are merged into one matrix that is used for downstream processing with scvitools to batch correct the data and Seurat for quality control, clustering, and 2D UMAP visualization.
+We use our previous xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamy_building_2021].
+
+The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300 Second, we use the DecontX algorithm [@yang_decontamination_2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination.
+Third, we altered the procedure for automatically aligning gene names across species (see below for details).
+Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process.
+
+Two custom cutoffs for the minimum required number of unique genes identified per cell were used for the E12 chicken data and the cornea dataset (SRP362101) as an abnormally high number of cells were returned from our default value of 250 (????).
+We set the cutoffs for these studies to be 1000 and 800, respectively, as these values resulted in approximately the same number of cells being returned as the authors reported in their papers.
+
+## Cross species gene alignment
+
+As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify, we must identify shared genes.
+We found homologs/orthologs by pulling from BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference.
+We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus Musculus*.
+This generates a table linked by Ensembl gene IDs and gene names.
+We then use a full join on non-duplicated Ensembl IDs between mouse and human.
+As we noticed a small number of gene names failed to be linked in this manner we ran another join on the remaining non-aligned genes, using gene name.
+This procedure gave us 17,769 shared human and mouse genes.
+To align the other species we again joined on Ensembl gene ID, removing genes where one chicken or macaque gene aligned to multiple human/mouse genes.
+In cases where multiple chicken or macaque genes aligned to one human/mouse gene, then we aggregated the counts by sum.
+In cases where there was no corresponding chicken or macaque gene, we filled in zeros.
+After these steps we had, as expected, 17,769 genes.
+
+## Metadata curation
+
+Every study brought into scEiaD was hand curated to identify unique biological samples, published paper ID (where available), organ, tissue (e.g.
+Cornea), source (iPSC based or tissue), single cell platform (e.g.
+10Xv2, DropSeq, etc.).
+We also, where possible, annotated retina region (e.g.
+fovea), sex, and age or developmental stage.
+To import individual cell type labels (e.g.
+Rod, Muller Glia), we wrote custom code for each study that made this table available to link their cell type assignments back to the matching barcoded cell.
+We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels.
+We normalized names for CellType, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).
+We did retain the original published (except for the broader name normalization) labels under SubCellType.
+
+## Differential gene testing
+
+## scEiaD structure and web app optimization
+
+The web app at plae.nei.nih.gov can display gene expression in only a few seconds across over a million cells.
+This speed is only possible due to custom data structures to optimize for data query efficiency.
+Like eyeIntegration's EiaD database, scEiaD is a sqlite database with tables for the core categories.
+This allows the app to initialize in about 30 seconds on a cloud server with memory usage under 8 GB.
+For the gene by cell expression matrix, the data was transformed into a "long" format where there are three columns: Gene, Cell Barcode, and Gene Count Value.
+This allows for cell - gene level queries to be completed, again with minimal memory usage, in under 0.5 seconds.
+To generate the aggregated information (in the Dot Plot, Expression Plot, and tables) we pre-calculated all queries by running a dplyr "group by" operation on all column fields.
+On user given query, the data is further aggregated to level request.
+This allows for complicated queries to complete in seconds rather than minutes.
+
+## Data reproducibility
+
+The codebase for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is XXXXXXX.
+There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which takes the optimal parameters from SnakePOP to build the scEiaD sqlite database for the app.
+
+# Results
+
+## New studies and improvements to the scEiaD database
+
+Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamy_building_2021].
+We updated the scEiaD resource for the PLAE v0.90 web app in six important ways.
+First, we added two chicken retina (*Gallus gallus*) studies.
+Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye.
+Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons.
+Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`).
+Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300 (CHECK THIS), to improve the database quality.
+Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons.
+
+After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (`r fig_cap(name = 'scEiaD_data_overview', display = 'cite')`).
+This resource is only possible due to the publication of `r meta_filter$PMID %>% unique %>% length()` single cell resources that scEiaD draws from (`r supTab_cap(name = 'studies_table', display = 'cite')`).
+To recognize these papers, we make the first data table on the loading page of plae.nei.nih.gov list each of these studies along with a direct link to their citation, where possible.
+
+```{r scEiaD_data_overview, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=scEiaD_data_overview_cap}
+
+num1 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(Tissue) %>% unique() %>% length()
+num2 <- meta_filter %>% filter(Organ != 'Eye') %>% pull(Tissue) %>% unique() %>% length()
+nun3 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique() %>% length()
+scEiaD_data_overview_cap <- fig_cap(name = 'scEiaD_data_overview', caption = glue::glue("Distribution of tissues and cell types in scEiaD.
+A.
+tabulation of the number of cells present across {num1} ocular and {num2} non-ocular tissues, B.
+Number of studies for each ocular tissue.
+C.
+Number of cells present across {num3} curated ocular-derived cell types."))
+
 # organ
@@
 
-# Sup fig 1 
-DecontX *in silico* ambient RNA contamination tool substantially removes Rhodopsin expression in non-rod cells while retaining high expression rods.
-First shown (a) is Rhodopsin expression across the scEiaD v0 dataset, without DecontX optimization.
-Note Rhodopsin expression is noticeable in many cell types beyond the rods.
-Next (b) is the scEiaD v1 dataset with DecontX optimization.
-Rhodopsin is nearly exclusively expressed in labelled rod cells.
-Despite the computational ambient RNA removal, Rhodopsin expression remains high in the Rods.
+## scRNA web resources containing ocular data
 
-```{r, fig.width=10, fig.height=12}
-source('src/make_exp_plot.R')
-scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vGiga}/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite'))
-scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vPLAE}/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20.sqlite'))
+Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources_cap', display = 'cite')`).
+The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are not ocular focused, but do contain some ocular data.
+The PanglaoDB and UCSC are resources which have ingress-ed a large number of published single cell datasets The UCSC Cell Browser has around 62 thousand ocular single cell transcriptomes while the PangloaDB has about 47 thousand.
+However, these resources provide no curation of the metadata and process the data for each study individually.
+Thus each study is effectively silo-ed off from the rest and cross comparisons are impractical.
+The Tabula Sapiens project is an effort to sequence all cells in the mouse and thus the ocular data present, though relatively small (at around 10 thousand), can be more readily cross compared to non-ocular tissues / cells.
+The Protein Atlas resource, which has aggregated protein labels (????) information has recently used one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g.
+one can see whether a gene of interest is enriched in Cones).
+The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has selected a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app.
+PLAE is distinguished from these resources by hand curation of the metadata, re-processing of the data from the raw sequence, and, most crucially, integrating the independent studies to deliver cross study comparisons.
 
-input <- list()
-input$exp_plot_groups <- c('organism')
-input$exp_plot_facet <- c('CellType_predict')
-input$exp_plot_genes <- c('RHO (ENSG00000163914)')
-input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
-input$exp_plot_col_num <- 16
-input$exp_filter_min_cell_number <- 50
+```{r web_resources, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=web_resources_cap}
 
-v0_rho <- make_exp_plot(input, scEiaD_2020_v00, meta_filter_v0)
-v1_rho <-  make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-input$exp_filter_min_cell_number <- 1000
-v1.1_rho <- make_exp_plot(input, scEiaD_2020_v01, meta_filter %>% filter(CellType_predict %in% unique(meta_filter_v0$CellType_predict)))
-plot_grid(v0_rho, v1_rho, ncol = 1, labels = c('a','b'), rel_heights = c(1,1.5))
-```
-<!-- ```{r, fig.height=5, fig.width=3} -->
-<!-- # Distribution of ML Retina Cell Types Between Mouse and Human in Pan-Retina Studies -->
-<!-- Drop this figure? -->
-<!-- Yes -->
-
-<!-- pan_retina_studies <- c('SRP255195','E-MTAB-7316','SRP151023','SRP222001','SRP222958','SRP194595','SRP223254','SRP050054','SRP269634','SRP269635','SRP158528', 'SRP257758','SRP212788') -->
-
-<!-- meta_filter %>%  -->
-<!--   filter(organism %in% c('Mus musculus','Homo sapiens'),  -->
-<!--          Stage == 'Mature',  -->
-<!--          study_accession %in% pan_retina_studies,  -->
-<!--          grepl('Amacr|Bipol|Cone|Epithel|Endoth|Horizon|Muller|Pericy|Ganglion|Rod', CellType_predict)) %>%  -->
-<!--   group_by(study_accession, CellType_predict) %>%  -->
-<!--   summarise(Count = n()) %>% mutate(Percentage = 100 * (Count / sum(Count)))  %>%  -->
-<!--   left_join(meta_filter %>% select(study_accession, Platform, organism) %>% unique()) %>%  -->
-<!--   ggplot(aes(x=organism, y=Percentage)) +  -->
-<!--   geom_boxplot() +  -->
-<!--   coord_flip() +  -->
-<!--   ggbeeswarm::geom_quasirandom(aes(color=study_accession)) +  -->
-<!--   facet_wrap(~CellType_predict, ncol = 1) +  -->
-<!--   cowplot::theme_cowplot() -->
-<!-- ``` -->
-
-# Introduction
-
-## Many published ocular atlases
-
-## Web atlases containing ocular data have substantial limitations
-
-There are several web app resources for querying gene expression across ocular cell types (table 1).
-
-Data difficult to access and cross comparisons impossible
-
-But important as seeing trends trends across different groups strongly demonstrates that 
-the signal is biological, not technical.
-
-## What we did
-
-Collated `r meta_filter$study_accession %>% unique() %>% length()` studies across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues.
-We hand-curated `r meta_filter$CellType %>% unique() %>% length()` cell types, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the remaining cells.
-Extensive quality control metrics were tuned and applied to remove lower quality cells.
-In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values.
-
-# Methods
-
-## scEiaD re-created and pipeline updated
-
-The scEiaD pipeline is structurally the same as in Swamy et al.
-[@swamy_building_2021].
-Very briefly, we collect the raw fastq files and quantify with the kallisto / bustools system [@melsted_barcode_2019; @melsted_modular_2019].
-The quantified counts are merged into one matrix that is used for downstream processing with scvitools to batch correct the data and Seurat for quality control, clustering, and 2D UMAP visualization.
-We use our previous xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamy_building_2021].
-
-The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300 Second, we use the DecontX algorithm [@yang_decontamination_2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination.
-Third, we altered the procedure for automatically aligning gene names across species (see below for details).
-Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process.
-
-Two custom cutoffs for number of unique genes identified were used for the E12 chicken data and the cornea dataset (SRP362101) as an abnormally high number of cells were returned from our pipeline.
-We set the cutoffs for these studies to be 1000 and 800, respectively, as these values resulted in approximately the same number of cells being returned as the authors reported in their papers.
-
-## Cross species gene alignment
-
-As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify, we must align shared genes.
-We found homologs/orthologs by pulling from BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference.
-We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus Musculus*.
-This generates a table linked by Ensembl gene IDs and gene names.
-We then use a full join on non-duplicated Ensembl IDs between mouse and human.
-As we noticed a small number of gene names failed to be linked in this manner we ran another join on the remaining non-aligned genes, using gene name.
-This procedure gave us 17,769 shared human and mouse genes.
-To align the other species we again joined on Ensembl gene ID, removing genes where one chicken or macaque gene aligned to multiple human/mouse genes.
-In cases where multiple chicken or macaque genes aligned to one human/mouse gene, then we aggregated the counts by sum.
-In cases where there was no corresponding chicken or macaque gene, we filled in zeros.
-After these steps we had, as expected, 17,769 genes.
-
-## Metadata curation
-
-Every study brought into scEiaD was hand curated to label biological sample, published paper ID (where available), organ, tissue (e.g.
-Cornea), source (iPSC based or tissue), single cell platform (e.g.
-10Xv1 or DropSeq).
-We also, where possible, annotated retina region (e.g.
-fovea), sex, and age or developmental stage.
-To bring across individual cell type labels (e.g.
-Rod, Muller Glia), we wrote custom code for each study that made this table available to link to the individual cells.
-We curated two types of cell type labels: "CellType" and "PubCellType" where the former is our normalized cell types taken from the published labels.
-In this field we normalized names, for example changin "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).
-We did retain the original published (except for the broader name normalization) labels under PubCellType.
-
-## Differential gene testing
-
-## scEiaD structure and web app optimization
-
-The web app at plae.nei.nih.gov can display million plus cell level gene expression in only a few seconds.
-This speed is only possible due to custom data structures to optimize for data querying.
-Like eyeIntegration's EiaD database, scEiaD is a sqlite database with tables for the core categories.
-This allows for app start up to complete in under 30 seconds on a modestly performant cloud server with memory usage under 8 GB.
-For the gene by cell expression matrix, the data was transformed into a "long" format where there are three columns: Gene, Cell Barcode, and Gene Count Value.
-This allows for cell - gene level queries to be completed, again with minimal memory usage, in under 0.5 seconds.
-To generate the aggregated information (in the Dot Plot, Expression Plot, and tables) we pre-calculated all queries by running a dplyr "group by" operation on all column fields.
-On user given query, the data is further aggregated to level request.
-This allows for complicated queries to complete in seconds rather than minutes.
-
-## Data reproducibility
-
-The codebase for scEiaD is available at github.com/davemcg/scEiaD.
-There are three relevent Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which takes the optimal parameters from SnakePOP to build the scEiaD sqlite database for the app.
-
-# Results
-
-## New studies and improvements to the scEiaD database
-Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamy_building_2021].
-We updated the scEiaD resource for the PLAE v0.90 web app in six important ways.
-First, we added two chicken retina (*Gallus gallus*) studies.
-Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye.
-Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons.
-Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX).
-Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300, which reduces the number of cells that pass QC but improves the quality.
-Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.
-
-After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).
-
-# Comparison of PLAE to "pan" scRNA web apps containing ocular data
-```{r}
+tab_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
 
@@
 
-While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently.
-PLAE is built around the scEiaD database, which runs an extensive multi-stage pipeline to build and identify a high performing scVI-based eye model that is used to integrate all datasets together.
+While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently [@noauthor_spectacle_nodate].
+PLAE is built around the scEiaD database, which runs an extensive multi-stage pipeline to build and identify a high performing scVI-based eye model that is used to integrate all datasets together [@swamy_building_2021].
 This allows for gene queries to be constructed and analyzed across studies.
 In contrast, Spectacle, as of March 2022, is a compilation of study-level datasets and thus you cannot run queries across studies in Spectacle.
-Also, Spectacle contains no non-ocular dataset and the cell types labels have not been harmonized between studies.
+Furthermore, Spectacle contains no non-ocular dataset and the cell types labels have not been harmonized between studies.
 Finally, as this Spectacle uses Seurat objects, the app takes many seconds to load an individual dataset and many more to display the data.
@@
 
-There are two interwoven reasons why PLAE is responsive, despite the huge amount of data it contains.
+There are two interwoven reasons why PLAE is fast, despite the huge amount of data it contains.
 First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables.
 A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage.
-The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous; the scEiaD v2022-03-21 sqlite database is about 368 GB.
-Second, the visualizations in PLAE (e.g.
-the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) along with customized R plotting code to reduce drawing time.
+The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous; the scEiaD v2022-03-22 sqlite database is about 368 GB.
+Second, the visualization code in PLAE (e.g.
+the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) to reduce drawing time.
 Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study (tested on 2022-03-23).
@@
 The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells.
-Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area.
+Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`a).
 
@@
 We also offer a "Dot Plot" view which can display a large number of genes in a space efficient manner.
-This allows a users to see whether a set of genes has related (or not) expression patterns across multiple cell types.
+This allows a users to see whether a set of genes has related (or not) expression patterns across multiple cell types (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`b and c).
 
@@
 The cell types are colored by the relative amounts of expression in the user selected gene.
-Like all visualizations in PLAE, the user can use powerful filtering options to only display data from certain metadata characteristics, like species or publication.
+Like all visualizations in PLAE, the user can use powerful filtering options to only display data from certain metadata characteristics, like species or publication (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`d).
+
+```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
+
+```
 
@@
 ```{r}
-scEiaD_2020_v01 %>% tbl('wilcox_diff_testing') %>% filter(Group == 'CellType (Predict)', FDR < 1e-5, mean_auc > 0.7 ) %>% 
+scEiaD_2020_v01 %>% tbl('wilcox_diff_testing') %>% 
+  filter(Group == 'CellType (Predict)', FDR < 1e-5, mean_auc > 0.7 ) %>% 
+  filter(Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique())) %>% 
   group_by(Base) %>% 
@@
 
+To identify
+
 Sanes proposed X genes (https://www.nature.com/articles/s41598-020-66092-9 --> Figure 9)
@@
 
-scEiaD is built using scvitool's scVI / scANVI batch correction tool.
-As part of the process a computational model of the reference datasets are built.
+scEiaD is built using scvi-tool's scANVI batch correction tool.
+As part of the process a computational transcriptome model of the reference datasets are built.
 This model is only a few mb in size and can be used on a modest computer to project private scRNA datasets onto it.
-This allows an ocular researcher several functions.
-First, the ability to check of quality control of an existing dataset by looking at the distribution of their cells overlapped with our labelled cell types.
+This allows an ocular researcher several important functionalities.
+First, the ability to check the quality control of an existing dataset by looking at the distribution of their cells overlapped with our labelled cell types.
 Second, in perturbation experiments, to identify unusual distributions of cells.
 Third, to leverage our extensive pre-labelled cell information to transfer cell type labels onto private data.
-# Conclusion
+
+We demonstrate some of these possibilities by projecting two organoid datasets from X and Y.
+
+# Conclusions
+
+## PLAE provides quick access to over one million transcriptomes across 35 studies
+
+## Projection of outside data onto scEiaD allows for ... (check lung atlas paper)
+
+##
+
+# Supplementary Figures and Tables
+
+```{r decontX, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=decontX_cap}
+decontX_cap <- supFig_cap(name='decontX', caption = 'DecontX *in silico* ambient RNA contamination tool substantially removes Rhodopsin expression in non-rod cells while retaining high expression rods.
+First shown (a) is Rhodopsin expression across the scEiaD v0 dataset, without DecontX optimization.
+Note Rhodopsin expression is noticeable in many cell types beyond the rods.
+Next (b) is the scEiaD v1 dataset with DecontX optimization.
+Rhodopsin is nearly exclusively expressed in labelled rod cells.
+Despite the computational ambient RNA removal, Rhodopsin expression remains high in the Rods.')
+source('src/make_exp_plot.R')
+
+input <- list()
+input$exp_plot_groups <- c('organism')
+input$exp_plot_facet <- c('CellType_predict')
+input$exp_plot_genes <- c('RHO (ENSG00000163914)')
+input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
+input$exp_plot_col_num <- 16
+input$exp_filter_min_cell_number <- 50
+
+v0_rho <- make_exp_plot(input, scEiaD_2020_v00, meta_filter_v0)
+v1_rho <-  make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+input$exp_filter_min_cell_number <- 1000
+v1.1_rho <- make_exp_plot(input, scEiaD_2020_v01, meta_filter %>% filter(CellType_predict %in% unique(meta_filter_v0$CellType_predict)))
+plot_grid(v0_rho, v1_rho, ncol = 1, labels = c('a','b'), rel_heights = c(1,1.5))
+```
+
+```{r studies_table, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=studies_table_cap}
+studies_table_cap <- supTab_cap(name = 'studies_table', caption = 'Studies, tissues, and number of cells across scEiaD')
+
+meta_filter %>% group_by(Citation, PMID, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable()
+```

### 2022-06-02 manuscript_00.Rmd
new diff tables

@@
 scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vGiga}/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite'))
-scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vPLAE}/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20.sqlite'))
+scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vPLAE}/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20__pointRelease01.sqlite'))
 
@@
 num2 <- meta_filter %>% filter(Organ != 'Eye') %>% pull(Tissue) %>% unique() %>% length()
-nun3 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique() %>% length()
+num3 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique() %>% length()
 scEiaD_data_overview_cap <- fig_cap(name = 'scEiaD_data_overview', caption = glue::glue("Distribution of tissues and cell types in scEiaD.
@@
 ```{r}
-scEiaD_2020_v01 %>% tbl('wilcox_diff_testing') %>% 
-  filter(Group == 'CellType (Predict)', FDR < 1e-5, mean_auc > 0.7 ) %>% 
-  filter(Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique())) %>% 
-  group_by(Base) %>% 
+diff_tab <- scEiaD_2020_v01 %>% tbl('diff_testing') %>% 
+  filter(Group == 'CellType (Predict)', padj < 1e-5, (log2FoldChange) > 2) %>% 
+  collect() %>% 
+  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique()))
+
+diff_tab %>% 
+  group_by(Base, Organism) %>% 
   summarise(Count = n()) %>% 
   rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
+  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
   as_tibble() %>% 
   flextable()
+```
+
+```{r}
+# top 100 genes in each organism/base category
+diff_tab %>%
+
+  group_by(Base, Organism) %>% 
+  top_n(., 100, -padj) %>% 
+  arrange(Base, Organism) %>% 
+  relocate(Base, Organism) %>% 
+  select(-Against, -Group)
+```
+
+```{r}
+# consistently diff genes
+consistent <- diff_tab %>% 
+  group_by(Gene, Base)  %>% 
+  summarise(Count = n()) %>% 
+  filter(Count > 1) %>% 
+  mutate(k = paste(Gene, Base, sep = '_'))
+
+diff_tab %>% mutate(k = paste(Gene, Base, sep = '_')) %>% 
+  filter(k %in% consistent$k) %>% 
+  group_by(Base, Organism) %>% 
+  summarise(Count = n()) %>% 
+  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
+  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
+  as_tibble() %>% 
+  flextable()
+
+```
+
+```{r}
+# flip flop diff genes
+down_tab <- scEiaD_2020_v01 %>% tbl('diff_testing') %>% 
+  filter(Group == 'CellType (Predict)', padj < 1e-5, (log2FoldChange) < -2) %>% 
+  collect() %>% 
+  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique()))
+
+diff_tab %>% mutate(k = paste(Gene, Base, sep = '_')) %>% 
+  filter(k %in% consistent$k) %>% 
+  group_by(Base, Organism) %>% 
+  summarise(Count = n()) %>% 
+  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
+  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
+  as_tibble() %>% 
+  flextable()
+
 ```

### 2022-06-02 manuscript_00.Rmd
diff updates

@@
 library(glue)
-data_dir_vGiga <- '/Volumes/McGaughey_S/scEiaD/'
-data_dir_vPLAE <- '/Volumes/McGaughey_S/scEiaD_2022_02/'
+data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 
@@
   collect() %>% 
-  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique()))
+  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique())) %>% mutate(k = paste(Gene, Base, sep = '_'))
 
 diff_tab %>% mutate(k = paste(Gene, Base, sep = '_')) %>% 
-  filter(k %in% consistent$k) %>% 
-  group_by(Base, Organism) %>% 
-  summarise(Count = n()) %>% 
-  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
-  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
-  as_tibble() %>% 
-  flextable()
+  filter(k %in% down_tab$k, Base %in% c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )) %>%
+  arrange(padj)
+```
 
+```{r diff_difff_exp_genes, fig.height = 3, fig.width = 8, echo = F, message = F}
+input <- list()
+input$exp_plot_groups <- c('organism')
+input$exp_plot_facet <- c('CellType_predict')
+input$exp_plot_genes <- c('GNAT2 (ENSG00000134183)')
+input$exp_filter_cat <- 'CellType_predict'
+input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
+input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
+input$exp_plot_col_num <- 16
+input$exp_filter_min_cell_number <- 50
+
+make_exp_plot(input, scEiaD_2020_v01, meta_filter)
 ```

### 2022-06-03 manuscript_00.Rmd
text updates

--- 2022-06-02 manuscript_00.Rmd
@@
 library(glue)
-data_dir_vGiga <- '/Volumes/McGaughey_S/scEiaD/'
-data_dir_vPLAE <- '/Volumes/McGaughey_S/scEiaD_2022_02/'
+data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
+data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
 
@@
 
-## Differential testing identifies thousands genes frequently enriched across various ocular cell types 
+## Differential testing identifies thousands genes enriched across various ocular cell types
+
+We used a "pseudo-bulk" based approach wherein the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping.
+The creates a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices.
+This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021].
+Once the aggregated matrix was constructured, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
+We tested each species (Human, Macaque, Mouse) separately.
+We had an insufficient number of chick studies to conduct differential testing.
+The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential expression.
+After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g.
+cones versus all remaining cell types) or in a pair-wise manner (cone against rod).
+We fin
 ```{r}
-diff_tab <- scEiaD_2020_v01 %>% tbl('diff_testing') %>% 
+diff_tab_big <- scEiaD_2020_v01 %>% tbl('diff_testing') %>% 
   filter(Group == 'CellType (Predict)', padj < 1e-5, (log2FoldChange) > 2) %>% 
-  collect() %>% 
-  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique()))
+  collect()
+
+diff_tab <- diff_tab_big %>% 
+  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique()))
 
@@
   flextable()
+```
+
+Many of these genes can be differentially expressed in more than one cell type.
+To identify genes that consistently differentially expressed we can use a more specific test.
+We define a consistent gene that is
+
+```{r}
+diff_tab_big %>% 
+  filter(Gene %in% diff_tab$Gene) %>% # only retain diff genes across "All"
+  group_by(Gene, Base, Organism) %>% 
+  summarise(Count = n()) %>% 
+  filter(Count > 1) %>% arrange(-Count)
 ```
@@
   collect() %>% 
-  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique()))
+  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique())) %>% mutate(k = paste(Gene, Base, sep = '_'))
 
 diff_tab %>% mutate(k = paste(Gene, Base, sep = '_')) %>% 
-  filter(k %in% consistent$k) %>% 
-  group_by(Base, Organism) %>% 
+  filter(k %in% down_tab$k, Base %in% c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )) %>%
+  arrange(padj)
+
+diff_tab_big %>% 
+    filter(Gene %in% diff_tab$Gene) %>% # only retain diff genes across "All"
+    group_by(Gene, Base, Organism) %>% 
     summarise(Count = n()) %>% 
-  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
-  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
-  as_tibble() %>% 
-  flextable()
+    filter(Count > 8, Base == 'Rod') %>% arrange(-Count)
+```
 
+```{r diff_difff_exp_genes, fig.height = 3, fig.width = 8, echo = F, message = F}
+source('src/make_exp_plot.R')
+
+input <- list()
+input$exp_plot_groups <- c('organism')
+input$exp_plot_facet <- c('CellType_predict')
+input$exp_plot_genes <- c('SCRG1 (ENSG00000164106)')
+input$exp_filter_cat <- 'CellType_predict'
+input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
+input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
+input$exp_plot_col_num <- 16
+input$exp_filter_min_cell_number <- 50
+
+make_exp_plot(input, scEiaD_2020_v01, meta_filter)
 ```

### 2022-06-06 manuscript_00.Rmd
diff table and gene symbol fixes

@@
 library(glue)
-data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
-data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-
+# data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
+# data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 # setup caption-ing
@@
 ```{r}
+# fix  occasional NA
+# data made from src/ensdb.R
+hs_v86_tab <- read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz')
+mm_v79_tab <- read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz')
+hs_mm <- bind_rows(hs_v86_tab, mm_v79_tab)
+
 diff_tab_big <- scEiaD_2020_v01 %>% tbl('diff_testing') %>% 
@@
   collect()
+
+diff_tab_big2 <- diff_tab_big %>% 
+  separate(Gene, into = c('Symbol', 'ID'), sep = ' ') %>% 
+  mutate(ID = gsub('\\(|\\)', '', ID)) %>% 
+  filter(Symbol == '-' | Symbol == 'NA')
 
@@
 Many of these genes can be differentially expressed in more than one cell type.
-To identify genes that consistently differentially expressed we can use a more specific test.
+To identify genes that are consistently differentially expressed we can use a more specific test.
 We define a consistent gene that is
@@
 ```{r}
+
+spec_diff <- diff_tab_big %>% 
+    filter(Gene %in% diff_tab$Gene) %>% # only retain diff genes across "All"
+    group_by(Gene, Organism, Base) %>% 
+    summarise(Count = n()) %>% 
+    filter(Count > 1) %>% summarise(Count = n()) %>% filter(Count < 3) %>% mutate(go = paste(Gene, '___', Organism, sep = '')) %>% pull(go)
+
 diff_tab_big %>% 
     filter(Gene %in% diff_tab$Gene) %>% # only retain diff genes across "All"
-  group_by(Gene, Base, Organism) %>% 
+    group_by(Gene, Organism, Base) %>% 
     summarise(Count = n()) %>% 
-  filter(Count > 1) %>% arrange(-Count)
+    filter(Count > 1) %>% mutate(go = paste(Gene, '___', Organism, sep = '')) %>% filter(go %in% (spec_diff))  %>% select(-go) %>% ungroup() %>% group_by(Organism, Base) %>% slice_max(n = 10, order_by = Count) %>% filter(Base == 'Rod')
 ```

### 2022-06-06 src/ensdb.R

@@ -1,0 +1,15 @@
+# fix  occasional NA
+library(org.Hs.eg.db)
+library(EnsDb.Hsapiens.v86)
+ensembl_id <- keys(EnsDb.Hsapiens.v86, keytype = "GENEID")
+ens_sym_table <- ensembldb::select(EnsDb.Hsapiens.v86, keys= ensembl_id, keytype = "GENEID", columns = c("SYMBOL","GENEID"))
+write_tsv(ens_sym_table, file = 'data/hs.ensdb.v86.symbol.geneid.tsv.gz')
+
+
+
+
+library(org.Mm.eg.db)
+library(EnsDb.Mmusculus.v79)
+ensembl_id <- keys(EnsDb.Mmusculus.v79, keytype = "GENEID")
+ens_sym_table <- ensembldb::select(EnsDb.Mmusculus.v79, keys= ensembl_id, keytype = "GENEID", columns = c("SYMBOL","GENEID"))
+write_tsv(ens_sym_table, file = 'data/mm.ensdb.v79.symbol.geneid.tsv.gz')

### 2022-06-06 src/make_exp_plot.R

@@ -95,7 +95,7 @@ make_exp_plot <- function(input, db, meta_filter){
   box_data %>%
     ggplot(aes(x=Gene, y = !!as.symbol(input$exp_plot_ylab), color = !!as.symbol(grouping_features))) +
     geom_boxplot(color = 'black', outlier.shape = NA) +
-    ggbeeswarm::geom_quasirandom(aes(size = `Total Cells`), groupOnX = TRUE) +
+    ggbeeswarm::geom_quasirandom(aes(size = 8), groupOnX = TRUE) +
     cowplot::theme_cowplot() +
     theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
     scale_radius(range=c(2, 6)) +

### 2022-06-07 manuscript_00.Rmd
replace gene

@@
   filter(Group == 'CellType (Predict)', padj < 1e-5, (log2FoldChange) > 2) %>% 
-  collect()
+  collect() %>% 
+  separate(Gene, into = c('Symbol', 'ID'), sep = ' ') %>% 
+  mutate(ID = gsub('\\(|\\)', '', ID))
 
 diff_tab_big2 <- diff_tab_big %>% 
-  separate(Gene, into = c('Symbol', 'ID'), sep = ' ') %>% 
-  mutate(ID = gsub('\\(|\\)', '', ID)) %>% 
   filter(Symbol == '-' | Symbol == 'NA')
+
+diff_tab_big2 <- 
+  diff_tab_big2 %>% 
+  left_join(hs_mm, by = c('ID' = 'GENEID')) %>% 
+  select(-Symbol) %>% 
+  rename(Symbol = SYMBOL) %>% 
+  relocate(Symbol)
+
+diff_tab_big <- bind_rows(
+  diff_tab_big %>% filter(!Symbol %in% c('-', 'NA')),
+  diff_tab_big2
+)
 
@@
 ```{r}
-
 spec_diff <- diff_tab_big %>% 
-    filter(Gene %in% diff_tab$Gene) %>% # only retain diff genes across "All"
-    group_by(Gene, Organism, Base) %>% 
+  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
+  group_by(Symbol, Organism, Base) %>% 
   summarise(Count = n()) %>% 
-    filter(Count > 1) %>% summarise(Count = n()) %>% filter(Count < 3) %>% mutate(go = paste(Gene, '___', Organism, sep = '')) %>% pull(go)
+  filter(Count > 1) %>% summarise(Count = n()) %>% filter(Count < 3) %>% mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% pull(go)
 
 diff_tab_big %>% 
-    filter(Gene %in% diff_tab$Gene) %>% # only retain diff genes across "All"
-    group_by(Gene, Organism, Base) %>% 
+  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
+  group_by(Symbol, Organism, Base) %>% 
   summarise(Count = n()) %>% 
-    filter(Count > 1) %>% mutate(go = paste(Gene, '___', Organism, sep = '')) %>% filter(go %in% (spec_diff))  %>% select(-go) %>% ungroup() %>% group_by(Organism, Base) %>% slice_max(n = 10, order_by = Count) %>% filter(Base == 'Rod')
+  filter(Count > 1) %>% mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% 
+  filter(go %in% (spec_diff))  %>% 
+  select(-go) %>% 
+  ungroup() %>% 
+  group_by(Organism, Base) %>% 
+  slice_max(n = 10, order_by = Count) %>% 
+  filter(Base == 'Rod')
 ```
@@
 consistent <- diff_tab %>% 
-  group_by(Gene, Base)  %>% 
+  group_by(Symbol, Base)  %>% 
   summarise(Count = n()) %>% 
   filter(Count > 1) %>% 
-  mutate(k = paste(Gene, Base, sep = '_'))
+  mutate(k = paste(Symbol, Base, sep = '_'))
 
-diff_tab %>% mutate(k = paste(Gene, Base, sep = '_')) %>% 
+diff_tab %>% mutate(k = paste(Symbol, Base, sep = '_')) %>% 
   filter(k %in% consistent$k) %>% 
@@
   collect() %>% 
-  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique())) %>% mutate(k = paste(Gene, Base, sep = '_'))
+  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique())) %>% mutate(k = paste(Symbol, Base, sep = '_'))
 
-diff_tab %>% mutate(k = paste(Gene, Base, sep = '_')) %>% 
+diff_tab %>% mutate(k = paste(Symbol, Base, sep = '_')) %>% 
   filter(k %in% down_tab$k, Base %in% c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )) %>%
@@
 diff_tab_big %>% 
-    filter(Gene %in% diff_tab$Gene) %>% # only retain diff genes across "All"
-    group_by(Gene, Base, Organism) %>% 
+  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
+  group_by(Symbol, Base, Organism) %>% 
   summarise(Count = n()) %>% 

### 2022-06-08 manuscript_00.Rmd
more diff tweaking

@@
   summarise(Count = n()) %>% 
-  filter(Count > 1) %>% summarise(Count = n()) %>% filter(Count < 3) %>% mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% pull(go)
+  summarise(Count = n())
+
+spec_diff_go <- spec_diff %>% 
+  filter(Count <= 3) %>% 
+  mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% 
+  pull(go)
 
@@
   group_by(Symbol, Organism, Base) %>% 
-  summarise(Count = n()) %>% 
-  filter(Count > 1) %>% mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% 
-  filter(go %in% (spec_diff))  %>% 
-  select(-go) %>% 
+  summarise(Count = n(),
+            baseMean = mean(baseMean),
+            log2FoldChange = mean(log2FoldChange),
+            padj = mean(padj)) %>% 
+  filter(Count > 1) %>% 
+  left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
+  mutate(Ratio = `Count.x` / `Count.y`) %>% 
+  select(-`Count.x`,-`Count.y`) %>% 
+  #mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% 
+  #filter(go %in% (spec_diff_go))  %>% 
+  #select(-go) %>% 
   ungroup() %>% 
   group_by(Organism, Base) %>% 
-  slice_max(n = 10, order_by = Count) %>% 
-  filter(Base == 'Rod')
+  slice_max(n = 10, order_by = log2FoldChange)
+
+```
+
+### heatmap
+```{r, fig.width=5, fig.height=20}
+diff_ratios <- diff_tab_big %>% 
+    filter(Against != 'All') %>% 
+    group_by(Symbol, ID, Organism, Base) %>% 
+    summarise(Count = n()) %>% 
+    filter(Count > 1) %>% 
+    left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
+    mutate(Ratio = `Count.x` / `Count.y`) %>% 
+    select(Symbol:Base, Ratio)
+
+gene_sets <- diff_tab %>% 
+  left_join(diff_ratios, by = c('Symbol', 'ID','Base','Organism')) %>% 
+  filter(Ratio > 0.2) %>% group_by(Base, Organism) %>% 
+  slice_max(n=20, order_by = log2FoldChange)
+
+g <- gene_sets %>% filter(Organism == 'Homo sapiens') %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
+
+x <- scEiaD_2020_v01 %>% tbl("diff_testing") %>% filter(Gene %in% g) %>% collect() %>% filter(Group == 'CellType (Predict)', Organism == 'Homo sapiens', Against == 'All') %>% select(Gene, log2FoldChange, Base) %>% pivot_wider(values_from=log2FoldChange, names_from = Base)
+celltype <- colnames(x[,-1])
+x <- x %>% data.frame()
+row.names(x) <- x$Gene
+x <- x[,-1]
+Heatmap(x, column_labels = celltype)
 ```

### 2022-06-09 manuscript_00.Rmd
text added

@@
 ## Differential gene testing
+
+The scater::aggregateAcrossCells function was used to aggregate gene counts across species, study accession, and [category] where category was either Cluster, CellType, or CellType (Predict) assignments for a cell.
+If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test.
+Any genes with a sum of 0 counts after aggregation were removed.
+The aggregated matrix was imported into DESeq2 using the DESeqDataSetFromMatrix function with the design given as "~study_accession + [category]".
+Contrasts were extracted either with the target category (e.g.
+"Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function.
+Genes were considered to be differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2.
+To enhance the specificity of the
 
@@
 
-We used a "pseudo-bulk" based approach wherein the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping.
+We used a "pseudo bulk" based approach wherein the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping.
 The creates a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices.
 This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021].
-Once the aggregated matrix was constructured, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
+Once the aggregated matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
 We tested each species (Human, Macaque, Mouse) separately.
-We had an insufficient number of chick studies to conduct differential testing.
+We had an insufficient number of chicken studies to conduct differential testing.
 The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential expression.

### 2022-06-10 manuscript_00.Rmd
output diff table for speed

@@
 # data made from src/ensdb.R
-hs_v86_tab <- read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz')
-mm_v79_tab <- read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz')
-hs_mm <- bind_rows(hs_v86_tab, mm_v79_tab)
-
-diff_tab_big <- scEiaD_2020_v01 %>% tbl('diff_testing') %>% 
-  filter(Group == 'CellType (Predict)', padj < 1e-5, (log2FoldChange) > 2) %>% 
-  collect() %>% 
-  separate(Gene, into = c('Symbol', 'ID'), sep = ' ') %>% 
-  mutate(ID = gsub('\\(|\\)', '', ID))
-
-diff_tab_big2 <- diff_tab_big %>% 
-  filter(Symbol == '-' | Symbol == 'NA')
-
-diff_tab_big2 <- 
-  diff_tab_big2 %>% 
-  left_join(hs_mm, by = c('ID' = 'GENEID')) %>% 
-  select(-Symbol) %>% 
-  rename(Symbol = SYMBOL) %>% 
-  relocate(Symbol)
-
-diff_tab_big <- bind_rows(
-  diff_tab_big %>% filter(!Symbol %in% c('-', 'NA')),
-  diff_tab_big2
-)
+diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
 
@@
   as_tibble() %>% 
+  flextable()
+
+# number of celltype (predict) - organism - study counts (with >=50 cells)
+meta_filter %>%  
+  filter(Organ == 'Eye') %>% 
+  group_by(organism, CellType_predict, study_accession) %>% 
+  count() %>% 
+  filter(n >= 50) %>% 
+  ungroup() %>% 
+  group_by(organism, CellType_predict) %>% 
+  count() %>% 
+  pivot_wider(values_from = n, names_from = organism) %>% 
+  filter(!is.na(CellType_predict)) %>% 
+  replace(is.na(.), values = 0) %>% 
   flextable()
@@
   filter(Ratio > 0.2) %>% group_by(Base, Organism) %>% 
-  slice_max(n=20, order_by = log2FoldChange)
+  slice_max(n=5, order_by = log2FoldChange)
 

### 2022-06-10 manuscript_00.Rmd
more diff adds

@@
 library(glue)
-# data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
-# data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
+data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 # setup caption-ing
@@
   as_tibble() %>% 
+  rename(`CellType (Predict)` = CellType_predict) %>% 
   flextable()
@@
   replace(is.na(.), values = 0) %>% 
+  rename(`CellType (Predict)` = CellType_predict) %>% 
   flextable()
@@
 
-### heatmap
+### heatmap human
 ```{r, fig.width=5, fig.height=20}
@@
 
-x <- scEiaD_2020_v01 %>% tbl("diff_testing") %>% filter(Gene %in% g) %>% collect() %>% filter(Group == 'CellType (Predict)', Organism == 'Homo sapiens', Against == 'All') %>% select(Gene, log2FoldChange, Base) %>% pivot_wider(values_from=log2FoldChange, names_from = Base)
+# only uses CT that exist in the organism <-> CellType 
+ct <- meta_filter %>% filter(organism == 'Homo sapiens') %>% pull(CellType) %>% unique()
+
+x <- scEiaD_2020_v01 %>% tbl("diff_testing") %>% filter(Gene %in% g, Base %in% ct) %>% collect() %>% filter(Group == 'CellType (Predict)', Organism == 'Homo sapiens', Against == 'All') %>% select(Gene, log2FoldChange, Base) %>% pivot_wider(values_from=log2FoldChange, names_from = Base)
+celltype <- colnames(x[,-1])
+x <- x %>% data.frame()
+row.names(x) <- x$Gene
+x <- x[,-1]
+Heatmap(x, column_labels = celltype)
+```
+
+### heatmap mouse
+```{r, fig.width=5, fig.height=20}
+
+g <- gene_sets %>% filter(Organism == 'Mus musculus') %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
+
+x <- scEiaD_2020_v01 %>% tbl("diff_testing") %>% filter(Gene %in% g) %>% collect() %>% filter(Group == 'CellType (Predict)', Organism == 'Mus musculus', Against == 'All') %>% select(Gene, log2FoldChange, Base) %>% pivot_wider(values_from=log2FoldChange, names_from = Base)
 celltype <- colnames(x[,-1])

### 2022-06-16 Untitled.Rmd
new files

--- 2022-06-10 manuscript_00.Rmd
@@
 ---
-title: 'PLAE enables quick and easy searching across one million ocular transcriptomes'
-author:
-  - Vinay Swamy:
-      institute: bg
-  - Zachary Batz:
-      institute: nnrl
-  - David McGaughey:
-      institute:
-        - bg
-      correspondence: "yes"
-      email: mcgaugheyd@mail.nih.gov
-institute:
-  - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
-  - nnrl: Neurobiology, Neurodegeneration & Repair Laboratory, National Eye Institute, National Institutes of Health
-  - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health
-
-date: '`r format(Sys.Date(), "%B %d, %Y")`'
-output:
-  word_document:
-    reference_docx: word-styles-reference-01.docx
-    fig_caption: yes
-    pandoc_args:
-      - '--lua-filter=scholarly-metadata.lua'
-      - '--lua-filter=author-info-blocks.lua'
-bibliography: zotero_plaeApp.bib
-csl: investigative-ophthalmology-and-visual-science.csl
-abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors.
-METHODS: We updated the content and structure of the underlying data (sc Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it.
-RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes, XX cell types, XX ocular tissues, XX body tissues.
-As a test of the value of this unified pan-eye dataset, we show XX.
-CONCLUSION: The PLAE v0.90 web app serves the pan-ocular and body dataset, scEiaD.
-This offers the eye community a powerful and quick means to test hypotheses on gene and transcript expression across 54 body and 19 eye tissues.
-"
-
-keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app"
+title: "R Notebook"
+output: html_notebook
 ---
@@
                             TRUE ~ Tissue),
-         Tissue = case_when(grepl('RPE|Choroid', Tissue) ~ 'RPE-Choroid',
+         Tissue = case_when(grepl('^RPE|^Choroid', Tissue) ~ 'RPE-Choroid',
                             TRUE ~ Tissue),
@@
 The codebase for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is XXXXXXX.
-There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which takes the optimal parameters from SnakePOP to build the scEiaD sqlite database for the app.
+There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app.
 
@@
 Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamy_building_2021].
-We updated the scEiaD resource for the PLAE v0.90 web app in six important ways.
-First, we added two chicken retina (*Gallus gallus*) studies.
+We updated the scEiaD resource for the PLAE v0.91 web app in six important ways.
+First, we added a large chicken retina (*Gallus gallus*) studies.
 Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye.
@@
 
-```{r scEiaD_data_overview, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=scEiaD_data_overview_cap}
+```{r scEiaD_data_overview, fig.height = 7, fig.width = 5, echo = F, message = F, fig.cap=scEiaD_data_overview_cap}
 
@@
 ```
-
-## scRNA web resources containing ocular data
-
-Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources_cap', display = 'cite')`).
-The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are not ocular focused, but do contain some ocular data.
-The PanglaoDB and UCSC are resources which have ingress-ed a large number of published single cell datasets The UCSC Cell Browser has around 62 thousand ocular single cell transcriptomes while the PangloaDB has about 47 thousand.
-However, these resources provide no curation of the metadata and process the data for each study individually.
-Thus each study is effectively silo-ed off from the rest and cross comparisons are impractical.
-The Tabula Sapiens project is an effort to sequence all cells in the mouse and thus the ocular data present, though relatively small (at around 10 thousand), can be more readily cross compared to non-ocular tissues / cells.
-The Protein Atlas resource, which has aggregated protein labels (????) information has recently used one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g.
-one can see whether a gene of interest is enriched in Cones).
-The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has selected a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app.
-PLAE is distinguished from these resources by hand curation of the metadata, re-processing of the data from the raw sequence, and, most crucially, integrating the independent studies to deliver cross study comparisons.
-
-```{r web_resources, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=web_resources_cap}
-
-tab_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
-
-feature_table <- rbind(
-  c('PLAE','Yes','Yes', 'Yes', '1136 ', 'https://plae.nei.nih.gov'),
-  c('Spectacle', 'Yes' ,'No', 'No', '1800', 'https://singlecell-eye.org'),
-  c('UCSC Cell Browser', 'Yes', 'No', 'Yes', '62', 'https://cells.ucsc.edu' ),
-  c('Tabula Sapiens', 'No', 'N/A', 'Yes', '10', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
-  c('PanglaoDB', 'Yes', 'No', 'Yes', '47', 'https://panglaodb.se'),
-  c('Protein Atlas', 'Yes*','No', 'Yes', '5',  'https://www.proteinatlas.org/humanproteome/celltype')) %>% 
-  as_tibble()
-colnames(feature_table) <- c('Resource Name', 'Independent Datasets', 'Harmonization', 'Ocular and non-Ocular Data', 'Ocular Cell Count\n(thousands)', 'URL')
-feature_table %>% flextable()
-# as of 2021 10 25
-```
-
-## PLAE differs from a complementary web app, Spectacle in several important ways
-
-While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently [@noauthor_spectacle_nodate].
-PLAE is built around the scEiaD database, which runs an extensive multi-stage pipeline to build and identify a high performing scVI-based eye model that is used to integrate all datasets together [@swamy_building_2021].
-This allows for gene queries to be constructed and analyzed across studies.
-In contrast, Spectacle, as of March 2022, is a compilation of study-level datasets and thus you cannot run queries across studies in Spectacle.
-Furthermore, Spectacle contains no non-ocular dataset and the cell types labels have not been harmonized between studies.
-Finally, as this Spectacle uses Seurat objects, the app takes many seconds to load an individual dataset and many more to display the data.
-
-## Extensive optimization makes the PLAE web app highly responsive despite containing huge amounts of data
-
-There are two interwoven reasons why PLAE is fast, despite the huge amount of data it contains.
-First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables.
-A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage.
-The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous; the scEiaD v2022-03-22 sqlite database is about 368 GB.
-Second, the visualization code in PLAE (e.g.
-the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) to reduce drawing time.
-Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study (tested on 2022-03-23).
-In the same amount of time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies.
-
-## Rich web-available visualizations
-
-PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values.
-So thus a user can see Crx expression across over a million cells and separately see cell type labels for those same million cells.
-The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells.
-Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`a).
-
-For users interested in seeing how expression of a gene relates to study or other covariates, the "Exp Plot" view displays expression by cell type, predicted cell type, or cluster.
-The values can then be split and colored by multiple covariates like study or organism.
-This allows a user to see whether a gene pattern is consistent across studies or species.
-We also offer a "Dot Plot" view which can display a large number of genes in a space efficient manner.
-This allows a users to see whether a set of genes has related (or not) expression patterns across multiple cell types (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`b and c).
-
-While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina.
-The structure of the cartoon retina was based on the human retina.
-This unique visualization displays the major cell types of the retina and the RPE / choroid layer behind it.
-The cell types are colored by the relative amounts of expression in the user selected gene.
-Like all visualizations in PLAE, the user can use powerful filtering options to only display data from certain metadata characteristics, like species or publication (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`d).
-
-```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
-
-```
-
-## Differential testing identifies thousands genes enriched across various ocular cell types
-
-We used a "pseudo bulk" based approach wherein the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping.
-The creates a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices.
-This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021].
-Once the aggregated matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
-We tested each species (Human, Macaque, Mouse) separately.
-We had an insufficient number of chicken studies to conduct differential testing.
-The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential expression.
-After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g.
-cones versus all remaining cell types) or in a pair-wise manner (cone against rod).
-We fin
-```{r}
-# fix  occasional NA
-# data made from src/ensdb.R
-diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
-
-diff_tab <- diff_tab_big %>% 
-  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique()))
-
-diff_tab %>% 
-  group_by(Base, Organism) %>% 
-  summarise(Count = n()) %>% 
-  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
-  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
-  as_tibble() %>% 
-  rename(`CellType (Predict)` = CellType_predict) %>% 
-  flextable()
-
-# number of celltype (predict) - organism - study counts (with >=50 cells)
-meta_filter %>%  
-  filter(Organ == 'Eye') %>% 
-  group_by(organism, CellType_predict, study_accession) %>% 
-  count() %>% 
-  filter(n >= 50) %>% 
-  ungroup() %>% 
-  group_by(organism, CellType_predict) %>% 
-  count() %>% 
-  pivot_wider(values_from = n, names_from = organism) %>% 
-  filter(!is.na(CellType_predict)) %>% 
-  replace(is.na(.), values = 0) %>% 
-  rename(`CellType (Predict)` = CellType_predict) %>% 
-  flextable()
-```
-
-Many of these genes can be differentially expressed in more than one cell type.
-To identify genes that are consistently differentially expressed we can use a more specific test.
-We define a consistent gene that is
-
-```{r}
-spec_diff <- diff_tab_big %>% 
-  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
-  group_by(Symbol, Organism, Base) %>% 
-  summarise(Count = n()) %>% 
-  summarise(Count = n())
-
-spec_diff_go <- spec_diff %>% 
-  filter(Count <= 3) %>% 
-  mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% 
-  pull(go)
-
-diff_tab_big %>% 
-  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
-  group_by(Symbol, Organism, Base) %>% 
-  summarise(Count = n(),
-            baseMean = mean(baseMean),
-            log2FoldChange = mean(log2FoldChange),
-            padj = mean(padj)) %>% 
-  filter(Count > 1) %>% 
-  left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
-  mutate(Ratio = `Count.x` / `Count.y`) %>% 
-  select(-`Count.x`,-`Count.y`) %>% 
-  #mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% 
-  #filter(go %in% (spec_diff_go))  %>% 
-  #select(-go) %>% 
-  ungroup() %>% 
-  group_by(Organism, Base) %>% 
-  slice_max(n = 10, order_by = log2FoldChange)
-
-```
-
-### heatmap human
-```{r, fig.width=5, fig.height=20}
-diff_ratios <- diff_tab_big %>% 
-    filter(Against != 'All') %>% 
-    group_by(Symbol, ID, Organism, Base) %>% 
-    summarise(Count = n()) %>% 
-    filter(Count > 1) %>% 
-    left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
-    mutate(Ratio = `Count.x` / `Count.y`) %>% 
-    select(Symbol:Base, Ratio)
-
-gene_sets <- diff_tab %>% 
-  left_join(diff_ratios, by = c('Symbol', 'ID','Base','Organism')) %>% 
-  filter(Ratio > 0.2) %>% group_by(Base, Organism) %>% 
-  slice_max(n=5, order_by = log2FoldChange)
-
-g <- gene_sets %>% filter(Organism == 'Homo sapiens') %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
-
-# only uses CT that exist in the organism <-> CellType 
-ct <- meta_filter %>% filter(organism == 'Homo sapiens') %>% pull(CellType) %>% unique()
-
-x <- scEiaD_2020_v01 %>% tbl("diff_testing") %>% filter(Gene %in% g, Base %in% ct) %>% collect() %>% filter(Group == 'CellType (Predict)', Organism == 'Homo sapiens', Against == 'All') %>% select(Gene, log2FoldChange, Base) %>% pivot_wider(values_from=log2FoldChange, names_from = Base)
-celltype <- colnames(x[,-1])
-x <- x %>% data.frame()
-row.names(x) <- x$Gene
-x <- x[,-1]
-Heatmap(x, column_labels = celltype)
-```
-
-### heatmap mouse
-```{r, fig.width=5, fig.height=20}
-
-g <- gene_sets %>% filter(Organism == 'Mus musculus') %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
-
-x <- scEiaD_2020_v01 %>% tbl("diff_testing") %>% filter(Gene %in% g) %>% collect() %>% filter(Group == 'CellType (Predict)', Organism == 'Mus musculus', Against == 'All') %>% select(Gene, log2FoldChange, Base) %>% pivot_wider(values_from=log2FoldChange, names_from = Base)
-celltype <- colnames(x[,-1])
-x <- x %>% data.frame()
-row.names(x) <- x$Gene
-x <- x[,-1]
-Heatmap(x, column_labels = celltype)
-```
-
-```{r}
-# top 100 genes in each organism/base category
-diff_tab %>%
-
-  group_by(Base, Organism) %>% 
-  top_n(., 100, -padj) %>% 
-  arrange(Base, Organism) %>% 
-  relocate(Base, Organism) %>% 
-  select(-Against, -Group)
-```
-
-```{r}
-# consistently diff genes
-consistent <- diff_tab %>% 
-  group_by(Symbol, Base)  %>% 
-  summarise(Count = n()) %>% 
-  filter(Count > 1) %>% 
-  mutate(k = paste(Symbol, Base, sep = '_'))
-
-diff_tab %>% mutate(k = paste(Symbol, Base, sep = '_')) %>% 
-  filter(k %in% consistent$k) %>% 
-  group_by(Base, Organism) %>% 
-  summarise(Count = n()) %>% 
-  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
-  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
-  as_tibble() %>% 
-  flextable()
-
-```
-
-```{r}
-# flip flop diff genes
-down_tab <- scEiaD_2020_v01 %>% tbl('diff_testing') %>% 
-  filter(Group == 'CellType (Predict)', padj < 1e-5, (log2FoldChange) < -2) %>% 
-  collect() %>% 
-  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique())) %>% mutate(k = paste(Symbol, Base, sep = '_'))
-
-diff_tab %>% mutate(k = paste(Symbol, Base, sep = '_')) %>% 
-  filter(k %in% down_tab$k, Base %in% c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )) %>%
-  arrange(padj)
-
-diff_tab_big %>% 
-  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
-  group_by(Symbol, Base, Organism) %>% 
-  summarise(Count = n()) %>% 
-  filter(Count > 8, Base == 'Rod') %>% arrange(-Count)
-```
-
-```{r diff_difff_exp_genes, fig.height = 3, fig.width = 8, echo = F, message = F}
-source('src/make_exp_plot.R')
-
-input <- list()
-input$exp_plot_groups <- c('organism')
-input$exp_plot_facet <- c('CellType_predict')
-input$exp_plot_genes <- c('SCRG1 (ENSG00000164106)')
-input$exp_filter_cat <- 'CellType_predict'
-input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
-input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
-input$exp_plot_col_num <- 16
-input$exp_filter_min_cell_number <- 50
-
-make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-```
-
-To aid in in the analysis of ocular gene expression patterns we provide pre-computed differential tests run
-
-## Pan-study fovea vs peripheral photoreceptor identifies N genes consistently differentially expressed
-
-To identify
-
-Sanes proposed X genes (https://www.nature.com/articles/s41598-020-66092-9 --> Figure 9)
-
-PRPH2 and RS1 for cones
-NR2E3, SLC24A1, PDE6H, GNAT1 for rods
-Also did RGC, Microglia, Astrocytes, Muller Glia
-
-## Projection of organoid data onto scEiaD demonstrates [SOMETHING]
-
-scEiaD is built using scvi-tool's scANVI batch correction tool.
-As part of the process a computational transcriptome model of the reference datasets are built.
-This model is only a few mb in size and can be used on a modest computer to project private scRNA datasets onto it.
-This allows an ocular researcher several important functionalities.
-First, the ability to check the quality control of an existing dataset by looking at the distribution of their cells overlapped with our labelled cell types.
-Second, in perturbation experiments, to identify unusual distributions of cells.
-Third, to leverage our extensive pre-labelled cell information to transfer cell type labels onto private data.
-
-We demonstrate some of these possibilities by projecting two organoid datasets from X and Y.
-
-# Conclusions
-
-## PLAE provides quick access to over one million transcriptomes across 35 studies
-
-## Projection of outside data onto scEiaD allows for ... (check lung atlas paper)
-
-##
-
-# Supplementary Figures and Tables
-
-```{r decontX, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=decontX_cap}
-decontX_cap <- supFig_cap(name='decontX', caption = 'DecontX *in silico* ambient RNA contamination tool substantially removes Rhodopsin expression in non-rod cells while retaining high expression rods.
-First shown (a) is Rhodopsin expression across the scEiaD v0 dataset, without DecontX optimization.
-Note Rhodopsin expression is noticeable in many cell types beyond the rods.
-Next (b) is the scEiaD v1 dataset with DecontX optimization.
-Rhodopsin is nearly exclusively expressed in labelled rod cells.
-Despite the computational ambient RNA removal, Rhodopsin expression remains high in the Rods.')
-source('src/make_exp_plot.R')
-
-input <- list()
-input$exp_plot_groups <- c('organism')
-input$exp_plot_facet <- c('CellType_predict')
-input$exp_plot_genes <- c('RHO (ENSG00000163914)')
-input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
-input$exp_plot_col_num <- 16
-input$exp_filter_min_cell_number <- 50
-
-v0_rho <- make_exp_plot(input, scEiaD_2020_v00, meta_filter_v0)
-v1_rho <-  make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-input$exp_filter_min_cell_number <- 1000
-v1.1_rho <- make_exp_plot(input, scEiaD_2020_v01, meta_filter %>% filter(CellType_predict %in% unique(meta_filter_v0$CellType_predict)))
-plot_grid(v0_rho, v1_rho, ncol = 1, labels = c('a','b'), rel_heights = c(1,1.5))
-```
-
-```{r studies_table, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=studies_table_cap}
-studies_table_cap <- supTab_cap(name = 'studies_table', caption = 'Studies, tissues, and number of cells across scEiaD')
-
-meta_filter %>% group_by(Citation, PMID, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable()
-```

### 2022-06-16 decontx.Rmd
new files

--- 2022-06-10 manuscript_00.Rmd
@@
 
-# Introduction
-
-## single cell RNA-seq allows for a higher resolution view of ocular transcriptomics
-
-Single cell RNA-sequencing dates back to . The first
-
-## Many published ocular single cell RNA-seq atlases
-
-## Web atlases containing ocular data have substantial limitations
-
-Data difficult to access and cross comparisons impossible
-
-But important as seeing trends trends across different groups strongly demonstrates that 
-the signal is biological, not technical.
-
-## What we did
-
-We published the ocular focused scEiaD v2021 resource in 2021, which was one of the first single cell meta-atlases.
-To broaden the usage of this powerful database, we restructured the data in a custom sqlite-based database and built a reactive web application around it.
-As many new studies were published since our last data build, we re-ran the scEiaD pipeline in March of 2022.
-In aggregate we have collated `r meta_filter$study_accession %>% unique() %>% length()` studies across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues.
-We hand-curated `r meta_filter$CellType %>% unique() %>% length()` cell types, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the remaining cells.
-Extensive quality control metrics were tuned and applied to remove lower quality cells.
-In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values.
-
-# Methods
-
-## scEiaD pipeline upgraded and the database re-built
-
-The scEiaD pipeline is structurally the same as in Swamy et al.
-[@swamy_building_2021].
-Very briefly, we collect the raw fastq files and quantify with the kallisto / bustools system [@melsted_barcode_2019; @melsted_modular_2019].
-The quantified counts are merged into one matrix that is used for downstream processing with scvitools to batch correct the data and Seurat for quality control, clustering, and 2D UMAP visualization.
-We use our previous xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamy_building_2021].
-
-The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300 Second, we use the DecontX algorithm [@yang_decontamination_2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination.
-Third, we altered the procedure for automatically aligning gene names across species (see below for details).
-Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process.
-
-Two custom cutoffs for the minimum required number of unique genes identified per cell were used for the E12 chicken data and the cornea dataset (SRP362101) as an abnormally high number of cells were returned from our default value of 250 (????).
-We set the cutoffs for these studies to be 1000 and 800, respectively, as these values resulted in approximately the same number of cells being returned as the authors reported in their papers.
-
-## Cross species gene alignment
-
-As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify, we must identify shared genes.
-We found homologs/orthologs by pulling from BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference.
-We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus Musculus*.
-This generates a table linked by Ensembl gene IDs and gene names.
-We then use a full join on non-duplicated Ensembl IDs between mouse and human.
-As we noticed a small number of gene names failed to be linked in this manner we ran another join on the remaining non-aligned genes, using gene name.
-This procedure gave us 17,769 shared human and mouse genes.
-To align the other species we again joined on Ensembl gene ID, removing genes where one chicken or macaque gene aligned to multiple human/mouse genes.
-In cases where multiple chicken or macaque genes aligned to one human/mouse gene, then we aggregated the counts by sum.
-In cases where there was no corresponding chicken or macaque gene, we filled in zeros.
-After these steps we had, as expected, 17,769 genes.
-
-## Metadata curation
-
-Every study brought into scEiaD was hand curated to identify unique biological samples, published paper ID (where available), organ, tissue (e.g.
-Cornea), source (iPSC based or tissue), single cell platform (e.g.
-10Xv2, DropSeq, etc.).
-We also, where possible, annotated retina region (e.g.
-fovea), sex, and age or developmental stage.
-To import individual cell type labels (e.g.
-Rod, Muller Glia), we wrote custom code for each study that made this table available to link their cell type assignments back to the matching barcoded cell.
-We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels.
-We normalized names for CellType, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).
-We did retain the original published (except for the broader name normalization) labels under SubCellType.
-
-## Differential gene testing
-
-The scater::aggregateAcrossCells function was used to aggregate gene counts across species, study accession, and [category] where category was either Cluster, CellType, or CellType (Predict) assignments for a cell.
-If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test.
-Any genes with a sum of 0 counts after aggregation were removed.
-The aggregated matrix was imported into DESeq2 using the DESeqDataSetFromMatrix function with the design given as "~study_accession + [category]".
-Contrasts were extracted either with the target category (e.g.
-"Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function.
-Genes were considered to be differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2.
-To enhance the specificity of the
-
-## scEiaD structure and web app optimization
-
-The web app at plae.nei.nih.gov can display gene expression in only a few seconds across over a million cells.
-This speed is only possible due to custom data structures to optimize for data query efficiency.
-Like eyeIntegration's EiaD database, scEiaD is a sqlite database with tables for the core categories.
-This allows the app to initialize in about 30 seconds on a cloud server with memory usage under 8 GB.
-For the gene by cell expression matrix, the data was transformed into a "long" format where there are three columns: Gene, Cell Barcode, and Gene Count Value.
-This allows for cell - gene level queries to be completed, again with minimal memory usage, in under 0.5 seconds.
-To generate the aggregated information (in the Dot Plot, Expression Plot, and tables) we pre-calculated all queries by running a dplyr "group by" operation on all column fields.
-On user given query, the data is further aggregated to level request.
-This allows for complicated queries to complete in seconds rather than minutes.
-
-## Data reproducibility
-
-The codebase for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is XXXXXXX.
-There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which takes the optimal parameters from SnakePOP to build the scEiaD sqlite database for the app.
-
-# Results
-
-## New studies and improvements to the scEiaD database
-
-Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamy_building_2021].
-We updated the scEiaD resource for the PLAE v0.90 web app in six important ways.
-First, we added two chicken retina (*Gallus gallus*) studies.
-Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye.
-Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons.
-Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`).
-Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300 (CHECK THIS), to improve the database quality.
-Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons.
-
-After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (`r fig_cap(name = 'scEiaD_data_overview', display = 'cite')`).
-This resource is only possible due to the publication of `r meta_filter$PMID %>% unique %>% length()` single cell resources that scEiaD draws from (`r supTab_cap(name = 'studies_table', display = 'cite')`).
-To recognize these papers, we make the first data table on the loading page of plae.nei.nih.gov list each of these studies along with a direct link to their citation, where possible.
-
-```{r scEiaD_data_overview, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=scEiaD_data_overview_cap}
-
-num1 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(Tissue) %>% unique() %>% length()
-num2 <- meta_filter %>% filter(Organ != 'Eye') %>% pull(Tissue) %>% unique() %>% length()
-num3 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique() %>% length()
-scEiaD_data_overview_cap <- fig_cap(name = 'scEiaD_data_overview', caption = glue::glue("Distribution of tissues and cell types in scEiaD.
-A.
-tabulation of the number of cells present across {num1} ocular and {num2} non-ocular tissues, B.
-Number of studies for each ocular tissue.
-C.
-Number of cells present across {num3} curated ocular-derived cell types."))
-
-# organ
-meta_filter <- meta_filter %>% mutate(organism = factor(organism, levels = meta_filter$organism %>% unique()))
-a <- meta_filter %>%
-  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
-                           TRUE ~ 'Body'),
-         Tissue = gsub('\\.|_', ' ', Tissue),
-         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
-         Tissue = gsub(" and Aorta","", Tissue)) %>% 
-  filter(Organ == 'Body') %>% 
-  filter(Tissue != 'Brain Choroid Plexus') %>% 
-  group_by(Organ, Tissue,organism, study_accession) %>%
-  summarise(Count = n()) %>%
-  ungroup() %>% 
-  mutate(organism = factor(organism, levels = meta_filter$organism %>% unique())) %>% 
-  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
-  ylab('Number of Cells') +
-  geom_bar(stat='identity', position = 'dodge') +
-  cowplot::theme_cowplot() +
-  coord_flip() +
-  # scale_y_continuous(breaks = c(2500, 5000, 7500, 10000, 12500, 84000)) +
-  # ggbreak::scale_y_break(c(13000, 82000)) +
-  scale_fill_discrete(drop=TRUE,
-                      limits = levels(meta_filter$organism)) +
-  theme(legend.position = "none") +
-  scale_y_continuous(expand = c(0, 0)) + 
-  ylab('') + xlab('')
-
-aa <- meta_filter %>% 
-  filter(Tissue == 'Brain Choroid Plexus') %>% 
-  group_by(Organ, Tissue,organism, study_accession) %>%
-  summarise(Count = n()) %>%
-  ungroup() %>% 
-  mutate(organism = factor(organism, levels = meta_filter$organism %>% unique())) %>% 
-  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
-  ylab('Number of Cells') +
-  geom_bar(stat='identity', position = 'dodge') +
-  cowplot::theme_cowplot() +
-  coord_flip() +
-  scale_fill_discrete(drop=TRUE,
-                      limits = levels(meta_filter$organism)) +
-  theme(legend.position = "none") +
-  scale_y_continuous(expand = c(0, 0)) + 
-  ylab('') + xlab('')
-
-# eye studies
-b <-  meta_filter %>%
-  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
-                           TRUE ~ 'Body'),
-         Tissue = gsub('\\.|_', ' ', Tissue),
-         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
-         Tissue = case_when(Tissue == "Choroid" ~ 'RPE-Choroid',
-                            TRUE ~ Tissue),
-         Tissue = gsub(" and Aorta","", Tissue)) %>% 
-  filter(Organ == 'Eye') %>% 
-  group_by(Organ, Tissue,organism) %>%
-  summarise(Count = n()) %>%
-  ungroup() %>% 
-  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
-  ylab('Number of Cells') +
-  geom_bar(stat='identity', position = 'dodge') +
-  cowplot::theme_cowplot() +
-  coord_flip() +
-  theme(legend.position = "none") +
-  scale_y_continuous(expand = c(0, 0)) +
-  xlab('')
-
-# eye studies samples
-c <-  meta_filter %>%
-  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
-                           TRUE ~ 'Body'),
-         Tissue = gsub('\\.|_', ' ', Tissue),
-         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
-         Tissue = gsub(" and Aorta","", Tissue)) %>% 
-  filter(Organ == 'Eye') %>% 
-  select(Organ, Tissue,organism, study_accession) %>%
-  unique() %>%
-  group_by(Organ, Tissue, organism) %>% 
-  summarise(Count = n()) %>%
-  ungroup() %>% 
-  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
-  ylab('Number of Studies') +
-  geom_bar(stat='identity', position = 'dodge') +
-  cowplot::theme_cowplot() +
-  coord_flip() +
-  theme(legend.position = "none") +
-  scale_y_continuous(expand = c(0, 0))  + xlab('')
-
-# retina cell types
-d <- meta_filter %>% filter(Organ == 'Eye') %>% 
-  mutate(CellType_predict = case_when(is.na(CellType_predict) ~ 'Unlabelled', TRUE ~ CellType_predict)) %>% 
-  group_by(CellType_predict, study_accession, organism) %>%
-  summarise(Count = n()) %>% filter(Count > 10) %>% ungroup() %>% 
-  ggplot(aes(x=CellType_predict,y=Count, fill = organism, group = organism)) +
-  ylab('Number of Cells') +
-  geom_bar(stat='identity', position = 'dodge') +
-  cowplot::theme_cowplot() +
-  coord_flip() + 
-  scale_y_continuous(expand = c(0, 0)) + xlab('')
-
-a_aa_b <- plot_grid(a,aa,b, ncol = 1, align = 'hv', rel_heights = c(10, 1.5,3.5), labels = 'a')
-c_d <- plot_grid(c,d, rel_heights = c(3,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))
-plot_grid(a_aa_b, c_d, ncol = 2, rel_widths = c(1,1.6))
-```
-
-## scRNA web resources containing ocular data
-
-Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources_cap', display = 'cite')`).
-The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are not ocular focused, but do contain some ocular data.
-The PanglaoDB and UCSC are resources which have ingress-ed a large number of published single cell datasets The UCSC Cell Browser has around 62 thousand ocular single cell transcriptomes while the PangloaDB has about 47 thousand.
-However, these resources provide no curation of the metadata and process the data for each study individually.
-Thus each study is effectively silo-ed off from the rest and cross comparisons are impractical.
-The Tabula Sapiens project is an effort to sequence all cells in the mouse and thus the ocular data present, though relatively small (at around 10 thousand), can be more readily cross compared to non-ocular tissues / cells.
-The Protein Atlas resource, which has aggregated protein labels (????) information has recently used one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g.
-one can see whether a gene of interest is enriched in Cones).
-The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has selected a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app.
-PLAE is distinguished from these resources by hand curation of the metadata, re-processing of the data from the raw sequence, and, most crucially, integrating the independent studies to deliver cross study comparisons.
-
-```{r web_resources, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=web_resources_cap}
-
-tab_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
-
-feature_table <- rbind(
-  c('PLAE','Yes','Yes', 'Yes', '1136 ', 'https://plae.nei.nih.gov'),
-  c('Spectacle', 'Yes' ,'No', 'No', '1800', 'https://singlecell-eye.org'),
-  c('UCSC Cell Browser', 'Yes', 'No', 'Yes', '62', 'https://cells.ucsc.edu' ),
-  c('Tabula Sapiens', 'No', 'N/A', 'Yes', '10', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
-  c('PanglaoDB', 'Yes', 'No', 'Yes', '47', 'https://panglaodb.se'),
-  c('Protein Atlas', 'Yes*','No', 'Yes', '5',  'https://www.proteinatlas.org/humanproteome/celltype')) %>% 
-  as_tibble()
-colnames(feature_table) <- c('Resource Name', 'Independent Datasets', 'Harmonization', 'Ocular and non-Ocular Data', 'Ocular Cell Count\n(thousands)', 'URL')
-feature_table %>% flextable()
-# as of 2021 10 25
-```
-
-## PLAE differs from a complementary web app, Spectacle in several important ways
-
-While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently [@noauthor_spectacle_nodate].
-PLAE is built around the scEiaD database, which runs an extensive multi-stage pipeline to build and identify a high performing scVI-based eye model that is used to integrate all datasets together [@swamy_building_2021].
-This allows for gene queries to be constructed and analyzed across studies.
-In contrast, Spectacle, as of March 2022, is a compilation of study-level datasets and thus you cannot run queries across studies in Spectacle.
-Furthermore, Spectacle contains no non-ocular dataset and the cell types labels have not been harmonized between studies.
-Finally, as this Spectacle uses Seurat objects, the app takes many seconds to load an individual dataset and many more to display the data.
-
-## Extensive optimization makes the PLAE web app highly responsive despite containing huge amounts of data
-
-There are two interwoven reasons why PLAE is fast, despite the huge amount of data it contains.
-First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables.
-A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage.
-The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous; the scEiaD v2022-03-22 sqlite database is about 368 GB.
-Second, the visualization code in PLAE (e.g.
-the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) to reduce drawing time.
-Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study (tested on 2022-03-23).
-In the same amount of time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies.
-
-## Rich web-available visualizations
-
-PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values.
-So thus a user can see Crx expression across over a million cells and separately see cell type labels for those same million cells.
-The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells.
-Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`a).
-
-For users interested in seeing how expression of a gene relates to study or other covariates, the "Exp Plot" view displays expression by cell type, predicted cell type, or cluster.
-The values can then be split and colored by multiple covariates like study or organism.
-This allows a user to see whether a gene pattern is consistent across studies or species.
-We also offer a "Dot Plot" view which can display a large number of genes in a space efficient manner.
-This allows a users to see whether a set of genes has related (or not) expression patterns across multiple cell types (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`b and c).
-
-While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina.
-The structure of the cartoon retina was based on the human retina.
-This unique visualization displays the major cell types of the retina and the RPE / choroid layer behind it.
-The cell types are colored by the relative amounts of expression in the user selected gene.
-Like all visualizations in PLAE, the user can use powerful filtering options to only display data from certain metadata characteristics, like species or publication (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`d).
-
-```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
-
-```
-
-## Differential testing identifies thousands genes enriched across various ocular cell types
-
-We used a "pseudo bulk" based approach wherein the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping.
-The creates a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices.
-This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021].
-Once the aggregated matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
-We tested each species (Human, Macaque, Mouse) separately.
-We had an insufficient number of chicken studies to conduct differential testing.
-The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential expression.
-After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g.
-cones versus all remaining cell types) or in a pair-wise manner (cone against rod).
-We fin
-```{r}
-# fix  occasional NA
-# data made from src/ensdb.R
-diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
-
-diff_tab <- diff_tab_big %>% 
-  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique()))
-
-diff_tab %>% 
-  group_by(Base, Organism) %>% 
-  summarise(Count = n()) %>% 
-  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
-  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
-  as_tibble() %>% 
-  rename(`CellType (Predict)` = CellType_predict) %>% 
-  flextable()
-
-# number of celltype (predict) - organism - study counts (with >=50 cells)
-meta_filter %>%  
-  filter(Organ == 'Eye') %>% 
-  group_by(organism, CellType_predict, study_accession) %>% 
-  count() %>% 
-  filter(n >= 50) %>% 
-  ungroup() %>% 
-  group_by(organism, CellType_predict) %>% 
-  count() %>% 
-  pivot_wider(values_from = n, names_from = organism) %>% 
-  filter(!is.na(CellType_predict)) %>% 
-  replace(is.na(.), values = 0) %>% 
-  rename(`CellType (Predict)` = CellType_predict) %>% 
-  flextable()
-```
-
-Many of these genes can be differentially expressed in more than one cell type.
-To identify genes that are consistently differentially expressed we can use a more specific test.
-We define a consistent gene that is
-
-```{r}
-spec_diff <- diff_tab_big %>% 
-  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
-  group_by(Symbol, Organism, Base) %>% 
-  summarise(Count = n()) %>% 
-  summarise(Count = n())
-
-spec_diff_go <- spec_diff %>% 
-  filter(Count <= 3) %>% 
-  mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% 
-  pull(go)
-
-diff_tab_big %>% 
-  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
-  group_by(Symbol, Organism, Base) %>% 
-  summarise(Count = n(),
-            baseMean = mean(baseMean),
-            log2FoldChange = mean(log2FoldChange),
-            padj = mean(padj)) %>% 
-  filter(Count > 1) %>% 
-  left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
-  mutate(Ratio = `Count.x` / `Count.y`) %>% 
-  select(-`Count.x`,-`Count.y`) %>% 
-  #mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% 
-  #filter(go %in% (spec_diff_go))  %>% 
-  #select(-go) %>% 
-  ungroup() %>% 
-  group_by(Organism, Base) %>% 
-  slice_max(n = 10, order_by = log2FoldChange)
-
-```
-
-### heatmap human
-```{r, fig.width=5, fig.height=20}
-diff_ratios <- diff_tab_big %>% 
-    filter(Against != 'All') %>% 
-    group_by(Symbol, ID, Organism, Base) %>% 
-    summarise(Count = n()) %>% 
-    filter(Count > 1) %>% 
-    left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
-    mutate(Ratio = `Count.x` / `Count.y`) %>% 
-    select(Symbol:Base, Ratio)
-
-gene_sets <- diff_tab %>% 
-  left_join(diff_ratios, by = c('Symbol', 'ID','Base','Organism')) %>% 
-  filter(Ratio > 0.2) %>% group_by(Base, Organism) %>% 
-  slice_max(n=5, order_by = log2FoldChange)
-
-g <- gene_sets %>% filter(Organism == 'Homo sapiens') %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
-
-# only uses CT that exist in the organism <-> CellType 
-ct <- meta_filter %>% filter(organism == 'Homo sapiens') %>% pull(CellType) %>% unique()
-
-x <- scEiaD_2020_v01 %>% tbl("diff_testing") %>% filter(Gene %in% g, Base %in% ct) %>% collect() %>% filter(Group == 'CellType (Predict)', Organism == 'Homo sapiens', Against == 'All') %>% select(Gene, log2FoldChange, Base) %>% pivot_wider(values_from=log2FoldChange, names_from = Base)
-celltype <- colnames(x[,-1])
-x <- x %>% data.frame()
-row.names(x) <- x$Gene
-x <- x[,-1]
-Heatmap(x, column_labels = celltype)
-```
-
-### heatmap mouse
-```{r, fig.width=5, fig.height=20}
-
-g <- gene_sets %>% filter(Organism == 'Mus musculus') %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
-
-x <- scEiaD_2020_v01 %>% tbl("diff_testing") %>% filter(Gene %in% g) %>% collect() %>% filter(Group == 'CellType (Predict)', Organism == 'Mus musculus', Against == 'All') %>% select(Gene, log2FoldChange, Base) %>% pivot_wider(values_from=log2FoldChange, names_from = Base)
-celltype <- colnames(x[,-1])
-x <- x %>% data.frame()
-row.names(x) <- x$Gene
-x <- x[,-1]
-Heatmap(x, column_labels = celltype)
-```
-
-```{r}
-# top 100 genes in each organism/base category
-diff_tab %>%
-
-  group_by(Base, Organism) %>% 
-  top_n(., 100, -padj) %>% 
-  arrange(Base, Organism) %>% 
-  relocate(Base, Organism) %>% 
-  select(-Against, -Group)
-```
-
-```{r}
-# consistently diff genes
-consistent <- diff_tab %>% 
-  group_by(Symbol, Base)  %>% 
-  summarise(Count = n()) %>% 
-  filter(Count > 1) %>% 
-  mutate(k = paste(Symbol, Base, sep = '_'))
-
-diff_tab %>% mutate(k = paste(Symbol, Base, sep = '_')) %>% 
-  filter(k %in% consistent$k) %>% 
-  group_by(Base, Organism) %>% 
-  summarise(Count = n()) %>% 
-  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
-  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
-  as_tibble() %>% 
-  flextable()
-
-```
-
-```{r}
-# flip flop diff genes
-down_tab <- scEiaD_2020_v01 %>% tbl('diff_testing') %>% 
-  filter(Group == 'CellType (Predict)', padj < 1e-5, (log2FoldChange) < -2) %>% 
-  collect() %>% 
-  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique())) %>% mutate(k = paste(Symbol, Base, sep = '_'))
-
-diff_tab %>% mutate(k = paste(Symbol, Base, sep = '_')) %>% 
-  filter(k %in% down_tab$k, Base %in% c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )) %>%
-  arrange(padj)
-
-diff_tab_big %>% 
-  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
-  group_by(Symbol, Base, Organism) %>% 
-  summarise(Count = n()) %>% 
-  filter(Count > 8, Base == 'Rod') %>% arrange(-Count)
-```
-
-```{r diff_difff_exp_genes, fig.height = 3, fig.width = 8, echo = F, message = F}
-source('src/make_exp_plot.R')
-
-input <- list()
-input$exp_plot_groups <- c('organism')
-input$exp_plot_facet <- c('CellType_predict')
-input$exp_plot_genes <- c('SCRG1 (ENSG00000164106)')
-input$exp_filter_cat <- 'CellType_predict'
-input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
-input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
-input$exp_plot_col_num <- 16
-input$exp_filter_min_cell_number <- 50
-
-make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-```
-
-To aid in in the analysis of ocular gene expression patterns we provide pre-computed differential tests run
-
-## Pan-study fovea vs peripheral photoreceptor identifies N genes consistently differentially expressed
-
-To identify
-
-Sanes proposed X genes (https://www.nature.com/articles/s41598-020-66092-9 --> Figure 9)
-
-PRPH2 and RS1 for cones
-NR2E3, SLC24A1, PDE6H, GNAT1 for rods
-Also did RGC, Microglia, Astrocytes, Muller Glia
-
-## Projection of organoid data onto scEiaD demonstrates [SOMETHING]
-
-scEiaD is built using scvi-tool's scANVI batch correction tool.
-As part of the process a computational transcriptome model of the reference datasets are built.
-This model is only a few mb in size and can be used on a modest computer to project private scRNA datasets onto it.
-This allows an ocular researcher several important functionalities.
-First, the ability to check the quality control of an existing dataset by looking at the distribution of their cells overlapped with our labelled cell types.
-Second, in perturbation experiments, to identify unusual distributions of cells.
-Third, to leverage our extensive pre-labelled cell information to transfer cell type labels onto private data.
-
-We demonstrate some of these possibilities by projecting two organoid datasets from X and Y.
-
-# Conclusions
-
-## PLAE provides quick access to over one million transcriptomes across 35 studies
-
-## Projection of outside data onto scEiaD allows for ... (check lung atlas paper)
-
-##
-
-# Supplementary Figures and Tables
-
-```{r decontX, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=decontX_cap}
+```{r decontX, fig.height = 25, fig.width = 20, echo = F, message = F, fig.cap=decontX_cap}
 decontX_cap <- supFig_cap(name='decontX', caption = 'DecontX *in silico* ambient RNA contamination tool substantially removes Rhodopsin expression in non-rod cells while retaining high expression rods.
@@
 ```
-
-```{r studies_table, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=studies_table_cap}
-studies_table_cap <- supTab_cap(name = 'studies_table', caption = 'Studies, tissues, and number of cells across scEiaD')
-
-meta_filter %>% group_by(Citation, PMID, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable()
-```

### 2022-06-16 heatmaps.Rmd
new files

---
title: "R Notebook"
output: html_notebook
---
### heatmap human
```{r, fig.width=12, fig.height=5}
diff_ratios <- diff_tab_big %>% 
  filter(Against != 'All') %>% 
  group_by(Symbol, ID, Organism, Base) %>% 
  summarise(Count = n()) %>% 
  filter(Count > 1) %>% 
  left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
  mutate(Ratio = `Count.x` / `Count.y`) %>% 
  select(Symbol:Base, Ratio)

gene_sets <- diff_tab %>% 
  left_join(diff_ratios, by = c('Symbol', 'ID','Base','Organism')) %>% 
  filter(Ratio > 0.2) %>% group_by(Base, Organism) %>% 
  slice_max(n=5, order_by = log2FoldChange)

org <- 'Homo sapiens'

g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()

# only uses CT that exist in the organism <-> CellType 
ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()

x <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Gene %in% g, Base %in% ct) %>% 
  collect() %>% 
  filter(Group == 'CellType (Predict)', 
         Organism == org, Against == 'All') %>% 
  select(Gene, log2FoldChange, Base) %>% 
  pivot_wider(values_from=log2FoldChange, names_from = Base) 
celltype <- colnames(x[,-1])
x <- x %>% data.frame()
row.names(x) <- x$Gene %>% gsub('\\(|\\)|ENSG\\d+','',.) %>% gsub(' ','',.)
x <- x[,-1]
Heatmap(t(x), row_labels = celltype)
```

# Rpe volcano
```{r, fig.width=3, fig.height=2}
rpe_diff <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Base == 'RPE') %>%  
  filter(Group == 'CellType (Predict)', 
         Organism == org, Against == 'All') %>% 
  select(Gene, log2FoldChange, pvalue, Base) %>% 
  collect()

genes <- bind_rows(rpe_diff %>% slice_max(log2FoldChange, n = 12),
                   rpe_diff %>% slice_min(log2FoldChange, n = 5),
                   rpe_diff %>% slice_min(pvalue, n = 7)) %>% 
  filter(!grepl('NA', Gene)) %>% pull(Gene) %>% unique()

rpe_diff %>% 
  ggplot(aes(x=log2FoldChange,y=-log2(pvalue))) + 
  geom_point() + cowplot::theme_cowplot() + ggrepel::geom_label_repel(data = rpe_diff %>% 
                                                                        filter(Gene %in% genes) %>% 
                                                                        mutate(Gene = gsub('\\(|\\)|ENSG\\d+','',Gene) %>% 
                                                                                 gsub(' ','',.)), 
                                                                      aes(label = Gene),
                                                                      max.overlaps = 100)
```

```{r, fig.width=6, fig.height=4}

x <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Gene %in% genes) %>% 
  collect() %>% 
  filter(Group == 'CellType (Predict)', 
         Organism == org, 
         Against == 'All') %>% 
  select(Gene, log2FoldChange, Base) %>% 
  pivot_wider(values_from=log2FoldChange, names_from = Base) 
celltype <- colnames(x[,-1])
x <- x %>% data.frame()
row.names(x) <- x$Gene %>% gsub('\\(|\\)|ENSG\\d+','',.) %>% gsub(' ','',.)
x <- x[,-1]
Heatmap(x, column_labels = celltype)
```

```{r, fig.width=6, fig.height=4}
org <- 'Mus musculus'
x <- scEiaD_2020_v01 %>% 
  tbl("diff_testing") %>% 
  filter(Gene %in% genes) %>% 
  collect() %>% 
  filter(Group == 'CellType (Predict)', 
         Organism == org, 
         Against == 'All') %>% 
  select(Gene, log2FoldChange, Base) %>% 
  pivot_wider(values_from=log2FoldChange, names_from = Base) 
celltype <- colnames(x[,-1])
x <- x %>% data.frame()
row.names(x) <- x$Gene %>% gsub('\\(|\\)|ENSG\\d+','',.) %>% gsub(' ','',.)
x <- x[,-1]
Heatmap(x, column_labels = celltype)
```
```

### 2022-06-16 manuscript_00.Rmd
new files

@@
 
-tab_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
+web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
 
@@
 
-```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
+#```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
 
-```
+#```
 
@@
   as_tibble() %>% 
-  rename(`CellType (Predict)` = CellType_predict) %>% 
   flextable()
@@
 
-g <- gene_sets %>% filter(Organism == 'Homo sapiens') %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
+org <- 'Homo sapiens'
+
+g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
 
 # only uses CT that exist in the organism <-> CellType 
-ct <- meta_filter %>% filter(organism == 'Homo sapiens') %>% pull(CellType) %>% unique()
+ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()
 
-x <- scEiaD_2020_v01 %>% tbl("diff_testing") %>% filter(Gene %in% g, Base %in% ct) %>% collect() %>% filter(Group == 'CellType (Predict)', Organism == 'Homo sapiens', Against == 'All') %>% select(Gene, log2FoldChange, Base) %>% pivot_wider(values_from=log2FoldChange, names_from = Base)
+x <- scEiaD_2020_v01 %>% 
+  tbl("diff_testing") %>% 
+  filter(Gene %in% g, Base %in% ct) %>% 
+  collect() %>% 
+  filter(Group == 'CellType (Predict)', 
+         Organism == org, Against == 'All') %>% 
+  select(Gene, log2FoldChange, Base) %>% 
+  pivot_wider(values_from=log2FoldChange, names_from = Base) 
 celltype <- colnames(x[,-1])
@@
 
-g <- gene_sets %>% filter(Organism == 'Mus musculus') %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
+org <- 'Mus musculus'
 
-x <- scEiaD_2020_v01 %>% tbl("diff_testing") %>% filter(Gene %in% g) %>% collect() %>% filter(Group == 'CellType (Predict)', Organism == 'Mus musculus', Against == 'All') %>% select(Gene, log2FoldChange, Base) %>% pivot_wider(values_from=log2FoldChange, names_from = Base)
+g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
+
+# only uses CT that exist in the organism <-> CellType 
+ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()
+
+x <- scEiaD_2020_v01 %>% 
+  tbl("diff_testing") %>% 
+  filter(Gene %in% g, Base %in% ct) %>% 
+  collect() %>% 
+  filter(Group == 'CellType (Predict)', 
+         Organism == org, Against == 'All') %>% 
+  select(Gene, log2FoldChange, Base) %>% 
+  pivot_wider(values_from=log2FoldChange, names_from = Base) 
+celltype <- colnames(x[,-1])
+x <- x %>% data.frame()
+row.names(x) <- x$Gene
+x <- x[,-1]
+Heatmap(x, column_labels = celltype)
+```
+
+### heatmap macaque
+```{r, fig.width=5, fig.height=20}
+
+org <- 'Macaca fascicularis'
+
+g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
+
+# only uses CT that exist in the organism <-> CellType 
+ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()
+
+x <- scEiaD_2020_v01 %>% 
+  tbl("diff_testing") %>% 
+  filter(Gene %in% g, Base %in% ct) %>% 
+  collect() %>% 
+  filter(Group == 'CellType (Predict)', 
+         Organism == org, Against == 'All') %>% 
+  select(Gene, log2FoldChange, Base) %>% 
+  pivot_wider(values_from=log2FoldChange, names_from = Base) 
 celltype <- colnames(x[,-1])
@@
 
-meta_filter %>% group_by(Citation, PMID, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable()
+meta_filter %>% group_by(Citation, PMID, organism, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable()
 ```

### 2022-06-16 ct_ctp_dot_plot.R

@@ -1,0 +1,115 @@
+# ARVO presentation
+# run "build_tables.R" to get data
+ct <- meta_filter %>%
+  mutate(CellType = gsub('AC/HC_Precurs', 'Amacrine/Horizontal Precursors', CellType)) %>%
+  #mutate(CellType = gsub("Rod Bipolar Cells", "Bipolar Cells", CellType)) %>%
+  filter(!is.na(CellType), Organ == 'Eye',
+         !is.na(study_accession),
+         !CellType %in% c('Doublet', 'Doublets'),
+         !grepl('Vascul', CellType)) %>%
+  mutate(organism = case_when(grepl('Homo', organism) ~ 'HS',
+                              grepl('Mus', organism) ~ 'MM',
+                              grepl('Gall', organism) ~ 'GG',
+                              TRUE ~ 'MF')) %>%
+  group_by(CellType,organism, study_accession) %>%
+  summarise(Count = n()) %>% filter(Count > 50) %>%
+  summarise(Studies = length(study_accession), Count = sum(Count)) %>%
+  summarise(Species = paste(organism, collapse = ', '), Studies = sum(Studies), Count = sum(Count)) %>%
+  arrange(-Count)
+
+ctp <- meta_filter %>%
+  mutate(CellType = gsub('AC/HC_Precurs', 'Amacrine/Horizontal Precursors', CellType_predict)) %>%
+  #mutate(CellType = gsub("Rod Bipolar Cells", "Bipolar Cells", CellType)) %>%
+  filter(!is.na(CellType),Organ == 'Eye',
+         !is.na(study_accession)) %>%
+  mutate(organism = case_when(grepl('Homo', organism) ~ 'HS',
+                              grepl('Mus', organism) ~ 'MM',
+                              grepl('Gall', organism) ~ 'GG',
+                              TRUE ~ 'MF')) %>%
+  group_by(CellType,organism, study_accession) %>%
+  summarise(Count = n()) %>% filter(Count > 50) %>%
+  summarise(Studies = length(study_accession), Count = sum(Count)) %>%
+  summarise(Species = paste(organism, collapse = ', '), Studies = sum(Studies), Count = sum(Count)) %>%
+  arrange(-Count)
+
+# ct study count (>50 cells)
+ct %>%
+  left_join(ctp, by = 'CellType') %>%
[…]

### 2022-06-16 manuscript_00.Rmd
updates

@@
 ```{r}
-spec_diff <- diff_tab_big %>% 
-  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
-  group_by(Symbol, Organism, Base) %>% 
-  summarise(Count = n()) %>% 
-  summarise(Count = n())
-
-spec_diff_go <- spec_diff %>% 
-  filter(Count <= 3) %>% 
-  mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% 
-  pull(go)
-
-diff_tab_big %>% 
-  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
-  group_by(Symbol, Organism, Base) %>% 
-  summarise(Count = n(),
-            baseMean = mean(baseMean),
-            log2FoldChange = mean(log2FoldChange),
-            padj = mean(padj)) %>% 
-  filter(Count > 1) %>% 
-  left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
-  mutate(Ratio = `Count.x` / `Count.y`) %>% 
-  select(-`Count.x`,-`Count.y`) %>% 
-  #mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% 
-  #filter(go %in% (spec_diff_go))  %>% 
-  #select(-go) %>% 
-  ungroup() %>% 
-  group_by(Organism, Base) %>% 
-  slice_max(n = 10, order_by = log2FoldChange)
+scEiaD_2020_v01 %>% 
+    tbl("diff_testing") %>% 
+    filter(Base == 'Cone') %>%  
+    filter(Group == 'CellType (Predict)', 
+           Against == 'All') %>% 
+  group_by(Gene) %>% 
+  summarise(Count = n(), log2FoldChange = mean(log2FoldChange), padj_count = sum(padj < 1e-4), padj = mean(padj)) %>% 
+  filter(!is.na(log2FoldChange), padj_count > 1) %>%  
+  arrange(-log2FoldChange)
 
@@
 ### heatmap human
-```{r, fig.width=5, fig.height=20}
-diff_ratios <- diff_tab_big %>% 
-    filter(Against != 'All') %>% 
-    group_by(Symbol, ID, Organism, Base) %>% 
-    summarise(Count = n()) %>% 
-    filter(Count > 1) %>% 
-    left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
-    mutate(Ratio = `Count.x` / `Count.y`) %>% 
-    select(Symbol:Base, Ratio)
+```{r, fig.width=6, fig.height=17}
+# diff_ratios <- diff_tab_big %>% 
+#     filter(Against != 'All') %>% 
+#     group_by(Symbol, ID, Organism, Base) %>% 
+#     summarise(Count = n()) %>% 
+#     filter(Count > 1) %>% 
+#     left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
+#     mutate(Ratio = `Count.x` / `Count.y`) %>% 
+#     select(Symbol:Base, Ratio)
 
 gene_sets <- diff_tab %>% 
-  left_join(diff_ratios, by = c('Symbol', 'ID','Base','Organism')) %>% 
-  filter(Ratio > 0.2) %>% group_by(Base, Organism) %>% 
+  #left_join(diff_ratios, by = c('Symbol', 'ID','Base','Organism')) %>% 
+  #filter(Ratio > 0.2) %>% 
+  group_by(Base, Organism) %>% 
   slice_max(n=5, order_by = log2FoldChange)
@@
   tbl("diff_testing") %>% 
-  filter(Gene %in% g, Base %in% ct) %>% 
+  filter(Gene %in% g) %>% 
   collect() %>% 
-  filter(Group == 'CellType (Predict)', 
+  filter(Base %in% ct,
+         Group == 'CellType (Predict)', 
          Organism == org, Against == 'All') %>% 
@@
   tbl("diff_testing") %>% 
-  filter(Gene %in% g, Base %in% ct) %>% 
+  filter(Gene %in% g) %>% 
   collect() %>% 
   filter(Group == 'CellType (Predict)', 
-         Organism == org, Against == 'All') %>% 
+         Organism == org, Against == 'All', Base %in% ct) %>% 
   select(Gene, log2FoldChange, Base) %>% 

### 2022-06-17 manuscript_00.Rmd
tweaks

@@
 cones versus all remaining cell types) or in a pair-wise manner (cone against rod).
-We fin
+We find
 ```{r}
@@
 
-Many of these genes can be differentially expressed in more than one cell type.
-To identify genes that are consistently differentially expressed we can use a more specific test.
-We define a consistent gene that is
+# Well supported ocular cell types
+
+Cell types with more than two independent datasets in both human and mouse.
+```{r}
+ws_oct <- meta_filter %>%  
+    filter(Organ == 'Eye') %>% 
+    group_by(organism, CellType_predict, study_accession) %>% 
+    count() %>% 
+    filter(n >= 50) %>% 
+    ungroup() %>% 
+    group_by(organism, CellType_predict) %>% 
+    count() %>% filter(n >= 3) %>% 
+  ungroup() %>% 
+  group_by(CellType_predict) %>% 
+  summarise(Count = n()) %>% 
+  filter(Count > 1, !is.na(CellType_predict)) %>% 
+  pull(CellType_predict)
+```
+
+To identify genes that are differentially expressed across species we
 
 ```{r}
-scEiaD_2020_v01 %>% 
+consist_diff <- scEiaD_2020_v01 %>% 
     tbl("diff_testing") %>% 
-    filter(Base == 'Cone') %>%  
+    filter(Base %in% ws_oct) %>%  
     filter(Group == 'CellType (Predict)', 
            Against == 'All') %>% 
-  group_by(Gene) %>% 
-  summarise(Count = n(), log2FoldChange = mean(log2FoldChange), padj_count = sum(padj < 1e-4), padj = mean(padj)) %>% 
-  filter(!is.na(log2FoldChange), padj_count > 1) %>%  
-  arrange(-log2FoldChange)
+    group_by(Base, Gene) %>% 
+    summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-4), `mean padj` = mean(padj)) %>% 
+    filter(`mean log2FC` > 2, `sig count` > 1, `mean padj` < 1e-5) %>%  
+    collect() %>% 
+    mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
+    left_join(gene_tab, by = 'GENEID') %>% 
+    select(-Gene) %>% 
+    relocate(SYMBOL, GENEID) %>% 
+    dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
+    arrange(-`mean log2FC`)
 

### 2022-06-17 manuscript_00.Rmd
updates

@@
 diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
-
+gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
 diff_tab <- diff_tab_big %>% 
@@
   replace(is.na(.), values = 0) %>% 
+  arrange(CellType_predict) %>% 
   rename(`CellType (Predict)` = CellType_predict) %>% 
-  flextable()
+  mutate(`Well Supported` = case_when(`Homo sapiens` > 2 & `Mus musculus` > 2 ~ 'Yes',
+                                      TRUE ~ 'No')) %>% 
+  flextable(set_caption())
 ```
 
-# Well supported ocular cell types
+# Well supported ocular cell types identify
 
-Cell types with more than two independent datasets in both human and mouse.
+We filter down to only assess cell types with more than two independent datasets in both human and mouse.
+This leaves .....
+These well supported cell types can be used to more strenuously identify genes that are consistently differentially expressed in a cell type across multiple species.
+To define these well supported marker genes we only ass
 ```{r}
@@
 
+# id diff genes only seen once across the cell types
+uniq_g <- consist_diff %>% group_by(Symbol) %>% count() %>% filter(n == 1)
+consist_diff %>% group_by(Base) %>% summarise(Count = n())
+consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n())
 ```

### 2022-06-17 manuscript_00.Rmd
new text

--- 2022-06-16 manuscript_00.Rmd
@@
 library(glue)
-data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
-data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+# data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
+# data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 # setup caption-ing
@@
                             TRUE ~ Tissue),
-         Tissue = case_when(grepl('^RPE|^Choroid', Tissue) ~ 'RPE-Choroid',
+         Tissue = case_when(grepl('RPE|Choroid', Tissue) ~ 'RPE-Choroid',
                             TRUE ~ Tissue),
@@
 
-The scater::aggregateAcrossCells function was used to aggregate gene counts across species, study accession, and [category] where category was either Cluster, CellType, or CellType (Predict) assignments for a cell.
-If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test.
-Any genes with a sum of 0 counts after aggregation were removed.
-The aggregated matrix was imported into DESeq2 using the DESeqDataSetFromMatrix function with the design given as "~study_accession + [category]".
-Contrasts were extracted either with the target category (e.g.
-"Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function.
-Genes were considered to be differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2.
-To enhance the specificity of the
-
 ## scEiaD structure and web app optimization
@@
 The codebase for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is XXXXXXX.
-There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app.
+There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which takes the optimal parameters from SnakePOP to build the scEiaD sqlite database for the app.
 
@@
 Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamy_building_2021].
-We updated the scEiaD resource for the PLAE v0.91 web app in six important ways.
-First, we added a large chicken retina (*Gallus gallus*) studies.
+We updated the scEiaD resource for the PLAE v0.90 web app in six important ways.
+First, we added two chicken retina (*Gallus gallus*) studies.
 Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye.
@@
 
-web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
+tab_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
 
@@
 
-#```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
+```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
 
-#```
+```
 
-## Differential testing identifies thousands genes enriched across various ocular cell types
+## Leveraging the high diversity of studies to identify thousands of consistently differentially expressed genes across ocular cell types
 
-We used a "pseudo bulk" based approach wherein the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping.
-The creates a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices.
+We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping.
+The created a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices and thus allowing for mature bulk RNA-seq tooling to be used.
 This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021].
-Once the aggregated matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
+Once the aggregated matrix was con structured, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
 We tested each species (Human, Macaque, Mouse) separately.
-We had an insufficient number of chicken studies to conduct differential testing.
-The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential expression.
+We had an insufficient number of chick studies to conduct differential testing.
+The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential gene expression.
 After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g.
 cones versus all remaining cell types) or in a pair-wise manner (cone against rod).
-We fin
+We find, in total, thousands of differentially expressed genes across the ocular cell types.
 ```{r}
@@
 # data made from src/ensdb.R
-diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
+hs_v86_tab <- read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz')
+mm_v79_tab <- read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz')
+hs_mm <- bind_rows(hs_v86_tab, mm_v79_tab)
+
+diff_tab_big <- scEiaD_2020_v01 %>% tbl('diff_testing') %>% 
+  filter(Group == 'CellType (Predict)', padj < 1e-5, (log2FoldChange) > 2) %>% 
+  collect() %>% 
+  separate(Gene, into = c('Symbol', 'ID'), sep = ' ') %>% 
+  mutate(ID = gsub('\\(|\\)', '', ID))
+
+diff_tab_big2 <- diff_tab_big %>% 
+  filter(Symbol == '-' | Symbol == 'NA')
+
+diff_tab_big2 <- 
+  diff_tab_big2 %>% 
+  left_join(hs_mm, by = c('ID' = 'GENEID')) %>% 
+  select(-Symbol) %>% 
+  rename(Symbol = SYMBOL) %>% 
+  relocate(Symbol)
+
+diff_tab_big <- bind_rows(
+  diff_tab_big %>% filter(!Symbol %in% c('-', 'NA')),
+  diff_tab_big2
+)
 
@@
   flextable()
-
-# number of celltype (predict) - organism - study counts (with >=50 cells)
-meta_filter %>%  
-  filter(Organ == 'Eye') %>% 
-  group_by(organism, CellType_predict, study_accession) %>% 
-  count() %>% 
-  filter(n >= 50) %>% 
-  ungroup() %>% 
-  group_by(organism, CellType_predict) %>% 
-  count() %>% 
-  pivot_wider(values_from = n, names_from = organism) %>% 
-  filter(!is.na(CellType_predict)) %>% 
-  replace(is.na(.), values = 0) %>% 
-  rename(`CellType (Predict)` = CellType_predict) %>% 
-  flextable()
 ```
@@
 ```{r}
-scEiaD_2020_v01 %>% 
-    tbl("diff_testing") %>% 
-    filter(Base == 'Cone') %>%  
-    filter(Group == 'CellType (Predict)', 
-           Against == 'All') %>% 
-  group_by(Gene) %>% 
-  summarise(Count = n(), log2FoldChange = mean(log2FoldChange), padj_count = sum(padj < 1e-4), padj = mean(padj)) %>% 
-  filter(!is.na(log2FoldChange), padj_count > 1) %>%  
-  arrange(-log2FoldChange)
+spec_diff <- diff_tab_big %>% 
+  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
+  group_by(Symbol, Organism, Base) %>% 
+  summarise(Count = n()) %>% 
+  summarise(Count = n())
+
+spec_diff_go <- spec_diff %>% 
+  filter(Count <= 3) %>% 
+  mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% 
+  pull(go)
+
+diff_tab_big %>% 
+  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
+  group_by(Symbol, Organism, Base) %>% 
+  summarise(Count = n(),
+            baseMean = mean(baseMean),
+            log2FoldChange = mean(log2FoldChange),
+            padj = mean(padj)) %>% 
+  filter(Count > 1) %>% 
+  left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
+  mutate(Ratio = `Count.x` / `Count.y`) %>% 
+  select(-`Count.x`,-`Count.y`) %>% 
+  #mutate(go = paste(Symbol, '___', Organism, sep = '')) %>% 
+  #filter(go %in% (spec_diff_go))  %>% 
+  #select(-go) %>% 
+  ungroup() %>% 
+  group_by(Organism, Base) %>% 
+  slice_max(n = 10, order_by = log2FoldChange)
 
@@
 
-### heatmap human
-```{r, fig.width=6, fig.height=17}
-# diff_ratios <- diff_tab_big %>% 
-#     filter(Against != 'All') %>% 
-#     group_by(Symbol, ID, Organism, Base) %>% 
-#     summarise(Count = n()) %>% 
-#     filter(Count > 1) %>% 
-#     left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
-#     mutate(Ratio = `Count.x` / `Count.y`) %>% 
-#     select(Symbol:Base, Ratio)
+### heatmap
+```{r, fig.width=5, fig.height=20}
+diff_ratios <- diff_tab_big %>% 
+    filter(Against != 'All') %>% 
+    group_by(Symbol, ID, Organism, Base) %>% 
+    summarise(Count = n()) %>% 
+    filter(Count > 1) %>% 
+    left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
+    mutate(Ratio = `Count.x` / `Count.y`) %>% 
+    select(Symbol:Base, Ratio)
 
 gene_sets <- diff_tab %>% 
-  #left_join(diff_ratios, by = c('Symbol', 'ID','Base','Organism')) %>% 
-  #filter(Ratio > 0.2) %>% 
-  group_by(Base, Organism) %>% 
-  slice_max(n=5, order_by = log2FoldChange)
+  left_join(diff_ratios, by = c('Symbol', 'ID','Base','Organism')) %>% 
+  filter(Ratio > 0.2) %>% group_by(Base, Organism) %>% 
+  slice_max(n=20, order_by = log2FoldChange)
 
-org <- 'Homo sapiens'
+g <- gene_sets %>% filter(Organism == 'Homo sapiens') %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
 
-g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
-
-# only uses CT that exist in the organism <-> CellType 
-ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()
-
-x <- scEiaD_2020_v01 %>% 
-  tbl("diff_testing") %>% 
-  filter(Gene %in% g) %>% 
-  collect() %>% 
-  filter(Base %in% ct,
-         Group == 'CellType (Predict)', 
-         Organism == org, Against == 'All') %>% 
-  select(Gene, log2FoldChange, Base) %>% 
-  pivot_wider(values_from=log2FoldChange, names_from = Base) 
-celltype <- colnames(x[,-1])
-x <- x %>% data.frame()
-row.names(x) <- x$Gene
-x <- x[,-1]
-Heatmap(x, column_labels = celltype)
-```
-
-### heatmap mouse
-```{r, fig.width=5, fig.height=20}
-
-org <- 'Mus musculus'
-
-g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
-
-# only uses CT that exist in the organism <-> CellType 
-ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()
-
-x <- scEiaD_2020_v01 %>% 
-  tbl("diff_testing") %>% 
-  filter(Gene %in% g) %>% 
-  collect() %>% 
-  filter(Base %in% ct,
-         Group == 'CellType (Predict)', 
-         Organism == org, Against == 'All') %>% 
-  select(Gene, log2FoldChange, Base) %>% 
-  pivot_wider(values_from=log2FoldChange, names_from = Base) 
-celltype <- colnames(x[,-1])
-x <- x %>% data.frame()
-row.names(x) <- x$Gene
-x <- x[,-1]
-Heatmap(x, column_labels = celltype)
-```
-
-### heatmap macaque
-```{r, fig.width=5, fig.height=20}
-
-org <- 'Macaca fascicularis'
-
-g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
-
-# only uses CT that exist in the organism <-> CellType 
-ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()
-
-x <- scEiaD_2020_v01 %>% 
-  tbl("diff_testing") %>% 
-  filter(Gene %in% g) %>% 
-  collect() %>% 
-  filter(Group == 'CellType (Predict)', 
-         Organism == org, Against == 'All', Base %in% ct) %>% 
-  select(Gene, log2FoldChange, Base) %>% 
-  pivot_wider(values_from=log2FoldChange, names_from = Base) 
+x <- scEiaD_2020_v01 %>% tbl("diff_testing") %>% filter(Gene %in% g) %>% collect() %>% filter(Group == 'CellType (Predict)', Organism == 'Homo sapiens', Against == 'All') %>% select(Gene, log2FoldChange, Base) %>% pivot_wider(values_from=log2FoldChange, names_from = Base)
 celltype <- colnames(x[,-1])
@@
 
-meta_filter %>% group_by(Citation, PMID, organism, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable()
+meta_filter %>% group_by(Citation, PMID, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable()
 ```

### 2022-06-17 heatmaps.Rmd

@@
   tbl("diff_testing") %>% 
-  filter(Base == 'RPE') %>%  
+  filter(Base == 'RPC') %>%  
   filter(Group == 'CellType (Predict)', 
          Organism == org, Against == 'All') %>% 
-  select(Gene, log2FoldChange, pvalue, Base) %>% 
+  select(Gene, log2FoldChange, pvalue, Base, baseMean) %>% 
   collect()
@@
   ggplot(aes(x=log2FoldChange,y=-log2(pvalue))) + 
-  geom_point() + cowplot::theme_cowplot() + ggrepel::geom_label_repel(data = rpe_diff %>% 
+  geom_point(aes(size=log2(baseMean))) + cowplot::theme_cowplot() + ggrepel::geom_label_repel(data = rpe_diff %>% 
                                                                         filter(Gene %in% genes) %>% 

### 2022-06-17 manuscript_00.Rmd

--- 2022-06-17 manuscript_00.Rmd
@@
 
-## Differential testing identifies thousands genes enriched across various ocular cell types
+## Leveraging the high diversity of studies to identify thousands of consistently differentially expressed genes across ocular cell types
 
-We used a "pseudo bulk" based approach wherein the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping.
-The creates a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices.
+We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping.
+The created a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices and thus allowing for mature bulk RNA-seq tooling to be used.
 This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021].
-Once the aggregated matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
+Once the aggregated matrix was con structured, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
 We tested each species (Human, Macaque, Mouse) separately.
-We had an insufficient number of chicken studies to conduct differential testing.
-The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential expression.
+We had an insufficient number of chick studies to conduct differential testing.
+The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential gene expression.
 After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g.
 cones versus all remaining cell types) or in a pair-wise manner (cone against rod).
-We find
+We find, in total, thousands of differentially expressed genes across the ocular cell types.
+
 ```{r}

### 2022-06-19 manuscript_00.Rmd
updates

@@
 
-diff_tab %>% 
+diff_tab_sup <- diff_tab %>% 
   group_by(Base, Organism) %>% 
@@
   pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
-  as_tibble() %>% 
+  as_tibble()
+write_csv(diff_tab_sup, file = 'supplemental_files/differential_expression_table.csv')
+diff_tab_sup %>% 
   flextable()
@@
                                       TRUE ~ 'No')) %>% 
-  flextable(set_caption())
+  flextable()
 ```
@@
 To define these well supported marker genes we only ass
+
+To identify genes that are differentially expressed across species we
+
 ```{r}
@@
   pull(CellType_predict)
-```
 
-To identify genes that are differentially expressed across species we
-
-```{r}
 consist_diff <- scEiaD_2020_v01 %>% 
@@
   left_join(gene_tab, by = 'GENEID') %>% 
-  select(-Gene) %>% 
   relocate(SYMBOL, GENEID) %>% 
@@
 consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n())
-```
-
-### heatmap human
-```{r, fig.width=6, fig.height=17}
-# diff_ratios <- diff_tab_big %>% 
-#     filter(Against != 'All') %>% 
-#     group_by(Symbol, ID, Organism, Base) %>% 
-#     summarise(Count = n()) %>% 
-#     filter(Count > 1) %>% 
-#     left_join(spec_diff, by = c("Symbol", "Organism")) %>% 
-#     mutate(Ratio = `Count.x` / `Count.y`) %>% 
-#     select(Symbol:Base, Ratio)
-
-gene_sets <- diff_tab %>% 
-  #left_join(diff_ratios, by = c('Symbol', 'ID','Base','Organism')) %>% 
-  #filter(Ratio > 0.2) %>% 
-  group_by(Base, Organism) %>% 
-  slice_max(n=5, order_by = log2FoldChange)
-
-org <- 'Homo sapiens'
-
-g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
-
-# only uses CT that exist in the organism <-> CellType 
-ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()
-
-x <- scEiaD_2020_v01 %>% 
-  tbl("diff_testing") %>% 
-  filter(Gene %in% g) %>% 
-  collect() %>% 
-  filter(Base %in% ct,
-         Group == 'CellType (Predict)', 
-         Organism == org, Against == 'All') %>% 
-  select(Gene, log2FoldChange, Base) %>% 
-  pivot_wider(values_from=log2FoldChange, names_from = Base) 
-celltype <- colnames(x[,-1])
-x <- x %>% data.frame()
-row.names(x) <- x$Gene
-x <- x[,-1]
-Heatmap(x, column_labels = celltype)
-```
-
-### heatmap mouse
-```{r, fig.width=5, fig.height=20}
-
-org <- 'Mus musculus'
-
-g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
-
-# only uses CT that exist in the organism <-> CellType 
-ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()
-
-x <- scEiaD_2020_v01 %>% 
-  tbl("diff_testing") %>% 
-  filter(Gene %in% g) %>% 
-  collect() %>% 
-  filter(Base %in% ct,
-         Group == 'CellType (Predict)', 
-         Organism == org, Against == 'All') %>% 
-  select(Gene, log2FoldChange, Base) %>% 
-  pivot_wider(values_from=log2FoldChange, names_from = Base) 
-celltype <- colnames(x[,-1])
-x <- x %>% data.frame()
-row.names(x) <- x$Gene
-x <- x[,-1]
-Heatmap(x, column_labels = celltype)
-```
-
-### heatmap macaque
-```{r, fig.width=5, fig.height=20}
-
-org <- 'Macaca fascicularis'
-
-g <- gene_sets %>% filter(Organism == org) %>%  mutate(gk = paste0(Symbol, ' (', ID, ')')) %>% pull(gk) %>% unique()
-
-# only uses CT that exist in the organism <-> CellType 
-ct <- meta_filter %>% filter(organism == org) %>% pull(CellType) %>% unique()
-
-x <- scEiaD_2020_v01 %>% 
-  tbl("diff_testing") %>% 
-  filter(Gene %in% g) %>% 
-  collect() %>% 
-  filter(Group == 'CellType (Predict)', 
-         Organism == org, Against == 'All', Base %in% ct) %>% 
-  select(Gene, log2FoldChange, Base) %>% 
-  pivot_wider(values_from=log2FoldChange, names_from = Base) 
-celltype <- colnames(x[,-1])
-x <- x %>% data.frame()
-row.names(x) <- x$Gene
-x <- x[,-1]
-Heatmap(x, column_labels = celltype)
 ```
@@
   collect() %>% 
-  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique())) %>% mutate(k = paste(Symbol, Base, sep = '_'))
+  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique())) %>% 
+  mutate(k = paste(Symbol, Base, sep = '_'))
 

### 2022-06-21 manuscript_00.Rmd
new fig

@@
 
+## Cell Type Proportions
+```{r, fig.width=5, fig.height=12}
+
+# map color as plae app does
+categorical_columns <- c("Phase","batch","study_accession","library_layout","organism","Platform",
+                         "Covariate","CellType","CellType_predict","TabulaMurisCellType","TabulaMurisCellType_predict",
+                         "GSE","Summary","Design","Citation","PMID","Stage","cluster",
+                         "Doublet","TechType", "SubCellType", 'subcluster', 'Age', "retina_region",
+                         'Tissue','Organ', 'Source','sample_accession')
+#"SubCellType" and subcluster are problems
+meta_filter <- meta_filter %>% mutate(Age = as.character(Age), SubCellType = tidyr::replace_na(SubCellType, 'None'),
+                                      subcluster = as.character(subcluster))
+
+map_color <- function(column, meta_filter){
+  master_colorlist <- c( pals::glasbey(), pals::alphabet(), pals::cols25()[1:23], pals::alphabet2()) %>% unique()
+  values <- meta_filter %>% pull(!!column) %>% unique %>% sort
+  if(length(values) > length(master_colorlist) ){
+    r= round(length(values) / length(master_colorlist)) +1
+    master_colorlist <- rep(master_colorlist, r)
+  }
+
+  colors <- master_colorlist[1:length(values)]
+  return(tibble(meta_category = column,value = values, color=colors))
+
+}
+
+cat_to_color_df <- lapply(categorical_columns, function(col) map_color(col, meta_filter)) %>% bind_rows()
+# now roll through and set each value to the same color across CellType and CellType (predict)
+# e.g. RPE will be one color in CellType and CellType (Predict)
+celltype_x_color_map <- cat_to_color_df %>% filter(meta_category %in% c("CellType", "CellType_predict")) %>% group_by(value) %>% summarise(c2 = head(color,1))
+cat_to_color_df <- cat_to_color_df %>% left_join(celltype_x_color_map) %>% mutate(color = case_when(!is.na(c2) ~ c2, TRUE ~ color)) %>% select(-c2)
+
+o_ct <- meta_filter %>% filter(Tissue == 'Retina') %>% pull(CellType)
+p_data <- meta_filter %>% 
+  #mutate(CellType_predict = Tissue) %>% 
+  #mutate(CellType_predict = case_when(!CellType_predict %in% ws_oct ~ 'Other',
+  #                                    TRUE ~ CellType_predict)) %>% 
+  filter(Organ == 'Eye') %>% 
+  mutate(Citation = str_trunc(Citation, width = 20)) %>% 
+  group_by(Citation, organism, Tissue, CellType_predict) %>% 
+  summarise(Count = n()) %>% 
+  mutate(Proportion = Count/sum(Count) * 100)
+
+meta_column <- 'CellType_predict'
+cur_color_df <- cat_to_color_df %>%
+  filter( meta_category %in%  meta_column,
+          value %in% p_data[[meta_column]]) %>% distinct
+
+color_list <- cur_color_df %>% pull(color)
+# replaced join with this for speed
+k <- cur_color_df$color
+names(k) <- cur_color_df$value
+np_color <- {k[p_data[[meta_column]] ]} %>% paste0(., '33')
+names(np_color) <- NULL
+color_data <- cur_color_df  %>%
+  select(value) %>%
+  mutate( x=0)
+names(color_list) <- color_data$value
+
+p_data %>% ggplot(aes(y=Citation,x=Proportion,fill=CellType_predict)) + geom_bar(stat= 'identity') +
+  scale_fill_manual(name= meta_column,
+                    values = color_list) + 
+  ggforce::facet_col(vars(organism, Tissue), scale = 'free', space = 'free') +
+  cowplot::theme_cowplot() +
+  guides(fill=guide_legend(ncol =1))
+```
 ## Leveraging the high diversity of studies to identify thousands of consistently differentially expressed genes across ocular cell types
@@
 
+```{r, fig.height=8, fig.width=6}
+g <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% slice_max(n=5, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
+
+test <- scEiaD_2020_v01 %>%
+  tbl("diff_testing") %>%
+  filter(Gene %in% g) %>%
+  collect() %>%
+  filter(Base %in% ws_oct,
+         Group == 'CellType (Predict)',
+         Against == 'All') %>%
+  select(Gene, Organism, log2FoldChange, Base)
+
+x <- test %>% 
+  mutate(GO = paste0(Base, '_', Organism)) %>% 
+  select(-Organism, -Base) %>% 
+  pivot_wider(values_from=log2FoldChange, names_from = GO)
+celltype <- colnames(x[,-1])
+x <- x %>% data.frame()
+row.names(x) <- x$Gene
+
+x <- x[,-1]
+
+organism <- str_extract(celltype, '_\\w+ \\w+') %>% gsub('_','',.)
+cts  <- str_extract(celltype, '.*_') %>% gsub('_','',.)
+genes <- row.names(x) %>% gsub(' \\(.*','',.)
+
+base_yo <- consist_diff %>% filter(Symbol %in% genes) %>% pull(Base)
+
+hm <- ComplexHeatmap::Heatmap(x, column_labels = organism, 
+                              column_split = cts, 
+                              row_split = base_yo, 
+                              row_labels = genes, 
+                              column_title_rot = 90, 
+                              name = 'log2FC', use_raster = TRUE,
+                              row_title_rot = 0)
+ComplexHeatmap::draw(hm, padding = unit(c(2, 2, 30, 2), "mm"))
+```
 ```{r}

### 2022-06-21 manuscript_00.Rmd
rejigger

@@
 library(glue)
-data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
-data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
+#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 # setup caption-ing
@@
 c_d <- plot_grid(c,d, rel_heights = c(3,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))
-plot_grid(a_aa_b, c_d, ncol = 2, rel_widths = c(1,1.6))
+
+# make E now
+
+# map color as plae app does
+categorical_columns <- c("Phase","batch","study_accession","library_layout","organism","Platform",
+                         "Covariate","CellType","CellType_predict","TabulaMurisCellType","TabulaMurisCellType_predict",
+                         "GSE","Summary","Design","Citation","PMID","Stage","cluster",
+                         "Doublet","TechType", "SubCellType", 'subcluster', 'Age', "retina_region",
+                         'Tissue','Organ', 'Source','sample_accession')
+#"SubCellType" and subcluster are problems
+meta_filter <- meta_filter %>% mutate(Age = as.character(Age), SubCellType = tidyr::replace_na(SubCellType, 'None'),
+                                      subcluster = as.character(subcluster))
+
+map_color <- function(column, meta_filter){
+  master_colorlist <- c( pals::glasbey(), pals::alphabet(), pals::cols25()[1:23], pals::alphabet2()) %>% unique()
+  values <- meta_filter %>% pull(!!column) %>% unique %>% sort
+  if(length(values) > length(master_colorlist) ){
+    r= round(length(values) / length(master_colorlist)) +1
+    master_colorlist <- rep(master_colorlist, r)
+  }
+
+  colors <- master_colorlist[1:length(values)]
+  return(tibble(meta_category = column,value = values, color=colors))
+
+}
+
+cat_to_color_df <- lapply(categorical_columns, function(col) map_color(col, meta_filter)) %>% bind_rows()
+# now roll through and set each value to the same color across CellType and CellType (predict)
+# e.g. RPE will be one color in CellType and CellType (Predict)
+celltype_x_color_map <- cat_to_color_df %>% filter(meta_category %in% c("CellType", "CellType_predict")) %>% group_by(value) %>% summarise(c2 = head(color,1))
+cat_to_color_df <- cat_to_color_df %>% left_join(celltype_x_color_map) %>% mutate(color = case_when(!is.na(c2) ~ c2, TRUE ~ color)) %>% select(-c2)
+
+o_ct <- meta_filter %>% filter(Tissue == 'Retina') %>% pull(CellType)
+p_data <- meta_filter %>% 
+  #mutate(CellType_predict = Tissue) %>% 
+  #mutate(CellType_predict = case_when(!CellType_predict %in% ws_oct ~ 'Other',
+  #                                    TRUE ~ CellType_predict)) %>% 
+  filter(Organ == 'Eye') %>% 
+  mutate(Citation = paste(str_trunc(Citation, width = 20), str_trunc(batch, side = "left", width = 15))) %>% 
+  group_by(Citation, organism, Tissue, CellType_predict) %>% 
+  summarise(Count = n()) %>% 
+  mutate(Proportion = Count/sum(Count) * 100)
+
+meta_column <- 'CellType_predict'
+cur_color_df <- cat_to_color_df %>%
+  filter( meta_category %in%  meta_column,
+          value %in% p_data[[meta_column]]) %>% distinct
+
+color_list <- cur_color_df %>% pull(color)
+# replaced join with this for speed
+k <- cur_color_df$color
+names(k) <- cur_color_df$value
+np_color <- {k[p_data[[meta_column]] ]} %>% paste0(., '33')
+names(np_color) <- NULL
+color_data <- cur_color_df  %>%
+  select(value) %>%
+  mutate( x=0)
+names(color_list) <- color_data$value
+
+proportion_bar_plot <- p_data %>% ggplot(aes(y=Citation,x=Proportion,fill=CellType_predict)) + geom_bar(stat= 'identity') +
+  scale_fill_manual(name= meta_column,
+                    values = color_list) + 
+  ggforce::facet_col(vars(organism, Tissue), scale = 'free', space = 'free') +
+  cowplot::theme_cowplot() +
+  guides(fill=guide_legend(ncol =1))
+
+plot_grid(a_aa_b, c_d, proportion_bar_plot, ncol = 3, rel_widths = c(1,1.6, 3))
 ```
@@
 
-## Cell Type Proportions
-```{r, fig.width=5, fig.height=12}
-
-# map color as plae app does
-categorical_columns <- c("Phase","batch","study_accession","library_layout","organism","Platform",
-                         "Covariate","CellType","CellType_predict","TabulaMurisCellType","TabulaMurisCellType_predict",
-                         "GSE","Summary","Design","Citation","PMID","Stage","cluster",
-                         "Doublet","TechType", "SubCellType", 'subcluster', 'Age', "retina_region",
-                         'Tissue','Organ', 'Source','sample_accession')
-#"SubCellType" and subcluster are problems
-meta_filter <- meta_filter %>% mutate(Age = as.character(Age), SubCellType = tidyr::replace_na(SubCellType, 'None'),
-                                      subcluster = as.character(subcluster))
-
-map_color <- function(column, meta_filter){
-  master_colorlist <- c( pals::glasbey(), pals::alphabet(), pals::cols25()[1:23], pals::alphabet2()) %>% unique()
-  values <- meta_filter %>% pull(!!column) %>% unique %>% sort
-  if(length(values) > length(master_colorlist) ){
-    r= round(length(values) / length(master_colorlist)) +1
-    master_colorlist <- rep(master_colorlist, r)
-  }
-
-  colors <- master_colorlist[1:length(values)]
-  return(tibble(meta_category = column,value = values, color=colors))
-
-}
-
-cat_to_color_df <- lapply(categorical_columns, function(col) map_color(col, meta_filter)) %>% bind_rows()
-# now roll through and set each value to the same color across CellType and CellType (predict)
-# e.g. RPE will be one color in CellType and CellType (Predict)
-celltype_x_color_map <- cat_to_color_df %>% filter(meta_category %in% c("CellType", "CellType_predict")) %>% group_by(value) %>% summarise(c2 = head(color,1))
-cat_to_color_df <- cat_to_color_df %>% left_join(celltype_x_color_map) %>% mutate(color = case_when(!is.na(c2) ~ c2, TRUE ~ color)) %>% select(-c2)
-
-o_ct <- meta_filter %>% filter(Tissue == 'Retina') %>% pull(CellType)
-p_data <- meta_filter %>% 
-  #mutate(CellType_predict = Tissue) %>% 
-  #mutate(CellType_predict = case_when(!CellType_predict %in% ws_oct ~ 'Other',
-  #                                    TRUE ~ CellType_predict)) %>% 
-  filter(Organ == 'Eye') %>% 
-  mutate(Citation = str_trunc(Citation, width = 20)) %>% 
-  group_by(Citation, organism, Tissue, CellType_predict) %>% 
-  summarise(Count = n()) %>% 
-  mutate(Proportion = Count/sum(Count) * 100)
-
-meta_column <- 'CellType_predict'
-cur_color_df <- cat_to_color_df %>%
-  filter( meta_category %in%  meta_column,
-          value %in% p_data[[meta_column]]) %>% distinct
-
-color_list <- cur_color_df %>% pull(color)
-# replaced join with this for speed
-k <- cur_color_df$color
-names(k) <- cur_color_df$value
-np_color <- {k[p_data[[meta_column]] ]} %>% paste0(., '33')
-names(np_color) <- NULL
-color_data <- cur_color_df  %>%
-  select(value) %>%
-  mutate( x=0)
-names(color_list) <- color_data$value
-
-p_data %>% ggplot(aes(y=Citation,x=Proportion,fill=CellType_predict)) + geom_bar(stat= 'identity') +
-  scale_fill_manual(name= meta_column,
-                    values = color_list) + 
-  ggforce::facet_col(vars(organism, Tissue), scale = 'free', space = 'free') +
-  cowplot::theme_cowplot() +
-  guides(fill=guide_legend(ncol =1))
-```
 ## Leveraging the high diversity of studies to identify thousands of consistently differentially expressed genes across ocular cell types
@@
 # number of celltype (predict) - organism - study counts (with >=50 cells)
-meta_filter %>%  
+ws_table <- meta_filter %>%  
   filter(Organ == 'Eye') %>% 
@@
   rename(`CellType (Predict)` = CellType_predict) %>% 
-  mutate(`Well Supported` = case_when(`Homo sapiens` > 2 & `Mus musculus` > 2 ~ 'Yes',
-                                      TRUE ~ 'No')) %>% 
+  mutate(`Well Supported` = case_when(`Homo sapiens` >= 2 & `Mus musculus` >= 2 ~ 'Yes',
+                                      TRUE ~ 'No'))
+ws_table %>% 
   flextable()
@@
 
-# Well supported ocular cell types identify
+## Well supported ocular cell types identify
 
@@
 ```{r}
-ws_oct <- meta_filter %>%  
-  filter(Organ == 'Eye') %>% 
-  group_by(organism, CellType_predict, study_accession) %>% 
-  count() %>% 
-  filter(n >= 50) %>% 
-  ungroup() %>% 
-  group_by(organism, CellType_predict) %>% 
-  count() %>% filter(n >= 3) %>% 
-  ungroup() %>% 
-  group_by(CellType_predict) %>% 
-  summarise(Count = n()) %>% 
-  filter(Count > 1, !is.na(CellType_predict)) %>% 
-  pull(CellType_predict)
+ws_oct <- ws_table %>% filter(`Well Supported` == 'Yes') %>% pull(`CellType (Predict)`)
 
@@
 
-```{r, fig.height=8, fig.width=6}
+```{r, fig.height=9, fig.width=7}
 g <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% slice_max(n=5, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
@@
 
-base_yo <- consist_diff %>% filter(Symbol %in% genes) %>% pull(Base)
+base_yo <- genes %>% as_tibble() %>% rename(Symbol = value) %>% left_join(consist_diff %>% select(Symbol, Base) %>% unique()) %>% pull(Base)
 
@@
                               column_split = cts, 
-                              row_split = base_yo, 
+                              row_split = base_yo, show_row_dend = FALSE, 
                               row_labels = genes, 
@@
 ```
+
+## Tease apart closely related cell types
 ```{r}
-# top 100 genes in each organism/base category
-diff_tab %>%
+x <- scEiaD_2020_v01 %>% 
+  tbl("diff_testing") %>% 
+  filter(Base %in% c("Early RPC", "Late RPC", "RPC")) %>%  
+  filter(Group == 'CellType (Predict)', 
+         Against %in% c("Early RPC", "Late RPC", "RPC")) %>% 
+  group_by(Base, Against, Gene) %>% 
+  summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-3)) %>% 
+  arrange(-`mean log2FC`) %>%   
+  collect() %>% 
+  filter(!is.na(`mean log2FC`))
 
-  group_by(Base, Organism) %>% 
-  top_n(., 100, -padj) %>% 
-  arrange(Base, Organism) %>% 
-  relocate(Base, Organism) %>% 
-  select(-Against, -Group)
+  mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
+  left_join(gene_tab, by = 'GENEID') %>% 
+  relocate(SYMBOL, GENEID) %>% 
+  dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
+  arrange(-`mean log2FC`)
 ```

### 2022-06-23 manuscript_00.Rmd
new plots

@@
 library(glue)
-#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
-#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
+data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 # setup caption-ing
@@
 a_aa_b <- plot_grid(a,aa,b, ncol = 1, align = 'hv', rel_heights = c(10, 1.5,3.5), labels = 'a')
-c_d <- plot_grid(c,d, rel_heights = c(3,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))
+c_d <- plot_grid(c,d + theme(legend.position = "none"), rel_heights = c(3,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))
 
@@
 
-plot_grid(a_aa_b, c_d, proportion_bar_plot, ncol = 3, rel_widths = c(1,1.6, 3))
+e <- plot_grid(proportion_bar_plot, labels = c('d'))
+
+plot_grid(a_aa_b, c_d , e, ncol = 3, rel_widths = c(1,1.6, 2.2))
 ```
@@
 diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
+down_tab_big <- read_tsv('data/down_tab_big.tsv.gz')
 gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
@@
 
-```{r, fig.height=9, fig.width=7}
-g <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% slice_max(n=5, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
+```{r, fig.height=11, fig.width=6}
+g <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% slice_max(n=8, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
 
@@
 ## Tease apart closely related cell types
+
+Neurogenic and RPC have very similar expression patterns.
+We leverage our more specific differential test to directly assess genes were are different between neurogenic cells and RPC.
+We show the top 5 genes that are both more highly expressed in RPC relative to neurogenic cell and vice versa.
 ```{r}
@@
     tbl("diff_testing") %>% 
-  filter(Base %in% c("Early RPC", "Late RPC", "RPC")) %>%  
-  filter(Group == 'CellType (Predict)', 
-         Against %in% c("Early RPC", "Late RPC", "RPC")) %>% 
+    filter(Against  == "RPC",   
+           Group == 'CellType (Predict)', 
+           Base == "Neurogenic Cell") %>% 
   group_by(Base, Against, Gene) %>% 
-  summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-3)) %>% 
+  summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
   arrange(-`mean log2FC`) %>%   
@@
 
-  mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
-  left_join(gene_tab, by = 'GENEID') %>% 
-  relocate(SYMBOL, GENEID) %>% 
-  dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
-  arrange(-`mean log2FC`)
+  # mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
+  # left_join(gene_tab, by = 'GENEID') %>% 
+  # relocate(SYMBOL, GENEID) %>% 
+  # dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
+  # arrange(-`mean log2FC`)
+
+down <- x %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` < 0) %>% head(5) %>% pull(Gene)
+up <-  x %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` > 0) %>% head(5) %>% pull(Gene)
+source('src/make_exp_plot.R')
+
+input <- list()
+input$exp_plot_groups <- c('study_accession')
+input$exp_plot_facet <- c('CellType_predict')
+input$exp_plot_genes <- down
+input$exp_filter_cat <- 'CellType_predict'
+#input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
+input$exp_filter_on <- c("Neurogenic Cell", "RPC" )
+input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
+#input$exp_plot_ylab <- '% of Cells Detected'
+input$exp_plot_col_num <- 16
+input$exp_filter_min_cell_number <- 50
+input$flip_facet <- TRUE
+
+a <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+
+input$exp_plot_genes <- up
+b <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) 
+plot_grid(a,b, labels = c('a','b'))
 ```
@@
 
-```{r}
+```{r, fig.width=3, fig.height=7}
 # flip flop diff genes
-down_tab <- scEiaD_2020_v01 %>% tbl('diff_testing') %>% 
-  filter(Group == 'CellType (Predict)', padj < 1e-5, (log2FoldChange) < -2) %>% 
+down_tab <- down_tab_big %>% 
+   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
+  filter(Base %in% ws_oct, 
+         Against == 'All')
+
+org_diff_diff_genes <- diff_tab_big %>% 
+  mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
+  filter(gene_ct_combo %in% down_tab$gene_ct_combo, 
+         Against == 'All') %>% 
+  arrange(padj) %>% mutate(x = paste0(Symbol, ' (', ID, ')'))
+
+genes <- scEiaD_2020_v01 %>% tbl('diff_testing_genes') %>% collect() %>% filter(grepl(org_diff_diff_genes %>% pull(ID) %>% paste0(., collapse = '|'), Gene)) %>% pull(Gene)
+test <- scEiaD_2020_v01 %>%
+  tbl("diff_testing") %>%
+  filter(Gene %in% genes) %>%
   collect() %>%
-  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType_predict) %>% unique())) %>% 
-  mutate(k = paste(Symbol, Base, sep = '_'))
+  mutate(Symbol = str_extract(Gene, '^\\w+ ') %>% gsub(' $', '', .)) %>% 
+  mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
+  filter(gene_ct_combo %in% org_diff_diff_genes$gene_ct_combo,
+         Group == 'CellType (Predict)',
+         Against == 'All') %>%
+  select(Gene, Organism, log2FoldChange, Base)
 
-diff_tab %>% mutate(k = paste(Symbol, Base, sep = '_')) %>% 
-  filter(k %in% down_tab$k, Base %in% c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )) %>%
-  arrange(padj)
+x <- test %>% pivot_wider(values_from = log2FoldChange, names_from = Organism)
+organism <- colnames(x)[c(3,4,5)]
+x <- x %>% data.frame()
+row.names(x) <- paste0(x$Gene,'_', x$Base)
 
-diff_tab_big %>% 
-  filter(Symbol %in% diff_tab$Symbol) %>% # only retain diff genes across "All"
-  group_by(Symbol, Base, Organism) %>% 
-  summarise(Count = n()) %>% 
-  filter(Count > 8, Base == 'Rod') %>% arrange(-Count)
+x <- x[,-c(1,2)]
+
+genes <- row.names(x) %>% gsub(' \\(.*','',.)
+
+base_yo <- row.names(x) %>% gsub('.*_','',.)
+
+hm <- ComplexHeatmap::Heatmap(x, column_labels = organism, 
+                              row_split = base_yo, show_row_dend = FALSE, 
+                              row_labels = genes, 
+                              column_title_rot = 90, 
+                              name = 'log2FC', use_raster = TRUE,
+                              row_title_rot = 0)
+hm
 ```
 
-```{r diff_difff_exp_genes, fig.height = 3, fig.width = 8, echo = F, message = F}
-source('src/make_exp_plot.R')
+```{r}
+library(enrichR)
+dbs <- c('GO_Biological_Process_2018', 'GO_Molecular_Function_2018')
+enriched_cone <- enrichr(org_diff_diff_genes %>% filter(Base == 'Rod') %>% pull(Symbol), dbs)
 
-input <- list()
-input$exp_plot_groups <- c('organism')
-input$exp_plot_facet <- c('CellType_predict')
-input$exp_plot_genes <- c('SCRG1 (ENSG00000164106)')
-input$exp_filter_cat <- 'CellType_predict'
-input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
-input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
-input$exp_plot_col_num <- 16
-input$exp_filter_min_cell_number <- 50
-
-make_exp_plot(input, scEiaD_2020_v01, meta_filter)
 ```
@@
 input$exp_filter_min_cell_number <- 50
+input$flip_facet <- FALSE
 

### 2022-06-23 src/ensdb.R

@@ -41,3 +41,29 @@ diff_tab_big <- bind_rows(
 )
 
 write_tsv(diff_tab_big, file = 'data/diff_tab_big.tsv.gz')
+
+
+# down tab
+
+down_tab_big <- scEiaD_2020_v01 %>% tbl('diff_testing') %>% 
+  filter(Group == 'CellType (Predict)', padj < 1e-5, (log2FoldChange) < -2) %>% 
+  collect() %>% 
+  separate(Gene, into = c('Symbol', 'ID'), sep = ' ') %>% 
+  mutate(ID = gsub('\\(|\\)', '', ID))
+
+down_tab_big2 <- down_tab_big %>% 
+  filter(Symbol == '-' | Symbol == 'NA')
+
+down_tab_big2 <- 
+  down_tab_big2 %>% 
+  left_join(hs_mm, by = c('ID' = 'GENEID')) %>% 
+  select(-Symbol) %>% 
+  rename(Symbol = SYMBOL) %>% 
+  relocate(Symbol)
+
+down_tab_big <- bind_rows(
+  down_tab_big %>% filter(!Symbol %in% c('-', 'NA')),
+  diff_tab_big2
+)
+
+write_tsv(down_tab_big, file = 'data/down_tab_big.tsv.gz')

### 2022-06-23 src/make_exp_plot.R

@@ -57,9 +57,9 @@ make_exp_plot <- function(input, db, meta_filter){
   #cat(input)
   box_data <- box_data %>%
     filter(!Platform %in% c('SCRBSeq')) %>%
-    mutate(CellType_predict = case_when(CellType_predict == 'RPC' ~ 'RPCs',
-                                        CellType_predict == 'Mesenchymal/RPE/Endothelial' ~ 'Endothelial',
-                                        TRUE ~ CellType_predict)) %>%
+    # mutate(CellType_predict = case_when(CellType_predict == 'RPC' ~ 'RPCs',
+    #                                     CellType_predict == 'Mesenchymal/RPE/Endothelial' ~ 'Endothelial',
+    #                                     TRUE ~ CellType_predict)) %>%
     mutate(Stage = factor(Stage, levels = c('Early Dev.', 'Late Dev.', 'Maturing', 'Mature'))) %>%
     #filter(!is.na(!!as.symbol(grouping_features))) %>%
     group_by_at(vars(one_of(c('Gene', input$exp_plot_facet, grouping_features)))) %>%
@@ -92,14 +92,30 @@ make_exp_plot <- function(input, db, meta_filter){
   }
   box_data$Group <- box_data[,c(2:(length(grouping_features)+1))] %>% tidyr::unite(x, sep = ' ') %>% pull(1)
   
-  box_data %>%
-    ggplot(aes(x=Gene, y = !!as.symbol(input$exp_plot_ylab), color = !!as.symbol(grouping_features))) +
-    geom_boxplot(color = 'black', outlier.shape = NA) +
-    ggbeeswarm::geom_quasirandom(aes(size = 8), groupOnX = TRUE) +
-    cowplot::theme_cowplot() +
-    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
-    scale_radius(range=c(2, 6)) +
-    scale_colour_manual(values = rep(c(pals::alphabet() %>% unname()), 20)) +
-    theme(legend.position="bottom") +
-    facet_wrap(ncol = as.numeric(input$exp_plot_col_num), scales = 'free_x', vars(!!as.symbol(input$exp_plot_facet)))
+  if (input$flip_facet){
+    box_data  %>% 
+      mutate(Gene = str_extract(Gene, '^\\w+ ') %>% gsub(" $", '', .)) %>%
[…]

### 2022-06-24 manuscript_00.Rmd
more

@@
   flextable()
+
+diff_tab_sup %>% pivot_longer(-`CellType (Predict)`, names_to = 'Organism', values_to = 'Diff Gene Count') %>% ggplot(aes(x=`CellType (Predict)`, y = `Diff Gene Count`, fill = Organism )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip()
 ```
@@
 consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n())
+
+consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n()) %>% ggplot(aes(x=`Base`, y = `Count` )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip() + xlab("CellTye (Predict)")
 ```
@@
                               row_labels = genes, 
-                              column_title_rot = 90, 
+                              column_title_rot = 90, row_split
                               name = 'log2FC', use_raster = TRUE,
                               row_title_rot = 0)
-ComplexHeatmap::draw(hm, padding = unit(c(2, 2, 30, 2), "mm"))
+ComplexHeatmap::draw(hm, padding = unit(c(2, 2, 32, 2), "mm"), row_title = "Marker Genes")
 ```
@@
 We show the top 5 genes that are both more highly expressed in RPC relative to neurogenic cell and vice versa.
-```{r}
+```{r, fig.width=7, fig.height=3}
 x <-  scEiaD_2020_v01 %>% 
@@
 
-```{r}
-# consistently diff genes
-consistent <- diff_tab %>% 
-  group_by(Symbol, Base)  %>% 
-  summarise(Count = n()) %>% 
-  filter(Count > 1) %>% 
-  mutate(k = paste(Symbol, Base, sep = '_'))
-
-diff_tab %>% mutate(k = paste(Symbol, Base, sep = '_')) %>% 
-  filter(k %in% consistent$k) %>% 
-  group_by(Base, Organism) %>% 
-  summarise(Count = n()) %>% 
-  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
-  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
-  as_tibble() %>% 
-  flextable()
-
-```
-
 ```{r, fig.width=3, fig.height=7}
@@
 
-hm <- ComplexHeatmap::Heatmap(x, column_labels = organism, 
+hm_flipflop <- ComplexHeatmap::Heatmap(x, column_labels = organism, 
                               row_split = base_yo, show_row_dend = FALSE, 
@@
                               row_title_rot = 0)
-hm
+hm_flipflop
 ```

### 2022-06-24 temp.rmd
more

---
title: "R Notebook"
output: html_notebook
---

This is an [R Markdown](http://rmarkdown.rstudio.com) Notebook. When you execute code within the notebook, the results appear beneath the code.

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*.

```{r scEiaD_data_overview, fig.height = 18, fig.width = 10, echo = F, message = F}
plot_grid(a_aa_b, c_d , e, ncol = 3, rel_widths = c(1,1.6, 2.2))
```

```{r, fig.height=11, fig.width=6}
ComplexHeatmap::draw(hm, padding = unit(c(2, 2, 32, 2), "mm"))
```

```{r, fig.width=3, fig.height=7}
hm_flipflop
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

### 2022-06-24 src/make_exp_plot.R

@@ -97,7 +97,7 @@ make_exp_plot <- function(input, db, meta_filter){
       mutate(Gene = str_extract(Gene, '^\\w+ ') %>% gsub(" $", '', .)) %>%
       ggplot(aes(x=!!as.symbol(input$exp_plot_facet), y = !!as.symbol(input$exp_plot_ylab), color = !!as.symbol(grouping_features))) +
       geom_boxplot(color = 'black', outlier.shape = NA) +
-      ggbeeswarm::geom_quasirandom(aes(size = 8), groupOnX = TRUE) +
+      ggbeeswarm::geom_quasirandom(size = 8, groupOnX = TRUE) +
       cowplot::theme_cowplot() +
       theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
       scale_radius(range=c(2, 6)) +
@@ -110,7 +110,7 @@ make_exp_plot <- function(input, db, meta_filter){
       mutate(Gene = str_extract(Gene, '^\\w+ ') %>% gsub(" $", '', .)) %>%
       ggplot(aes(x=Gene, y = !!as.symbol(input$exp_plot_ylab), color = !!as.symbol(grouping_features))) +
       geom_boxplot(color = 'black', outlier.shape = NA) +
-      ggbeeswarm::geom_quasirandom(aes(size = 8), groupOnX = TRUE) +
+      ggbeeswarm::geom_quasirandom(size = 8, groupOnX = TRUE) +
       cowplot::theme_cowplot() +
       theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
       scale_radius(range=c(2, 6)) +

### 2022-06-25 manuscript_00.Rmd
swap in covariate for plot

@@
   filter(Organ == 'Eye') %>% 
-  mutate(Citation = paste(str_trunc(Citation, width = 20), str_trunc(batch, side = "left", width = 15))) %>% 
+  mutate(Citation = paste(str_trunc(Citation, width = 20), Covariate)) %>% 
   group_by(Citation, organism, Tissue, CellType_predict) %>% 

### 2022-06-25 src/make_exp_plot.R
dot size

@@ -97,7 +97,7 @@ make_exp_plot <- function(input, db, meta_filter){
       mutate(Gene = str_extract(Gene, '^\\w+ ') %>% gsub(" $", '', .)) %>%
       ggplot(aes(x=!!as.symbol(input$exp_plot_facet), y = !!as.symbol(input$exp_plot_ylab), color = !!as.symbol(grouping_features))) +
       geom_boxplot(color = 'black', outlier.shape = NA) +
-      ggbeeswarm::geom_quasirandom(size = 8, groupOnX = TRUE) +
+      ggbeeswarm::geom_quasirandom(size = 4, groupOnX = TRUE) +
       cowplot::theme_cowplot() +
       theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
       scale_radius(range=c(2, 6)) +
@@ -110,7 +110,7 @@ make_exp_plot <- function(input, db, meta_filter){
       mutate(Gene = str_extract(Gene, '^\\w+ ') %>% gsub(" $", '', .)) %>%
       ggplot(aes(x=Gene, y = !!as.symbol(input$exp_plot_ylab), color = !!as.symbol(grouping_features))) +
       geom_boxplot(color = 'black', outlier.shape = NA) +
-      ggbeeswarm::geom_quasirandom(size = 8, groupOnX = TRUE) +
+      ggbeeswarm::geom_quasirandom(size = 4, groupOnX = TRUE) +
       cowplot::theme_cowplot() +
       theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
       scale_radius(range=c(2, 6)) +

### 2022-07-05 manuscript_00.Rmd
updates

@@
 As a test of the value of this unified pan-eye dataset, we show XX.
-CONCLUSION: The PLAE v0.90 web app serves the pan-ocular and body dataset, scEiaD.
+CONCLUSION: The PLAE v0.91 web app serves the pan-ocular and body dataset, scEiaD.
 This offers the eye community a powerful and quick means to test hypotheses on gene and transcript expression across 54 body and 19 eye tissues.
@@
 library(glue)
+library(tictoc)
+tic()
 data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
@@
 
+toc()
 ```
@@
 
-## single cell RNA-seq allows for a higher resolution view of ocular transcriptomics
-
-Single cell RNA-sequencing dates back to . The first
-
-## Many published ocular single cell RNA-seq atlases
+  Examples of single cell transcriptomics the retina date begin in the early 2000s where researchers [@hagstrom_cone_2000; @jabs_evidence_2000; @li_localization_2004; @lindqvist_single_2002-1; @reid_retinoschisin_2003] generally used RT-PCR approaches to amplify select transcripts in individual cells.
+Later, researchers used microarray platforms to analyze a broader array of transcripts researchers [@cherry_development_2009; @gustincich_gene_2004-1; @kim_identification_2008-1; @roesch_transcriptome_2008; @trimarchi_individual_2008; @trimarchi_molecular_2007-1; @wahlin_method_2004].
+The modern era of single cell transcriptomics is denoted by the introduction of droplet-based technologies , which enabled quantitation of thousands of cells in a single experiment; in Macoksko et al.
+the adult mouse retina was profiled [@klein_droplet_2015; @macosko_highly_2015].
+The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by producing over 30 single cell transcriptome based studies of the eye since the Macosko publication.
+  These publications's foci includes, but is not limited to, studying developmental transcriptomic dynamics, identifying genes to distinguish core cell types, proposing finer gradations of cell types within existing categories, and comparing ocular transcriptomes between organisms.
+Outside querying of these datasets often requires a long series of steps.
+Briefly, one would have to identify the relevant publication, find the associated data deposit, download the cell - count matrix, find the cell label table (and optionally the t-SNE or UMAP coordinates), and then load the data into R or python to run the desired queries.
+If one desired to compare *across* publications, then one would also have to normalize the cell type labels and optionally run a batch-effect removal algorithm.
+It is desirable, perhaps even necessary, to use multiple datasets to assess gene expression to confirm whether a gene expression is consistent and reproducible.
+Reliance on a single dataset can be problematic as there can be technical issues - for example the large consortium dataset GTEx has persistant contamination from highly expressed genes across tissues [@nieuwenhuis_consistent_2020].
+  Web-based resources to query gene expression are powerful means to democratize access to large datasets as the computationally intensive steps above can be run once and then shared to all.
+Single cell RNA-seq resources that contain ocular tissues include the UCSC Cell Browser, Tabula Muris, PanglaoDB, and the Protein Atlas [@franzen_panglaodb_2019; @karlsson_singlecell_2021; @noauthor_single-cell_nodate; @speir_ucsc_2021].
+Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging.
+The Spectacle project is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigt_spectacle_2020].
+Each data deposit is independently presented, which makes cross study queries difficult.
+We recently created a unified ocular resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space [@swamy_building_2021].
+We have built upon this dataset, adding more datasets, improving the quality control, removing ambient noise, improved the differential gene testing structure, and finally created a powerful, reactive, and fast web app to query the database.
+We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, genes which have species-specific expression cell type patterning, propose genes XYZ differentially expressed between the fovea and peripheral cones, and
 
@@
 [@swamy_building_2021].
-Very briefly, we collect the raw fastq files and quantify with the kallisto / bustools system [@melsted_barcode_2019; @melsted_modular_2019].
-The quantified counts are merged into one matrix that is used for downstream processing with scvitools to batch correct the data and Seurat for quality control, clustering, and 2D UMAP visualization.
+Very briefly, we collected the raw fastq files and quantified with the kallisto / bustools system [@melsted_barcode_2019; @melsted_modular_2019].
+The quantified counts are merged into one matrix that is used for downstream processing with scvitools create a batch corrected lower dimensional space.
+and then into Seurat for quality control, clustering, and 2D UMAP visualization.
 We use our previous xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamy_building_2021].
@@
 ```{r scEiaD_data_overview, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=scEiaD_data_overview_cap}
-
+tic()
 num1 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(Tissue) %>% unique() %>% length()
@@
   filter(Organ == 'Eye') %>% 
-  mutate(Citation = paste(str_trunc(Citation, width = 20), Covariate)) %>% 
+  mutate(Citation = paste(str_trunc(Citation, width = 20), str_trunc(Covariate, width = 12, side = 'left',ellipsis = ''))) %>% 
   group_by(Citation, organism, Tissue, CellType_predict) %>% 
@@
 plot_grid(a_aa_b, c_d , e, ncol = 3, rel_widths = c(1,1.6, 2.2))
+toc()
 ```
@@
 ```{r web_resources, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=web_resources_cap}
-
+tic()
 web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
@@
 # as of 2021 10 25
+toc()
 ```
@@
 
-While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently [@noauthor_spectacle_nodate].
-PLAE is built around the scEiaD database, which runs an extensive multi-stage pipeline to build and identify a high performing scVI-based eye model that is used to integrate all datasets together [@swamy_building_2021].
+While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently [@voigt_spectacle_2020].
+PLAE is built around the scEiaD database, which runs an multi-stage pipeline to build and identify a high performing scVI-based eye model that is used to integrate all datasets together [@swamy_building_2021].
 This allows for gene queries to be constructed and analyzed across studies.
-In contrast, Spectacle, as of March 2022, is a compilation of study-level datasets and thus you cannot run queries across studies in Spectacle.
-Furthermore, Spectacle contains no non-ocular dataset and the cell types labels have not been harmonized between studies.
+In contrast, Spectacle, as of March 2022, is a compilation of study-level datasets and thus you cannot run queries across studies.
+Furthermore, Spectacle contains no non-ocular dataset and the cell types labels have not been curated or harmonized between studies.
 Finally, as this Spectacle uses Seurat objects, the app takes many seconds to load an individual dataset and many more to display the data.
@@
 
-There are two interwoven reasons why PLAE is fast, despite the huge amount of data it contains.
+There are two interwoven reasons why PLAE is user responsive, despite the huge amount of data it contains.
 First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables.
 A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage.
-The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous; the scEiaD v2022-03-22 sqlite database is about 368 GB.
+The major downside is that this approach requires a large amount of disk storage, which makes local (on computer) usage onerous; the scEiaD v2022-03-22 sqlite database is about 368 GB.
 Second, the visualization code in PLAE (e.g.
@@
 cones versus all remaining cell types) or in a pair-wise manner (cone against rod).
-We find, in total, thousands of differentially expressed genes across the ocular cell types.
+For the one against all testing, we
 
 ```{r}
-# fix  occasional NA
-# data made from src/ensdb.R
-diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
-down_tab_big <- read_tsv('data/down_tab_big.tsv.gz')
-gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
-diff_tab <- diff_tab_big %>% 
-  filter(Against == 'All', Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique()))
-
-diff_tab_sup <- diff_tab %>% 
-  group_by(Base, Organism) %>% 
-  summarise(Count = n()) %>% 
-  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
-  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
-  as_tibble()
-write_csv(diff_tab_sup, file = 'supplemental_files/differential_expression_table.csv')
-diff_tab_sup %>% 
-  flextable()
-
+tic()
 # number of celltype (predict) - organism - study counts (with >=50 cells)
+# ws == Well Supported
 ws_table <- meta_filter %>%  
@@
                                       TRUE ~ 'No'))
+
+# fix  occasional NA
+# data made from src/ensdb.R
+diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
+down_tab_big <- read_tsv('data/down_tab_big.tsv.gz')
+gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
+diff_tab <- diff_tab_big %>% 
+  filter(Against == 'All', 
+         Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% 
+                      pull(CellType) %>% unique()))
+
+diff_tab_sup <- diff_tab %>% 
+  group_by(Base, Organism) %>% 
+  summarise(Count = n()) %>% 
+  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
+  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
+  as_tibble()
+
+diff_tab_sup <- bind_rows(
+  diff_tab_sup,
+  ws_table$`CellType (Predict)`[!ws_table$`CellType (Predict)` %in% 
+                                  diff_tab_sup$`CellType (Predict)`] %>% 
+    enframe() %>% 
+    select(`CellType (Predict)` = value) %>%
+    mutate(`Homo sapiens` = 0, `Mus musculus` = 0, `Macaca fascicularis` = 0) %>% 
+    arrange(`CellType (Predict)`)
+)
+
+write_csv(diff_tab, file = 'supplemental_files/differential_expression_table.csv.gz')
+
+diff_tab_sup %>% 
+  flextable()
+
 ws_table %>% 
@@
 
-diff_tab_sup %>% pivot_longer(-`CellType (Predict)`, names_to = 'Organism', values_to = 'Diff Gene Count') %>% ggplot(aes(x=`CellType (Predict)`, y = `Diff Gene Count`, fill = Organism )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip()
+diff_tab_barplot <- diff_tab_sup %>% 
+  pivot_longer(-`CellType (Predict)`, names_to = 'Organism', values_to = 'Diff Gene Count') %>% 
+  ggplot(aes(x=`CellType (Predict)`, y = `Diff Gene Count`, fill = Organism )) + 
+  geom_bar(stat='identity', position = 'dodge') + 
+  cowplot::theme_cowplot() + coord_flip()
+
+toc()
 ```
 
-## Well supported ocular cell types identify
+## Well supported ocular cell types identify high confidence cell type markers
 
@@
 ```{r}
+tic()
 ws_oct <- ws_table %>% filter(`Well Supported` == 'Yes') %>% pull(`CellType (Predict)`)
@@
 consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n()) %>% ggplot(aes(x=`Base`, y = `Count` )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip() + xlab("CellTye (Predict)")
+toc()
 ```
@@
 ```{r, fig.height=11, fig.width=6}
+tic()
 g <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% slice_max(n=8, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
@@
                               row_labels = genes, 
-                              column_title_rot = 90, row_split
+                              column_title_rot = 90, 
                               name = 'log2FC', use_raster = TRUE,
@@
 ComplexHeatmap::draw(hm, padding = unit(c(2, 2, 32, 2), "mm"), row_title = "Marker Genes")
+toc()
 ```
@@
 We show the top 5 genes that are both more highly expressed in RPC relative to neurogenic cell and vice versa.
-```{r, fig.width=7, fig.height=3}
+```{r, fig.width=9, fig.height=3}
+tic()
 x <-  scEiaD_2020_v01 %>% 
@@
 b <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-plot_grid(a,b, labels = c('a','b'))
+a_legend <- get_legend(a)
+plot_grid(plot_grid(a + theme(legend.position="none"),
+          b + theme(legend.position="none") + ylab(''), labels = c('a','b')),
+          a_legend, ncol = 1, rel_heights = c(1,0.2))
+toc()
 ```
@@
 ```{r, fig.width=3, fig.height=7}
+tic()
 # flip flop diff genes
@@
 hm_flipflop
+toc()
 ```

### 2022-07-05 manuscript_00.Rmd
updates

@@
 tic()
-data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
-data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
+#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 # setup caption-ing
@@
 dbs <- c('GO_Biological_Process_2018', 'GO_Molecular_Function_2018')
-enriched_cone <- enrichr(org_diff_diff_genes %>% filter(Base == 'Rod') %>% pull(Symbol), dbs)
+enriched <- enrichr(org_diff_diff_genes %>% filter(Base == 'Bipolar Cell') %>% pull(Symbol), dbs)
 
@@
 
-## Projection of organoid data onto scEiaD demonstrates [SOMETHING]
+<!-- ## Projection of organoid data onto scEiaD demonstrates [SOMETHING] -->
 
-scEiaD is built using scvi-tool's scANVI batch correction tool.
+<!-- scEiaD is built using scvi-tool's scANVI batch correction tool.
 As part of the process a computational transcriptome model of the reference datasets are built.
@@
 Third, to leverage our extensive pre-labelled cell information to transfer cell type labels onto private data.
+-->
 
-We demonstrate some of these possibilities by projecting two organoid datasets from X and Y.
+<!-- We demonstrate some of these possibilities by projecting two organoid datasets from X and Y.  -->
 

### 2022-07-07 manuscript_00.Rmd
move diff data pulls to sep script

--- 2022-07-05 manuscript_00.Rmd
@@
 
+load('data/diff_data_pull.Rdata') # from src/diff_data_pull.R
 toc()
@@
 
-```{r scEiaD_data_overview, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=scEiaD_data_overview_cap}
+```{r scEiaD_data_overview, fig.height = 35, fig.width = 25, echo = F, message = F, fig.cap=scEiaD_data_overview_cap}
 tic()
@@
 tic()
+i = 16.5 # width of the side borders in the word_document output (in centimeters)
+ w = i*0.3937 # width of the side borders in the word_document output (in inches)
+
 web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
@@
   c('PLAE','Yes','Yes', 'Yes', '1136 ', 'https://plae.nei.nih.gov'),
-  c('Spectacle', 'Yes' ,'No', 'No', '1800', 'https://singlecell-eye.org'),
+  c('Spectacle', 'Yes' ,'No', 'No', '1500', 'https://singlecell-eye.org'),
   c('UCSC Cell Browser', 'Yes', 'No', 'Yes', '62', 'https://cells.ucsc.edu' ),
@@
 colnames(feature_table) <- c('Resource Name', 'Independent Datasets', 'Harmonization', 'Ocular and non-Ocular Data', 'Ocular Cell Count\n(thousands)', 'URL')
-feature_table %>% flextable()
+feature_table %>% flextable() %>% width(., width = (w/(ncol(feature_table)))) 
 # as of 2021 10 25
@@
 
-```{r}
-tic()
-# number of celltype (predict) - organism - study counts (with >=50 cells)
-# ws == Well Supported
-ws_table <- meta_filter %>%  
-  filter(Organ == 'Eye') %>% 
-  group_by(organism, CellType_predict, study_accession) %>% 
-  count() %>% 
-  filter(n >= 50) %>% 
-  ungroup() %>% 
-  group_by(organism, CellType_predict) %>% 
-  count() %>% 
-  pivot_wider(values_from = n, names_from = organism) %>% 
-  filter(!is.na(CellType_predict)) %>% 
-  replace(is.na(.), values = 0) %>% 
-  arrange(CellType_predict) %>% 
-  rename(`CellType (Predict)` = CellType_predict) %>% 
-  mutate(`Well Supported` = case_when(`Homo sapiens` >= 2 & `Mus musculus` >= 2 ~ 'Yes',
-                                      TRUE ~ 'No'))
-
-# fix  occasional NA
-# data made from src/ensdb.R
-diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
-down_tab_big <- read_tsv('data/down_tab_big.tsv.gz')
-gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
-diff_tab <- diff_tab_big %>% 
-  filter(Against == 'All', 
-         Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% 
-                      pull(CellType) %>% unique()))
-
-diff_tab_sup <- diff_tab %>% 
-  group_by(Base, Organism) %>% 
-  summarise(Count = n()) %>% 
-  rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
-  pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
-  as_tibble()
-
-diff_tab_sup <- bind_rows(
-  diff_tab_sup,
-  ws_table$`CellType (Predict)`[!ws_table$`CellType (Predict)` %in% 
-                                  diff_tab_sup$`CellType (Predict)`] %>% 
-    enframe() %>% 
-    select(`CellType (Predict)` = value) %>%
-    mutate(`Homo sapiens` = 0, `Mus musculus` = 0, `Macaca fascicularis` = 0) %>% 
-    arrange(`CellType (Predict)`)
-)
-
-write_csv(diff_tab, file = 'supplemental_files/differential_expression_table.csv.gz')
+```{r, fig.width=12}
+# tic()
+# # number of celltype (predict) - organism - study counts (with >=50 cells)
+# # ws == Well Supported
+# ws_table <- meta_filter %>%  
+#   filter(Organ == 'Eye') %>% 
+#   group_by(organism, CellType_predict, study_accession) %>% 
+#   count() %>% 
+#   filter(n >= 50) %>% 
+#   ungroup() %>% 
+#   group_by(organism, CellType_predict) %>% 
+#   count() %>% 
+#   pivot_wider(values_from = n, names_from = organism) %>% 
+#   filter(!is.na(CellType_predict)) %>% 
+#   replace(is.na(.), values = 0) %>% 
+#   arrange(CellType_predict) %>% 
+#   rename(`CellType (Predict)` = CellType_predict) %>% 
+#   mutate(`Well Supported` = case_when(`Homo sapiens` >= 2 & `Mus musculus` >= 2 ~ 'Yes',
+#                                       TRUE ~ 'No'))
+# 
+# # fix  occasional NA
+# # data made from src/ensdb.R
+# diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
+# down_tab_big <- read_tsv('data/down_tab_big.tsv.gz')
+# gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
+# diff_tab <- diff_tab_big %>% 
+#   filter(Against == 'All', 
+#          Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% 
+#                       pull(CellType) %>% unique()))
+# 
+# 
+# diff_tab_sup <- diff_tab %>% 
+#   group_by(Base, Organism) %>% 
+#   summarise(Count = n()) %>% 
+#   rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
+#   pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
+#   as_tibble()
+# 
+# diff_tab_sup <- bind_rows(
+#   diff_tab_sup,
+#   ws_table$`CellType (Predict)`[!ws_table$`CellType (Predict)` %in% 
+#                                   diff_tab_sup$`CellType (Predict)`] %>% 
+#     enframe() %>% 
+#     select(`CellType (Predict)` = value) %>%
+#     mutate(`Homo sapiens` = 0, `Mus musculus` = 0, `Macaca fascicularis` = 0) %>% 
+#     arrange(`CellType (Predict)`)
+# )
+# 
+# write_csv(diff_tab, file = 'supplemental_files/differential_expression_table.csv.gz')
 
@@
 
-toc()
+#toc()
 ```
@@
 ```{r}
-tic()
-ws_oct <- ws_table %>% filter(`Well Supported` == 'Yes') %>% pull(`CellType (Predict)`)
-
-consist_diff <- scEiaD_2020_v01 %>% 
-  tbl("diff_testing") %>% 
-  filter(Base %in% ws_oct) %>%  
-  filter(Group == 'CellType (Predict)', 
-         Against == 'All') %>% 
-  group_by(Base, Gene) %>% 
-  summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-4), `mean padj` = mean(padj)) %>% 
-  filter(`mean log2FC` > 2, `sig count` > 1, `mean padj` < 1e-5) %>%  
-  collect() %>% 
-  mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
-  left_join(gene_tab, by = 'GENEID') %>% 
-  relocate(SYMBOL, GENEID) %>% 
-  dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
-  arrange(-`mean log2FC`)
-
-# id diff genes only seen once across the cell types
-uniq_g <- consist_diff %>% group_by(Symbol) %>% count() %>% filter(n == 1)
+# tic()
+# ws_oct <- ws_table %>% filter(`Well Supported` == 'Yes') %>% pull(`CellType (Predict)`)
+# 
+# consist_diff <- scEiaD_2020_v01 %>% 
+#   tbl("diff_testing") %>% 
+#   filter(Base %in% ws_oct) %>%  
+#   filter(Group == 'CellType (Predict)', 
+#          Against == 'All') %>% 
+#   group_by(Base, Gene) %>% 
+#   summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-4), `mean padj` = mean(padj)) %>% 
+#   filter(`mean log2FC` > 2, `sig count` > 1, `mean padj` < 1e-5) %>%  
+#   collect() %>% 
+#   mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
+#   left_join(gene_tab, by = 'GENEID') %>% 
+#   relocate(SYMBOL, GENEID) %>% 
+#   dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
+#   arrange(-`mean log2FC`)
+# 
+# # id diff genes only seen once across the cell types
+# uniq_g <- consist_diff %>% group_by(Symbol) %>% count() %>% filter(n == 1)
 consist_diff %>% group_by(Base) %>% summarise(Count = n())
@@
 consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n()) %>% ggplot(aes(x=`Base`, y = `Count` )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip() + xlab("CellTye (Predict)")
-toc()
+#toc()
 ```
@@
 ```{r, fig.height=11, fig.width=6}
-tic()
-g <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% slice_max(n=8, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
+# tic()
+# g <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% slice_max(n=8, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
+# 
+# consist_diff_long <- scEiaD_2020_v01 %>%
+#   tbl("diff_testing") %>%
+#   filter(Gene %in% g) %>%
+#   collect() %>%
+#   filter(Base %in% ws_oct,
+#          Group == 'CellType (Predict)',
+#          Against == 'All') %>%
+#   select(Gene, Organism, log2FoldChange, Base)
+# 
+# consist_diff_wide <- consist_diff_long %>% 
+#   mutate(GO = paste0(Base, '_', Organism)) %>% 
+#   select(-Organism, -Base) %>% 
+#   pivot_wider(values_from=log2FoldChange, names_from = GO)
+# consist_diff_celltype <- colnames(consist_diff_wide[,-1])
+# consist_diff_wide <- consist_diff_wide %>% data.frame()
+# row.names(consist_diff_wide) <- consist_diff_wide$Gene
+# 
+# consist_diff_wide <- consist_diff_wide[,-1]
+# 
+# 
+# 
+# consist_diff_organism <- str_extract(consist_diff_celltype, '_\\w+ \\w+') %>% gsub('_','',.)
+# consist_diff_cts  <- str_extract(consist_diff_celltype, '.*_') %>% gsub('_','',.)
+# consist_diff_genes <- row.names(consist_diff_wide) %>% gsub(' \\(.*','',.)
+# 
+# consist_diff_base <- consist_diff_genes %>% as_tibble() %>% rename(Symbol = value) %>% left_join(consist_diff %>% select(Symbol, Base) %>% unique()) %>% pull(Base)
 
-test <- scEiaD_2020_v01 %>%
-  tbl("diff_testing") %>%
-  filter(Gene %in% g) %>%
-  collect() %>%
-  filter(Base %in% ws_oct,
-         Group == 'CellType (Predict)',
-         Against == 'All') %>%
-  select(Gene, Organism, log2FoldChange, Base)
-
-x <- test %>% 
-  mutate(GO = paste0(Base, '_', Organism)) %>% 
-  select(-Organism, -Base) %>% 
-  pivot_wider(values_from=log2FoldChange, names_from = GO)
-celltype <- colnames(x[,-1])
-x <- x %>% data.frame()
-row.names(x) <- x$Gene
-
-x <- x[,-1]
-
-organism <- str_extract(celltype, '_\\w+ \\w+') %>% gsub('_','',.)
-cts  <- str_extract(celltype, '.*_') %>% gsub('_','',.)
-genes <- row.names(x) %>% gsub(' \\(.*','',.)
-
-base_yo <- genes %>% as_tibble() %>% rename(Symbol = value) %>% left_join(consist_diff %>% select(Symbol, Base) %>% unique()) %>% pull(Base)
-
-hm <- ComplexHeatmap::Heatmap(x, column_labels = organism, 
-                              column_split = cts, 
-                              row_split = base_yo, show_row_dend = FALSE, 
-                              row_labels = genes, 
+consist_diff_hm <- ComplexHeatmap::Heatmap(consist_diff_wide, column_labels = consist_diff_organism, 
+                              column_split = consist_diff_cts, 
+                              row_split = consist_diff_base, show_row_dend = FALSE, 
+                              row_labels = consist_diff_genes, 
                               column_title_rot = 90, 
@@
                               row_title_rot = 0)
-ComplexHeatmap::draw(hm, padding = unit(c(2, 2, 32, 2), "mm"), row_title = "Marker Genes")
-toc()
+ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(2, 2, 32, 2), "mm"), row_title = "Marker Genes")
+#toc()
 ```
@@
 ```{r, fig.width=9, fig.height=3}
-tic()
-x <-  scEiaD_2020_v01 %>% 
-  tbl("diff_testing") %>% 
-  filter(Against  == "RPC",   
-         Group == 'CellType (Predict)', 
-         Base == "Neurogenic Cell") %>% 
-  group_by(Base, Against, Gene) %>% 
-  summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
-  arrange(-`mean log2FC`) %>%   
-  collect() %>% 
-  filter(!is.na(`mean log2FC`))
-
-# mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
-# left_join(gene_tab, by = 'GENEID') %>% 
-# relocate(SYMBOL, GENEID) %>% 
-# dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
-# arrange(-`mean log2FC`)
-
-down <- x %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` < 0) %>% head(5) %>% pull(Gene)
-up <-  x %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` > 0) %>% head(5) %>% pull(Gene)
+# tic()
+# pairwise_neurogenic_test<-  scEiaD_2020_v01 %>% 
+#   tbl("diff_testing") %>% 
+#   filter(Against  == "RPC",   
+#          Group == 'CellType (Predict)', 
+#          Base == "Neurogenic Cell") %>% 
+#   group_by(Base, Against, Gene) %>% 
+#   summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
+#   arrange(-`mean log2FC`) %>%   
+#   collect() %>% 
+#   filter(!is.na(`mean log2FC`))
+# 
+# 
+# # mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
+# # left_join(gene_tab, by = 'GENEID') %>% 
+# # relocate(SYMBOL, GENEID) %>% 
+# # dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
+# # arrange(-`mean log2FC`)
+# 
+# down <- pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` < 0) %>% head(5) %>% pull(Gene)
+# up <-  pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` > 0) %>% head(5) %>% pull(Gene)
 source('src/make_exp_plot.R')
@@
 
-a <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+exp_plot_giga<- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
 
 input$exp_plot_genes <- up
-b <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-a_legend <- get_legend(a)
-plot_grid(plot_grid(a + theme(legend.position="none"),
-          b + theme(legend.position="none") + ylab(''), labels = c('a','b')),
+exp_plot_iovs <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+a_legend <- get_legend(exp_plot_giga)
+plot_grid(plot_grid(exp_plot_giga + theme(legend.position="none"),
+          exp_plot_iovs + theme(legend.position="none") + ylab(''), labels = c('a','b')),
           a_legend, ncol = 1, rel_heights = c(1,0.2))
-toc()
+#toc()
 ```
@@
 ```{r, fig.width=3, fig.height=7}
-tic()
-# flip flop diff genes
-down_tab <- down_tab_big %>% 
-  mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
-  filter(Base %in% ws_oct, 
-         Against == 'All')
+# tic()
+# # flip_flop diff genes
+# down_tab <- down_tab_big %>% 
+#   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
+#   filter(Base %in% ws_oct, 
+#          Against == 'All')
+# 
+# org_diff_diff_genes <- diff_tab_big %>% 
+#   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
+#   filter(gene_ct_combo %in% down_tab$gene_ct_combo, 
+#          Against == 'All') %>% 
+#   arrange(padj) %>% mutate(x = paste0(Symbol, ' (', ID, ')')) 
+# 
+# flip_flop_genes <- scEiaD_2020_v01 %>% tbl('diff_testing_genes') %>% collect() %>% filter(grepl(org_diff_diff_genes %>% pull(ID) %>% paste0(., collapse = '|'), Gene)) %>% pull(Gene)
+# flip_flop_long <- scEiaD_2020_v01 %>%
+#   tbl("diff_testing") %>%
+#   filter(Gene %in% genes) %>%
+#   collect() %>%
+#   mutate(Symbol = str_extract(Gene, '^\\w+ ') %>% gsub(' $', '', .)) %>% 
+#   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
+#   filter(gene_ct_combo %in% org_diff_diff_genes$gene_ct_combo,
+#          Group == 'CellType (Predict)',
+#          Against == 'All') %>%
+#   select(Gene, Organism, log2FoldChange, Base)
+# 
+# flip_flop_wide <- flip_flop_long %>% pivot_wider(values_from = log2FoldChange, names_from = Organism)
+# flip_flop_organism <- colnames(flip_flop_wide)[c(3,4,5)]
+# flip_flop_wide <- flip_flop_wide %>% data.frame()
+# row.names(flip_flop_wide) <- paste0(flip_flop_wide$Gene,'_', flip_flop_wide$Base)
+# 
+# flip_flop_wide <- flip_flop_wide[,-c(1,2)]
+# 
+# 
+# 
+# flip_flop_genes <- row.names(flip_flop_wide) %>% gsub(' \\(.*','',.)
+# 
+# flip_flop_base <- row.names(flip_flop_wide) %>% gsub('.*_','',.)
 
-org_diff_diff_genes <- diff_tab_big %>% 
-  mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
-  filter(gene_ct_combo %in% down_tab$gene_ct_combo, 
-         Against == 'All') %>% 
-  arrange(padj) %>% mutate(x = paste0(Symbol, ' (', ID, ')'))
-
-genes <- scEiaD_2020_v01 %>% tbl('diff_testing_genes') %>% collect() %>% filter(grepl(org_diff_diff_genes %>% pull(ID) %>% paste0(., collapse = '|'), Gene)) %>% pull(Gene)
-test <- scEiaD_2020_v01 %>%
-  tbl("diff_testing") %>%
-  filter(Gene %in% genes) %>%
-  collect() %>%
-  mutate(Symbol = str_extract(Gene, '^\\w+ ') %>% gsub(' $', '', .)) %>% 
-  mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
-  filter(gene_ct_combo %in% org_diff_diff_genes$gene_ct_combo,
-         Group == 'CellType (Predict)',
-         Against == 'All') %>%
-  select(Gene, Organism, log2FoldChange, Base)
-
-x <- test %>% pivot_wider(values_from = log2FoldChange, names_from = Organism)
-organism <- colnames(x)[c(3,4,5)]
-x <- x %>% data.frame()
-row.names(x) <- paste0(x$Gene,'_', x$Base)
-
-x <- x[,-c(1,2)]
-
-genes <- row.names(x) %>% gsub(' \\(.*','',.)
-
-base_yo <- row.names(x) %>% gsub('.*_','',.)
-
-hm_flipflop <- ComplexHeatmap::Heatmap(x, column_labels = organism, 
-                                       row_split = base_yo, show_row_dend = FALSE, 
-                                       row_labels = genes, 
+hm_flipflop <- ComplexHeatmap::Heatmap(flip_flop_wide, column_labels = flip_flop_organism, 
+                                       row_split = flip_flop_base, show_row_dend = FALSE, 
+                                       row_labels = flip_flop_genes, 
                                        column_title_rot = 90, 
@@
 hm_flipflop
-toc()
+#toc()
 ```

### 2022-07-07 src/diff_data_pull.R

@@ -1,0 +1,293 @@
+library(tidyverse)
+
+library(cowplot)
+library(ggrepel)
+
+
+
+library(glue)
+library(tictoc)
+
+tic()
+data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
+data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+
+
+library(pool)
+library(RSQLite)
+meta_filter_v0 <- fst::read_fst(glue("{data_dir_vGiga}/paper_data/2021_03_17_meta_filter.fst")) %>% 
+  
+  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
+                                      TRUE ~ CellType_predict),
+         CellType= case_when(CellType == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
+                             TRUE ~ CellType))
+
+meta_filter <- fst::read_fst(glue("{data_dir_vPLAE}/meta_filter.fst")) %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>% 
+  
+  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
+                                      TRUE ~ CellType_predict),
+         CellType= case_when(CellType == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
+                             TRUE ~ CellType),  
+         Tissue = case_when(grepl('Iris', Tissue) ~ 'Iris',
+                            TRUE ~ Tissue),
+         Tissue = case_when(grepl('^RPE|^Choroid', Tissue) ~ 'RPE-Choroid',
+                            TRUE ~ Tissue),
+         Citation = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'Swamy VS, Fufa TD, Hufnagel RB, McGaughey DM Building the Mega Single Cell Transcriptome Ocular Meta-Atlas Gigascience 2021 Oct 13;10(10)',
+                              TRUE ~ Citation),
+         PMID = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ '34651173',
[…]

### 2022-07-07 manuscript_00.Rmd
tweak fig sizes

@@
 
-```{r, fig.height=11, fig.width=6}
+```{r, fig.height=24, fig.width=12}
 # tic()
@@
                               row_title_rot = 0)
-ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(2, 2, 32, 2), "mm"), row_title = "Marker Genes")
+ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(0, 0, 1.5, 0), "in"), row_title = "Marker Genes")
 #toc()
@@
 We show the top 5 genes that are both more highly expressed in RPC relative to neurogenic cell and vice versa.
-```{r, fig.width=9, fig.height=3}
+
+```{r, fig.width=15, fig.height=6, echo = F, message = F}
 # tic()
@@
 a_legend <- get_legend(exp_plot_giga)
-plot_grid(plot_grid(exp_plot_giga + theme(legend.position="none"),
-          exp_plot_iovs + theme(legend.position="none") + ylab(''), labels = c('a','b')),
+plot_grid(plot_grid(exp_plot_giga + theme(legend.position="none") + xlab(''),
+          exp_plot_iovs + theme(legend.position="none") + ylab('') + xlab(''), labels = c('a','b')),
           a_legend, ncol = 1, rel_heights = c(1,0.2))
@@
 
-```{r, fig.width=3, fig.height=7}
+```{r, fig.width=5, fig.height=15, echo = F, message = F}
 # tic()

### 2022-07-07 tester.Rmd
tweak fig sizes

--- 2022-07-07 manuscript_00.Rmd
@@
 
-# Introduction
-
-  Examples of single cell transcriptomics the retina date begin in the early 2000s where researchers [@hagstrom_cone_2000; @jabs_evidence_2000; @li_localization_2004; @lindqvist_single_2002-1; @reid_retinoschisin_2003] generally used RT-PCR approaches to amplify select transcripts in individual cells.
-Later, researchers used microarray platforms to analyze a broader array of transcripts researchers [@cherry_development_2009; @gustincich_gene_2004-1; @kim_identification_2008-1; @roesch_transcriptome_2008; @trimarchi_individual_2008; @trimarchi_molecular_2007-1; @wahlin_method_2004].
-The modern era of single cell transcriptomics is denoted by the introduction of droplet-based technologies , which enabled quantitation of thousands of cells in a single experiment; in Macoksko et al.
-the adult mouse retina was profiled [@klein_droplet_2015; @macosko_highly_2015].
-The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by producing over 30 single cell transcriptome based studies of the eye since the Macosko publication.
-  These publications's foci includes, but is not limited to, studying developmental transcriptomic dynamics, identifying genes to distinguish core cell types, proposing finer gradations of cell types within existing categories, and comparing ocular transcriptomes between organisms.
-Outside querying of these datasets often requires a long series of steps.
-Briefly, one would have to identify the relevant publication, find the associated data deposit, download the cell - count matrix, find the cell label table (and optionally the t-SNE or UMAP coordinates), and then load the data into R or python to run the desired queries.
-If one desired to compare *across* publications, then one would also have to normalize the cell type labels and optionally run a batch-effect removal algorithm.
-It is desirable, perhaps even necessary, to use multiple datasets to assess gene expression to confirm whether a gene expression is consistent and reproducible.
-Reliance on a single dataset can be problematic as there can be technical issues - for example the large consortium dataset GTEx has persistant contamination from highly expressed genes across tissues [@nieuwenhuis_consistent_2020].
-  Web-based resources to query gene expression are powerful means to democratize access to large datasets as the computationally intensive steps above can be run once and then shared to all.
-Single cell RNA-seq resources that contain ocular tissues include the UCSC Cell Browser, Tabula Muris, PanglaoDB, and the Protein Atlas [@franzen_panglaodb_2019; @karlsson_singlecell_2021; @noauthor_single-cell_nodate; @speir_ucsc_2021].
-Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging.
-The Spectacle project is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigt_spectacle_2020].
-Each data deposit is independently presented, which makes cross study queries difficult.
-We recently created a unified ocular resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space [@swamy_building_2021].
-We have built upon this dataset, adding more datasets, improving the quality control, removing ambient noise, improved the differential gene testing structure, and finally created a powerful, reactive, and fast web app to query the database.
-We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, genes which have species-specific expression cell type patterning, propose genes XYZ differentially expressed between the fovea and peripheral cones, and
-
-## Web atlases containing ocular data have substantial limitations
-
-Data difficult to access and cross comparisons impossible
-
-But important as seeing trends trends across different groups strongly demonstrates that 
-the signal is biological, not technical.
-
-## What we did
-
-We published the ocular focused scEiaD v2021 resource in 2021, which was one of the first single cell meta-atlases.
-To broaden the usage of this powerful database, we restructured the data in a custom sqlite-based database and built a reactive web application around it.
-As many new studies were published since our last data build, we re-ran the scEiaD pipeline in March of 2022.
-In aggregate we have collated `r meta_filter$study_accession %>% unique() %>% length()` studies across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues.
-We hand-curated `r meta_filter$CellType %>% unique() %>% length()` cell types, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the remaining cells.
-Extensive quality control metrics were tuned and applied to remove lower quality cells.
-In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values.
-
-# Methods
-
-## scEiaD pipeline upgraded and the database re-built
-
-The scEiaD pipeline is structurally the same as in Swamy et al.
-[@swamy_building_2021].
-Very briefly, we collected the raw fastq files and quantified with the kallisto / bustools system [@melsted_barcode_2019; @melsted_modular_2019].
-The quantified counts are merged into one matrix that is used for downstream processing with scvitools create a batch corrected lower dimensional space.
-and then into Seurat for quality control, clustering, and 2D UMAP visualization.
-We use our previous xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamy_building_2021].
-
-The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300 Second, we use the DecontX algorithm [@yang_decontamination_2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination.
-Third, we altered the procedure for automatically aligning gene names across species (see below for details).
-Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process.
-
-Two custom cutoffs for the minimum required number of unique genes identified per cell were used for the E12 chicken data and the cornea dataset (SRP362101) as an abnormally high number of cells were returned from our default value of 250 (????).
-We set the cutoffs for these studies to be 1000 and 800, respectively, as these values resulted in approximately the same number of cells being returned as the authors reported in their papers.
-
-## Cross species gene alignment
-
-As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify, we must identify shared genes.
-We found homologs/orthologs by pulling from BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference.
-We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus Musculus*.
-This generates a table linked by Ensembl gene IDs and gene names.
-We then use a full join on non-duplicated Ensembl IDs between mouse and human.
-As we noticed a small number of gene names failed to be linked in this manner we ran another join on the remaining non-aligned genes, using gene name.
-This procedure gave us 17,769 shared human and mouse genes.
-To align the other species we again joined on Ensembl gene ID, removing genes where one chicken or macaque gene aligned to multiple human/mouse genes.
-In cases where multiple chicken or macaque genes aligned to one human/mouse gene, then we aggregated the counts by sum.
-In cases where there was no corresponding chicken or macaque gene, we filled in zeros.
-After these steps we had, as expected, 17,769 genes.
-
-## Metadata curation
-
-Every study brought into scEiaD was hand curated to identify unique biological samples, published paper ID (where available), organ, tissue (e.g.
-Cornea), source (iPSC based or tissue), single cell platform (e.g.
-10Xv2, DropSeq, etc.).
-We also, where possible, annotated retina region (e.g.
-fovea), sex, and age or developmental stage.
-To import individual cell type labels (e.g.
-Rod, Muller Glia), we wrote custom code for each study that made this table available to link their cell type assignments back to the matching barcoded cell.
-We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels.
-We normalized names for CellType, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).
-We did retain the original published (except for the broader name normalization) labels under SubCellType.
-
-## Differential gene testing
-
-The scater::aggregateAcrossCells function was used to aggregate gene counts across species, study accession, and [category] where category was either Cluster, CellType, or CellType (Predict) assignments for a cell.
-If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test.
-Any genes with a sum of 0 counts after aggregation were removed.
-The aggregated matrix was imported into DESeq2 using the DESeqDataSetFromMatrix function with the design given as "~study_accession + [category]".
-Contrasts were extracted either with the target category (e.g.
-"Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function.
-Genes were considered to be differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2.
-To enhance the specificity of the
-
-## scEiaD structure and web app optimization
-
-The web app at plae.nei.nih.gov can display gene expression in only a few seconds across over a million cells.
-This speed is only possible due to custom data structures to optimize for data query efficiency.
-Like eyeIntegration's EiaD database, scEiaD is a sqlite database with tables for the core categories.
-This allows the app to initialize in about 30 seconds on a cloud server with memory usage under 8 GB.
-For the gene by cell expression matrix, the data was transformed into a "long" format where there are three columns: Gene, Cell Barcode, and Gene Count Value.
-This allows for cell - gene level queries to be completed, again with minimal memory usage, in under 0.5 seconds.
-To generate the aggregated information (in the Dot Plot, Expression Plot, and tables) we pre-calculated all queries by running a dplyr "group by" operation on all column fields.
-On user given query, the data is further aggregated to level request.
-This allows for complicated queries to complete in seconds rather than minutes.
-
-## Data reproducibility
-
-The codebase for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is XXXXXXX.
-There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app.
-
-# Results
-
-## New studies and improvements to the scEiaD database
-
-Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamy_building_2021].
-We updated the scEiaD resource for the PLAE v0.91 web app in six important ways.
-First, we added a large chicken retina (*Gallus gallus*) studies.
-Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye.
-Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons.
-Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`).
-Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300 (CHECK THIS), to improve the database quality.
-Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons.
-
-After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (`r fig_cap(name = 'scEiaD_data_overview', display = 'cite')`).
-This resource is only possible due to the publication of `r meta_filter$PMID %>% unique %>% length()` single cell resources that scEiaD draws from (`r supTab_cap(name = 'studies_table', display = 'cite')`).
-To recognize these papers, we make the first data table on the loading page of plae.nei.nih.gov list each of these studies along with a direct link to their citation, where possible.
-
-```{r scEiaD_data_overview, fig.height = 35, fig.width = 25, echo = F, message = F, fig.cap=scEiaD_data_overview_cap}
-tic()
-num1 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(Tissue) %>% unique() %>% length()
-num2 <- meta_filter %>% filter(Organ != 'Eye') %>% pull(Tissue) %>% unique() %>% length()
-num3 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique() %>% length()
-scEiaD_data_overview_cap <- fig_cap(name = 'scEiaD_data_overview', caption = glue::glue("Distribution of tissues and cell types in scEiaD.
-A.
-tabulation of the number of cells present across {num1} ocular and {num2} non-ocular tissues, B.
-Number of studies for each ocular tissue.
-C.
-Number of cells present across {num3} curated ocular-derived cell types."))
-
-# organ
-meta_filter <- meta_filter %>% mutate(organism = factor(organism, levels = meta_filter$organism %>% unique()))
-a <- meta_filter %>%
-  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
-                           TRUE ~ 'Body'),
-         Tissue = gsub('\\.|_', ' ', Tissue),
-         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
-         Tissue = gsub(" and Aorta","", Tissue)) %>% 
-  filter(Organ == 'Body') %>% 
-  filter(Tissue != 'Brain Choroid Plexus') %>% 
-  group_by(Organ, Tissue,organism, study_accession) %>%
-  summarise(Count = n()) %>%
-  ungroup() %>% 
-  mutate(organism = factor(organism, levels = meta_filter$organism %>% unique())) %>% 
-  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
-  ylab('Number of Cells') +
-  geom_bar(stat='identity', position = 'dodge') +
-  cowplot::theme_cowplot() +
-  coord_flip() +
-  # scale_y_continuous(breaks = c(2500, 5000, 7500, 10000, 12500, 84000)) +
-  # ggbreak::scale_y_break(c(13000, 82000)) +
-  scale_fill_discrete(drop=TRUE,
-                      limits = levels(meta_filter$organism)) +
-  theme(legend.position = "none") +
-  scale_y_continuous(expand = c(0, 0)) + 
-  ylab('') + xlab('')
-
-aa <- meta_filter %>% 
-  filter(Tissue == 'Brain Choroid Plexus') %>% 
-  group_by(Organ, Tissue,organism, study_accession) %>%
-  summarise(Count = n()) %>%
-  ungroup() %>% 
-  mutate(organism = factor(organism, levels = meta_filter$organism %>% unique())) %>% 
-  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
-  ylab('Number of Cells') +
-  geom_bar(stat='identity', position = 'dodge') +
-  cowplot::theme_cowplot() +
-  coord_flip() +
-  scale_fill_discrete(drop=TRUE,
-                      limits = levels(meta_filter$organism)) +
-  theme(legend.position = "none") +
-  scale_y_continuous(expand = c(0, 0)) + 
-  ylab('') + xlab('')
-
-# eye studies
-b <-  meta_filter %>%
-  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
-                           TRUE ~ 'Body'),
-         Tissue = gsub('\\.|_', ' ', Tissue),
-         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
-         Tissue = case_when(Tissue == "Choroid" ~ 'RPE-Choroid',
-                            TRUE ~ Tissue),
-         Tissue = gsub(" and Aorta","", Tissue)) %>% 
-  filter(Organ == 'Eye') %>% 
-  group_by(Organ, Tissue,organism) %>%
-  summarise(Count = n()) %>%
-  ungroup() %>% 
-  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
-  ylab('Number of Cells') +
-  geom_bar(stat='identity', position = 'dodge') +
-  cowplot::theme_cowplot() +
-  coord_flip() +
-  theme(legend.position = "none") +
-  scale_y_continuous(expand = c(0, 0)) +
-  xlab('')
-
-# eye studies samples
-c <-  meta_filter %>%
-  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
-                           TRUE ~ 'Body'),
-         Tissue = gsub('\\.|_', ' ', Tissue),
-         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
-         Tissue = gsub(" and Aorta","", Tissue)) %>% 
-  filter(Organ == 'Eye') %>% 
-  select(Organ, Tissue,organism, study_accession) %>%
-  unique() %>%
-  group_by(Organ, Tissue, organism) %>% 
-  summarise(Count = n()) %>%
-  ungroup() %>% 
-  ggplot(aes(x=Tissue,y=Count, fill = organism, group = organism)) +
-  ylab('Number of Studies') +
-  geom_bar(stat='identity', position = 'dodge') +
-  cowplot::theme_cowplot() +
-  coord_flip() +
-  theme(legend.position = "none") +
-  scale_y_continuous(expand = c(0, 0))  + xlab('')
-
-# retina cell types
-d <- meta_filter %>% filter(Organ == 'Eye') %>% 
-  mutate(CellType_predict = case_when(is.na(CellType_predict) ~ 'Unlabelled', TRUE ~ CellType_predict)) %>% 
-  group_by(CellType_predict, study_accession, organism) %>%
-  summarise(Count = n()) %>% filter(Count > 10) %>% ungroup() %>% 
-  ggplot(aes(x=CellType_predict,y=Count, fill = organism, group = organism)) +
-  ylab('Number of Cells') +
-  geom_bar(stat='identity', position = 'dodge') +
-  cowplot::theme_cowplot() +
-  coord_flip() + 
-  scale_y_continuous(expand = c(0, 0)) + xlab('')
-
-a_aa_b <- plot_grid(a,aa,b, ncol = 1, align = 'hv', rel_heights = c(10, 1.5,3.5), labels = 'a')
-c_d <- plot_grid(c,d + theme(legend.position = "none"), rel_heights = c(3,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))
-
-# make E now
-
-# map color as plae app does
-categorical_columns <- c("Phase","batch","study_accession","library_layout","organism","Platform",
-                         "Covariate","CellType","CellType_predict","TabulaMurisCellType","TabulaMurisCellType_predict",
-                         "GSE","Summary","Design","Citation","PMID","Stage","cluster",
-                         "Doublet","TechType", "SubCellType", 'subcluster', 'Age', "retina_region",
-                         'Tissue','Organ', 'Source','sample_accession')
-#"SubCellType" and subcluster are problems
-meta_filter <- meta_filter %>% mutate(Age = as.character(Age), SubCellType = tidyr::replace_na(SubCellType, 'None'),
-                                      subcluster = as.character(subcluster))
-
-map_color <- function(column, meta_filter){
-  master_colorlist <- c( pals::glasbey(), pals::alphabet(), pals::cols25()[1:23], pals::alphabet2()) %>% unique()
-  values <- meta_filter %>% pull(!!column) %>% unique %>% sort
-  if(length(values) > length(master_colorlist) ){
-    r= round(length(values) / length(master_colorlist)) +1
-    master_colorlist <- rep(master_colorlist, r)
-  }
-
-  colors <- master_colorlist[1:length(values)]
-  return(tibble(meta_category = column,value = values, color=colors))
-
-}
-
-cat_to_color_df <- lapply(categorical_columns, function(col) map_color(col, meta_filter)) %>% bind_rows()
-# now roll through and set each value to the same color across CellType and CellType (predict)
-# e.g. RPE will be one color in CellType and CellType (Predict)
-celltype_x_color_map <- cat_to_color_df %>% filter(meta_category %in% c("CellType", "CellType_predict")) %>% group_by(value) %>% summarise(c2 = head(color,1))
-cat_to_color_df <- cat_to_color_df %>% left_join(celltype_x_color_map) %>% mutate(color = case_when(!is.na(c2) ~ c2, TRUE ~ color)) %>% select(-c2)
-
-o_ct <- meta_filter %>% filter(Tissue == 'Retina') %>% pull(CellType)
-p_data <- meta_filter %>% 
-  #mutate(CellType_predict = Tissue) %>% 
-  #mutate(CellType_predict = case_when(!CellType_predict %in% ws_oct ~ 'Other',
-  #                                    TRUE ~ CellType_predict)) %>% 
-  filter(Organ == 'Eye') %>% 
-  mutate(Citation = paste(str_trunc(Citation, width = 20), str_trunc(Covariate, width = 12, side = 'left',ellipsis = ''))) %>% 
-  group_by(Citation, organism, Tissue, CellType_predict) %>% 
-  summarise(Count = n()) %>% 
-  mutate(Proportion = Count/sum(Count) * 100)
-
-meta_column <- 'CellType_predict'
-cur_color_df <- cat_to_color_df %>%
-  filter( meta_category %in%  meta_column,
-          value %in% p_data[[meta_column]]) %>% distinct
-
-color_list <- cur_color_df %>% pull(color)
-# replaced join with this for speed
-k <- cur_color_df$color
-names(k) <- cur_color_df$value
-np_color <- {k[p_data[[meta_column]] ]} %>% paste0(., '33')
-names(np_color) <- NULL
-color_data <- cur_color_df  %>%
-  select(value) %>%
-  mutate( x=0)
-names(color_list) <- color_data$value
-
-proportion_bar_plot <- p_data %>% ggplot(aes(y=Citation,x=Proportion,fill=CellType_predict)) + geom_bar(stat= 'identity') +
-  scale_fill_manual(name= meta_column,
-                    values = color_list) + 
-  ggforce::facet_col(vars(organism, Tissue), scale = 'free', space = 'free') +
-  cowplot::theme_cowplot() +
-  guides(fill=guide_legend(ncol =1))
-
-e <- plot_grid(proportion_bar_plot, labels = c('d'))
-
-plot_grid(a_aa_b, c_d , e, ncol = 3, rel_widths = c(1,1.6, 2.2))
-toc()
-```
-
-## scRNA web resources containing ocular data
-
-Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources_cap', display = 'cite')`).
-The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are not ocular focused, but do contain some ocular data.
-The PanglaoDB and UCSC are resources which have ingress-ed a large number of published single cell datasets The UCSC Cell Browser has around 62 thousand ocular single cell transcriptomes while the PangloaDB has about 47 thousand.
-However, these resources provide no curation of the metadata and process the data for each study individually.
-Thus each study is effectively silo-ed off from the rest and cross comparisons are impractical.
-The Tabula Sapiens project is an effort to sequence all cells in the mouse and thus the ocular data present, though relatively small (at around 10 thousand), can be more readily cross compared to non-ocular tissues / cells.
-The Protein Atlas resource, which has aggregated protein labels (????) information has recently used one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g.
-one can see whether a gene of interest is enriched in Cones).
-The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has selected a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app.
-PLAE is distinguished from these resources by hand curation of the metadata, re-processing of the data from the raw sequence, and, most crucially, integrating the independent studies to deliver cross study comparisons.
-
-```{r web_resources, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=web_resources_cap}
-tic()
-i = 16.5 # width of the side borders in the word_document output (in centimeters)
- w = i*0.3937 # width of the side borders in the word_document output (in inches)
-
-web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
-
-feature_table <- rbind(
-  c('PLAE','Yes','Yes', 'Yes', '1136 ', 'https://plae.nei.nih.gov'),
-  c('Spectacle', 'Yes' ,'No', 'No', '1500', 'https://singlecell-eye.org'),
-  c('UCSC Cell Browser', 'Yes', 'No', 'Yes', '62', 'https://cells.ucsc.edu' ),
-  c('Tabula Sapiens', 'No', 'N/A', 'Yes', '10', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
-  c('PanglaoDB', 'Yes', 'No', 'Yes', '47', 'https://panglaodb.se'),
-  c('Protein Atlas', 'Yes*','No', 'Yes', '5',  'https://www.proteinatlas.org/humanproteome/celltype')) %>% 
-  as_tibble()
-colnames(feature_table) <- c('Resource Name', 'Independent Datasets', 'Harmonization', 'Ocular and non-Ocular Data', 'Ocular Cell Count\n(thousands)', 'URL')
-feature_table %>% flextable() %>% width(., width = (w/(ncol(feature_table)))) 
-# as of 2021 10 25
-toc()
-```
-
-## PLAE differs from a complementary web app, Spectacle in several important ways
-
-While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently [@voigt_spectacle_2020].
-PLAE is built around the scEiaD database, which runs an multi-stage pipeline to build and identify a high performing scVI-based eye model that is used to integrate all datasets together [@swamy_building_2021].
-This allows for gene queries to be constructed and analyzed across studies.
-In contrast, Spectacle, as of March 2022, is a compilation of study-level datasets and thus you cannot run queries across studies.
-Furthermore, Spectacle contains no non-ocular dataset and the cell types labels have not been curated or harmonized between studies.
-Finally, as this Spectacle uses Seurat objects, the app takes many seconds to load an individual dataset and many more to display the data.
-
-## Extensive optimization makes the PLAE web app highly responsive despite containing huge amounts of data
-
-There are two interwoven reasons why PLAE is user responsive, despite the huge amount of data it contains.
-First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables.
-A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage.
-The major downside is that this approach requires a large amount of disk storage, which makes local (on computer) usage onerous; the scEiaD v2022-03-22 sqlite database is about 368 GB.
-Second, the visualization code in PLAE (e.g.
-the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) to reduce drawing time.
-Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study (tested on 2022-03-23).
-In the same amount of time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies.
-
-## Rich web-available visualizations
-
-PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values.
-So thus a user can see Crx expression across over a million cells and separately see cell type labels for those same million cells.
-The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells.
-Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`a).
-
-For users interested in seeing how expression of a gene relates to study or other covariates, the "Exp Plot" view displays expression by cell type, predicted cell type, or cluster.
-The values can then be split and colored by multiple covariates like study or organism.
-This allows a user to see whether a gene pattern is consistent across studies or species.
-We also offer a "Dot Plot" view which can display a large number of genes in a space efficient manner.
-This allows a users to see whether a set of genes has related (or not) expression patterns across multiple cell types (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`b and c).
-
-While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina.
-The structure of the cartoon retina was based on the human retina.
-This unique visualization displays the major cell types of the retina and the RPE / choroid layer behind it.
-The cell types are colored by the relative amounts of expression in the user selected gene.
-Like all visualizations in PLAE, the user can use powerful filtering options to only display data from certain metadata characteristics, like species or publication (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`d).
-
-#```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
-
-#```
-
-## Leveraging the high diversity of studies to identify thousands of consistently differentially expressed genes across ocular cell types
-
-We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping.
-The created a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices and thus allowing for mature bulk RNA-seq tooling to be used.
-This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021].
-Once the aggregated matrix was con structured, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
-We tested each species (Human, Macaque, Mouse) separately.
-We had an insufficient number of chick studies to conduct differential testing.
-The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential gene expression.
-After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g.
-cones versus all remaining cell types) or in a pair-wise manner (cone against rod).
-For the one against all testing, we
-
-```{r, fig.width=12}
-# tic()
-# # number of celltype (predict) - organism - study counts (with >=50 cells)
-# # ws == Well Supported
-# ws_table <- meta_filter %>%  
-#   filter(Organ == 'Eye') %>% 
-#   group_by(organism, CellType_predict, study_accession) %>% 
-#   count() %>% 
-#   filter(n >= 50) %>% 
-#   ungroup() %>% 
-#   group_by(organism, CellType_predict) %>% 
-#   count() %>% 
-#   pivot_wider(values_from = n, names_from = organism) %>% 
-#   filter(!is.na(CellType_predict)) %>% 
-#   replace(is.na(.), values = 0) %>% 
-#   arrange(CellType_predict) %>% 
-#   rename(`CellType (Predict)` = CellType_predict) %>% 
-#   mutate(`Well Supported` = case_when(`Homo sapiens` >= 2 & `Mus musculus` >= 2 ~ 'Yes',
-#                                       TRUE ~ 'No'))
-# 
-# # fix  occasional NA
-# # data made from src/ensdb.R
-# diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
-# down_tab_big <- read_tsv('data/down_tab_big.tsv.gz')
-# gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
-# diff_tab <- diff_tab_big %>% 
-#   filter(Against == 'All', 
-#          Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% 
-#                       pull(CellType) %>% unique()))
-# 
-# 
-# diff_tab_sup <- diff_tab %>% 
-#   group_by(Base, Organism) %>% 
-#   summarise(Count = n()) %>% 
-#   rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
-#   pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
-#   as_tibble()
-# 
-# diff_tab_sup <- bind_rows(
-#   diff_tab_sup,
-#   ws_table$`CellType (Predict)`[!ws_table$`CellType (Predict)` %in% 
-#                                   diff_tab_sup$`CellType (Predict)`] %>% 
-#     enframe() %>% 
-#     select(`CellType (Predict)` = value) %>%
-#     mutate(`Homo sapiens` = 0, `Mus musculus` = 0, `Macaca fascicularis` = 0) %>% 
-#     arrange(`CellType (Predict)`)
-# )
-# 
-# write_csv(diff_tab, file = 'supplemental_files/differential_expression_table.csv.gz')
-
-diff_tab_sup %>% 
-  flextable()
-
-ws_table %>% 
-  flextable()
-
-diff_tab_barplot <- diff_tab_sup %>% 
-  pivot_longer(-`CellType (Predict)`, names_to = 'Organism', values_to = 'Diff Gene Count') %>% 
-  ggplot(aes(x=`CellType (Predict)`, y = `Diff Gene Count`, fill = Organism )) + 
-  geom_bar(stat='identity', position = 'dodge') + 
-  cowplot::theme_cowplot() + coord_flip()
-
-#toc()
-```
-
-## Well supported ocular cell types identify high confidence cell type markers
-
-We filter down to only assess cell types with more than two independent datasets in both human and mouse.
-This leaves .....
-These well supported cell types can be used to more strenuously identify genes that are consistently differentially expressed in a cell type across multiple species.
-To define these well supported marker genes we only ass
-
-To identify genes that are differentially expressed across species we
-
-```{r}
-# tic()
-# ws_oct <- ws_table %>% filter(`Well Supported` == 'Yes') %>% pull(`CellType (Predict)`)
-# 
-# consist_diff <- scEiaD_2020_v01 %>% 
-#   tbl("diff_testing") %>% 
-#   filter(Base %in% ws_oct) %>%  
-#   filter(Group == 'CellType (Predict)', 
-#          Against == 'All') %>% 
-#   group_by(Base, Gene) %>% 
-#   summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-4), `mean padj` = mean(padj)) %>% 
-#   filter(`mean log2FC` > 2, `sig count` > 1, `mean padj` < 1e-5) %>%  
-#   collect() %>% 
-#   mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
-#   left_join(gene_tab, by = 'GENEID') %>% 
-#   relocate(SYMBOL, GENEID) %>% 
-#   dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
-#   arrange(-`mean log2FC`)
-# 
-# # id diff genes only seen once across the cell types
-# uniq_g <- consist_diff %>% group_by(Symbol) %>% count() %>% filter(n == 1)
-consist_diff %>% group_by(Base) %>% summarise(Count = n())
-consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n())
-
-consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n()) %>% ggplot(aes(x=`Base`, y = `Count` )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip() + xlab("CellTye (Predict)")
-#toc()
-```
-
-```{r, fig.height=11, fig.width=6}
-# tic()
-# g <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% slice_max(n=8, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
-# 
-# consist_diff_long <- scEiaD_2020_v01 %>%
-#   tbl("diff_testing") %>%
-#   filter(Gene %in% g) %>%
-#   collect() %>%
-#   filter(Base %in% ws_oct,
-#          Group == 'CellType (Predict)',
-#          Against == 'All') %>%
-#   select(Gene, Organism, log2FoldChange, Base)
-# 
-# consist_diff_wide <- consist_diff_long %>% 
-#   mutate(GO = paste0(Base, '_', Organism)) %>% 
-#   select(-Organism, -Base) %>% 
-#   pivot_wider(values_from=log2FoldChange, names_from = GO)
-# consist_diff_celltype <- colnames(consist_diff_wide[,-1])
-# consist_diff_wide <- consist_diff_wide %>% data.frame()
-# row.names(consist_diff_wide) <- consist_diff_wide$Gene
-# 
-# consist_diff_wide <- consist_diff_wide[,-1]
-# 
-# 
-# 
-# consist_diff_organism <- str_extract(consist_diff_celltype, '_\\w+ \\w+') %>% gsub('_','',.)
-# consist_diff_cts  <- str_extract(consist_diff_celltype, '.*_') %>% gsub('_','',.)
-# consist_diff_genes <- row.names(consist_diff_wide) %>% gsub(' \\(.*','',.)
-# 
-# consist_diff_base <- consist_diff_genes %>% as_tibble() %>% rename(Symbol = value) %>% left_join(consist_diff %>% select(Symbol, Base) %>% unique()) %>% pull(Base)
-
-consist_diff_hm <- ComplexHeatmap::Heatmap(consist_diff_wide, column_labels = consist_diff_organism, 
-                              column_split = consist_diff_cts, 
-                              row_split = consist_diff_base, show_row_dend = FALSE, 
-                              row_labels = consist_diff_genes, 
-                              column_title_rot = 90, 
-                              name = 'log2FC', use_raster = TRUE,
-                              row_title_rot = 0)
-ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(2, 2, 32, 2), "mm"), row_title = "Marker Genes")
-#toc()
-```
-
-## Tease apart closely related cell types
-
-Neurogenic and RPC have very similar expression patterns.
-We leverage our more specific differential test to directly assess genes were are different between neurogenic cells and RPC.
-We show the top 5 genes that are both more highly expressed in RPC relative to neurogenic cell and vice versa.
-```{r, fig.width=9, fig.height=3}
-# tic()
-# pairwise_neurogenic_test<-  scEiaD_2020_v01 %>% 
-#   tbl("diff_testing") %>% 
-#   filter(Against  == "RPC",   
-#          Group == 'CellType (Predict)', 
-#          Base == "Neurogenic Cell") %>% 
-#   group_by(Base, Against, Gene) %>% 
-#   summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
-#   arrange(-`mean log2FC`) %>%   
-#   collect() %>% 
-#   filter(!is.na(`mean log2FC`))
-# 
-# 
-# # mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
-# # left_join(gene_tab, by = 'GENEID') %>% 
-# # relocate(SYMBOL, GENEID) %>% 
-# # dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
-# # arrange(-`mean log2FC`)
-# 
-# down <- pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` < 0) %>% head(5) %>% pull(Gene)
-# up <-  pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` > 0) %>% head(5) %>% pull(Gene)
-source('src/make_exp_plot.R')
-
-input <- list()
-input$exp_plot_groups <- c('study_accession')
-input$exp_plot_facet <- c('CellType_predict')
-input$exp_plot_genes <- down
-input$exp_filter_cat <- 'CellType_predict'
-#input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
-input$exp_filter_on <- c("Neurogenic Cell", "RPC" )
-input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
-#input$exp_plot_ylab <- '% of Cells Detected'
-input$exp_plot_col_num <- 16
-input$exp_filter_min_cell_number <- 50
-input$flip_facet <- TRUE
-
-exp_plot_giga<- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-
-input$exp_plot_genes <- up
-exp_plot_iovs <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-a_legend <- get_legend(exp_plot_giga)
-plot_grid(plot_grid(exp_plot_giga + theme(legend.position="none"),
-          exp_plot_iovs + theme(legend.position="none") + ylab(''), labels = c('a','b')),
-          a_legend, ncol = 1, rel_heights = c(1,0.2))
-#toc()
-```
-
-```{r, fig.width=3, fig.height=7}
+```{r, fig.width=4, fig.height=15, echo = F, message = F}
 # tic()
@@
 ```
-
-```{r}
-library(enrichR)
-dbs <- c('GO_Biological_Process_2018', 'GO_Molecular_Function_2018')
-enriched <- enrichr(org_diff_diff_genes %>% filter(Base == 'Bipolar Cell') %>% pull(Symbol), dbs)
-
-```
-
-To aid in in the analysis of ocular gene expression patterns we provide pre-computed differential tests run
-
-## Pan-study fovea vs peripheral photoreceptor identifies N genes consistently differentially expressed
-
-To identify
-
-Sanes proposed X genes (https://www.nature.com/articles/s41598-020-66092-9 --> Figure 9)
-
-PRPH2 and RS1 for cones
-NR2E3, SLC24A1, PDE6H, GNAT1 for rods
-Also did RGC, Microglia, Astrocytes, Muller Glia
-
-<!-- ## Projection of organoid data onto scEiaD demonstrates [SOMETHING] -->
-
-<!-- scEiaD is built using scvi-tool's scANVI batch correction tool.
-As part of the process a computational transcriptome model of the reference datasets are built.
-This model is only a few mb in size and can be used on a modest computer to project private scRNA datasets onto it.
-This allows an ocular researcher several important functionalities.
-First, the ability to check the quality control of an existing dataset by looking at the distribution of their cells overlapped with our labelled cell types.
-Second, in perturbation experiments, to identify unusual distributions of cells.
-Third, to leverage our extensive pre-labelled cell information to transfer cell type labels onto private data.
--->
-
-<!-- We demonstrate some of these possibilities by projecting two organoid datasets from X and Y.  -->
-
-# Conclusions
-
-## PLAE provides quick access to over one million transcriptomes across 35 studies
-
-## Projection of outside data onto scEiaD allows for ... (check lung atlas paper)
-
-##
-
-# Supplementary Figures and Tables
-
-```{r decontX, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=decontX_cap}
-decontX_cap <- supFig_cap(name='decontX', caption = 'DecontX *in silico* ambient RNA contamination tool substantially removes Rhodopsin expression in non-rod cells while retaining high expression rods.
-First shown (a) is Rhodopsin expression across the scEiaD v0 dataset, without DecontX optimization.
-Note Rhodopsin expression is noticeable in many cell types beyond the rods.
-Next (b) is the scEiaD v1 dataset with DecontX optimization.
-Rhodopsin is nearly exclusively expressed in labelled rod cells.
-Despite the computational ambient RNA removal, Rhodopsin expression remains high in the Rods.')
-source('src/make_exp_plot.R')
-
-input <- list()
-input$exp_plot_groups <- c('organism')
-input$exp_plot_facet <- c('CellType_predict')
-input$exp_plot_genes <- c('RHO (ENSG00000163914)')
-input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
-input$exp_plot_col_num <- 16
-input$exp_filter_min_cell_number <- 50
-input$flip_facet <- FALSE
-
-v0_rho <- make_exp_plot(input, scEiaD_2020_v00, meta_filter_v0)
-v1_rho <-  make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-input$exp_filter_min_cell_number <- 1000
-v1.1_rho <- make_exp_plot(input, scEiaD_2020_v01, meta_filter %>% filter(CellType_predict %in% unique(meta_filter_v0$CellType_predict)))
-plot_grid(v0_rho, v1_rho, ncol = 1, labels = c('a','b'), rel_heights = c(1,1.5))
-```
-
-```{r studies_table, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=studies_table_cap}
-studies_table_cap <- supTab_cap(name = 'studies_table', caption = 'Studies, tissues, and number of cells across scEiaD')
-
-meta_filter %>% group_by(Citation, PMID, organism, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable()
-```

### 2022-07-07 src/diff_data_pull.R
fix bug, save image

@@ -9,10 +9,10 @@ library(glue)
 library(tictoc)
 
 tic()
-data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
-data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
+#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 
 
 library(pool)
@@ -259,7 +259,7 @@ org_diff_diff_genes <- diff_tab_big %>%
 flip_flop_genes <- scEiaD_2020_v01 %>% tbl('diff_testing_genes') %>% collect() %>% filter(grepl(org_diff_diff_genes %>% pull(ID) %>% paste0(., collapse = '|'), Gene)) %>% pull(Gene)
 flip_flop_long <- scEiaD_2020_v01 %>%
   tbl("diff_testing") %>%
-  filter(Gene %in% genes) %>%
+  filter(Gene %in% flip_flop_genes) %>%
   collect() %>%
   mutate(Symbol = str_extract(Gene, '^\\w+ ') %>% gsub(' $', '', .)) %>% 
   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
@@ -291,3 +291,4 @@ flip_flop_base <- row.names(flip_flop_wide) %>% gsub('.*_','',.)
 # hm_flipflop
 toc()
 
+save.image(file = 'data/diff_data_pull.Rdata')

### 2022-07-07 manuscript_00.Rmd
bug fix

@@
 
+load('data/diff_data_pull.Rdata') # from src/diff_data_pull.R
+
 scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vGiga}/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite'))
@@
 
-load('data/diff_data_pull.Rdata') # from src/diff_data_pull.R
 toc()
@@
 ```{r scEiaD_data_overview, fig.height = 35, fig.width = 25, echo = F, message = F, fig.cap=scEiaD_data_overview_cap}
-tic()
+
 num1 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(Tissue) %>% unique() %>% length()
@@
 
-exp_plot_giga<- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+exp_plot_down <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
 
 input$exp_plot_genes <- up
-exp_plot_iovs <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-a_legend <- get_legend(exp_plot_giga)
-plot_grid(plot_grid(exp_plot_giga + theme(legend.position="none") + xlab(''),
-          exp_plot_iovs + theme(legend.position="none") + ylab('') + xlab(''), labels = c('a','b')),
+exp_plot_up <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+a_legend <- get_legend(exp_plot_down)
+plot_grid(plot_grid(exp_plot_down + theme(legend.position="none") + xlab(''),
+          exp_plot_up + theme(legend.position="none") + ylab('') + xlab(''), labels = c('a','b')),
           a_legend, ncol = 1, rel_heights = c(1,0.2))

### 2022-07-12 manuscript_00.Rmd
fig and text updates

@@
 library(tictoc)
-tic()
-#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
-#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+#tic()()
+data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
+data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 # setup caption-ing
@@
 
-  Examples of single cell transcriptomics the retina date begin in the early 2000s where researchers [@hagstrom_cone_2000; @jabs_evidence_2000; @li_localization_2004; @lindqvist_single_2002-1; @reid_retinoschisin_2003] generally used RT-PCR approaches to amplify select transcripts in individual cells.
-Later, researchers used microarray platforms to analyze a broader array of transcripts researchers [@cherry_development_2009; @gustincich_gene_2004-1; @kim_identification_2008-1; @roesch_transcriptome_2008; @trimarchi_individual_2008; @trimarchi_molecular_2007-1; @wahlin_method_2004].
-The modern era of single cell transcriptomics is denoted by the introduction of droplet-based technologies , which enabled quantitation of thousands of cells in a single experiment; in Macoksko et al.
+  Single cell transcriptomics of the retina date to the early 2000s when researchers [@hagstrom_cone_2000; @jabs_evidence_2000; @li_localization_2004; @lindqvist_single_2002-1; @reid_retinoschisin_2003] used RT-PCR approaches to amplify select transcripts in small numbers of cells.
+Later, researchers used microarray platforms to analyze a broader array of transcripts [@cherry_development_2009; @gustincich_gene_2004-1; @kim_identification_2008-1; @roesch_transcriptome_2008; @trimarchi_individual_2008; @trimarchi_molecular_2007-1; @wahlin_method_2004].
+The modern era of single cell transcriptomics is marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment; in Macoksko et al.
 the adult mouse retina was profiled [@klein_droplet_2015; @macosko_highly_2015].
-The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by producing over 30 single cell transcriptome based studies of the eye since the Macosko publication.
+The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since Macosko publication.
   These publications's foci includes, but is not limited to, studying developmental transcriptomic dynamics, identifying genes to distinguish core cell types, proposing finer gradations of cell types within existing categories, and comparing ocular transcriptomes between organisms.
-Outside querying of these datasets often requires a long series of steps.
+Independent querying of these datasets requires a labrious series of steps.
 Briefly, one would have to identify the relevant publication, find the associated data deposit, download the cell - count matrix, find the cell label table (and optionally the t-SNE or UMAP coordinates), and then load the data into R or python to run the desired queries.
-If one desired to compare *across* publications, then one would also have to normalize the cell type labels and optionally run a batch-effect removal algorithm.
+If one desired to compare *across* publications, then one would also have to normalize the cell type labels and would likely have to re-quantify the counts with consistent bioinformatic tooling and run a batch-effect removal algorithm.
 It is desirable, perhaps even necessary, to use multiple datasets to assess gene expression to confirm whether a gene expression is consistent and reproducible.
-Reliance on a single dataset can be problematic as there can be technical issues - for example the large consortium dataset GTEx has persistant contamination from highly expressed genes across tissues [@nieuwenhuis_consistent_2020].
-  Web-based resources to query gene expression are powerful means to democratize access to large datasets as the computationally intensive steps above can be run once and then shared to all.
-Single cell RNA-seq resources that contain ocular tissues include the UCSC Cell Browser, Tabula Muris, PanglaoDB, and the Protein Atlas [@franzen_panglaodb_2019; @karlsson_singlecell_2021; @noauthor_single-cell_nodate; @speir_ucsc_2021].
+Reliance on a single dataset can be problematic as there can be technical issues - for example the large consortium dataset GTEx has persistent contamination from highly expressed genes across tissues [@nieuwenhuis_consistent_2020].
+  Web-based resources to query gene expression are powerful means to democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device.
+Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Muris, PanglaoDB, and the Protein Atlas [@franzen_panglaodb_2019; @karlsson_singlecell_2021; @noauthor_single-cell_nodate; @speir_ucsc_2021].
 Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging.
-The Spectacle project is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigt_spectacle_2020].
-Each data deposit is independently presented, which makes cross study queries difficult.
-We recently created a unified ocular resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space [@swamy_building_2021].
-We have built upon this dataset, adding more datasets, improving the quality control, removing ambient noise, improved the differential gene testing structure, and finally created a powerful, reactive, and fast web app to query the database.
-We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, genes which have species-specific expression cell type patterning, propose genes XYZ differentially expressed between the fovea and peripheral cones, and
-
-## Web atlases containing ocular data have substantial limitations
-
-Data difficult to access and cross comparisons impossible
-
-But important as seeing trends trends across different groups strongly demonstrates that 
-the signal is biological, not technical.
-
-## What we did
-
-We published the ocular focused scEiaD v2021 resource in 2021, which was one of the first single cell meta-atlases.
-To broaden the usage of this powerful database, we restructured the data in a custom sqlite-based database and built a reactive web application around it.
+In contrast to these, the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigt_spectacle_2020].
+However each data deposit is independently processed, which makes cross study queries difficult.
+We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space [@swamy_building_2021].
+  To broaden the usage of this powerful database, we restructured the data into a custom sqlite-based database and built a reactive web application around it.
 As many new studies were published since our last data build, we re-ran the scEiaD pipeline in March of 2022.
@@
 In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values.
+We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, genes which have species-specific expression cell type patterning, and finally propose genes XYZ that differentially expressed between the fovea and peripheral cones.
 
@@
 
-```{r scEiaD_data_overview, fig.height = 35, fig.width = 25, echo = F, message = F, fig.cap=scEiaD_data_overview_cap}
+```{r scEiaD_data_overview, fig.height = 16, fig.width = 20, echo = F, message = F, fig.cap=scEiaD_data_overview_cap}
 
@@
 C.
-Number of cells present across {num3} curated ocular-derived cell types."))
+Number of cells present across {num3} curated ocular-derived cell types.
+D.
+Proportion of cell types (predicted) across each study (split by species and tissue)."))
 
@@
 
-a_aa_b <- plot_grid(a,aa,b, ncol = 1, align = 'hv', rel_heights = c(10, 1.5,3.5), labels = 'a')
+a_aa_b <- plot_grid(a + scale_y_continuous(guide = guide_axis(check.overlap = TRUE)),
+                    aa + scale_y_continuous(guide = guide_axis(check.overlap = TRUE)),
+                    b + scale_y_continuous(guide = guide_axis(check.overlap = TRUE)), 
+                    ncol = 1, align = 'hv', rel_heights = c(10, 1.5,3.5), labels = 'a')
+org_legend <- get_legend(d)
 c_d <- plot_grid(c,d + theme(legend.position = "none"), rel_heights = c(3,10), ncol = 1, align= 'hv', axis = 'l',labels = c('b','c'))
@@
   filter(Organ == 'Eye') %>% 
-  mutate(Citation = paste(str_trunc(Citation, width = 20), str_trunc(Covariate, width = 12, side = 'left',ellipsis = ''))) %>% 
+  #mutate(Citation = paste(str_trunc(Citation, width = 20), str_trunc(Covariate, width = 12, side = 'left',ellipsis = ''))) %>% 
+  mutate(Citation = str_trunc(Citation, width = 20)) %>% 
   group_by(Citation, organism, Tissue, CellType_predict) %>% 
@@
 
-e <- plot_grid(proportion_bar_plot, labels = c('d'))
-
-plot_grid(a_aa_b, c_d , e, ncol = 3, rel_widths = c(1,1.6, 2.2))
-toc()
+e <- plot_grid(proportion_bar_plot+ theme(legend.position = "none"), labels = c('d'))
+ct_legend <- get_legend(proportion_bar_plot)
+fig_overview_legends <- plot_grid(org_legend, ct_legend,ncol=1)
+plot_grid(a_aa_b, c_d , e ,
+          fig_overview_legends, ncol =4, rel_widths = c(1,1.6, 2.2, 0.8))
+#toc()()
 ```
@@
 The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are not ocular focused, but do contain some ocular data.
-The PanglaoDB and UCSC are resources which have ingress-ed a large number of published single cell datasets The UCSC Cell Browser has around 62 thousand ocular single cell transcriptomes while the PangloaDB has about 47 thousand.
+The PanglaoDB and UCSC are resources which have ingressed a large number of published single cell datasets The UCSC Cell Browser has around sixty thousand ocular single cell transcriptomes while the PangloaDB has about 47 thousand.
 However, these resources provide no curation of the metadata and process the data for each study individually.
 Thus each study is effectively silo-ed off from the rest and cross comparisons are impractical.
-The Tabula Sapiens project is an effort to sequence all cells in the mouse and thus the ocular data present, though relatively small (at around 10 thousand), can be more readily cross compared to non-ocular tissues / cells.
-The Protein Atlas resource, which has aggregated protein labels (????) information has recently used one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g.
+The Tabula Sapiens project is an effort to sequence all cells in the mouse; while it only contains around around 10 thousand ocular cells it can be more readily cross compared to non-ocular tissues / cells.
+The Protein Atlas resource has recently used one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g.
 one can see whether a gene of interest is enriched in Cones).
 The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has selected a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app.
-PLAE is distinguished from these resources by hand curation of the metadata, re-processing of the data from the raw sequence, and, most crucially, integrating the independent studies to deliver cross study comparisons.
+PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence, and, most crucially, integrating the independent studies to deliver cross study comparisons.
 
 ```{r web_resources, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=web_resources_cap}
-tic()
+#tic()()
 i = 16.5 # width of the side borders in the word_document output (in centimeters)
@@
   as_tibble()
-colnames(feature_table) <- c('Resource Name', 'Independent Datasets', 'Harmonization', 'Ocular and non-Ocular Data', 'Ocular Cell Count\n(thousands)', 'URL')
-feature_table %>% flextable() %>% width(., width = (w/(ncol(feature_table)))) 
+colnames(feature_table) <- c('Resource Name', 'Independent\nDatasets', 'Harmonization', 'Ocular and\nnon-Ocular\n Data', 'Ocular Cell\nCount\n(thousands)', 'URL')
+feature_table %>% flextable() %>%  fontsize(size = 10, part = 'all') %>% autofit() %>% fit_to_width(7)#width(., width = c(1,0.75,0.75,0.75,0.75,3))
 # as of 2021 10 25
-toc()
+#toc()()
 ```
@@
 
-#```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
-
-#```
+```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
+PLAE_screenshots_cap <- fig_cap(name='PLAE_screenshots', caption = 'Screenshots from PLAE app.
+A.
+UMAP view showing CRX expression with cell type labels, B.
+Information table showing expression of Crx across celltypes and organism.
+C.
+Box plot of selected genes in a user selected cell types across different studies.
+D.
+In silico in-situ of Rho.
+E.
+Dotplot demonstating efficient display of gene expression across many cell types.')
+knitr::include_graphics('fig/fig1.png')
+```
 
@@
 This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021].
-Once the aggregated matrix was con structured, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
+Once the aggregated matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
 We tested each species (Human, Macaque, Mouse) separately.
@@
 
-## Projection of outside data onto scEiaD allows for ... (check lung atlas paper)
+## Improvements in data processing ...
 
-##
+## The data underlying PLAE is available in several forms to enable custom queries
 

### 2022-07-12 tester.Rmd
fig and text updates

@@
 
-load('data/diff_data_pull.Rdata') # from src/diff_data_pull.R
+#load('data/diff_data_pull.Rdata') # from src/diff_data_pull.R
 toc()
@@
 
-```{r, fig.width=4, fig.height=15, echo = F, message = F}
-# tic()
-# # flip_flop diff genes
-# down_tab <- down_tab_big %>% 
-#   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
-#   filter(Base %in% ws_oct, 
-#          Against == 'All')
-# 
-# org_diff_diff_genes <- diff_tab_big %>% 
-#   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
-#   filter(gene_ct_combo %in% down_tab$gene_ct_combo, 
-#          Against == 'All') %>% 
-#   arrange(padj) %>% mutate(x = paste0(Symbol, ' (', ID, ')')) 
-# 
-# flip_flop_genes <- scEiaD_2020_v01 %>% tbl('diff_testing_genes') %>% collect() %>% filter(grepl(org_diff_diff_genes %>% pull(ID) %>% paste0(., collapse = '|'), Gene)) %>% pull(Gene)
-# flip_flop_long <- scEiaD_2020_v01 %>%
-#   tbl("diff_testing") %>%
-#   filter(Gene %in% genes) %>%
-#   collect() %>%
-#   mutate(Symbol = str_extract(Gene, '^\\w+ ') %>% gsub(' $', '', .)) %>% 
-#   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
-#   filter(gene_ct_combo %in% org_diff_diff_genes$gene_ct_combo,
-#          Group == 'CellType (Predict)',
-#          Against == 'All') %>%
-#   select(Gene, Organism, log2FoldChange, Base)
-# 
-# flip_flop_wide <- flip_flop_long %>% pivot_wider(values_from = log2FoldChange, names_from = Organism)
-# flip_flop_organism <- colnames(flip_flop_wide)[c(3,4,5)]
-# flip_flop_wide <- flip_flop_wide %>% data.frame()
-# row.names(flip_flop_wide) <- paste0(flip_flop_wide$Gene,'_', flip_flop_wide$Base)
-# 
-# flip_flop_wide <- flip_flop_wide[,-c(1,2)]
-# 
-# 
-# 
-# flip_flop_genes <- row.names(flip_flop_wide) %>% gsub(' \\(.*','',.)
-# 
-# flip_flop_base <- row.names(flip_flop_wide) %>% gsub('.*_','',.)
+```{r web_resources, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=web_resources_cap}
+#tic()()
+i = 16.5 # width of the side borders in the word_document output (in centimeters)
+w = i*0.3937 # width of the side borders in the word_document output (in inches)
 
-hm_flipflop <- ComplexHeatmap::Heatmap(flip_flop_wide, column_labels = flip_flop_organism, 
-                                       row_split = flip_flop_base, show_row_dend = FALSE, 
-                                       row_labels = flip_flop_genes, 
-                                       column_title_rot = 90, 
-                                       name = 'log2FC', use_raster = TRUE,
-                                       row_title_rot = 0)
-hm_flipflop
-#toc()
+web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
+
+feature_table <- rbind(
+  c('PLAE','Yes','Yes', 'Yes', '1136 ', 'https://plae.nei.nih.gov'),
+  c('Spectacle', 'Yes' ,'No', 'No', '1500', 'https://singlecell-eye.org'),
+  c('UCSC Cell Browser', 'Yes', 'No', 'Yes', '62', 'https://cells.ucsc.edu' ),
+  c('Tabula Sapiens', 'No', 'N/A', 'Yes', '10', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
+  c('PanglaoDB', 'Yes', 'No', 'Yes', '47', 'https://panglaodb.se'),
+  c('Protein Atlas', 'Yes*','No', 'Yes', '5',  'https://www.proteinatlas.org/humanproteome/celltype')) %>% 
+  as_tibble()
+colnames(feature_table) <- c('Resource Name', 'Independent\nDatasets', 'Harmonization', 'Ocular and\nnon-Ocular\n Data', 'Ocular Cell\nCount\n(thousands)', 'URL')
+feature_table %>% flextable() %>%  fontsize(size = 10, part = 'all') %>% autofit() %>% fit_to_width(7)#width(., width = c(1,0.75,0.75,0.75,0.75,3))
+# as of 2021 10 25
+#toc()()
 ```

### 2022-07-14 manuscript_00.Rmd
text updates

@@
 "Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function.
-Genes were considered to be differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2.
-To enhance the specificity of the
+Genes were considered to be significantly differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2.
 
@@
 
-We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping.
-The created a counts matrix which has statistical properties matching that of "bulk" RNA seq experiment matrices and thus allowing for mature bulk RNA-seq tooling to be used.
+We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping (see Methods).
+This created a counts matrix which has statistical properties approximating "bulk" RNA seq experiment matrices.
+This allowed use to use the more mature bulk RNA-seq tooling.
 This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021].
-Once the aggregated matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
+Once the pseudo-bulk matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
 We tested each species (Human, Macaque, Mouse) separately.
-We had an insufficient number of chick studies to conduct differential testing.
+Chicken was not included as we had an insufficient number of independent studies to conduct differential testing.
 The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential gene expression.
@@
 cones versus all remaining cell types) or in a pair-wise manner (cone against rod).
-For the one against all testing, we
+Unless stated otherwise, we will be discussing the results of the "against all" contrast for the remainder of the paper.
 
-```{r, fig.width=12}
-# tic()
-# # number of celltype (predict) - organism - study counts (with >=50 cells)
-# # ws == Well Supported
-# ws_table <- meta_filter %>%  
-#   filter(Organ == 'Eye') %>% 
-#   group_by(organism, CellType_predict, study_accession) %>% 
-#   count() %>% 
-#   filter(n >= 50) %>% 
-#   ungroup() %>% 
-#   group_by(organism, CellType_predict) %>% 
-#   count() %>% 
-#   pivot_wider(values_from = n, names_from = organism) %>% 
-#   filter(!is.na(CellType_predict)) %>% 
-#   replace(is.na(.), values = 0) %>% 
-#   arrange(CellType_predict) %>% 
-#   rename(`CellType (Predict)` = CellType_predict) %>% 
-#   mutate(`Well Supported` = case_when(`Homo sapiens` >= 2 & `Mus musculus` >= 2 ~ 'Yes',
-#                                       TRUE ~ 'No'))
-# 
-# # fix  occasional NA
-# # data made from src/ensdb.R
-# diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
-# down_tab_big <- read_tsv('data/down_tab_big.tsv.gz')
-# gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
-# diff_tab <- diff_tab_big %>% 
-#   filter(Against == 'All', 
-#          Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% 
-#                       pull(CellType) %>% unique()))
-# 
-# 
-# diff_tab_sup <- diff_tab %>% 
-#   group_by(Base, Organism) %>% 
-#   summarise(Count = n()) %>% 
-#   rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
-#   pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
-#   as_tibble()
-# 
-# diff_tab_sup <- bind_rows(
-#   diff_tab_sup,
-#   ws_table$`CellType (Predict)`[!ws_table$`CellType (Predict)` %in% 
-#                                   diff_tab_sup$`CellType (Predict)`] %>% 
-#     enframe() %>% 
-#     select(`CellType (Predict)` = value) %>%
-#     mutate(`Homo sapiens` = 0, `Mus musculus` = 0, `Macaca fascicularis` = 0) %>% 
-#     arrange(`CellType (Predict)`)
-# )
-# 
-# write_csv(diff_tab, file = 'supplemental_files/differential_expression_table.csv.gz')
-
-diff_tab_sup %>% 
-  flextable()
-
-ws_table %>% 
-  flextable()
-
-diff_tab_barplot <- diff_tab_sup %>% 
-  pivot_longer(-`CellType (Predict)`, names_to = 'Organism', values_to = 'Diff Gene Count') %>% 
-  ggplot(aes(x=`CellType (Predict)`, y = `Diff Gene Count`, fill = Organism )) + 
-  geom_bar(stat='identity', position = 'dodge') + 
-  cowplot::theme_cowplot() + coord_flip()
-
-#toc()
-```
+Across the cell types of the eye, we find `r diff_tab %>% nrow()` gene - celltype combinations that are significantly differentially expressed.
+We define significant as a log2 fold change greater than two as well as a p-value (corrected) less than $1 x 10^-5.
+We find a mean of
+`r diff_tab_sup$`Homo sapiens` %>% mean() %>% round()`, 
+`r diff_tab_sup$`Mus musculus` %>% mean() %>% round()`, and 
+`r diff_tab_sup$`Macaca fascicularis` %>% mean() %>% round()` differentially expressed genes in human, macaque, and mouse respectively (`r supTab_cap(name = 'diff_tab_sup_tab', display = 'cite')`).
+These variable results directly reflect the differing number of independent studies - mouse and human have substantially more than macaque (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
+The full differential table is made available as supplementary file "differential_expression_table.csv.gz." While this differential expression table is large, it can be mined further to give more precise results.
 
 ## Well supported ocular cell types identify high confidence cell type markers
+
+Cell type marker genes can be proposed on the basis of existing knowledge about their function and/or based on high expression differences when comparing to other cell types.
+We can further narrow down the list of markers by using the  diersity of studies and organisms in scEiaD.
+We first identify a set of "well supported" cell types which we define as cell types which have two or more independent studies across both human and mouse (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
+These
 
@@
 
+```{r diff_tab_sup_tab, echo = F, message = F, fig.cap=diff_tab_sup_cap}
+diff_tab_sup_cap <- supTab_cap(name='diff_tab_sup', caption = 'Counts of signifantly differentially expressed genes for each cell type across the organisms')
+
+# tic()
+# # number of celltype (predict) - organism - study counts (with >=50 cells)
+# # ws == Well Supported
+# ws_table <- meta_filter %>%  
+#   filter(Organ == 'Eye') %>% 
+#   group_by(organism, CellType_predict, study_accession) %>% 
+#   count() %>% 
+#   filter(n >= 50) %>% 
+#   ungroup() %>% 
+#   group_by(organism, CellType_predict) %>% 
+#   count() %>% 
+#   pivot_wider(values_from = n, names_from = organism) %>% 
+#   filter(!is.na(CellType_predict)) %>% 
+#   replace(is.na(.), values = 0) %>% 
+#   arrange(CellType_predict) %>% 
+#   rename(`CellType (Predict)` = CellType_predict) %>% 
+#   mutate(`Well Supported` = case_when(`Homo sapiens` >= 2 & `Mus musculus` >= 2 ~ 'Yes',
+#                                       TRUE ~ 'No'))
+# 
+# # fix  occasional NA
+# # data made from src/ensdb.R
+# diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
+# down_tab_big <- read_tsv('data/down_tab_big.tsv.gz')
+# gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
+# diff_tab <- diff_tab_big %>% 
+#   filter(Against == 'All', 
+#          Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% 
+#                       pull(CellType) %>% unique()))
+# 
+# 
+# diff_tab_sup <- diff_tab %>% 
+#   group_by(Base, Organism) %>% 
+#   summarise(Count = n()) %>% 
+#   rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
+#   pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
+#   as_tibble()
+# 
+# diff_tab_sup <- bind_rows(
+#   diff_tab_sup,
+#   ws_table$`CellType (Predict)`[!ws_table$`CellType (Predict)` %in% 
+#                                   diff_tab_sup$`CellType (Predict)`] %>% 
+#     enframe() %>% 
+#     select(`CellType (Predict)` = value) %>%
+#     mutate(`Homo sapiens` = 0, `Mus musculus` = 0, `Macaca fascicularis` = 0) %>% 
+#     arrange(`CellType (Predict)`)
+# )
+# 
+# write_csv(diff_tab, file = 'supplemental_files/differential_expression_table.csv.gz')
+
+diff_tab_sup %>% 
+  flextable()
+
+ws_table %>% 
+  flextable()
+
+diff_tab_barplot <- diff_tab_sup %>% 
+  pivot_longer(-`CellType (Predict)`, names_to = 'Organism', values_to = 'Diff Gene Count') %>% 
+  ggplot(aes(x=`CellType (Predict)`, y = `Diff Gene Count`, fill = Organism )) + 
+  geom_bar(stat='identity', position = 'dodge') + 
+  cowplot::theme_cowplot() + coord_flip()
+
+#toc()
+
+diff_tab_sup_cap %>% flextable()
+```
+
+```{r well_supported_celltypes_tab, echo = F, message = F, fig.cap=well_supported_celltypes_tab_cap}
+well_supported_celltypes_tab_cap <- supTab_cap(name='well_supported_celltypes_tab', caption = 'Counts of independent studies for each cell type across the organisms')
+
+ws_table %>% flextable()
+```
+
 ```{r decontX, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=decontX_cap}

### 2022-07-14 manuscript_00.Rmd
updates

@@
 Cell type marker genes can be proposed on the basis of existing knowledge about their function and/or based on high expression differences when comparing to other cell types.
-We can further narrow down the list of markers by using the  diersity of studies and organisms in scEiaD.
-We first identify a set of "well supported" cell types which we define as cell types which have two or more independent studies across both human and mouse (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
-These
+We can further narrow down the list of markers by using the  diversity of studies and organisms in scEiaD to propose a set of community supported cell type markers.
+We first identify a set of well supported cell types which we define as cell types which have two or more independent studies across both human and mouse (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
+These cell types were then assessed to identify differentially expressed genes which met the following criteria: 1.
+padj less than $1 x 10^-4
+in both human and mouse, 2. mean log2 fold change greater than two, and 3. mean padj less than $1 x 10^-5
 

### 2022-07-14 manuscript_00.Rmd
test tweaks

@@
 #tic()()
-data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
-data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
+#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 # setup caption-ing
@@
 Across the cell types of the eye, we find `r diff_tab %>% nrow()` gene - celltype combinations that are significantly differentially expressed.
-We define significant as a log2 fold change greater than two as well as a p-value (corrected) less than $1 x 10^-5.
+We define significant as a log2 fold change greater than two as well as a p-value (corrected) less than $1 x 10^-5#.
 We find a mean of
-`r diff_tab_sup$`Homo sapiens` %>% mean() %>% round()`, 
-`r diff_tab_sup$`Mus musculus` %>% mean() %>% round()`, and 
-`r diff_tab_sup$`Macaca fascicularis` %>% mean() %>% round()` differentially expressed genes in human, macaque, and mouse respectively (`r supTab_cap(name = 'diff_tab_sup_tab', display = 'cite')`).
+`r diff_tab_sup$"Homo sapiens" %>% mean() %>% round()`, 
+`r diff_tab_sup$"Mus musculus" %>% mean() %>% round()`, and 
+`r diff_tab_sup$"Macaca fascicularis" %>% mean() %>% round()` differentially expressed genes in human, macaque, and mouse respectively (`r supTab_cap(name = 'diff_tab_sup_tab', display = 'cite')`).
 These variable results directly reflect the differing number of independent studies - mouse and human have substantially more than macaque (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
@@
 We can further narrow down the list of markers by using the  diversity of studies and organisms in scEiaD to propose a set of community supported cell type markers.
-We first identify a set of well supported cell types which we define as cell types which have two or more independent studies across both human and mouse (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
+We first identify a set of well supported cell types which we define as cell types which have two or more independent studies across both human mouse (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
 These cell types were then assessed to identify differentially expressed genes which met the following criteria: 1.
-padj less than $1 x 10^-4
-in both human and mouse, 2. mean log2 fold change greater than two, and 3. mean padj less than $1 x 10^-5
+padj less than $1 x 10^-4$
+in both human and mouse, 2. mean log2 fold change greater than two, and 3. mean padj less than $1 x 10^-5$
 
@@
 # 
-# consist_diff <- scEiaD_2020_v01 %>% 
-#   tbl("diff_testing") %>% 
-#   filter(Base %in% ws_oct) %>%  
-#   filter(Group == 'CellType (Predict)', 
-#          Against == 'All') %>% 
-#   group_by(Base, Gene) %>% 
-#   summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-4), `mean padj` = mean(padj)) %>% 
-#   filter(`mean log2FC` > 2, `sig count` > 1, `mean padj` < 1e-5) %>%  
-#   collect() %>% 
-#   mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
-#   left_join(gene_tab, by = 'GENEID') %>% 
-#   relocate(SYMBOL, GENEID) %>% 
-#   dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
-#   arrange(-`mean log2FC`)
-# 
+consist_diff2 <- scEiaD_2020_v01 %>%
+  tbl("diff_testing") %>%
+  filter(Base %in% ws_oct) %>%
+  filter(Group == 'CellType (Predict)',
+         Against == 'All',
+         Organism %in% c('Homo sapiens','Mus musculus')) %>%
+  group_by(Base, Gene) %>%
+  summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-5), `mean padj` = mean(padj)) %>%
+  filter(`mean log2FC` > 2, `sig count` > 1, `mean padj` < 1e-5) %>%
+  collect() %>%
+  mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>%
+  left_join(gene_tab, by = 'GENEID') %>%
+  relocate(SYMBOL, GENEID) %>%
+  dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>%
+  arrange(-`mean log2FC`)
+
 # # id diff genes only seen once across the cell types
@@
 
-consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n()) %>% ggplot(aes(x=`Base`, y = `Count` )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip() + xlab("CellTye (Predict)")
+consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n()) %>% ggplot(aes(x=`Base`, y = `Count` )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip() + xlab("CellType (Predict)")
 #toc()
@@
 
-diff_tab_sup %>% 
-  flextable()
-
-ws_table %>% 
-  flextable()
+# diff_tab_sup %>% 
+#   flextable()
+# 
+# ws_table %>% 
+#   flextable()
 
@@
 
-diff_tab_sup_cap %>% flextable()
+diff_tab_sup %>% flextable()
 ```

### 2022-07-17 manuscript_00.Rmd
jpdates

@@
 #tic()()
-#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
-#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
+data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 # setup caption-ing
@@
 We can further narrow down the list of markers by using the  diversity of studies and organisms in scEiaD to propose a set of community supported cell type markers.
-We first identify a set of well supported cell types which we define as cell types which have two or more independent studies across both human mouse (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
-These cell types were then assessed to identify differentially expressed genes which met the following criteria: 1.
+We first identify a set of well supported cell types which we define as cell types which have two or more independent studies across both human and mouse (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
+These cell types were then assessed to identify differentially expressed genes which met the following criteria across human, mouse, and macaque: 1.
 padj less than $1 x 10^-4$
-in both human and mouse, 2. mean log2 fold change greater than two, and 3. mean padj less than $1 x 10^-5$
-
-We filter down to only assess cell types with more than two independent datasets in both human and mouse.
-This leaves .....
-These well supported cell types can be used to more strenuously identify genes that are consistently differentially expressed in a cell type across multiple species.
-To define these well supported marker genes we only ass
-
-To identify genes that are differentially expressed across species we
+in two or more species, 2. mean log2 fold change greater than two, and 3. mean padj less than $1 x 10^-5$
 

### 2022-07-17 manuscript_00.Rmd
updates

@@
 # 
-consist_diff2 <- scEiaD_2020_v01 %>%
+consist_diff3 <- scEiaD_2020_v01 %>%
   tbl("diff_testing") %>%
@@
   summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-5), `mean padj` = mean(padj)) %>%
-  filter(`mean log2FC` > 2, `sig count` > 1, `mean padj` < 1e-5) %>%
+  filter(`mean log2FC` > 4, `sig count` > 1, `mean padj` < 1e-5) %>%
   collect() %>%
@@
 
-consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n()) %>% ggplot(aes(x=`Base`, y = `Count` )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip() + xlab("CellType (Predict)")
+abc <- consist_diff3 %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n()) %>% ggplot(aes(x=`Base`, y = `Count` )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip() + xlab("CellType (Predict)")
 #toc()

### 2022-07-19 manuscript_00.Rmd
tweak diff heatmap

--- 2022-07-17 manuscript_00.Rmd
@@
 # 
-consist_diff2 <- scEiaD_2020_v01 %>%
-  tbl("diff_testing") %>%
-  filter(Base %in% ws_oct) %>%
-  filter(Group == 'CellType (Predict)',
-         Against == 'All',
-         Organism %in% c('Homo sapiens','Mus musculus')) %>%
-  group_by(Base, Gene) %>%
-  summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-5), `mean padj` = mean(padj)) %>%
-  filter(`mean log2FC` > 2, `sig count` > 1, `mean padj` < 1e-5) %>%
-  collect() %>%
-  mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>%
-  left_join(gene_tab, by = 'GENEID') %>%
-  relocate(SYMBOL, GENEID) %>%
-  dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>%
-  arrange(-`mean log2FC`)
+# consist_diff <- scEiaD_2020_v01 %>%
+#   tbl("diff_testing") %>%
+#   filter(Base %in% ws_oct) %>%
+#   filter(Group == 'CellType (Predict)',
+#          Against == 'All',
+#          Organism %in% c('Homo sapiens','Mus musculus')) %>%
+#   group_by(Base, Gene) %>%
+#   summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-5), `mean padj` = mean(padj)) %>%
+#   filter(`mean log2FC` > 2.463, `sig count` > 1, `mean padj` < 1e-5) %>%
+#   collect() %>%
+#   mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>%
+#   left_join(gene_tab, by = 'GENEID') %>%
+#   relocate(SYMBOL, GENEID) %>%
+#   dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>%
+#   arrange(-`mean log2FC`)
 
@@
 
-diff_tab_sup %>% flextable()
+diff_tab_sup %>% flextable() %>% autofit() %>% fit_to_width(7)
 ```
@@
 
-ws_table %>% flextable()
+ws_table %>% flextable() %>% autofit() %>% fit_to_width(7)
 ```
@@
 
-meta_filter %>% group_by(Citation, PMID, organism, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable()
+meta_filter %>% group_by(Citation, PMID, organism, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable() %>% autofit() %>% fit_to_width(7)
 ```

### 2022-07-19 src/diff_data_pull.R

@@ -9,10 +9,10 @@ library(glue)
 library(tictoc)
 
 tic()
-#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
-#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
+data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 
 
 library(pool)
@@ -118,23 +118,24 @@ toc()
 tic()
 ws_oct <- ws_table %>% filter(`Well Supported` == 'Yes') %>% pull(`CellType (Predict)`)
 
-consist_diff <- scEiaD_2020_v01 %>% 
-  tbl("diff_testing") %>% 
-  filter(Base %in% ws_oct) %>%  
-  filter(Group == 'CellType (Predict)', 
-         Against == 'All') %>% 
-  group_by(Base, Gene) %>% 
-  summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-4), `mean padj` = mean(padj)) %>% 
-  filter(`mean log2FC` > 2, `sig count` > 1, `mean padj` < 1e-5) %>%  
-  collect() %>% 
-  mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
-  left_join(gene_tab, by = 'GENEID') %>% 
-  relocate(SYMBOL, GENEID) %>% 
-  dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
+consist_diff <- scEiaD_2020_v01 %>%
+  tbl("diff_testing") %>%
+  filter(Base %in% ws_oct) %>%
+  filter(Group == 'CellType (Predict)',
+         Against == 'All',
+         Organism %in% c('Homo sapiens','Mus musculus')) %>%
+  group_by(Base, Gene) %>%
+  summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-5), `mean padj` = mean(padj)) %>%
[…]

### 2022-07-19 manuscript_00.Rmd
text add, update consist diff

@@
 These variable results directly reflect the differing number of independent studies - mouse and human have substantially more than macaque (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
-The full differential table is made available as supplementary file "differential_expression_table.csv.gz." While this differential expression table is large, it can be mined further to give more precise results.
+The full differential table is made available as supplementary file "differential_expression_table.csv.gz." While this differential expression table is large, it can be further processed to give more precise results.
 
@@
 
-Cell type marker genes can be proposed on the basis of existing knowledge about their function and/or based on high expression differences when comparing to other cell types.
-We can further narrow down the list of markers by using the  diversity of studies and organisms in scEiaD to propose a set of community supported cell type markers.
+Cell type marker genes can be proposed on the basis of existing knowledge about their function and on high expression differences when comparing to other cell types.
+We can further narrow down the list of markers by using the high diversity of studies and organisms in scEiaD to propose a set of community supported cell type markers.
 We first identify a set of well supported cell types which we define as cell types which have two or more independent studies across both human and mouse (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
@@
 padj less than $1 x 10^-4$
-in two or more species, 2. mean log2 fold change greater than two, and 3. mean padj less than $1 x 10^-5$
+in two or more species, 2.
+mean log2 fold change greater than 2, and 3.
+mean padj less than $1 x 10^-5$.
+A final filter was applied to remove genes as candidates if they were differentially expressed (using the criteria above) in more than three different cell types.
+This left us with `r top_genes %>% nrow()` gene - cell type markers, a substantial reduction from the `r diff_tab %>% nrow()` we had initially.
+We provide a table of these genes as supplementary file "consistently_differentially_expressed_genes.csv.gz."
+
+To visualize these top markers we select up to eight genes (ordered by the mean log2 fold change across organisms) per well supported cell type for each and plot the log2 fold change of the gene expression relative to all other cell types (`r fig_cap(name = 'consist_genes_heatmap', displah = 'cite')`).
+We see that human, mouse, and macaque have similar expression pattterning of the proposed markers across the well supported cell types.
 
@@
 # uniq_g <- consist_diff %>% group_by(Symbol) %>% count() %>% filter(n == 1)
-consist_diff %>% group_by(Base) %>% summarise(Count = n())
-consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n())
+#consist_diff %>% group_by(Base) %>% summarise(Count = n())
+#consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n())
 
-consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n()) %>% ggplot(aes(x=`Base`, y = `Count` )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip() + xlab("CellType (Predict)")
+sup_tab_consist_diff <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n()) %>% ggplot(aes(x=`Base`, y = `Count` )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip() + xlab("CellType (Predict)")
 #toc()
@@
 
-```{r, fig.height=24, fig.width=12}
+```{r consist_genes_heatmap, fig.height=24, fig.width=12, fig.cap=consist_genes_heatmap_cap}
+consist_genes_heatmap_cap <- fig_cap(name='consist_genes_heatmap', caption = 'Rows are genes, columns are cell types and organisms.
+Rows are split by what cell type the gene is a marker for.
+Columns are split by the cell types.
+More red is a higher log2 fold change of the gene - cell type relative to all other cell types and blue is more negative.')
+
 # tic()
@@
 
-## Tease apart closely related cell types
+## Tease apart closely neurogenic and RPC cell types
 
-Neurogenic and RPC have very similar expression patterns.
-We leverage our more specific differential test to directly assess genes were are different between neurogenic cells and RPC.
+As we see in Neurogenic and RPC have very similar expression patterns.
+We leverage our more specific pairwise differential test to directly assess genes that are different between neurogenic cells and RPC.
 We show the top 5 genes that are both more highly expressed in RPC relative to neurogenic cell and vice versa.

### 2022-07-19 src/diff_data_pull.R

@@ -126,7 +126,7 @@ consist_diff <- scEiaD_2020_v01 %>%
          Organism %in% c('Homo sapiens','Mus musculus')) %>%
   group_by(Base, Gene) %>%
   summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-5), `mean padj` = mean(padj)) %>%
-  filter(`mean log2FC` > 2.463, `sig count` > 1, `mean padj` < 1e-5) %>% # quartile marking
+  filter(`mean log2FC` > 2, `sig count` > 1, `mean padj` < 1e-5) %>% 
   collect() %>%
   mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>%
   left_join(gene_tab, by = 'GENEID') %>%
@@ -135,7 +135,8 @@ consist_diff <- scEiaD_2020_v01 %>%
   arrange(-`mean log2FC`)
 
 # set genes to one cell types (if there are multiple, then pick the one with the highest mean log2FC)
-top_genes <- consist_diff %>% group_by(Symbol) %>% slice_max(., n=1, order_by = `mean log2FC`)
+# remove genes that are diff expressed in more than 3 diff cell types
+top_genes <- consist_diff %>% group_by(Symbol) %>% summarise(Count = n()) %>% filter(Count <= 3) %>% left_join(consist_diff %>%  group_by(Symbol) %>% slice_max(., n=1, order_by = `mean log2FC`), by = 'Symbol')
 # consist_diff %>% group_by(Base) %>% summarise(Count = n())
 # consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n())
 
@@ -182,6 +183,10 @@ consist_diff_base <- consist_diff_genes %>% as_tibble() %>% rename(Symbol = valu
 #                                            name = 'log2FC', use_raster = TRUE,
 #                                            row_title_rot = 0)
 # ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(2, 2, 32, 2), "mm"), row_title = "Marker Genes")
+
+# write consist diff file
[…]

### 2022-07-21 manuscript_00.Rmd
neuro vs rpc diff testing updates

@@
 
-```{r consist_genes_heatmap, fig.height=24, fig.width=12, fig.cap=consist_genes_heatmap_cap}
+```{r consist_genes_heatmap, fig.height=20, fig.width=10, fig.cap=consist_genes_heatmap_cap}
 consist_genes_heatmap_cap <- fig_cap(name='consist_genes_heatmap', caption = 'Rows are genes, columns are cell types and organisms.
@@
 
-```{r, fig.width=15, fig.height=6, echo = F, message = F}
+```{r, fig.width=15, fig.height=5, echo = F, message = F}
 # tic()
@@
 # 
-# down <- pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` < 0) %>% head(5) %>% pull(Gene)
-# up <-  pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` > 0) %>% head(5) %>% pull(Gene)
+down <- pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` < 0, `mean padj` < 0.05) %>% head(5) %>% pull(Gene)
+up <-  pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` > 0) %>% head(9) %>% pull(Gene)
 source('src/make_exp_plot.R')
@@
 input$exp_plot_groups <- c('study_accession')
-input$exp_plot_facet <- c('CellType_predict')
+input$exp_plot_facet <- c('CellType_predict', 'organism')
 input$exp_plot_genes <- down
@@
 #input$exp_plot_ylab <- '% of Cells Detected'
+#input$exp_plot_groups <- c('organism')
 input$exp_plot_col_num <- 16
-input$exp_filter_min_cell_number <- 50
-input$flip_facet <- TRUE
+input$exp_filter_min_cell_number <- 25
+input$flip_facet <- FALSE
+input$figure <- TRUE
 
@@
 a_legend <- get_legend(exp_plot_down)
-plot_grid(plot_grid(exp_plot_down + theme(legend.position="none") + xlab(''),
-          exp_plot_up + theme(legend.position="none") + ylab('') + xlab(''), labels = c('a','b')),
+neuro_plot <- plot_grid(plot_grid(exp_plot_down + theme(legend.position="none") + xlab(''),
+          exp_plot_up + theme(legend.position="none") + ylab('') + xlab(''), labels = c('b','c'),
+
+          rel_widths = c(0.1,1)),
           a_legend, ncol = 1, rel_heights = c(1,0.2))
-#toc()
+
+lu_et_al <- c("PTTG1", "SCG3", "SHD", "SOX11", "ONECUT2", "VSX1", "ASCL1", "TCF4", "SOX4", "RORB", "GADD45G", "MYBL1", "MFAP4", "HES6", "NEUROD1", "C8orf46", "OTX2", "OLIG2", "INSM1", "PCBP4", "NEUROD4", "FAM131C", "HES4", "HES5", "HMGA1", "YBX1", "GADD45A", "ATOH7", "DLX2", "DLX1", "SFRP1", "MYB", "NFIB", "SOX9", "ZFP36L1", "FOS", "JUN", "ZFP36L2", "SOX2", "HES1", "EGR1", "JUNB", "NR4A1", "ETV5", "FOXP1", "ID1", "ID2", "ID3", "ID4", "FOXP2")
+
+g_lu <- scEiaD_2020_v01 %>% tbl('genes') %>% collect() %>% filter(grepl(paste(paste0('^',lu_et_al), collapse = ' |'), Gene)) %>% pull(Gene)
+
+lu_long <-  scEiaD_2020_v01 %>% 
+  tbl("diff_testing") %>% 
+  filter(Gene %in% g_lu,
+         Against  == "RPC",   
+         Group == 'CellType (Predict)', 
+         Base == "Neurogenic Cell") %>% 
+  #group_by(Base, Against, Gene) %>% 
+  #summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
+  #arrange(-`mean log2FC`) %>%   
+  collect() %>% 
+  mutate(Gene = str_extract(Gene, '\\w+ ') %>% 
+           gsub(' ', '', .)) %>% 
+  mutate(Gene = factor(Gene, levels = rev(lu_et_al)))
+
+#order <- lu_et_al %>% enframe() %>% mutate(Gene = value) %>% left_join(g_lu %>% enframe() %>% mutate(full = value, Gene = str_extract(value, '\\w+ ') %>% gsub(' ', '', .)) %>% select(-value), by = 'Gene') %>% data.frame() %>% filter(!is.na(full)) %>% pull(full)
+
+lu_rep_plot <- lu_long %>% 
+  mutate(Significance = case_when(padj < 0.01 ~ '**', padj < 0.1 ~ '*', TRUE ~ '')) %>% 
+  ggplot(aes(x=Gene,y=log2FoldChange,  label = Significance)) + 
+  geom_bar(stat='identity', position = position_dodge2(), fill = 'gray') + 
+  geom_errorbar(aes(ymin = log2FoldChange - lfcSE,
+                    ymax = log2FoldChange + lfcSE), width = 0.5,
+                position = position_dodge2()) +
+  coord_flip() + 
+  cowplot::theme_cowplot()  + geom_text(aes(y=7), position = position_dodge2()) + facet_wrap(~Organism)
+#g_lu[!g_lu %in% lu_long$Gene]
+
+cowplot::plot_grid(lu_rep_plot, neuro_plot, rel_widths = c(0.7,1), labels = c('a'))
 ```
 
-```{r, fig.width=5, fig.height=15, echo = F, message = F}
+```{r, fig.width=4, fig.height=12, echo = F, message = F}
 # tic()
@@
 
+gene_anno <- ComplexHeatmap::rowAnnotation(Duplicated = flip_flop_genes %in% (flip_flop_genes[duplicated(flip_flop_genes)]),
+                                           col = list(Duplicated = c("TRUE" = "green", "FALSE" = "gray")))
 hm_flipflop <- ComplexHeatmap::Heatmap(flip_flop_wide, column_labels = flip_flop_organism, 
@@
                                        column_title_rot = 90, 
+                                       right_annotation = gene_anno,
                                        name = 'log2FC', use_raster = TRUE,

### 2022-07-21 src/diff_data_pull.R

@@ -145,13 +145,14 @@ toc()
 #######################################
 
 tic()
-g <- top_genes %>% select(Symbol, Base) %>% left_join(consist_diff, by = c("Symbol", "Base"))  %>% group_by(Base) %>% slice_max(n=8, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
+g <- top_genes %>% select(Symbol, Base) %>% filter(!Base %in% c('Early RPC','Late RPC')) %>% left_join(consist_diff, by = c("Symbol", "Base"))  %>% group_by(Base) %>% slice_max(n=8, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
 
 consist_diff_long <- scEiaD_2020_v01 %>%
   tbl("diff_testing") %>%
   filter(Gene %in% g) %>%
   collect() %>%
   filter(Base %in% ws_oct,
+         !Base %in% c('Early RPC','Late RPC'), 
          Group == 'CellType (Predict)',
          Against == 'All') %>%
   select(Gene, Organism, log2FoldChange, Base)

### 2022-07-21 src/make_exp_plot.R

@@ -77,7 +77,7 @@ make_exp_plot <- function(input, db, meta_filter){
            `Total Cells` = Count,
            `Mean log2(Counts + 1)` = Expression,
            `% of Cells Detected` = `%`) %>%
-    filter(!is.na(CellType_predict)) %>% 
+    filter(!is.na(!!as.symbol(input$exp_filter_cat))) %>% 
     mutate(`Mean log2(Counts + 1)` = case_when(is.na(`Mean log2(Counts + 1)`) ~ 0, TRUE ~ `Mean log2(Counts + 1)`)) %>% 
     filter(`Total Cells` > as.integer(input$exp_filter_min_cell_number))
   # expand missing genes (nothing detected) to all genes used
@@ -103,9 +103,20 @@ make_exp_plot <- function(input, db, meta_filter){
       scale_radius(range=c(2, 6)) +
       scale_colour_manual(values = rep(c(pals::alphabet() %>% unname()), 20)) +
       theme(legend.position="bottom") +
-      facet_wrap(ncol = as.numeric(input$exp_plot_col_num), scales = 'free', 'Gene')
-  }
-  else {
+      facet_wrap(ncol = as.numeric(input$exp_plot_col_num), scales = 'free', 'Gene') 
+  } else if (input$figure){
+    box_data  %>% 
+      mutate(Gene = str_extract(Gene, '^\\w+ ') %>% gsub(" $", '', .)) %>%
+      ggplot(aes(x=!!as.symbol(input$exp_plot_facet), y = !!as.symbol(input$exp_plot_ylab), color = !!as.symbol(grouping_features), group = CellType_predict, shape = organism)) +
+      geom_boxplot(color = 'black', outlier.shape = NA) +
+      ggbeeswarm::geom_quasirandom(size = 4, groupOnX = TRUE) +
+      cowplot::theme_cowplot() +
+      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
+      scale_radius(range=c(2, 6)) +
+      scale_colour_manual(values = rep(c(pals::alphabet() %>% unname()), 20)) +
+      theme(legend.position="bottom") +
+      facet_wrap(ncol = as.numeric(input$exp_plot_col_num), scales = 'free', 'Gene') 
+  }  else {
     box_data %>%
       mutate(Gene = str_extract(Gene, '^\\w+ ') %>% gsub(" $", '', .)) %>%
[…]

### 2022-07-21 manuscript_00.Rmd
tweaking diff neuro rpc test

@@
 
-To visualize these top markers we select up to eight genes (ordered by the mean log2 fold change across organisms) per well supported cell type for each and plot the log2 fold change of the gene expression relative to all other cell types (`r fig_cap(name = 'consist_genes_heatmap', displah = 'cite')`).
+To visualize these top markers we select up to eight genes (ordered by the mean log2 fold change across organisms) per well supported cell type for each and plot the log2 fold change of the gene expression relative to all other cell types (`r fig_cap(name = 'consist_genes_heatmap', display = 'cite')`).
 We see that human, mouse, and macaque have similar expression pattterning of the proposed markers across the well supported cell types.
@@
 
-## Tease apart closely neurogenic and RPC cell types
+## Tease apart the closely neurogenic and RPC cell types
 
-As we see in Neurogenic and RPC have very similar expression patterns.
-We leverage our more specific pairwise differential test to directly assess genes that are different between neurogenic cells and RPC.
-We show the top 5 genes that are both more highly expressed in RPC relative to neurogenic cell and vice versa.
+We see that the neurogenic cells and RPCs have very similar expression patterns.
+These cell types are closely related as the neurogenic cells are derived from the multi-potent RPC to later specify neural cell types.
+First we use the set of genes identified by Lu et al.
+[@lu_single-cell_2020] in their study of human retinal differentiation to determine whether our broader scEiaD retains the same trends (`r fig_cap(name = 'neuro', display = 'cite')`A).
+As expected, we see several confirming patterns.
+First, we see most of the gene expression changes between RPC and neurogenic are matched between mouse and human.
+Next we see that the direction of the changes largely matches Lu et al (`r supFig_cap(name='lu_et_al', display = 'cite'`).
+Third, we see a small number of genes are confirmed as consistently differentially expressed with a padj < 0.1.
 
-```{r, fig.width=15, fig.height=5, echo = F, message = F}
+To determine whether we can propose any new genes that distinguish RPC and neurogenic states we use our DESeq2 differential expression contrast that directly compares these two cell types.
+We find `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Homo sapiens') %>% nrow()` human genes and `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Mus musculus') %>% nrow()` mouse genes that have an absolute log2 fold change greater than two and a adjusted p-value below 0.05 (`r supTab_cap(name = 'neuro_table', display = 'cite')`).
+To identify candidate genes that are differentially expressed at this transitional state between mouse and human we apply similar logic that we used in identifying well supported marker genes across the retina cell types, with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two.
+
+We leverage our more specific pairwise differential test to directly assess genes that are consisntently different between neurogenic cells and RPC in both mouse and human.
+
+```{r neuro, fig.width=15, fig.height=5, echo = F, message = F, fig.cap=neuro_cap}
+neuro_cap <- fig_cap(name='neuro', caption = 'A.
+Genes identified in Figure 2K of Lu et al.
+as being differently expressed during the RPC to neurogenic differentiation in human.
+We plot the differential expression (log2(fold change) directly between RPC and neurogenic cells.
+The standard error (calculated by DESeq2) is plotted.
+Genes with a padj < 0.1 and padj < 0.01 are labelled with "*" and "**", respectively.
+B.
+ZBTB38 .....')
 # tic()
@@
                 position = position_dodge2()) +
+  ylab('Neurogenic / RPC Gene Expression (log2 Fold Change)') +
   coord_flip() + 
@@
 
+```{r neuro_table, echo = F, message = F, fig.cap=neuro_table_cap}
+neuro_table_cap <- supTab_cap(name='neuro_table', caption = 'Differentially expressed genes between RPC and neurogenic cells in human and mouse.
+Genes filtered with a abs(log2FoldChange) > 2 and a padj < 0.01.')
+
+pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05) %>% flextable() %>% autofit() %>% fit_to_width(7)
+```
+
 ```{r decontX, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=decontX_cap}

### 2022-07-21 src/diff_data_pull.R

@@ -197,13 +197,25 @@ toc()
 ###############
 
 tic()
+
+pairwise_neurogenic_test_ungrouped <-  scEiaD_2020_v01 %>% 
+  tbl("diff_testing") %>% 
+  filter(Against  == "RPC",   
+         Group == 'CellType (Predict)', 
+         Base == "Neurogenic Cell") %>% 
+  #group_by(Base, Against, Gene) %>% 
+  #summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
+  #arrange(-`mean log2FC`) %>%   
+  collect() 
+
+
 pairwise_neurogenic_test<-  scEiaD_2020_v01 %>% 
   tbl("diff_testing") %>% 
   filter(Against  == "RPC",   
          Group == 'CellType (Predict)', 
          Base == "Neurogenic Cell") %>% 
   group_by(Base, Against, Gene) %>% 
-  summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
+  summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 0.05), `na count` = sum(is.na(log2FoldChange)), baseMean = mean(baseMean), lfcSE = mean(lfcSE)) %>% 
   arrange(-`mean log2FC`) %>%   
   collect() %>% 
   filter(!is.na(`mean log2FC`))

### 2022-07-21 manuscript_00.Rmd
neuro sectoin

@@
 We find `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Homo sapiens') %>% nrow()` human genes and `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Mus musculus') %>% nrow()` mouse genes that have an absolute log2 fold change greater than two and a adjusted p-value below 0.05 (`r supTab_cap(name = 'neuro_table', display = 'cite')`).
-To identify candidate genes that are differentially expressed at this transitional state between mouse and human we apply similar logic that we used in identifying well supported marker genes across the retina cell types, with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two.
+To identify candidate genes that are differentially expressed at this transitional state between mouse and human we apply similar logic that we used in identifying well supported marker genes across the retina cell types, with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two or less than one.
+We find one gene that
 
@@
 # 
-down <- pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` < 0, `mean padj` < 0.05) %>% head(5) %>% pull(Gene)
-up <-  pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` > 0) %>% head(9) %>% pull(Gene)
+down <- pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` < 0, `mean padj` < 0.05) %>% mutate(d = `mean log2FC` - lfcSE) %>% arrange(-d) %>% pull(Gene)
+up <-  pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` > 2, `mean padj` < 0.05) %>% mutate(d = `mean log2FC` - lfcSE) %>% arrange(-d) %>% pull(Gene)
 source('src/make_exp_plot.R')

### 2022-07-21 manuscript_00.Rmd
beginning last section of resultsj

--- 2022-07-21 manuscript_00.Rmd
@@
 
-sup_tab_consist_diff <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n()) %>% ggplot(aes(x=`Base`, y = `Count` )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip() + xlab("CellType (Predict)")
+sup_tab_consist_diff <- top_genes %>% select(Symbol, Base) %>% left_join(consist_diff, by = c('Symbol','Base')) %>% group_by(Base) %>% summarise(Count = n()) %>% ggplot(aes(x=`Base`, y = `Count` )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip() + xlab("CellType (Predict)")
 #toc()
@@
 
-We see that the neurogenic cells and RPCs have very similar expression patterns.
-These cell types are closely related as the neurogenic cells are derived from the multi-potent RPC to later specify neural cell types.
-First we use the set of genes identified by Lu et al.
-[@lu_single-cell_2020] in their study of human retinal differentiation to determine whether our broader scEiaD retains the same trends (`r fig_cap(name = 'neuro', display = 'cite')`A).
-As expected, we see several confirming patterns.
-First, we see most of the gene expression changes between RPC and neurogenic are matched between mouse and human.
-Next we see that the direction of the changes largely matches Lu et al (`r supFig_cap(name='lu_et_al', display = 'cite'`).
-Third, we see a small number of genes are confirmed as consistently differentially expressed with a padj < 0.1.
-
-To determine whether we can propose any new genes that distinguish RPC and neurogenic states we use our DESeq2 differential expression contrast that directly compares these two cell types.
-We find `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Homo sapiens') %>% nrow()` human genes and `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Mus musculus') %>% nrow()` mouse genes that have an absolute log2 fold change greater than two and a adjusted p-value below 0.05 (`r supTab_cap(name = 'neuro_table', display = 'cite')`).
-To identify candidate genes that are differentially expressed at this transitional state between mouse and human we apply similar logic that we used in identifying well supported marker genes across the retina cell types, with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two.
-
-We leverage our more specific pairwise differential test to directly assess genes that are consisntently different between neurogenic cells and RPC in both mouse and human.
-
-```{r neuro, fig.width=15, fig.height=5, echo = F, message = F, fig.cap=neuro_cap}
+```{r neuro, fig.width=25, fig.height=8, echo = F, message = F, fig.cap=neuro_cap}
 neuro_cap <- fig_cap(name='neuro', caption = 'A.
@@
 #
-down <- pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` < 0, `mean padj` < 0.05) %>% head(5) %>% pull(Gene)
-up <-  pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` > 0) %>% head(9) %>% pull(Gene)
+
+lu_et_al <- c("PTTG1", "SCG3", "SHD", "SOX11", "ONECUT2", "VSX1", "ASCL1", "TCF4", "SOX4", "RORB", "GADD45G", "MYBL1", "MFAP4", "HES6", "NEUROD1", "C8orf46", "OTX2", "OLIG2", "INSM1", "PCBP4", "NEUROD4", "FAM131C", "HES4", "HES5", "HMGA1", "YBX1", "GADD45A", "ATOH7", "DLX2", "DLX1", "SFRP1", "MYB", "NFIB", "SOX9", "ZFP36L1", "FOS", "JUN", "ZFP36L2", "SOX2", "HES1", "EGR1", "JUNB", "NR4A1", "ETV5", "FOXP1", "ID1", "ID2", "ID3", "ID4", "FOXP2")
+
+g_lu <- scEiaD_2020_v01 %>% tbl('genes') %>% collect() %>% filter(grepl(paste(paste0('^',lu_et_al), collapse = ' |'), Gene)) %>% pull(Gene)
+
+down <- pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` < 0, `mean padj` < 0.05) %>% mutate(d = `mean log2FC` - lfcSE) %>% arrange(-d) %>% pull(Gene)
+up <-  pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` > 2, `mean padj` < 0.05) %>% mutate(d = `mean log2FC` - lfcSE) %>% arrange(-d) %>% pull(Gene)
+
 source('src/make_exp_plot.R')
@@
 input$exp_plot_facet <- c('CellType_predict', 'organism')
-input$exp_plot_genes <- down
+input$exp_plot_genes <- c(down, up[!up %in% g_lu])
 input$exp_filter_cat <- 'CellType_predict'
@@
 
-exp_plot_down <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+exp_plot_new <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
 
-input$exp_plot_genes <- up
-exp_plot_up <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-a_legend <- get_legend(exp_plot_down)
-neuro_plot <- plot_grid(plot_grid(exp_plot_down + theme(legend.position="none") + xlab(''),
-          exp_plot_up + theme(legend.position="none") + ylab('') + xlab(''), labels = c('b','c'),
-
-          rel_widths = c(0.1,1)),
+input$exp_plot_genes <- up[up %in% g_lu]
+exp_plot_existing <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+a_legend <- get_legend(exp_plot_new)
+neuro_plot <- plot_grid(plot_grid(exp_plot_existing + theme(legend.position="none") + xlab(''),
+          exp_plot_new + theme(legend.position="none") + ylab('') + xlab(''), 
+          labels = c('b','c'),
+          rel_widths = c(0.6,1)),
           a_legend, ncol = 1, rel_heights = c(1,0.2))
-
-lu_et_al <- c("PTTG1", "SCG3", "SHD", "SOX11", "ONECUT2", "VSX1", "ASCL1", "TCF4", "SOX4", "RORB", "GADD45G", "MYBL1", "MFAP4", "HES6", "NEUROD1", "C8orf46", "OTX2", "OLIG2", "INSM1", "PCBP4", "NEUROD4", "FAM131C", "HES4", "HES5", "HMGA1", "YBX1", "GADD45A", "ATOH7", "DLX2", "DLX1", "SFRP1", "MYB", "NFIB", "SOX9", "ZFP36L1", "FOS", "JUN", "ZFP36L2", "SOX2", "HES1", "EGR1", "JUNB", "NR4A1", "ETV5", "FOXP1", "ID1", "ID2", "ID3", "ID4", "FOXP2")
-
-g_lu <- scEiaD_2020_v01 %>% tbl('genes') %>% collect() %>% filter(grepl(paste(paste0('^',lu_et_al), collapse = ' |'), Gene)) %>% pull(Gene)
 
@@
 
-cowplot::plot_grid(lu_rep_plot, neuro_plot, rel_widths = c(0.7,1), labels = c('a'))
+cowplot::plot_grid(lu_rep_plot, neuro_plot, rel_widths = c(0.6,1), labels = c('a'))
 ```
+
+We see that the neurogenic cells and RPCs have very similar expression patterns.
+These cell types are closely related as the neurogenic cells are derived from the multi-potent RPC to later specify neural cell types.
+First we use the set of genes identified by Lu et al.
+[@lu_single-cell_2020] in their study of human retinal differentiation to determine whether our broader scEiaD retains the same trends (`r fig_cap(name = 'neuro', display = 'cite')`A).
+As expected, we see several confirming patterns.
+First, we see most of the gene expression changes between RPC and neurogenic are matched between mouse and human.
+Next we see that the direction of the changes largely matches Lu et al (`r supFig_cap(name='lu_et_al', display = 'cite')`).
+Third, we see a small number of genes are confirmed as consistently differentially expressed with a padj < 0.1.
+
+To determine whether we can propose any new genes that distinguish RPC and neurogenic states we use our DESeq2 differential expression contrast that directly compares these two cell types.
+We find `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Homo sapiens') %>% nrow()` human genes and `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Mus musculus') %>% nrow()` mouse genes that have an absolute log2 fold change greater than two and a adjusted p-value below 0.05 (`r supTab_cap(name = 'neuro_table', display = 'cite')`).
+To identify candidate genes that are consistently differentially expressed at this transitional state in mouse and human we apply similar logic that we used in identifying well supported marker genes across the retina cell types.
+We filter to genes with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two or less than negative one (`r fig_cap(name = 'neuro', display = 'cite')`B and C).
+We find one gene, ZBTB38, that meets this criteria and drops in expression from RPC to neurogenic.
+We find `r up %>% length()` genes that increase in expression from RPC to neurogenic, `r up[!up %in% g_lu] %>% length()` of which are not previously identified by Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`C).
+We replicate `r up[up %in% g_lu] %>% str_extract(., '\\w+') %>% paste(., collapse = ', ')` from Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`B).
 
@@
 
-```{r}
-library(enrichR)
-dbs <- c('GO_Biological_Process_2018', 'GO_Molecular_Function_2018')
-enriched <- enrichr(org_diff_diff_genes %>% filter(Base == 'Bipolar Cell') %>% pull(Symbol), dbs)
+## Pan-study fovea vs peripheral cone test identifies 16 genes consistently differentially expressed
 
-```
-
-To aid in in the analysis of ocular gene expression patterns we provide pre-computed differential tests run
-
-## Pan-study fovea vs peripheral photoreceptor identifies N genes consistently differentially expressed
-
-To identify
+Yan et al.
+proposed a PRPH2 and RS1 as markers to distinguish fovea and peripheral located cones [@yan_cell_2020].
+To see whether we could replicate these findings and propose new markers we perfomed a custom differntial test in human cones that were derived from studies that took retina punches from the fovea or peripheral regions.
+As this custom test was not
 
@@
 
-pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05) %>% flextable() %>% autofit() %>% fit_to_width(7)
+pairwise_neurogenic_test_ungrouped %>% filter(padj < 0.05) %>% filter(log2FoldChange > 2 | log2FoldChange < 0) %>% flextable() %>% autofit() %>% fit_to_width(7)
 ```
@@
 input$flip_facet <- FALSE
+input$figure <- FALSE
 

### 2022-07-21 src/make_exp_plot.R

@@ -42,7 +42,8 @@ make_exp_plot <- function(input, db, meta_filter){
         filter(!!as.symbol(input$exp_filter_cat) >= input$exp_filter_on[1],
                !!as.symbol(input$exp_filter_cat) <= input$exp_filter_on[2])
     }
-    
+    box_data <- box_data %>% 
+      filter(!is.na(!!as.symbol(input$exp_filter_cat))) 
   } else {
     box_data <- db %>% tbl('grouped_stats') %>%
       filter(Gene %in% gene) %>%
@@ -77,7 +78,7 @@ make_exp_plot <- function(input, db, meta_filter){
            `Total Cells` = Count,
            `Mean log2(Counts + 1)` = Expression,
            `% of Cells Detected` = `%`) %>%
-    filter(!is.na(!!as.symbol(input$exp_filter_cat))) %>% 
+    
     mutate(`Mean log2(Counts + 1)` = case_when(is.na(`Mean log2(Counts + 1)`) ~ 0, TRUE ~ `Mean log2(Counts + 1)`)) %>% 
     filter(`Total Cells` > as.integer(input$exp_filter_min_cell_number))
   # expand missing genes (nothing detected) to all genes used

### 2022-07-21 manuscript_00.Rmd
text update

@@
 
+While we have provided data in many forms via the web app at plae.nei.nih.gov, there are many more dissections of the data that can be performed.
+To facilitate custom testing of scEiaD we provide as an example a human fovea versus peripheral cone gene differential testing document as supplementary file "pseudobulk_cone_region.freeze01.html".
 Yan et al.
-proposed a PRPH2 and RS1 as markers to distinguish fovea and peripheral located cones [@yan_cell_2020].
-To see whether we could replicate these findings and propose new markers we perfomed a custom differntial test in human cones that were derived from studies that took retina punches from the fovea or peripheral regions.
+proposed PRPH2 and RS1 as markers to distinguish fovea and peripheral located cones [@yan_cell_2020].
+In our analysis document we use the Seurat object containing the full scEiaD dataset we provide at plae.nei.nih.gov -> Data to build a custom pseudobulk counts matrix containing only cones from human studies that were taken in a region specific manner.
+This left us with
+
+To see whether we could replicate these findings and whether we can propose new markers we performed a custom differential test in human cones that were derived from studies that took retina punches from the fovea or peripheral regions.
 As this custom test was not

### 2022-07-26 manuscript_00.Rmd
update text

--- 2022-07-21 manuscript_00.Rmd
@@
 abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors.
-METHODS: We updated the content and structure of the underlying data (sc Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it.
-RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes, XX cell types, XX ocular tissues, XX body tissues.
-As a test of the value of this unified pan-eye dataset, we show XX.
+METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it.
+RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encomopassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 37 publications.
+As a test of the value of this unified pan-eye dataset, we propose consistent cell type markers, replicate neurogenic and cone fovea markers, and propose six new cone fovea and peripheral markers .
 CONCLUSION: The PLAE v0.91 web app serves the pan-ocular and body dataset, scEiaD.
-This offers the eye community a powerful and quick means to test hypotheses on gene and transcript expression across 54 body and 19 eye tissues.
-"
+This provides the eye community a powerful and quick means to test hypotheses on gene expression across a highly diverse community derived database."
 
@@
 
-  Single cell transcriptomics of the retina date to the early 2000s when researchers [@hagstrom_cone_2000; @jabs_evidence_2000; @li_localization_2004; @lindqvist_single_2002-1; @reid_retinoschisin_2003] used RT-PCR approaches to amplify select transcripts in small numbers of cells.
-Later, researchers used microarray platforms to analyze a broader array of transcripts [@cherry_development_2009; @gustincich_gene_2004-1; @kim_identification_2008-1; @roesch_transcriptome_2008; @trimarchi_individual_2008; @trimarchi_molecular_2007-1; @wahlin_method_2004].
+Single cell transcriptomics of the retina date to the early 2000s when researchers [@hagstromConePigmentGene2000b; @jabsEvidenceP2XP2X2000; @liLocalizationDopamineD1receptor2004; @lindqvistSingleCellRTPCR2002a; @reidRetinoschisinPhotoreceptorsecretedProtein2003]  used RT-PCR approaches to amplify select transcripts in small numbers of cells.
+Later, researchers used microarray platforms to analyze a broader array of transcripts[@cherryDevelopmentDiversificationRetinal2009; @gustincichGeneDiscoveryGenetically2004a; @kimIdentificationMolecularMarkers2008a; @roeschTranscriptomeRetinalMuller2008; @trimarchiIndividualRetinalProgenitor2008; @trimarchiMolecularHeterogeneityDeveloping2007b; @wahlinMethodAnalysisGene2004a]
+
 The modern era of single cell transcriptomics is marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment; in Macoksko et al.
-the adult mouse retina was profiled [@klein_droplet_2015; @macosko_highly_2015].
+the adult mouse retina was profiled[@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a].
 The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since Macosko publication.
+
 These publications's foci includes, but is not limited to, studying developmental transcriptomic dynamics, identifying genes to distinguish core cell types, proposing finer gradations of cell types within existing categories, and comparing ocular transcriptomes between organisms.
@@
 Briefly, one would have to identify the relevant publication, find the associated data deposit, download the cell - count matrix, find the cell label table (and optionally the t-SNE or UMAP coordinates), and then load the data into R or python to run the desired queries.
-If one desired to compare *across* publications, then one would also have to normalize the cell type labels and would likely have to re-quantify the counts with consistent bioinformatic tooling and run a batch-effect removal algorithm.
-It is desirable, perhaps even necessary, to use multiple datasets to assess gene expression to confirm whether a gene expression is consistent and reproducible.
-Reliance on a single dataset can be problematic as there can be technical issues - for example the large consortium dataset GTEx has persistent contamination from highly expressed genes across tissues [@nieuwenhuis_consistent_2020].
+If one desired to compare across publications, then one would also have to align the cell type labels and should re-create the count quantification with consistent bioinformatic tooling.
+It is desirable, perhaps even necessary, to use multiple datasets to assess gene expression to confirm whether a gene expression pattern is consistent and reproducible.
+Reliance on a single dataset can be problematic as there can be technical issues - for example the large consortium dataset GTEx has persistent contamination from highly expressed genes across tissues [@nieuwenhuisConsistentRNASequencing2020].
+
 Web-based resources to query gene expression are powerful means to democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device.
-Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Muris, PanglaoDB, and the Protein Atlas [@franzen_panglaodb_2019; @karlsson_singlecell_2021; @noauthor_single-cell_nodate; @speir_ucsc_2021].
+Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas[@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021].
 Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging.
-In contrast to these, the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigt_spectacle_2020].
+In contrast to these, the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020].
 However each data deposit is independently processed, which makes cross study queries difficult.
-We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space [@swamy_building_2021].
+We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space [@swamyBuildingMegaSinglecell2021].
+
 To broaden the usage of this powerful database, we restructured the data into a custom sqlite-based database and built a reactive web application around it.
 As many new studies were published since our last data build, we re-ran the scEiaD pipeline in March of 2022.
-In aggregate we have collated `r meta_filter$study_accession %>% unique() %>% length()` studies across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues.
-We hand-curated `r meta_filter$CellType %>% unique() %>% length()` cell types, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the remaining cells.
-Extensive quality control metrics were tuned and applied to remove lower quality cells.
+In aggregate we have collated `r meta_filter$study_accession %>% unique() %>% length()` datasets across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues.
+We hand-curated `r meta_filter$CellType %>% unique() %>% length()` cell type labels, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the remaining cells.
+Quality control metrics were tuned and applied to remove lower quality cells.
 In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values.
-We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, genes which have species-specific expression cell type patterning, and finally propose genes XYZ that differentially expressed between the fovea and peripheral cones.
+We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, genes which have species-specific expression cell type patterning, demonstrate how our pre-computed differential testing can identify markers between similar progenitor cell types, and finally provide a R-based analysis document which end-to-end demonstrates how we use scEiaD to run a custom differential gene expression test to validate and propose new genes that distinguish fovea from peripheral cones in humans.
 
@@
 
-The scEiaD pipeline is structurally the same as in Swamy et al.
-[@swamy_building_2021].
-Very briefly, we collected the raw fastq files and quantified with the kallisto / bustools system [@melsted_barcode_2019; @melsted_modular_2019].
+The scEiaD pipeline is largely the same as in Swamy et al.
+[@swamyBuildingMegaSinglecell2021].
+Very briefly, we collected the raw fastq files and quantified with the kallisto / bustools system [@melstedBarcodeUMISet2019; @melstedModularEfficientPreprocessing2019].
 The quantified counts are merged into one matrix that is used for downstream processing with scvitools create a batch corrected lower dimensional space.
 and then into Seurat for quality control, clustering, and 2D UMAP visualization.
-We use our previous xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamy_building_2021].
+We use our previous xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamyBuildingMegaSinglecell2021].
 
-The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300 Second, we use the DecontX algorithm [@yang_decontamination_2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination.
-Third, we altered the procedure for automatically aligning gene names across species (see below for details).
-Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process.
+The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300.
+Second, we use the DecontX algorithm[@yangDecontaminationAmbientRNA2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination.
+Third, we altered the procedure for aligning gene names across species (see below for details).
+Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process [@xuProbabilisticHarmonizationAnnotation2021].
 
-Two custom cutoffs for the minimum required number of unique genes identified per cell were used for the E12 chicken data and the cornea dataset (SRP362101) as an abnormally high number of cells were returned from our default value of 250 (????).
-We set the cutoffs for these studies to be 1000 and 800, respectively, as these values resulted in approximately the same number of cells being returned as the authors reported in their papers.
+A custom cutoffs for the minimum required number of unique genes identified per cell were used for the SRP362101 cornea dataset as an abnormally high number of cells were returned from our default value of 200.
+We set the cutoff for this studies to be 800 as this value resulted in approximately the same number of cells being returned as the authors reported in their papers.
 
@@
 
-The scater::aggregateAcrossCells function was used to aggregate gene counts across species, study accession, and [category] where category was either Cluster, CellType, or CellType (Predict) assignments for a cell.
+The scater::aggregateAcrossCells function was used to aggregate gene counts across species, study accession, and [category] where category was either Cluster, CellType, or CellType (Predict) assignments for a cell [@mccarthyScaterPreprocessingQuality2017].
 If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test.
@@
 Contrasts were extracted either with the target category (e.g.
-"Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function.
-Genes were considered to be significantly differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2.
+"Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function [@loveModeratedEstimationFold2014a].
+<!-- Genes were considered to be significantly differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2.  -->
 
@@
 This speed is only possible due to custom data structures to optimize for data query efficiency.
-Like eyeIntegration's EiaD database, scEiaD is a sqlite database with tables for the core categories.
+Like eyeIntegration's EiaD database [@swamyEyeDiskEyeIntegration2019a], scEiaD is a sqlite database with tables for the core categories.
 This allows the app to initialize in about 30 seconds on a cloud server with memory usage under 8 GB.
-For the gene by cell expression matrix, the data was transformed into a "long" format where there are three columns: Gene, Cell Barcode, and Gene Count Value.
+For the gene by cell expression matrix, the data was transformed into a "long" format where there are three columns: gene, cell barcode, and gene count value.
 This allows for cell - gene level queries to be completed, again with minimal memory usage, in under 0.5 seconds.
@@
 
-Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamy_building_2021].
-We updated the scEiaD resource for the PLAE v0.91 web app in six important ways.
-First, we added a large chicken retina (*Gallus gallus*) studies.
-Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye.
-Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons.
+Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021].
+We updated the scEiaD database for the PLAE v0.91 web app in six important ways.
+First, we added a new species, chicken (*Gallus gallus*) [@yamagataCellAtlasChick2021].
+Second we added two ocular outflow tract datasets, one brain choroid plexus, and three cornea datasets to enhance coverage across the eye [@collinSingleCellAtlas2021; @gautamMultispeciesSinglecellTranscriptomic2021; @daniCellularSpatialMap2021; @ligockiMolecularCharacteristicsSpatial2021; @patelMolecularTaxonomyHuman2020; @vanzylCellAtlasAqueous2020].
+Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons [@heSinglecellTranscriptomeProfiling2020].
 Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`).
-Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300 (CHECK THIS), to improve the database quality.
-Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons.
+Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300.
+Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons [@SinglecellTranscriptomics20].
 
@@
 This resource is only possible due to the publication of `r meta_filter$PMID %>% unique %>% length()` single cell resources that scEiaD draws from (`r supTab_cap(name = 'studies_table', display = 'cite')`).
-To recognize these papers, we make the first data table on the loading page of plae.nei.nih.gov list each of these studies along with a direct link to their citation, where possible.
+To recognize these papers, we prominently feature each of these studies along with a direct link to their citation, where possible, on the loading page of plae.nei.nih.gov.
 
@@
 The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has selected a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app.
-PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence, and, most crucially, integrating the independent studies to deliver cross study comparisons.
+PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence, and, crucially, full availability of the underlying data for outside usage and integrating the independent studies to deliver cross study comparisons.
 
@@
 
-While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently [@voigt_spectacle_2020].
-PLAE is built around the scEiaD database, which runs an multi-stage pipeline to build and identify a high performing scVI-based eye model that is used to integrate all datasets together [@swamy_building_2021].
+While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently[@voigtSpectacleInteractiveResource2020].
+PLAE is built around the scEiaD database, which use a multi-stage pipeline to build and identify a high performing scVI-based single cell RNA eye model that is used to integrate all datasets together [@swamyBuildingMegaSinglecell2021].
 This allows for gene queries to be constructed and analyzed across studies.
@@
 Furthermore, Spectacle contains no non-ocular dataset and the cell types labels have not been curated or harmonized between studies.
-Finally, as this Spectacle uses Seurat objects, the app takes many seconds to load an individual dataset and many more to display the data.
+Finally, the Spectacle web app takes many seconds to load an individual dataset and many more to display the data.
 
@@
 
-There are two interwoven reasons why PLAE is user responsive, despite the huge amount of data it contains.
+There are two interwoven reasons why the PLAE app is quick, despite the huge amount of data it contains.
 First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables.
 A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage.
-The major downside is that this approach requires a large amount of disk storage, which makes local (on computer) usage onerous; the scEiaD v2022-03-22 sqlite database is about 368 GB.
+The major downside is that this approach requires a large amount of disk storage, which makes local (on personal computer) usage onerous; the scEiaD v2022-03-22 sqlite database is about 368 GB.
 Second, the visualization code in PLAE (e.g.
@@
 
-For users interested in seeing how expression of a gene relates to study or other covariates, the "Exp Plot" view displays expression by cell type, predicted cell type, or cluster.
+For users interested in seeing how expression of a gene relates to study or other covariates, the "Expression Plot" view displays expression by cell type, predicted cell type, or cluster.
 The values can then be split and colored by multiple covariates like study or organism.
@@
 This created a counts matrix which has statistical properties approximating "bulk" RNA seq experiment matrices.
-This allowed use to use the more mature bulk RNA-seq tooling.
-This approach has been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021].
+This allowed the use of the more mature bulk RNA-seq tooling.
+This approach has also been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments[@squairConfrontingFalseDiscoveries2021].
 Once the pseudo-bulk matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
@@
 These variable results directly reflect the differing number of independent studies - mouse and human have substantially more than macaque (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
-The full differential table is made available as supplementary file "differential_expression_table.csv.gz." While this differential expression table is large, it can be further processed to give more precise results.
+The full differential table is made available as supplementary file "differential_expression_table.csv.gz."
 
@@
 To visualize these top markers we select up to eight genes (ordered by the mean log2 fold change across organisms) per well supported cell type for each and plot the log2 fold change of the gene expression relative to all other cell types (`r fig_cap(name = 'consist_genes_heatmap', display = 'cite')`).
-We see that human, mouse, and macaque have similar expression pattterning of the proposed markers across the well supported cell types.
+We see that human, mouse, and macaque have similar expression patterning of the proposed markers across the well supported cell types.
+The cell types are arranged by the columns by how related the expression patterns are.
+We see how the differentiated cell types of the retina are grouped together with the exception of the Müller glia, which are next to the RPC from which they derive from.
+The photoreceptors are also next to each other
 
-```{r}
+```{r echo = F, message = F, }
 # tic()
@@
 
-```{r consist_genes_heatmap, fig.height=20, fig.width=10, fig.cap=consist_genes_heatmap_cap}
+```{r consist_genes_heatmap, fig.height=20, fig.width=10, echo = F, message = F,  fig.cap=consist_genes_heatmap_cap}
 consist_genes_heatmap_cap <- fig_cap(name='consist_genes_heatmap', caption = 'Rows are genes, columns are cell types and organisms.
@@
 
-## Tease apart the closely neurogenic and RPC cell types
+## Tease apart the similar neurogenic and RPC cell types
 
-```{r neuro, fig.width=25, fig.height=8, echo = F, message = F, fig.cap=neuro_cap}
+```{r neuro, fig.width=30, fig.height=8, echo = F, message = F, fig.cap=neuro_cap}
 neuro_cap <- fig_cap(name='neuro', caption = 'A.
@@
   coord_flip() + 
-  cowplot::theme_cowplot()  + geom_text(aes(y=7), position = position_dodge2()) + facet_wrap(~Organism)
+  cowplot::theme_cowplot()  + geom_text(aes(y=7), position = position_dodge2(width = 1)) + facet_wrap(~Organism)
 #g_lu[!g_lu %in% lu_long$Gene]
@@
 
-We see that the neurogenic cells and RPCs have very similar expression patterns.
+While the well supported markers we propose clearly define most of the cell types, we see that the neurogenic cells and RPCs have very similar expression patterns across the gene sets.
 These cell types are closely related as the neurogenic cells are derived from the multi-potent RPC to later specify neural cell types.
-First we use the set of genes identified by Lu et al.
-[@lu_single-cell_2020] in their study of human retinal differentiation to determine whether our broader scEiaD retains the same trends (`r fig_cap(name = 'neuro', display = 'cite')`A).
-As expected, we see several confirming patterns.
-First, we see most of the gene expression changes between RPC and neurogenic are matched between mouse and human.
+We use the set of genes identified by Lu et al.
+[@luSingleCellAnalysisHuman2020b] in their study of human retinal differentiation to determine whether our scEiaD database retains the same trends (`r fig_cap(name = 'neuro', display = 'cite')`A).
+As expected, we see several patterns.
+First, we see most of the gene expression changes between RPC and neurogenic are matched between human.
+We further show that these genes also show the same trends in mouse datasets.
 Next we see that the direction of the changes largely matches Lu et al (`r supFig_cap(name='lu_et_al', display = 'cite')`).
-Third, we see a small number of genes are confirmed as consistently differentially expressed with a padj < 0.1.
+Finally, we see a small number of genes are confirmed in our database as consistently differentially expressed with a padj < 0.1.
 
@@
 We filter to genes with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two or less than negative one (`r fig_cap(name = 'neuro', display = 'cite')`B and C).
-We find one gene, ZBTB38, that meets this criteria and drops in expression from RPC to neurogenic.
+We find one gene, ZBTB38, that meets this criteria and drops in expression when comparing RPC to neurogenic.
 We find `r up %>% length()` genes that increase in expression from RPC to neurogenic, `r up[!up %in% g_lu] %>% length()` of which are not previously identified by Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`C).
@@
 
-```{r, fig.width=4, fig.height=12, echo = F, message = F}
-# tic()
-# # flip_flop diff genes
-# down_tab <- down_tab_big %>% 
-#   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
-#   filter(Base %in% ws_oct, 
-#          Against == 'All')
+# ```{r, fig.width=8, fig.height=12, echo = F, message = F}
+# # tic()
+# # # flip_flop diff genes
+# # down_tab <- down_tab_big %>% 
+# #   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
+# #   filter(Base %in% ws_oct, 
+# #          Against == 'All')
+# # 
+# # org_diff_diff_genes <- diff_tab_big %>% 
+# #   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
+# #   filter(gene_ct_combo %in% down_tab$gene_ct_combo, 
+# #          Against == 'All') %>% 
+# #   arrange(padj) %>% mutate(x = paste0(Symbol, ' (', ID, ')')) 
+# # 
+# # flip_flop_genes <- scEiaD_2020_v01 %>% tbl('diff_testing_genes') %>% collect() %>% filter(grepl(org_diff_diff_genes %>% pull(ID) %>% paste0(., collapse = '|'), Gene)) %>% pull(Gene)
+# # flip_flop_long <- scEiaD_2020_v01 %>%
+# #   tbl("diff_testing") %>%
+# #   filter(Gene %in% genes) %>%
+# #   collect() %>%
+# #   mutate(Symbol = str_extract(Gene, '^\\w+ ') %>% gsub(' $', '', .)) %>% 
+# #   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
+# #   filter(gene_ct_combo %in% org_diff_diff_genes$gene_ct_combo,
+# #          Group == 'CellType (Predict)',
+# #          Against == 'All') %>%
+# #   select(Gene, Organism, log2FoldChange, Base)
+# # 
+# # flip_flop_wide <- flip_flop_long %>% pivot_wider(values_from = log2FoldChange, names_from = Organism)
+# # flip_flop_organism <- colnames(flip_flop_wide)[c(3,4,5)]
+# # flip_flop_wide <- flip_flop_wide %>% data.frame()
+# # row.names(flip_flop_wide) <- paste0(flip_flop_wide$Gene,'_', flip_flop_wide$Base)
+# # 
+# # flip_flop_wide <- flip_flop_wide[,-c(1,2)]
+# # 
+# # 
+# # 
+# # flip_flop_genes <- row.names(flip_flop_wide) %>% gsub(' \\(.*','',.)
+# # 
+# # flip_flop_base <- row.names(flip_flop_wide) %>% gsub('.*_','',.) 
 # 
-# org_diff_diff_genes <- diff_tab_big %>% 
-#   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
-#   filter(gene_ct_combo %in% down_tab$gene_ct_combo, 
-#          Against == 'All') %>% 
-#   arrange(padj) %>% mutate(x = paste0(Symbol, ' (', ID, ')')) 
-# 
-# flip_flop_genes <- scEiaD_2020_v01 %>% tbl('diff_testing_genes') %>% collect() %>% filter(grepl(org_diff_diff_genes %>% pull(ID) %>% paste0(., collapse = '|'), Gene)) %>% pull(Gene)
-# flip_flop_long <- scEiaD_2020_v01 %>%
-#   tbl("diff_testing") %>%
-#   filter(Gene %in% genes) %>%
-#   collect() %>%
-#   mutate(Symbol = str_extract(Gene, '^\\w+ ') %>% gsub(' $', '', .)) %>% 
-#   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
-#   filter(gene_ct_combo %in% org_diff_diff_genes$gene_ct_combo,
-#          Group == 'CellType (Predict)',
-#          Against == 'All') %>%
-#   select(Gene, Organism, log2FoldChange, Base)
-# 
-# flip_flop_wide <- flip_flop_long %>% pivot_wider(values_from = log2FoldChange, names_from = Organism)
-# flip_flop_organism <- colnames(flip_flop_wide)[c(3,4,5)]
-# flip_flop_wide <- flip_flop_wide %>% data.frame()
-# row.names(flip_flop_wide) <- paste0(flip_flop_wide$Gene,'_', flip_flop_wide$Base)
-# 
-# flip_flop_wide <- flip_flop_wide[,-c(1,2)]
-# 
-# 
-# 
-# flip_flop_genes <- row.names(flip_flop_wide) %>% gsub(' \\(.*','',.)
-# 
-# flip_flop_base <- row.names(flip_flop_wide) %>% gsub('.*_','',.)
+# gene_anno <- ComplexHeatmap::rowAnnotation(Duplicated = flip_flop_genes %in% (flip_flop_genes[duplicated(flip_flop_genes)]),
+#                                            col = list(Duplicated = c("TRUE" = "green", "FALSE" = "gray")))
+# hm_flipflop <- ComplexHeatmap::Heatmap(flip_flop_wide, column_labels = flip_flop_organism, 
+#                                        row_split = flip_flop_base, show_row_dend = FALSE, 
+#                                        row_labels = flip_flop_genes, 
+#                                        column_title_rot = 90, 
+#                                        right_annotation = gene_anno,
+#                                        name = 'log2FC', use_raster = TRUE,
+#                                        row_title_rot = 0)
+# hm_flipflop
+# #toc()
+# ```
 
-gene_anno <- ComplexHeatmap::rowAnnotation(Duplicated = flip_flop_genes %in% (flip_flop_genes[duplicated(flip_flop_genes)]),
-                                           col = list(Duplicated = c("TRUE" = "green", "FALSE" = "gray")))
-hm_flipflop <- ComplexHeatmap::Heatmap(flip_flop_wide, column_labels = flip_flop_organism, 
-                                       row_split = flip_flop_base, show_row_dend = FALSE, 
-                                       row_labels = flip_flop_genes, 
-                                       column_title_rot = 90, 
-                                       right_annotation = gene_anno,
-                                       name = 'log2FC', use_raster = TRUE,
-                                       row_title_rot = 0)
-hm_flipflop
-#toc()
+## Our pan-study macula vs peripheral cone test replicates a published macaque finding and proposes six new markers
+
+```{r region_fig, fig.width=14, fig.height=3,  echo = F, message = F, fig.cap = region_fig_cap}
+region_fig_cap <- fig_cap(name='regino_fig', caption = 'Plot of markers distinguishing human macula or peripheral punch derived cones.')
+
+source('src/make_exp_plot.R')
+#load('data/pseudoBulk_cone_region_files.Rdata')
+region_table <- read_csv('supplemental_files/pseudoBulk_cone_region_table.csv.gz')
+input <- list()
+input$exp_plot_groups <- c('study_accession')
+input$exp_plot_facet <- c('CellType_predict', 'organism', 'retina_region')
+input$exp_plot_genes <- c('PRPH2 (ENSG00000112619)')
+input$exp_filter_cat <- c('CellType_predict', 'organism')
+#input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
+input$exp_filter_on <- c("Cone", "Homo sapiens" )
+input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
+#input$exp_plot_ylab <- '% of Cells Detected'
+#input$exp_plot_groups <- c('organism')
+input$exp_plot_col_num <- 16
+input$exp_filter_min_cell_number <- 10
+input$flip_facet <- FALSE
+input$figure <- FALSE
+input$figure_fovea <- TRUE
+
+prph2 <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') +  theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() ) 
+input$exp_plot_genes <-  'RS1 (ENSG00000102104)';rs1 <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') +  theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+
+input$exp_plot_genes <-  'GNGT1 (ENSG00000127928)';gngt1 <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') +  theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+#input$exp_plot_genes <-  'RS1 (ENSG00000102104)';rs1 <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+
+diff_region_cone_genes <- region_table %>% filter(padj < 0.01) %>% 
+  arrange(log2FoldChange) %>% 
+  mutate(id = paste0(Gene, ' (', Ensembl, ')')) %>% 
+  pull(id)
+input$exp_filter_on <- c("Cone", "Homo sapiens" )
+input$exp_plot_genes <- diff_region_cone_genes[1]; a <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') +  theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[2]; b <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[3]; c <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[4]; d <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + ylab('') + xlab('') + theme(axis.ticks.x=element_blank(), axis.text.x=element_blank() ) 
+input$exp_plot_genes <- diff_region_cone_genes[5]; e <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[6]; f <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[7]; g <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+
+region_legend <- get_legend(d)
+
+plot_grid(plot_grid(a,b,c,d + theme(legend.position="none"),e,f,g, nrow = 1),
+          region_legend, 
+          rel_heights = c(1,0.2), 
+          nrow = 2)
+
 ```
 
-## Pan-study fovea vs peripheral cone test identifies 16 genes consistently differentially expressed
+While we have provided data in many forms via the web app at plae.nei.nih.gov, there are many more dissections of the data that can be performed.
+To facilitate custom testing with the scEiaD database we provide, as an example, a human-specific macula versus peripheral cone differential testing analysis document as supplementary file "pseudobulk_cone_region.freeze01.html".
+In our analysis document we use the Seurat object containing the full scEiaD dataset we provide at plae.nei.nih.gov to build a custom pseudobulk counts matrix containing only cones from human studies that were taken in a region specific manner.
+This left us with 26678 cells across six studies.
+Yan et al.
+previously proposed PRPH2 and RS1 as markers to distinguish macula and peripheral located human cones[@yanCellAtlasHuman2020a] and Peng et al.
+proposed CALB1 and GNGT1 as cone region markers in macaque [@pengMolecularClassificationComparative2019a].
 
-Yan et al.
-proposed a PRPH2 and RS1 as markers to distinguish fovea and peripheral located cones [@yan_cell_2020].
-To see whether we could replicate these findings and propose new markers we perfomed a custom differntial test in human cones that were derived from studies that took retina punches from the fovea or peripheral regions.
-As this custom test was not
-
-Sanes proposed X genes (https://www.nature.com/articles/s41598-020-66092-9 --> Figure 9)
-
-PRPH2 and RS1 for cones
-NR2E3, SLC24A1, PDE6H, GNAT1 for rods
-Also did RGC, Microglia, Astrocytes, Muller Glia
-
-<!-- ## Projection of organoid data onto scEiaD demonstrates [SOMETHING] -->
-
-<!-- scEiaD is built using scvi-tool's scANVI batch correction tool.
-As part of the process a computational transcriptome model of the reference datasets are built.
-This model is only a few mb in size and can be used on a modest computer to project private scRNA datasets onto it.
-This allows an ocular researcher several important functionalities.
-First, the ability to check the quality control of an existing dataset by looking at the distribution of their cells overlapped with our labelled cell types.
-Second, in perturbation experiments, to identify unusual distributions of cells.
-Third, to leverage our extensive pre-labelled cell information to transfer cell type labels onto private data.
--->
-
-<!-- We demonstrate some of these possibilities by projecting two organoid datasets from X and Y.  -->
+Our differential testing found that while PRPH2, RS1, and GNGT1 had expression differences matching that the findings Yan et al.
+and Peng et al.
+published, they failed to replicate in our test with pvalue and padj being above 0.1 for all three.
+In contrast CALB1 did replicate strongly with the sixth lowest padj (0.00721) across our entire test and a log2 fold change of 2.87.
+We find six more genes with a padj < 0.01: PDE6G, AKAP12, HSPB1, GADD45B, CC2D2A, and RP1L1 (`r fig_cap(name = 'region_fig', display = 'cite')`).
+None of these candidate cone region marker genes display regionally different expression in rods (`r supFig_cap(name = 'rod_region_fig', display = 'cite')`).
 
@@
 
-## PLAE provides quick access to over one million transcriptomes across 35 studies
+## High feature web app provides quick access to over one million transcriptomes across `r meta_filter$study_accession %>% unique %>% length` datasets
 
-## Improvements in data processing ...
+## Wealth of data avaibility formats enables custom querying
 
-## The data underlying PLAE is available in several forms to enable custom queries
+## High diversity of studies enables reproduction and extension of cell type gene expression findings
 
@@
 
-```{r diff_tab_sup_tab, echo = F, message = F, fig.cap=diff_tab_sup_cap}
+```{r diff_tab_sup_tab, echo = F, message = F,  echo = F, message = F, fig.cap=diff_tab_sup_cap}
 diff_tab_sup_cap <- supTab_cap(name='diff_tab_sup', caption = 'Counts of signifantly differentially expressed genes for each cell type across the organisms')
@@
 
-```{r neuro_table, echo = F, message = F, fig.cap=neuro_table_cap}
+```{r neuro_table, echo = F, message = F,  echo = F, message = F, fig.cap=neuro_table_cap}
 neuro_table_cap <- supTab_cap(name='neuro_table', caption = 'Differentially expressed genes between RPC and neurogenic cells in human and mouse.
@@
 input$figure <- FALSE
+input$figure_fovea <- FALSE
 
@@
 ```
+
+```{r region_table,  echo = F, message = F, fig.cap = region_table_cap}
+region_table_cap <- supTab_cap(name = 'region_table', caption = 'Fovea versus peripheral cone differential gene testing table')
+
+ region_table %>% filter(padj < 0.1) %>% 
+  arrange(padj) %>% flextable() %>% autofit() %>% fit_to_width(7)
+```
+
+```{r cone_region_rep, fig.width=8, fig.height=10,  echo = F, message = F, fig.cap = cone_region_rep_cap}
+cone_region_rep_cap <- supFig_cap(name = 'rod_region_fig', caption = 'Genes identified by Peng et al. and Yan et al. as macula / peripheral differentially expressed in cones.')
+cowplot::plot_grid(prph2, rs1,gngt1, ncol = 1)
+```
+
+```{r rod_region_fig, fig.width=20, fig.height=10,  echo = F, message = F, fig.cap = rod_region_fig_cap}
+rod_region_fig_cap <- supFig_cap(name = 'rod_region_fig', caption = 'Human cone fovea - peripheral differentially expressed genes are not differentially expressed in human rods.')
+
+# 
+# rods
+input <- list()
+input$exp_plot_groups <- c('study_accession')
+input$exp_plot_facet <- c('CellType_predict', 'organism', 'retina_region')
+input$exp_plot_genes <- c('PRPH2 (ENSG00000112619)')
+input$exp_filter_cat <- c('CellType_predict', 'organism')
+#input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
+input$exp_filter_on <- c("Rod", "Homo sapiens" )
+input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
+#input$exp_plot_ylab <- '% of Cells Detected'
+#input$exp_plot_groups <- c('organism')
+input$exp_plot_col_num <- 16
+input$exp_filter_min_cell_number <- 10
+input$flip_facet <- FALSE
+input$figure <- FALSE
+input$figure_fovea <- TRUE
+input$exp_plot_genes <- diff_region_cone_genes[1]; a <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') +  theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[2]; b <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[3]; c <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[4]; d <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + ylab('') + xlab('') + theme(axis.ticks.x=element_blank(), axis.text.x=element_blank() ) 
+input$exp_plot_genes <- diff_region_cone_genes[5]; e <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[6]; f <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[7]; g <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+
+region_legend <- get_legend(d)
+
+plot_grid(plot_grid(a,b,c,d + theme(legend.position="none"),e,f,g, nrow = 1),
+          region_legend, 
+          rel_heights = c(1,0.2), 
+          nrow = 2)
+```
+
+# Works Cited

### 2022-07-26 src/diff_data_pull.R

@@ -151,7 +151,7 @@ consist_diff_long <- scEiaD_2020_v01 %>%
   tbl("diff_testing") %>%
   filter(Gene %in% g) %>%
   collect() %>%
-  filter(Base %in% ws_oct,
+  filter(Base %in% top_genes$Base,
          !Base %in% c('Early RPC','Late RPC'), 
          Group == 'CellType (Predict)',
          Against == 'All') %>%

### 2022-07-26 src/make_exp_plot.R

@@ -56,41 +56,46 @@ make_exp_plot <- function(input, db, meta_filter){
   
   
   #cat(input)
-  box_data <- box_data %>%
-    filter(!Platform %in% c('SCRBSeq')) %>%
-    # mutate(CellType_predict = case_when(CellType_predict == 'RPC' ~ 'RPCs',
-    #                                     CellType_predict == 'Mesenchymal/RPE/Endothelial' ~ 'Endothelial',
-    #                                     TRUE ~ CellType_predict)) %>%
-    mutate(Stage = factor(Stage, levels = c('Early Dev.', 'Late Dev.', 'Maturing', 'Mature'))) %>%
-    #filter(!is.na(!!as.symbol(grouping_features))) %>%
-    group_by_at(vars(one_of(c('Gene', input$exp_plot_facet, grouping_features)))) %>%
-    summarise(counts = sum(counts * cell_exp_ct) / sum(cell_exp_ct),
-              cell_exp_ct = sum(cell_exp_ct, na.rm = TRUE)) %>%
-    full_join(., meta_filter_EXP %>%
-                group_by_at(vars(one_of(input$exp_plot_facet, grouping_features))) %>%
-                summarise(Count = n())) %>%
-    mutate(cell_exp_ct = ifelse(is.na(cell_exp_ct), 0, cell_exp_ct)) %>%
-    mutate(`%` = round((cell_exp_ct / Count) * 100, 2),
-           Expression = round(counts * (`%` / 100), 2)) %>%
-    select_at(vars(one_of(c('Gene', grouping_features, 'cell_exp_ct', 'Count', '%', 'Expression')))) %>%
-    arrange(-Expression) %>%
-    rename(`Cell # Detected` = cell_exp_ct,
-           `Total Cells` = Count,
-           `Mean log2(Counts + 1)` = Expression,
-           `% of Cells Detected` = `%`) %>%
-    
-    mutate(`Mean log2(Counts + 1)` = case_when(is.na(`Mean log2(Counts + 1)`) ~ 0, TRUE ~ `Mean log2(Counts + 1)`)) %>% 
-    filter(`Total Cells` > as.integer(input$exp_filter_min_cell_number))
-  # expand missing genes (nothing detected) to all genes used
-  box_data_NA <- box_data %>% filter(is.na(Gene))
-  if (nrow(box_data_NA > 0)){
-    box_data_NA_list <- list()
-    for (i in gene){
[…]

### 2022-07-26 manuscript_00.Rmd
caption fix

@@
 
-```{r web_resources, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=web_resources_cap}
+```{r web_resources, fig.height = 15, fig.width = 20, echo = F, message = F, results = 'asis', fig.cap=web_resources_cap}
 #tic()()
@@
 web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
+
+cat("<div id=\"tbl:web_resources\">")
 
@@
 #toc()()
+
+#Output Caption (* * for italics)
+cat(paste0("\n*",web_resources_cap,"*\n"))
+# Output </div>
+cat("</div>")
 ```
@@
 
-# ```{r, fig.width=8, fig.height=12, echo = F, message = F}
-# # tic()
-# # # flip_flop diff genes
-# # down_tab <- down_tab_big %>% 
-# #   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
-# #   filter(Base %in% ws_oct, 
-# #          Against == 'All')
-# # 
-# # org_diff_diff_genes <- diff_tab_big %>% 
-# #   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
-# #   filter(gene_ct_combo %in% down_tab$gene_ct_combo, 
-# #          Against == 'All') %>% 
-# #   arrange(padj) %>% mutate(x = paste0(Symbol, ' (', ID, ')')) 
-# # 
-# # flip_flop_genes <- scEiaD_2020_v01 %>% tbl('diff_testing_genes') %>% collect() %>% filter(grepl(org_diff_diff_genes %>% pull(ID) %>% paste0(., collapse = '|'), Gene)) %>% pull(Gene)
-# # flip_flop_long <- scEiaD_2020_v01 %>%
-# #   tbl("diff_testing") %>%
-# #   filter(Gene %in% genes) %>%
-# #   collect() %>%
-# #   mutate(Symbol = str_extract(Gene, '^\\w+ ') %>% gsub(' $', '', .)) %>% 
-# #   mutate(gene_ct_combo = paste(Symbol, Base, sep = '_')) %>% 
-# #   filter(gene_ct_combo %in% org_diff_diff_genes$gene_ct_combo,
-# #          Group == 'CellType (Predict)',
-# #          Against == 'All') %>%
-# #   select(Gene, Organism, log2FoldChange, Base)
-# # 
-# # flip_flop_wide <- flip_flop_long %>% pivot_wider(values_from = log2FoldChange, names_from = Organism)
-# # flip_flop_organism <- colnames(flip_flop_wide)[c(3,4,5)]
-# # flip_flop_wide <- flip_flop_wide %>% data.frame()
-# # row.names(flip_flop_wide) <- paste0(flip_flop_wide$Gene,'_', flip_flop_wide$Base)
-# # 
-# # flip_flop_wide <- flip_flop_wide[,-c(1,2)]
-# # 
-# # 
-# # 
-# # flip_flop_genes <- row.names(flip_flop_wide) %>% gsub(' \\(.*','',.)
-# # 
-# # flip_flop_base <- row.names(flip_flop_wide) %>% gsub('.*_','',.) 
-# 
-# gene_anno <- ComplexHeatmap::rowAnnotation(Duplicated = flip_flop_genes %in% (flip_flop_genes[duplicated(flip_flop_genes)]),
-#                                            col = list(Duplicated = c("TRUE" = "green", "FALSE" = "gray")))
-# hm_flipflop <- ComplexHeatmap::Heatmap(flip_flop_wide, column_labels = flip_flop_organism, 
-#                                        row_split = flip_flop_base, show_row_dend = FALSE, 
-#                                        row_labels = flip_flop_genes, 
-#                                        column_title_rot = 90, 
-#                                        right_annotation = gene_anno,
-#                                        name = 'log2FC', use_raster = TRUE,
-#                                        row_title_rot = 0)
-# hm_flipflop
-# #toc()
-# ```
-
 ## Our pan-study macula vs peripheral cone test replicates a published macaque finding and proposes six new markers
@@
 
-```{r diff_tab_sup_tab, echo = F, message = F,  echo = F, message = F, fig.cap=diff_tab_sup_cap}
+```{r diff_tab_sup_tab, echo = F, message = F,  echo = F, message = F, results = 'asis', fig.cap=diff_tab_sup_cap}
 diff_tab_sup_cap <- supTab_cap(name='diff_tab_sup', caption = 'Counts of signifantly differentially expressed genes for each cell type across the organisms')
-
+cat("<div id=\"tbl:diff_tab_sup\">")
 # tic()
@@
 diff_tab_sup %>% flextable() %>% autofit() %>% fit_to_width(7)
-```
-
-```{r well_supported_celltypes_tab, echo = F, message = F, fig.cap=well_supported_celltypes_tab_cap}
+
+#Output Caption (* * for italics)
+cat(paste0("\n*",diff_tab_sup_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
+```{r well_supported_celltypes_tab, echo = F, message = F, results = 'asis', fig.cap=well_supported_celltypes_tab_cap}
 well_supported_celltypes_tab_cap <- supTab_cap(name='well_supported_celltypes_tab', caption = 'Counts of independent studies for each cell type across the organisms')
-
+cat("<div id=\"tbl:well_supported_celltypes_tab\">")
 ws_table %>% flextable() %>% autofit() %>% fit_to_width(7)
-```
-
-```{r neuro_table, echo = F, message = F,  echo = F, message = F, fig.cap=neuro_table_cap}
+
+#Output Caption (* * for italics)
+cat(paste0("\n*",well_supported_celltypes_tab_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
+```{r neuro_table, echo = F, message = F,  echo = F, message = F, results = 'asis',fig.cap=neuro_table_cap}
 neuro_table_cap <- supTab_cap(name='neuro_table', caption = 'Differentially expressed genes between RPC and neurogenic cells in human and mouse.
 Genes filtered with a abs(log2FoldChange) > 2 and a padj < 0.01.')
-
+cat("<div id=\"tbl:neuro_table\">")
 pairwise_neurogenic_test_ungrouped %>% filter(padj < 0.05) %>% filter(log2FoldChange > 2 | log2FoldChange < 0) %>% flextable() %>% autofit() %>% fit_to_width(7)
+
+#Output Caption (* * for italics)
+cat(paste0("\n*",neuro_table_cap,"*\n"))
+# Output </div>
+cat("</div>")
 ```
@@
 
-```{r studies_table, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=studies_table_cap}
+```{r studies_table, fig.height = 15, fig.width = 20, echo = F, message = F, results = 'asis',fig.cap=studies_table_cap}
 studies_table_cap <- supTab_cap(name = 'studies_table', caption = 'Studies, tissues, and number of cells across scEiaD')
-
+cat("<div id=\"tbl:studies_table\">")
 meta_filter %>% group_by(Citation, PMID, organism, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable() %>% autofit() %>% fit_to_width(7)
-```
-
-```{r region_table,  echo = F, message = F, fig.cap = region_table_cap}
+
+#Output Caption (* * for italics)
+cat(paste0("\n*",studies_table_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
+```{r region_table,  echo = F, message = F, results = 'asis',fig.cap = region_table_cap}
 region_table_cap <- supTab_cap(name = 'region_table', caption = 'Fovea versus peripheral cone differential gene testing table')
-
+cat("<div id=\"tbl:region_table\">")
 region_table %>% filter(padj < 0.1) %>% 
   arrange(padj) %>% flextable() %>% autofit() %>% fit_to_width(7)
+
+#Output Caption (* * for italics)
+cat(paste0("\n*",region_table_cap,"*\n"))
+# Output </div>
+cat("</div>")
 ```

### 2022-08-02 manuscript_00.Rmd
text updates

@@
 
-For users interested in seeing how expression of a gene relates to study or other covariates, the "Expression Plot" view displays expression by cell type, predicted cell type, or cluster.
+For users interested in seeing how expression of a gene relates to study or other covariates, the Expression Plot view displays expression by cell type, predicted cell type, or cluster.
 The values can then be split and colored by multiple covariates like study or organism.
 This allows a user to see whether a gene pattern is consistent across studies or species.
-We also offer a "Dot Plot" view which can display a large number of genes in a space efficient manner.
+We also offer a Dot Plot view which can display a large number of genes in a space efficient manner.
 This allows a users to see whether a set of genes has related (or not) expression patterns across multiple cell types (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`b and c).
@@
 Like all visualizations in PLAE, the user can use powerful filtering options to only display data from certain metadata characteristics, like species or publication (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`d).
+
+Finally we have also added a Heatmap visualization which summaries the relative expression differences between user selected cell types or clusters across human, macaque, and mouse.
+This plot differs from the Dot Plot view as the expression values are relative instead of absolute.
+That is to say that while the Dot Plot view is proportionate to counts, the Heatmap view is showing the difference in expression between a given cell type (or cluster) and all other cells.
+This view is potentially useful if a user is especially interested in gene expression patterns across a variety of cell types.
 
@@
 In contrast CALB1 did replicate strongly with the sixth lowest padj (0.00721) across our entire test and a log2 fold change of 2.87.
-We find six more genes with a padj < 0.01: PDE6G, AKAP12, HSPB1, GADD45B, CC2D2A, and RP1L1 (`r fig_cap(name = 'region_fig', display = 'cite')`).
+We find six more genes with a padj < 0.01: PDE6G, AKAP12, HSPB1, GADD45B, CC2D2A, and RP1L1 (`r fig_cap(name = 'region_fig', display = 'cite')`and `r supTab_cap(name = 'region_table', display = 'cite')`).
 None of these candidate cone region marker genes display regionally different expression in rods (`r supFig_cap(name = 'rod_region_fig', display = 'cite')`).
@@
 # Supplementary Figures and Tables
-
-```{r diff_tab_sup_tab, echo = F, message = F,  echo = F, message = F, results = 'asis', fig.cap=diff_tab_sup_cap}
-diff_tab_sup_cap <- supTab_cap(name='diff_tab_sup', caption = 'Counts of signifantly differentially expressed genes for each cell type across the organisms')
-cat("<div id=\"tbl:diff_tab_sup\">")
-# tic()
-# # number of celltype (predict) - organism - study counts (with >=50 cells)
-# # ws == Well Supported
-# ws_table <- meta_filter %>%  
-#   filter(Organ == 'Eye') %>% 
-#   group_by(organism, CellType_predict, study_accession) %>% 
-#   count() %>% 
-#   filter(n >= 50) %>% 
-#   ungroup() %>% 
-#   group_by(organism, CellType_predict) %>% 
-#   count() %>% 
-#   pivot_wider(values_from = n, names_from = organism) %>% 
-#   filter(!is.na(CellType_predict)) %>% 
-#   replace(is.na(.), values = 0) %>% 
-#   arrange(CellType_predict) %>% 
-#   rename(`CellType (Predict)` = CellType_predict) %>% 
-#   mutate(`Well Supported` = case_when(`Homo sapiens` >= 2 & `Mus musculus` >= 2 ~ 'Yes',
-#                                       TRUE ~ 'No'))
-# 
-# # fix  occasional NA
-# # data made from src/ensdb.R
-# diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
-# down_tab_big <- read_tsv('data/down_tab_big.tsv.gz')
-# gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
-# diff_tab <- diff_tab_big %>% 
-#   filter(Against == 'All', 
-#          Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% 
-#                       pull(CellType) %>% unique()))
-# 
-# 
-# diff_tab_sup <- diff_tab %>% 
-#   group_by(Base, Organism) %>% 
-#   summarise(Count = n()) %>% 
-#   rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
-#   pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
-#   as_tibble()
-# 
-# diff_tab_sup <- bind_rows(
-#   diff_tab_sup,
-#   ws_table$`CellType (Predict)`[!ws_table$`CellType (Predict)` %in% 
-#                                   diff_tab_sup$`CellType (Predict)`] %>% 
-#     enframe() %>% 
-#     select(`CellType (Predict)` = value) %>%
-#     mutate(`Homo sapiens` = 0, `Mus musculus` = 0, `Macaca fascicularis` = 0) %>% 
-#     arrange(`CellType (Predict)`)
-# )
-# 
-# write_csv(diff_tab, file = 'supplemental_files/differential_expression_table.csv.gz')
-
-# diff_tab_sup %>% 
-#   flextable()
-# 
-# ws_table %>% 
-#   flextable()
-
-diff_tab_barplot <- diff_tab_sup %>% 
-  pivot_longer(-`CellType (Predict)`, names_to = 'Organism', values_to = 'Diff Gene Count') %>% 
-  ggplot(aes(x=`CellType (Predict)`, y = `Diff Gene Count`, fill = Organism )) + 
-  geom_bar(stat='identity', position = 'dodge') + 
-  cowplot::theme_cowplot() + coord_flip()
-
-#toc()
-
-diff_tab_sup %>% flextable() %>% autofit() %>% fit_to_width(7)
-
-#Output Caption (* * for italics)
-cat(paste0("\n*",diff_tab_sup_cap,"*\n"))
-# Output </div>
-cat("</div>")
-```
-
-```{r well_supported_celltypes_tab, echo = F, message = F, results = 'asis', fig.cap=well_supported_celltypes_tab_cap}
-well_supported_celltypes_tab_cap <- supTab_cap(name='well_supported_celltypes_tab', caption = 'Counts of independent studies for each cell type across the organisms')
-cat("<div id=\"tbl:well_supported_celltypes_tab\">")
-ws_table %>% flextable() %>% autofit() %>% fit_to_width(7)
-
-#Output Caption (* * for italics)
-cat(paste0("\n*",well_supported_celltypes_tab_cap,"*\n"))
-# Output </div>
-cat("</div>")
-```
-
-```{r neuro_table, echo = F, message = F,  echo = F, message = F, results = 'asis',fig.cap=neuro_table_cap}
-neuro_table_cap <- supTab_cap(name='neuro_table', caption = 'Differentially expressed genes between RPC and neurogenic cells in human and mouse.
-Genes filtered with a abs(log2FoldChange) > 2 and a padj < 0.01.')
-cat("<div id=\"tbl:neuro_table\">")
-pairwise_neurogenic_test_ungrouped %>% filter(padj < 0.05) %>% filter(log2FoldChange > 2 | log2FoldChange < 0) %>% flextable() %>% autofit() %>% fit_to_width(7)
-
-#Output Caption (* * for italics)
-cat(paste0("\n*",neuro_table_cap,"*\n"))
-# Output </div>
-cat("</div>")
-```
 
@@
 
-```{r studies_table, fig.height = 15, fig.width = 20, echo = F, message = F, results = 'asis',fig.cap=studies_table_cap}
-studies_table_cap <- supTab_cap(name = 'studies_table', caption = 'Studies, tissues, and number of cells across scEiaD')
-cat("<div id=\"tbl:studies_table\">")
-meta_filter %>% group_by(Citation, PMID, organism, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable() %>% autofit() %>% fit_to_width(7)
-
-#Output Caption (* * for italics)
-cat(paste0("\n*",studies_table_cap,"*\n"))
-# Output </div>
-cat("</div>")
-```
-
-```{r region_table,  echo = F, message = F, results = 'asis',fig.cap = region_table_cap}
-region_table_cap <- supTab_cap(name = 'region_table', caption = 'Fovea versus peripheral cone differential gene testing table')
-cat("<div id=\"tbl:region_table\">")
-region_table %>% filter(padj < 0.1) %>% 
-  arrange(padj) %>% flextable() %>% autofit() %>% fit_to_width(7)
-
-#Output Caption (* * for italics)
-cat(paste0("\n*",region_table_cap,"*\n"))
-# Output </div>
-cat("</div>")
-```
-
 ```{r cone_region_rep, fig.width=8, fig.height=10,  echo = F, message = F, fig.cap = cone_region_rep_cap}
-cone_region_rep_cap <- supFig_cap(name = 'rod_region_fig', caption = 'Genes identified by Peng et al. and Yan et al. as macula / peripheral differentially expressed in cones.')
+cone_region_rep_cap <- supFig_cap(name = 'cone_region_rep', caption = 'Genes identified by Peng et al. and Yan et al. as macula / peripheral differentially expressed in cones.')
 cowplot::plot_grid(prph2, rs1,gngt1, ncol = 1)
@@
 
+```{r studies_table, fig.height = 15, fig.width = 20, echo = F, message = F, results = 'asis',fig.cap=studies_table_cap}
+# 1
+
+studies_table_cap <- supTab_cap(name = 'studies_table', caption = 'Studies, tissues, and number of cells across scEiaD')
+cat("<div id=\"tbl:studies_table\">")
+meta_filter %>% group_by(Citation, PMID, organism, study_accession) %>% summarise(Count = n(), Tissues = paste(unique(Tissue), collapse = ', ')) %>% flextable() %>% autofit() %>% fit_to_width(7)
+
+#Output Caption (* * for italics)
+cat(paste0("\n*",studies_table_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
+```{r diff_tab_sup_tab, echo = F, message = F,  echo = F, message = F, results = 'asis', fig.cap=diff_tab_sup_tab_cap}
+# 2
+diff_tab_sup_tab_cap <- supTab_cap(name='diff_tab_sup_tab', caption = 'Counts of significantly differentially expressed genes for each cell type across the organisms')
+cat("<div id=\"tbl:diff_tab_sup_tab\">")
+# tic()
+# # number of celltype (predict) - organism - study counts (with >=50 cells)
+# # ws == Well Supported
+# ws_table <- meta_filter %>%  
+#   filter(Organ == 'Eye') %>% 
+#   group_by(organism, CellType_predict, study_accession) %>% 
+#   count() %>% 
+#   filter(n >= 50) %>% 
+#   ungroup() %>% 
+#   group_by(organism, CellType_predict) %>% 
+#   count() %>% 
+#   pivot_wider(values_from = n, names_from = organism) %>% 
+#   filter(!is.na(CellType_predict)) %>% 
+#   replace(is.na(.), values = 0) %>% 
+#   arrange(CellType_predict) %>% 
+#   rename(`CellType (Predict)` = CellType_predict) %>% 
+#   mutate(`Well Supported` = case_when(`Homo sapiens` >= 2 & `Mus musculus` >= 2 ~ 'Yes',
+#                                       TRUE ~ 'No'))
+# 
+# # fix  occasional NA
+# # data made from src/ensdb.R
+# diff_tab_big <- read_tsv('data/diff_tab_big.tsv.gz')
+# down_tab_big <- read_tsv('data/down_tab_big.tsv.gz')
+# gene_tab <- bind_rows(read_tsv('data/hs.ensdb.v86.symbol.geneid.tsv.gz'), read_tsv('data/mm.ensdb.v79.symbol.geneid.tsv.gz'))
+# diff_tab <- diff_tab_big %>% 
+#   filter(Against == 'All', 
+#          Base %in% (meta_filter %>% filter(Organ == 'Eye') %>% 
+#                       pull(CellType) %>% unique()))
+# 
+# 
+# diff_tab_sup <- diff_tab %>% 
+#   group_by(Base, Organism) %>% 
+#   summarise(Count = n()) %>% 
+#   rename(`CellType (Predict)` = Base, `Differentially Expressed Gene Count` = Count) %>% 
+#   pivot_wider(values_from = `Differentially Expressed Gene Count`, names_from = Organism) %>% replace(is.na(.),0) %>% 
+#   as_tibble()
+# 
+# diff_tab_sup <- bind_rows(
+#   diff_tab_sup,
+#   ws_table$`CellType (Predict)`[!ws_table$`CellType (Predict)` %in% 
+#                                   diff_tab_sup$`CellType (Predict)`] %>% 
+#     enframe() %>% 
+#     select(`CellType (Predict)` = value) %>%
+#     mutate(`Homo sapiens` = 0, `Mus musculus` = 0, `Macaca fascicularis` = 0) %>% 
+#     arrange(`CellType (Predict)`)
+# )
+# 
+# write_csv(diff_tab, file = 'supplemental_files/differential_expression_table.csv.gz')
+
+# diff_tab_sup %>% 
+#   flextable()
+# 
+# ws_table %>% 
+#   flextable()
+
+diff_tab_barplot <- diff_tab_sup %>% 
+  pivot_longer(-`CellType (Predict)`, names_to = 'Organism', values_to = 'Diff Gene Count') %>% 
+  ggplot(aes(x=`CellType (Predict)`, y = `Diff Gene Count`, fill = Organism )) + 
+  geom_bar(stat='identity', position = 'dodge') + 
+  cowplot::theme_cowplot() + coord_flip()
+
+#toc()
+
+diff_tab_sup %>% flextable() %>% autofit() %>% fit_to_width(7)
+
+#Output Caption (* * for italics)
+cat(paste0("\n*",diff_tab_sup_tab_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
+```{r well_supported_celltypes_tab, echo = F, message = F, results = 'asis', fig.cap=well_supported_celltypes_tab_cap}
+# 2
+well_supported_celltypes_tab_cap <- supTab_cap(name='well_supported_celltypes_tab', caption = 'Counts of independent studies for each cell type across the organisms')
+cat("<div id=\"tbl:well_supported_celltypes_tab\">")
+ws_table %>% flextable() %>% autofit() %>% fit_to_width(7)
+
+#Output Caption (* * for italics)
+cat(paste0("\n*",well_supported_celltypes_tab_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
+```{r neuro_table, echo = F, message = F,  echo = F, message = F, results = 'asis',fig.cap=neuro_table_cap}
+# 4
+neuro_table_cap <- supTab_cap(name='neuro_table', caption = 'Differentially expressed genes between RPC and neurogenic cells in human and mouse.
+Genes filtered with a abs(log2FoldChange) > 2 and a padj < 0.01.')
+cat("<div id=\"tbl:neuro_table\">")
+pairwise_neurogenic_test_ungrouped %>% filter(padj < 0.05) %>% filter(log2FoldChange > 2 | log2FoldChange < 0) %>% flextable() %>% autofit() %>% fit_to_width(7)
+
+#Output Caption (* * for italics)
+cat(paste0("\n*",neuro_table_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
+```{r region_table,  echo = F, message = F, results = 'asis',fig.cap = region_table_cap}
+# 5
+region_table_cap <- supTab_cap(name = 'region_table', caption = 'Fovea versus peripheral cone differential gene testing table')
+cat("<div id=\"tbl:region_table\">")
+region_table %>% filter(padj < 0.1) %>% 
+  arrange(padj) %>% flextable() %>% autofit() %>% fit_to_width(7)
+
+#Output Caption (* * for italics)
+cat(paste0("\n*",region_table_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
 # Works Cited

### 2022-08-02 supplemental_files/cone_fovea_vs_peripheral_diff_testing.Rmd
sup files

---
title: "Test Fovea vs Peripheral Cones"
output: html_notebook
---

# Load
```{r}
library(Seurat)
library(tidyverse)
library(SingleCellExperiment)
library(scran)
library(org.Hs.eg.db)
#system('wget -O ~/Downloads/scEiaD_all_seurat_v3.Rdata https://hpc.nih.gov/~mcgaugheyd/scEiaD/2022_03_22/scEiaD_all_seurat_v3.Rdata)
# the -O renames the file and puts it in the ~/data/scEiaD directory
load('~/Downloads/scEiaD_all_seurat_v3.Rdata')
```
## Compare macula cones vs peripheral cones
Let us now demonstrate how having a large annotated set of data can be of value in asking more specific questions, like "how do macula/fovea cones differ relative to peripheral cones."

First, we need to see whether this is even a tractable question.
```{r}
scEiaD@meta.data %>% 
  as_tibble() %>% 
  group_by(retina_region, organism, study_accession, CellType_predict) %>% 
  summarise(Count = n()) %>% 
  filter(CellType_predict %in% c("Cone")) %>% 
  filter(Count > 10, !is.na(retina_region))
```

Yes, we have several human studies with macula / peripheral cone cells.

Let's demonstrate how we can quickly run a conservative differential expression test the leverages the many studies we have. The "traditional" scRNA diff tests often return insane numbers of differentially expressed genes, which is  annoying as then you have to plot each result individually to confirm that it does not look wacky. We are going to build a "pseudo bulk" experiment here where we sum all the counts by gene, cell type (in this case we are doing macula cones *versus* peripheral cones), and study. This "pseudo" count table can then be used in a classic differential testing system like edgeR, limma, or DESeq2 (we will use the later).

```{r}
# Create new column with a pasted together retina region and a CellType
scEiaD <- AddMetaData(scEiaD, 
                      metadata = 
                        paste0(
                          scEiaD@meta.data$retina_region, 
                          '_', 
                          scEiaD@meta.data$CellType_predict,
                          '_',
                          scEiaD@meta.data$organism
                        ), 
                      col.name = 'regionCT')

# filter down scEiaD to only human and in semi-supported (more than 10 cones cells) studies
Idents(scEiaD) <- scEiaD@meta.data$regionCT
scEiaD__subset <- subset(scEiaD, idents = c('Macula_Cone_Homo sapiens','Peripheral_Cone_Homo sapiens'))
scEiaD__subset <- subset(scEiaD__subset, 
                         subset = study_accession %in% 
                           (scEiaD__subset@meta.data %>% group_by(study_accession) %>% 
                              summarise(Count = n()) %>% filter(Count > 10) %>% 
                              pull(study_accession)))
# create SCE object so can create a pseudobulk matrix to run DESeq2 on
sce <- as.SingleCellExperiment(scEiaD__subset)
summed <- scater::aggregateAcrossCells(sce, 
                                       ids=colData(sce)[,c("regionCT", "study_accession")])
# pull out the counts to build the DESeq2 object
mat <- assay(summed, 'counts')
colnames(mat) <-colData(summed) %>% as_tibble() %>% 
  mutate(names = glue::glue("{study_accession}_{retina_region}")) %>% pull(names)

library(DESeq2)
dds <- DESeqDataSetFromMatrix(countData = round(mat),
                              colData = colData(summed),
                              design= ~ study_accession + retina_region) # study as covariate, testing against fovea / not fovea
dds <- DESeq(dds)
deseq_res <- results(dds)
```

### Wrinkle

scEiaD is based on the Ensembl gene id (like ENSG00000185527). That's not very friendly to humans so I added a bit of code to link the Ensembl ID to the gene name and plot the top differentially expressed genes (padj < 0.05).

```{r}
# get human gene names 
library(org.Hs.eg.db)
symbols <- mapIds(org.Hs.eg.db, keys = row.names(deseq_res), keytype = "ENSEMBL", column="SYMBOL")

deseq_res %>% as_tibble(rownames = 'name') %>%  
  left_join(symbols %>% enframe()) %>% 
  dplyr::rename(Ensembl = name, Gene = value) %>% 
  filter(padj < 0.05) %>% 
  arrange(pvalue) %>% DT::datatable()
```

### Always look at the plots

#### PDE6G 
Top gene more highly expressed in the macula compared to the periphery. Seems pretty consistently higher across all the studies. 
```{r}
scEiaD__subset <- ScaleData(scEiaD__subset)
VlnPlot(scEiaD__subset, features = c('ENSG00000185527'), log = TRUE)

VlnPlot(scEiaD__subset, c('ENSG00000185527'), split.by = 'retina_region', group.by='study_accession', log = TRUE) 
```

#### HSPB1
Now we see the opposite side - the top gene that is higher in peripheral vs macula. Again looks fairly consistent across all the studies.
```{r}
VlnPlot(scEiaD__subset, features = c('ENSG00000099860'))

VlnPlot(scEiaD__subset, c('ENSG00000099860'), split.by = 'regionCT', group.by='study_accession', ) + aes(color=scEiaD__subset$regionCT)
```

#### PRPH2
Yan et al (https://www.nature.com/articles/s41598-020-66092-9) proposed PRPH2 as a cone fovea/peripheral specific gene. We can see whether that effect is still shown across multiple studies in scEiaD. We see the padj is not impressive in our diff test.

```{r}
# first grab the ensembl id
deseq_res %>% as_tibble(rownames = 'name') %>%  
  left_join(symbols %>% enframe()) %>% 
  dplyr::rename(Ensembl = name, Gene = value) %>% 
  filter(Gene == 'PRPH2')
```

The plots certainly do not DISPROVE the Yan et al hypothesis - you still see that the trend is (mostly) the same across all the studies. However the p value is likely low for because the effect size (fold change) is modest across most studies (and in one case, the effect is opposite). As a exercise to the reader, you could re-do the pseudobulk test, removing SRP194595 and seeing whether padj value you get returned is substantially different.

```{r}
VlnPlot(scEiaD__subset, features = c('ENSG00000112619'))

VlnPlot(scEiaD__subset, c('ENSG00000112619'), split.by = 'regionCT', group.by='study_accession', ) + aes(color=scEiaD__subset$regionCT)

```

# Save our file
```{r}
save(scEiaD__subset, file = 'scEiaD_fovea_vs_peripheral_cone.seuratV3.Rdata')
```

```{r}
devtools::session_info()
```

### 2022-08-02 manuscript_00.Rmd
tweak sizing

@@
 
-```{r consist_genes_heatmap, fig.height=20, fig.width=10, echo = F, message = F,  fig.cap=consist_genes_heatmap_cap}
+```{r consist_genes_heatmap, fig.height=22, fig.width=10, echo = F, message = F,  fig.cap=consist_genes_heatmap_cap}
 consist_genes_heatmap_cap <- fig_cap(name='consist_genes_heatmap', caption = 'Rows are genes, columns are cell types and organisms.
@@
 
-```{r region_fig, fig.width=14, fig.height=3,  echo = F, message = F, fig.cap = region_fig_cap}
+```{r region_fig, fig.width=18, fig.height=3,  echo = F, message = F, fig.cap = region_fig_cap}
 region_fig_cap <- fig_cap(name='regino_fig', caption = 'Plot of markers distinguishing human macula or peripheral punch derived cones.')

### 2022-08-02 manuscript_00.Rmd
table num fix

@@
 
-Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources_cap', display = 'cite')`).
+Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources', display = 'cite')`).
 The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are not ocular focused, but do contain some ocular data.

### 2022-08-11 manuscript_00.Rmd
text updates

--- 2022-08-02 manuscript_00.Rmd
@@
 RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encomopassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 37 publications.
-As a test of the value of this unified pan-eye dataset, we propose consistent cell type markers, replicate neurogenic and cone fovea markers, and propose six new cone fovea and peripheral markers .
+As a test of the value of this unified pan-eye dataset, we propose consistent cell type markers, replicate neurogenic and cone macula markers, and propose six new cone macula and peripheral markers .
 CONCLUSION: The PLAE v0.91 web app serves the pan-ocular and body dataset, scEiaD.
@@
 The modern era of single cell transcriptomics is marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment; in Macoksko et al.
-the adult mouse retina was profiled[@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a].
+the adult mouse retina was profiled [@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a].
 The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since Macosko publication.
@@
 Web-based resources to query gene expression are powerful means to democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device.
-Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas[@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021].
+Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas [@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021].
 Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging.
@@
 In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values.
-We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, genes which have species-specific expression cell type patterning, demonstrate how our pre-computed differential testing can identify markers between similar progenitor cell types, and finally provide a R-based analysis document which end-to-end demonstrates how we use scEiaD to run a custom differential gene expression test to validate and propose new genes that distinguish fovea from peripheral cones in humans.
+We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, genes which have species-specific expression cell type patterning, demonstrate how our pre-computed differential testing can identify markers between similar progenitor cell types, and finally provide a R-based analysis document which end-to-end demonstrates how we use scEiaD to run a custom differential gene expression test to validate and propose new genes that distinguish macula from peripheral cones in humans.
 
@@
 
-A custom cutoffs for the minimum required number of unique genes identified per cell were used for the SRP362101 cornea dataset as an abnormally high number of cells were returned from our default value of 200.
+A custom cutoff for the minimum required number of unique genes identified per cell were used for the SRP362101 cornea dataset as an abnormally high number of cells were returned from our default value of 200.
 We set the cutoff for this studies to be 800 as this value resulted in approximately the same number of cells being returned as the authors reported in their papers.
@@
 We also, where possible, annotated retina region (e.g.
-fovea), sex, and age or developmental stage.
+macula), sex, and age or developmental stage.
 To import individual cell type labels (e.g.
@@
 
-The web app at plae.nei.nih.gov can display gene expression in only a few seconds across over a million cells.
-This speed is only possible due to custom data structures to optimize for data query efficiency.
+The web app at plae.nei.nih.gov can display gene expression across over a million cells in only a few seconds.
+This speed is only possible due to custom data structures that were optimized for data query efficiency.
 Like eyeIntegration's EiaD database [@swamyEyeDiskEyeIntegration2019a], scEiaD is a sqlite database with tables for the core categories.
@@
 There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app.
+The codebase for the web app (along with local installation instructins) is available at github.com/davemcg/scEiaD/plaeApp; the commit corresponding to this manuscript is XXXXXXXXXXXX.
 
@@
 one can see whether a gene of interest is enriched in Cones).
-The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has selected a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app.
+The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has collated a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app.
 PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence, and, crucially, full availability of the underlying data for outside usage and integrating the independent studies to deliver cross study comparisons.
@@
 
-Finally we have also added a Heatmap visualization which summaries the relative expression differences between user selected cell types or clusters across human, macaque, and mouse.
+We have also added a Heatmap visualization which summaries the relative expression differences between user selected cell types or clusters across human, macaque, and mouse.
 This plot differs from the Dot Plot view as the expression values are relative instead of absolute.
@@
 These variable results directly reflect the differing number of independent studies - mouse and human have substantially more than macaque (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
-The full differential table is made available as supplementary file "differential_expression_table.csv.gz."
+The significant filtered differential table is made available as supplementary file "differential_expression_table.csv.gz" and the full differential tables are available for download at plae.nei.nih.gov -> Data.
 
@@
 The photoreceptors are also next to each other
+
+Of course other possibilities for ranking genes are possible.
+We show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`b and c how you can directly use our powerful web app to select candidate bipolar cell differentially expressed genes from the differential testing table and then quickly display the expression pattern of those genes in a compact heatmap visualization.
+The wealth of visualization options in plae allow for quick prototyping and hypothesis testing of gene expression patterning across a large number of cells and studies.
 
@@
 neuro_cap <- fig_cap(name='neuro', caption = 'A.
-Genes identified in Figure 2K of Lu et al.
-as being differently expressed during the RPC to neurogenic differentiation in human.
+Genes proposed by Lu et al.
+as being significantly differentially expressed during the RPC to neurogenic differentiation in human.
 We plot the differential expression (log2(fold change) directly between RPC and neurogenic cells.
@@
 B.
-ZBTB38 .....')
+Genes from Lu et al.
+that we replicated with the scEiaD data.
+C.
+Novel genes we propose as being differentially expressed between RPC and neurogenic cells')
 # tic()
@@
 We find `r up %>% length()` genes that increase in expression from RPC to neurogenic, `r up[!up %in% g_lu] %>% length()` of which are not previously identified by Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`C).
-We replicate `r up[up %in% g_lu] %>% str_extract(., '\\w+') %>% paste(., collapse = ', ')` from Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`B).
+We replicate `r up[up %in% g_lu] %>% str_extract(., '\\w+') %>% paste(., collapse = ', ') %>% str_to_title()` from Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`B) as differentially expressed between neurogenic and RPC.
+Futhermore we show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`a how this figure can be replicated directly in our web app.
 
@@
           nrow = 2)
-
 ```
@@
 Yan et al.
-previously proposed PRPH2 and RS1 as markers to distinguish macula and peripheral located human cones[@yanCellAtlasHuman2020a] and Peng et al.
-proposed CALB1 and GNGT1 as cone region markers in macaque [@pengMolecularClassificationComparative2019a].
-
-Our differential testing found that while PRPH2, RS1, and GNGT1 had expression differences matching that the findings Yan et al.
+previously proposed Prph2 and Rs1 as markers to distinguish macula and peripheral located human cones[@yanCellAtlasHuman2020a] and Peng et al.
+proposed Calb1 and Gngt1 as cone region markers in macaque [@pengMolecularClassificationComparative2019a].
+
+Our differential testing found that while Prph2, Rs1, and Gngt1 had expression differences matching that the findings Yan et al.
 and Peng et al.
 published, they failed to replicate in our test with pvalue and padj being above 0.1 for all three.
-In contrast CALB1 did replicate strongly with the sixth lowest padj (0.00721) across our entire test and a log2 fold change of 2.87.
-We find six more genes with a padj < 0.01: PDE6G, AKAP12, HSPB1, GADD45B, CC2D2A, and RP1L1 (`r fig_cap(name = 'region_fig', display = 'cite')`and `r supTab_cap(name = 'region_table', display = 'cite')`).
+In contrast Calb1 did replicate strongly with the sixth lowest padj (0.00721) across our entire test and a log2 fold change of 2.87.
+We find six more genes with a padj < 0.01: Pde6g, Akap12, Hspb1, Gadd45b, Cc2d2a, and RP1l1 (`r fig_cap(name = 'region_fig', display = 'cite')` and `r supTab_cap(name = 'region_table', display = 'cite')`).
 None of these candidate cone region marker genes display regionally different expression in rods (`r supFig_cap(name = 'rod_region_fig', display = 'cite')`).
@@
 
+## Limitations of scEiaD
+
+The longer term goal is to continue adding datasets to the scEiaD at plae.nei.nih.gov until we have three or more independent datasets per species and nearly all cell types.
+Earlier we proposed well supported cell types.
+Right now, most of these well supported cell types are in the retina.
+The RPE behind the retina, the cornea and lens in front, and the outflow tract and supporting musculature still could use more independent studies.
+Many common human diseases underlie the RPE (e.g.
+AMD) and the cornea (e.g.
+Keratoconus) and these tissues will remain a focus for the next update of our database.
+While we have curated a large number of cell types across the eye, many of these cell types have "sub" cell types.
+For examplethe human has three types of cones that are optimized for short, medium, and long wavelengths.
+More dramatically, Yan et al.
+have proposed over sixty types of mouse amacrine cell types [@yanMouseRetinalCell2020a].
+While we have found it fairly straightforward to use machine learning to transfer the broad (e.g.
+rod, cone, amacrine) cell type labels across all datasets, we have found it very challenging to transfer the "sub" cell type labels.
+This is largely because while we have high diversity of studies for the broad cell types, this is not true for most of the sub celltypes.
+We believe more datasets that carefully dissect the sub-celltypes may be necessary to confidently propose the sub-celltypes of the retina.
+Finally, while our scVI based model can theoretically be used by an outside group to transfer their internal data into the scEiaD latent space, it is very challenging to implement as the the software landscape is still rapidly changing and it is very difficult to architect consistent outputs over time with regularly changing software versions.
+We are investigating simpler methods to enable transfer of knowledge from scEiaD to outside data.
+
 ## High feature web app provides quick access to over one million transcriptomes across `r meta_filter$study_accession %>% unique %>% length` datasets
 
-## Wealth of data avaibility formats enables custom querying
+We have improved on the scEiaD dataset we published in Swamy et al.
+[@swamyBuildingMegaSinglecell2021] by adding another species (chicken), new ocular tissues, a human non-ocular reference set; in total `r meta_filter$Citation %>% unique() %>% length() - meta_filter_v0$Citation %>% unique() %>% length()` new single cell datasets.
+We re-structured the data into a custom database to enable our custom built high performance web app to quickly access the data.
+This democratizes access to over a million cells across `r meta_filter$study_accession %>% unique() %>% length()` studies to anyone with a web browser.
+We recognize that while we have built a powerful visualization platform we cannot anticipate all uses and for those researchers who wish to make custom queries of the data, we provide the data underlying plae in a wide variety of formats, including anndata and Seurat objects for each study, count tables, cell level metadata, and the DESeq2 based differential testing across three species.
+We provide a short example of how a custom query could work by running a custom differential test between macula and peripheral derived human cones and demonstrate how this can both validate existing proposed genes and suggest new ones.
 
 ## High diversity of studies enables reproduction and extension of cell type gene expression findings
+
+A great many groups have proposed cell atlases.
+These range from efforts focusing on a subset of cells from a single layer of a tissue (for example [@tranSingleCellProfilesRetinal2019a] ) to a huge consortium encompassing multiple organs [@consortiumTabulaSapiensSingle2021].
+We propose a complementary approach wherein we use the research data from the community at large to create a meta-atlas.
+This approach enables a small group to produce an atlas which rivals the size from massive multi-group consortiums.
+Furthermore the inherent variability between the data from independent groups can be used as a feature to get insight into whether a gene cell type signal is stable across an entire community's datasets.
+We briefly demonstrate how the high study diversity in scEiaD can be used to validate proposed RPC to neurogenic gene expression changes.
 
@@
 
-```{r cone_region_rep, fig.width=8, fig.height=10,  echo = F, message = F, fig.cap = cone_region_rep_cap}
-cone_region_rep_cap <- supFig_cap(name = 'cone_region_rep', caption = 'Genes identified by Peng et al. and Yan et al. as macula / peripheral differentially expressed in cones.')
-cowplot::plot_grid(prph2, rs1,gngt1, ncol = 1)
+```{r PLAE_screenshots_supplemental, echo = F, message = F, fig.cap = PLAE_screenshots_supplemental_cap}
+PLAE_screenshots_supplemental_cap <- supFig_cap(name='PLAE_screenshots_supplemental', caption = 'a.
+Recreation of Figure 4 directly in the plae.nei.nih.gov web app.
+b.
+Identify potential bipolar cell markers with the precomputed differential testing and in c.
+visualize the top five genes with a Heatmap.')
+knitr::include_graphics('fig/plae_sup_fig_screenshots.png')
+
+```
+
+```{r lu_et_al, echo = F, message = F, fig.cap = lu_et_al_cap}
+lu_et_al_cap <- supFig_cap(name='lu_et_al', caption = 'Supplemental figure 2K from Lu et al showing their candidate RPC / Neurogenic differentially expressed genes')
+knitr::include_graphics('fig/lu_et_al_fig2k_screenshot.png')
+
 ```
@@
 
+```{r cone_region_rep, fig.width=8, fig.height=10,  echo = F, message = F, fig.cap = cone_region_rep_cap}
+cone_region_rep_cap <- supFig_cap(name = 'cone_region_rep', caption = 'Genes identified by Peng et al. and Yan et al. as macula / peripheral differentially expressed in cones.')
+cowplot::plot_grid(prph2, rs1,gngt1, ncol = 1)
+```
+
 ```{r studies_table, fig.height = 15, fig.width = 20, echo = F, message = F, results = 'asis',fig.cap=studies_table_cap}
@@
 # 5
-region_table_cap <- supTab_cap(name = 'region_table', caption = 'Fovea versus peripheral cone differential gene testing table')
+region_table_cap <- supTab_cap(name = 'region_table', caption = 'Macula versus peripheral cone differential gene testing table')
 cat("<div id=\"tbl:region_table\">")

### 2022-08-25 manuscript_00.Rmd
many text updates

@@
 ---
-title: 'PLAE enables quick and easy searching across one million ocular transcriptomes'
+title: 'PLAE web app enables powerful searching and multiple visualizations across one million unified single-cell ocular transcriptomes'
 author:
@@
 csl: investigative-ophthalmology-and-visual-science.csl
-abstract: "PURPOSE: To create a high performance reactive web application to query gene expression across cell type, species, study, and other factors.
+abstract: "PURPOSE: To create a high performance reactive web application to query single cell gene expression data across cell type, species, study, and other factors.
 METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it.
 RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encomopassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 37 publications.
-As a test of the value of this unified pan-eye dataset, we propose consistent cell type markers, replicate neurogenic and cone macula markers, and propose six new cone macula and peripheral markers .
-CONCLUSION: The PLAE v0.91 web app serves the pan-ocular and body dataset, scEiaD.
-This provides the eye community a powerful and quick means to test hypotheses on gene expression across a highly diverse community derived database."
+To demonstrate the value of this unified pan-eye dataset, we replicate known neurogenic and cone macula markers as well as propose six????
+new cone human region markers.
+CONCLUSION: The PLAE web application provides the eye community a powerful and quick means to test hypotheses related to gene expression across a highly diverse, community-derived database."
 
@@
 
+data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
+data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
 scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vGiga}/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite'))
@@
 
+# Celltypes in celltype but not celltype_predict
+dropped <- meta_filter$CellType[!meta_filter$CellType %in% meta_filter$CellType_predict]
+
 #toc()
@@
 
-Single cell transcriptomics of the retina date to the early 2000s when researchers [@hagstromConePigmentGene2000b; @jabsEvidenceP2XP2X2000; @liLocalizationDopamineD1receptor2004; @lindqvistSingleCellRTPCR2002a; @reidRetinoschisinPhotoreceptorsecretedProtein2003]  used RT-PCR approaches to amplify select transcripts in small numbers of cells.
+The retina is composed of six major cell populations: the photoreceptors, horizontal, bipolar, amacrine, ganglion, and non-neuronal cells [@masland_cell_2011].
+These six populations can be further divided into dozens of cell types [@yan_cell_2020].
+Furthermore, human disease of the eye can derive from damage to specific cell types.
+Glaucoma is characterized by damage to the retinal ganglion cells while cone and cone-rod dystrophies are characterized by photoreceptor degeneration.
+Efforts to untangle the transcriptomics of these diverse cell types with single cell transcriptomics date to the early 2000s when researchers [@hagstromConePigmentGene2000b; @jabsEvidenceP2XP2X2000; @liLocalizationDopamineD1receptor2004; @lindqvistSingleCellRTPCR2002a; @reidRetinoschisinPhotoreceptorsecretedProtein2003]  used RT-PCR approaches to amplify select transcripts in small numbers of cells.
 Later, researchers used microarray platforms to analyze a broader array of transcripts[@cherryDevelopmentDiversificationRetinal2009; @gustincichGeneDiscoveryGenetically2004a; @kimIdentificationMolecularMarkers2008a; @roeschTranscriptomeRetinalMuller2008; @trimarchiIndividualRetinalProgenitor2008; @trimarchiMolecularHeterogeneityDeveloping2007b; @wahlinMethodAnalysisGene2004a]
 
-The modern era of single cell transcriptomics is marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment; in Macoksko et al.
-the adult mouse retina was profiled [@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a].
+The modern era of single cell transcriptomics ha been marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment.
+This approach was pioneered by Macoksko et al.
+and enabled the profiling of gene expression in 44,808 single cells in the adult mouse retina.
+[@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a].
 The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since Macosko publication.
 
-These publications's foci includes, but is not limited to, studying developmental transcriptomic dynamics, identifying genes to distinguish core cell types, proposing finer gradations of cell types within existing categories, and comparing ocular transcriptomes between organisms.
-Independent querying of these datasets requires a labrious series of steps.
+These publications's foci includes, but are not limited to, studying developmental transcriptomic dynamics, identifying genes to distinguish core cell types, proposing finer gradations of cell types within existing categories, and comparing ocular transcriptomes between organisms.
+Independent querying of these datasets requires a laborious series of steps.
+
+  Unfortunately manuscripts can only present a sliver of the underlying data and further exploration is laborious, technically challenging, and computationally expensive.
 Briefly, one would have to identify the relevant publication, find the associated data deposit, download the cell - count matrix, find the cell label table (and optionally the t-SNE or UMAP coordinates), and then load the data into R or python to run the desired queries.
@@
 
-Web-based resources to query gene expression are powerful means to democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device.
+Web-based resources to query gene expression can democratize access to large datasets as the computationally intensive steps above can be run once and then shared to anyone with a internet connected device.
 Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas [@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021].
@@
 In contrast to these, the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020].
-However each data deposit is independently processed, which makes cross study queries difficult.
+However each data deposit is independently processed, which makes cross study queries impossible.
 We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space [@swamyBuildingMegaSinglecell2021].
 
-To broaden the usage of this powerful database, we restructured the data into a custom sqlite-based database and built a reactive web application around it.
+To broaden the usage of this powerful meta-atlas, we restructured the data into a custom sqlite-based database and built a reactive web application to access it.
 As many new studies were published since our last data build, we re-ran the scEiaD pipeline in March of 2022.
 In aggregate we have collated `r meta_filter$study_accession %>% unique() %>% length()` datasets across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues.
-We hand-curated `r meta_filter$CellType %>% unique() %>% length()` cell type labels, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the remaining cells.
+We hand-curated `r meta_filter$CellType_predict %>% unique() %>% length()` cell type labels, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the `r length(meta_filter$CellType_predict[!is.na(meta_filter$CellType_predict)]) - length(meta_filter$CellType[!is.na(meta_filter$CellType)])` more cells.
 Quality control metrics were tuned and applied to remove lower quality cells.
 In the end, the single cell Eye in a Disk (scEiaD) database contains `r meta_filter %>% nrow()` cells and 1,509,179,347 non-zero gene/cell expression values.
-We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, genes which have species-specific expression cell type patterning, demonstrate how our pre-computed differential testing can identify markers between similar progenitor cell types, and finally provide a R-based analysis document which end-to-end demonstrates how we use scEiaD to run a custom differential gene expression test to validate and propose new genes that distinguish macula from peripheral cones in humans.
+We demonstrate how this unified resource can be used to identify genes that consistently distinguish cell types across studies and species, find genes which have species-specific expression cell type patterning, demonstrate how our pre-computed differential testing can identify markers between similar progenitor cell types, and finally provide a R-based analysis document which end-to-end demonstrates how we use scEiaD to run a custom differential gene expression test to validate and propose new genes that distinguish macula from peripheral cones in humans.
 
@@
 Very briefly, we collected the raw fastq files and quantified with the kallisto / bustools system [@melstedBarcodeUMISet2019; @melstedModularEfficientPreprocessing2019].
-The quantified counts are merged into one matrix that is used for downstream processing with scvitools create a batch corrected lower dimensional space.
-and then into Seurat for quality control, clustering, and 2D UMAP visualization.
-We use our previous xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamyBuildingMegaSinglecell2021].
+The quantified counts are merged into one matrix that is used for the downstream integration with scvitools create a batch corrected lower dimensional space.
+The raw counts and scANVI batch corrected lower dimensional space is sent into Seurat for quality control, clustering, and 2D UMAP visualization.
+We use our previously published xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamyBuildingMegaSinglecell2021].
 
 The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300.
-Second, we use the DecontX algorithm[@yangDecontaminationAmbientRNA2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination.
+Second, we use the DecontX algorithm[@yangDecontaminationAmbientRNA2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination on a per-study basis.
 Third, we altered the procedure for aligning gene names across species (see below for details).
@@
 
-A custom cutoff for the minimum required number of unique genes identified per cell were used for the SRP362101 cornea dataset as an abnormally high number of cells were returned from our default value of 200.
+A custom cutoff for the minimum required number of unique genes identified per cell were used for the SRP362101 cornea dataset as an abnormally high number of cells were returned from our default value of 300.
 We set the cutoff for this studies to be 800 as this value resulted in approximately the same number of cells being returned as the authors reported in their papers.
 
+We found that while nearly all datasets we curated could be integrated as a whole, a small number had strange behavior in the two dimensional UMAP space and/or nearly all cells from a study were machine labelled as a single cell type.
+These studies were hand-removed from the resource.
+
 ## Cross species gene alignment
@@
 As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify, we must identify shared genes.
-We found homologs/orthologs by pulling from BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference.
-We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus Musculus*.
+We found homologs/orthologs by pulling from the BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference.
+We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus musculus*.
 This generates a table linked by Ensembl gene IDs and gene names.
@@
 macula), sex, and age or developmental stage.
+Samples with perturbations (genetic or treatment) were not included.
 To import individual cell type labels (e.g.
@@
 
-The scater::aggregateAcrossCells function was used to aggregate gene counts across species, study accession, and [category] where category was either Cluster, CellType, or CellType (Predict) assignments for a cell [@mccarthyScaterPreprocessingQuality2017].
+The aggregateAcrossCells function from the scuttle R package was used to aggregate gene counts across species, study accession, and category where category was either Cluster, CellType, or CellType (Predict) assignments for a cell [@mccarthyScaterPreprocessingQuality2017].
 If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test.
 Any genes with a sum of 0 counts after aggregation were removed.
-The aggregated matrix was imported into DESeq2 using the DESeqDataSetFromMatrix function with the design given as "~study_accession + [category]".
+The aggregated matrix was imported into DESeq2 using the DESeqDataSetFromMatrix function with the design given as "~study_accession + category".
 Contrasts were extracted either with the target category (e.g.
@@
 Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons [@SinglecellTranscriptomics20].
+As before, the pipeline's scVI's integration parameters were chosen rigorously by testing inegration performance across a wide range of latent dimensions and number of highly variable genes in our scPOP framework [@swamyBuildingMegaSinglecell2021].
+Each study in the resource were hand assessed to determine whether they had any unusual properties in regards to machine learnt cell type proportions or position in the 2D UMAP space.
+The small number of outlier studies were removed from the resource.
 
@@
 
-## scRNA web resources containing ocular data
+## PLAE resource contains several crucial features relative to other single cell resources
 
@@
 ```
-
-## PLAE differs from a complementary web app, Spectacle in several important ways
 

### 2022-08-26 src/new_fig1_test.R
tester fig

@@ -1,0 +1,98 @@
+study_count_by_tissue <- meta_filter %>%
+  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
+                           TRUE ~ 'Body'),
+         Tissue = gsub('\\.|_', ' ', Tissue),
+         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
+         Tissue = gsub(" and Aorta","", Tissue)) %>% 
+  #filter(Organ == 'Eye') %>% 
+  select(Organ, Tissue,organism, study_accession) %>%
+  unique() %>%
+  group_by(Organ, Tissue, organism) %>% 
+  summarise(Count = n()) %>%
+  ungroup() %>% 
+  ggplot(aes(x=Tissue,y=Count, fill = organism, label = Count, group = organism)) +
+  ylab('Number of Studies') +
+  geom_bar(stat='identity', position='stack') +
+  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5)) +
+  cowplot::theme_cowplot() +
+  coord_flip() +
+  theme(legend.position = "none") +
+  scale_y_continuous(expand = c(0, 0))  + xlab('') +
+  ggforce::facet_col(~Organ, scales = 'free_y', space = 'free')
+
+# retina cell types
+retina_ct <- meta_filter %>% filter(Organ == 'Eye') %>% 
+  pull(CellType) %>% unique()
+
+cell_counts <- meta_filter %>% filter(Organ == 'Eye', CellType_predict %in% retina_ct) %>% 
+  mutate(CellType_predict = case_when(is.na(CellType_predict) ~ 'Unlabelled', TRUE ~ CellType_predict)) %>% 
+  group_by(CellType_predict, organism) %>%
+  summarise(Count = n()) %>% 
+  mutate(sCount = log2(Count)) %>% 
+  ggplot(aes(x=CellType_predict,y=sCount, label = Count, fill = organism, group = organism)) +
+  ylab('log2(Number of Cells)') +
+  geom_bar(stat='identity', position='stack') +
+  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5)) + 
+  cowplot::theme_cowplot() +
+  coord_flip() + 
+  scale_y_continuous(expand = c(0, 0)) + xlab('')
+
+study_counts_by_cell <- meta_filter %>% filter(Organ == 'Eye', CellType_predict %in% retina_ct) %>% 
[…]

### 2022-08-26 manuscript_00.Rmd
reworked fig1

@@
 
-```{r scEiaD_data_overview, fig.height = 16, fig.width = 20, echo = F, message = F, fig.cap=scEiaD_data_overview_cap}
-
+```{r, fig.height=14, fig.width=12, fig.cap = scEiaD_data_overview_cap}
 num1 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(Tissue) %>% unique() %>% length()
@@
 num3 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique() %>% length()
+
 scEiaD_data_overview_cap <- fig_cap(name = 'scEiaD_data_overview', caption = glue::glue("Distribution of tissues and cell types in scEiaD.
+A.
+Number of cells for each ocular cell type.
+B.
+Number of studies for each ocular cell type (over 10 cells), colored by species.
+C.
+Tabulation of the number of studies present across {num1} ocular and {num2} non-ocular tissues, D.
+Web app performance (lower is better, missing functionality is shown as blank) for important functions compared between PLAE, Spectacle, and Tabula Sapiens (see methods)."))
+
+study_count_by_tissue <- meta_filter %>%
+  mutate(Organ = case_when(Organ == 'Eye' ~ 'Eye',
+                           TRUE ~ 'Body'),
+         Tissue = gsub('\\.|_', ' ', Tissue),
+         Tissue = gsub("RPE$","RPE-Choroid", Tissue),
+         Tissue = gsub(" and Aorta","", Tissue)) %>% 
+  #filter(Organ == 'Eye') %>% 
+  select(Organ, Tissue,organism, study_accession) %>%
+  unique() %>%
+  group_by(Organ, Tissue, organism) %>% 
+  summarise(Count = n()) %>%
+  ungroup() %>% 
+  ggplot(aes(x=Tissue,y=Count, fill = organism, label = Count, group = organism)) +
+  ylab('Number of Studies') +
+  geom_bar(stat='identity', position='stack') +
+  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5)) +
+  cowplot::theme_cowplot() +
+  coord_flip() +
+  theme(legend.position = "none") +
+  scale_y_continuous(expand = c(0, 0))  + xlab('') +
+  ggforce::facet_col(~Organ, scales = 'free_y', space = 'free')
+
+# retina cell types
+retina_ct <- meta_filter %>% filter(Organ == 'Eye') %>% 
+  pull(CellType) %>% unique()
+
+cell_counts <- meta_filter %>% filter(Organ == 'Eye', CellType_predict %in% retina_ct) %>% 
+  mutate(CellType_predict = case_when(is.na(CellType_predict) ~ 'Unlabelled', TRUE ~ CellType_predict)) %>% 
+  group_by(CellType_predict, organism) %>%
+  summarise(Count = n()) %>% 
+  mutate(sCount = log2(Count)) %>% 
+  ggplot(aes(x=CellType_predict,y=sCount, label = Count, fill = organism, group = organism)) +
+  ylab('log2(Number of Cells)') +
+  geom_bar(stat='identity', position='stack') +
+  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5)) + 
+  cowplot::theme_cowplot() +
+  coord_flip() + 
+  scale_y_continuous(expand = c(0, 0)) + xlab('')
+
+study_counts_by_cell <- meta_filter %>% filter(Organ == 'Eye', CellType_predict %in% retina_ct) %>% 
+  mutate(CellType_predict = case_when(is.na(CellType_predict) ~ 'Unlabelled', TRUE ~ CellType_predict)) %>% 
+  group_by(organism, CellType_predict, study_accession) %>%
+  summarise(Count = n()) %>% 
+  filter(Count > 10) %>% 
+  summarise(Count = n()) %>% 
+  mutate(Count = case_when(CellType_predict != 'Unlabelled' ~ Count)) %>% 
+  mutate(sCount = log2(Count)) %>% 
+  ggplot(aes(x=CellType_predict,y=Count, label = Count, fill = organism, group = organism)) +
+  ylab('Number of Studies') +
+  geom_bar(stat='identity', position='stack') +
+  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5)) + 
+  cowplot::theme_cowplot() +
+  coord_flip() + 
+  scale_y_continuous(expand = c(0, 0)) + xlab('')
+
+timings <- rbind(
+  c(1,7,1),
+  c(9,13,8),
+  c(3,NA,2),
+  c(2,NA,NA),
+  c(2,1,NA),
+  c(2,NA,NA),
+  c(2,1,NA),
+  c(4,41,NA)
+) %>% data.frame()
+
+row.names(timings) <- c('Initial Page',
+                        'Cell Type Colored UMAP',
+                        'Abca4 Expression UMAP',
+                        'Boxplot of Abca4 Expression By CellType Across All Studies',
+                        'Boxplot of Abca4 Expression By CellType Across One Study',
+                        'Dotplot of Abca4, Pax6, Nrl, and Atoh7',
+                        'Heatmap of Abca4, Pax6, Nrl, and Atoh7',
+                        'Table of Differentially Expressed Genes between Bipolar and Cone Cells'
+
+) %>% str_wrap(., width = 32)
+
+colnames(timings) <- c('PLAE',"Spectacle",'Tabula Sapiens\n(CELLxGENE)')
+
+breaks_fun <- function(x) {
+  if (max(x) > 30) {
+    seq(0,40,20)
+  } else if (max(x) > 9) {
+    seq(0, round(max(x)), 5)
+  } else if (max(x) >= 6) {
+    seq(0, round(max(x)), 2)
+  } else if (max(x) >= 3) {
+    seq(0, round(max(x)), 1)
+  } else {
+    seq(0,2,1)
+  }
+}
+
+timing_plot <- timings %>% as_tibble(rownames = 'Test') %>% 
+  pivot_longer(-Test) %>% 
+  mutate(Test = factor(Test, levels = row.names(timings))) %>% 
+  ggplot(aes(x=name,y=value)) + 
+  geom_bar(stat= 'identity') + 
+  facet_wrap(~Test, scales = 'free_y', ncol = 2) + 
+  xlab('Resource') + 
+  ylab('Seconds to Load') + 
+  cowplot::theme_cowplot() +
+  scale_y_continuous(breaks = breaks_fun)
+
+ab <- cowplot::plot_grid(cell_counts + theme(legend.position = "none"), 
+                         study_counts_by_cell, 
+                         nrow = 1, 
+                         rel_widths = c(1,1), labels = c('a','b'))
+
+cd <- cowplot::plot_grid(study_count_by_tissue, 
+                         timing_plot, 
+                         nrow = 1, 
+                         rel_widths = c(0.7,1),
+                         labels = c('c','d'))
+
+cowplot::plot_grid(ab, cd, ncol = 1, rel_heights = c(1,0.8))
+
+```
+
+```{r sup_scEiaD_data_overview, fig.height = 16, fig.width = 20, echo = F, message = F, fig.cap=sup_scEiaD_data_overview_cap}
+
+num1 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(Tissue) %>% unique() %>% length()
+num2 <- meta_filter %>% filter(Organ != 'Eye') %>% pull(Tissue) %>% unique() %>% length()
+num3 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique() %>% length()
+sup_scEiaD_data_overview_cap <- supFig_cap(name = '_supscEiaD_data_overview', caption = glue::glue("Distribution of tissues and cell types in scEiaD.
 A.

### 2022-08-26 manuscript_00.Rmd
fig update into doc

@@
 To generate the aggregated information (in the Dot Plot, Expression Plot, and tables) we pre-calculated all queries by running a dplyr "group by" operation on all column fields.
-On user given query, the data is further aggregated to level request.
+On user given query, the data is further aggregated to the user request.
 This allows for complicated queries to complete in seconds rather than minutes.
+To compare loading times for useful tasks we queried https://plae.nei.nih.gov, http://singlecell-eye.org, and http://tabula-sapiens-portal.ds.czbiohub.org, on 2022-08-24.
+For Spectacle, we used the "Single-Cell RNA-Seq Analysis of Retinal Development..." 10X dataset from Clark et al.
+for all timings.
+Timings are accurate to the second.
+If the functionality tested was not available for a certain web app, it was represented by a blank timing.
 
@@
 
-The codebase for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is XXXXXXX.
+The code base for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is XXXXXXX.
 There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app.
-The codebase for the web app (along with local installation instructins) is available at github.com/davemcg/scEiaD/plaeApp; the commit corresponding to this manuscript is XXXXXXXXXXXX.
+The code base for the web app (along with local installation instructions) is available at github.com/davemcg/scEiaD/plaeApp; the commit corresponding to this manuscript is XXXXXXXXXXXX.
 
@@
 
-After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (`r fig_cap(name = 'scEiaD_data_overview', display = 'cite')`).
+After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (`r fig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`).
+The scEiaD database contains `r num2` non-ocular human and mouse tissues and `r num1` ocular tissues across human, chicken, macaque, and mouse (`r fig_cap(name = 'scEiaD_data_overview', display = 'cite')`).
 This resource is only possible due to the publication of `r meta_filter$PMID %>% unique %>% length()` single cell resources that scEiaD draws from (`r supTab_cap(name = 'studies_table', display = 'cite')`).
@@
 
+```
+
+## PLAE resource contains several crucial features relative to other single cell resources
+
+Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources', display = 'cite')`).
+The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are not ocular focused, but do contain some ocular data.
+The PanglaoDB and UCSC are resources which have ingressed a large number of published single cell datasets The UCSC Cell Browser has around sixty thousand ocular single cell transcriptomes while the PangloaDB has about 47 thousand.
+However, these resources provide no curation of the metadata and process the data for each study individually.
+Thus each study is effectively silo-ed off from the rest and cross comparisons are impractical.
+The Tabula Sapiens project is an effort to sequence all cells in the mouse; while it only contains around around 10 thousand ocular cells it can be more readily cross compared to non-ocular tissues / cells.
+The Protein Atlas resource has recently used one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g.
+one can see whether a gene of interest is enriched in Cones).
+The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has collated a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app.
+PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence, and, crucially, full availability of the underlying data for outside usage and integrating the independent studies to deliver cross study comparisons.
+
+```{r web_resources, fig.height = 15, fig.width = 20, echo = F, message = F, results = 'asis', fig.cap=web_resources_cap}
+#tic()()
+i = 16.5 # width of the side borders in the word_document output (in centimeters)
+w = i*0.3937 # width of the side borders in the word_document output (in inches)
+
+web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
+
+cat("<div id=\"tbl:web_resources\">")
+
+feature_table <- rbind(
+  c('PLAE','Yes','Yes', 'Yes', '1136 ', 'https://plae.nei.nih.gov'),
+  c('Spectacle', 'Yes' ,'No', 'No', '1500', 'https://singlecell-eye.org'),
+  c('UCSC Cell Browser', 'Yes', 'No', 'Yes', '62', 'https://cells.ucsc.edu' ),
+  c('Tabula Sapiens', 'No', 'N/A', 'Yes', '10', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
+  c('PanglaoDB', 'Yes', 'No', 'Yes', '47', 'https://panglaodb.se'),
+  c('Protein Atlas', 'Yes*','No', 'Yes', '5',  'https://www.proteinatlas.org/humanproteome/celltype')) %>% 
+  as_tibble()
+colnames(feature_table) <- c('Resource Name', 'Independent\nDatasets', 'Harmonization', 'Ocular and\nnon-Ocular\n Data', 'Ocular Cell\nCount\n(thousands)', 'URL')
+feature_table %>% flextable() %>%  fontsize(size = 10, part = 'all') %>% autofit() %>% fit_to_width(7)#width(., width = c(1,0.75,0.75,0.75,0.75,3))
+# as of 2021 10 25
+#toc()()
+
+#Output Caption (* * for italics)
+cat(paste0("\n*",web_resources_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
+While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently[@voigtSpectacleInteractiveResource2020].
+PLAE is built around the scEiaD database, which use a multi-stage pipeline to build and identify a high performing scVI-based single cell RNA eye model that is used to integrate all datasets together [@swamyBuildingMegaSinglecell2021].
+This allows for gene queries to be constructed and analyzed across studies.
+In contrast, Spectacle, as of March 2022, is a compilation of study-level datasets and thus you cannot run queries across studies.
+Furthermore, Spectacle contains no non-ocular dataset and the cell types labels have not been curated or harmonized between studies.
+Finally, the Spectacle web app takes many seconds to load an individual dataset and many more to display the data.
+
+## Extensive optimization makes the PLAE web app highly responsive despite containing huge amounts of data
+
+There are two interwoven reasons why the PLAE app is quick, despite the huge amount of data it contains.
+First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables.
+A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage.
+The major downside is that this approach requires a large amount of disk storage, which can make local (on personal computer) usage difficult; the scEiaD v2022-03-22 sqlite database is about 368 GB.
+Second, the visualization code in PLAE (e.g.
+the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) to reduce drawing time.
+In contrast, starting from the loading page of Spectacle, it takes approximately eleven seconds to plot the cell type metadata for a several thousand cells across across a single study (tested on 2022-08-26).
+In the same amount of time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies (`r fig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')d`).
+The speed of the PLAE app is also comparable in several ways to the CELLXGENE browser while providing additional bar and dot plot functionality.
+
+## Rich web-available visualizations
+
+PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values.
+So thus a user can see Crx expression across over a million cells and separately see cell type labels for those same million cells.
+The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells.
+Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`a).
+
+For users interested in seeing how expression of a gene relates to study or other covariates, the Expression Plot view displays expression by cell type, predicted cell type, or cluster.
+The values can then be split and colored by multiple covariates like study or organism.
+This allows a user to see whether a gene pattern is consistent across studies or species.
+We also offer a Dot Plot view which can display a large number of genes in a space efficient manner.
+This allows a users to see whether a set of genes has related (or not) expression patterns across multiple cell types (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`b and c).
+
+While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina.
+The structure of the cartoon retina was based on the human retina.
+This unique visualization displays the major cell types of the retina and the RPE / choroid layer behind it.
+The cell types are colored by the relative amounts of expression in the user selected gene.
+Like all visualizations in PLAE, the user can use powerful filtering options to only display data from certain metadata characteristics, like species or publication (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`d).
+
+We have also added a Heatmap visualization which summaries the relative expression differences between user selected cell types or clusters across human, macaque, and mouse.
+This plot differs from the Dot Plot view as the expression values are relative instead of absolute.
+That is to say that while the Dot Plot view is proportionate to counts, the Heatmap view is showing the difference in expression between a given cell type (or cluster) and all other cells.
+This view is potentially useful if a user is especially interested in gene expression patterns across a variety of cell types.
+
+```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
+PLAE_screenshots_cap <- fig_cap(name='PLAE_screenshots', caption = 'Screenshots from PLAE app.
+A.
+UMAP view showing CRX expression with cell type labels, B.
+Information table showing expression of Crx across celltypes and organism.
+C.
+Box plot of selected genes in a user selected cell types across different studies.
+D.
+In silico in-situ of Rho.
+E.
+Dotplot demonstating efficient display of gene expression across many cell types.')
+knitr::include_graphics('fig/fig1.png')
+```
+
+## Leveraging the high diversity of studies to identify thousands of consistently differentially expressed genes across ocular cell types
+
+We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping (see Methods).
+This created a counts matrix which has statistical properties approximating "bulk" RNA seq experiment matrices.
+This allowed the use of the more mature bulk RNA-seq tooling.
+This approach has also been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments[@squairConfrontingFalseDiscoveries2021].
+Once the pseudo-bulk matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
+We tested each species (Human, Macaque, Mouse) separately.
+Chicken was not included as we had an insufficient number of independent studies to conduct differential testing.
+The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential gene expression.
+After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g.
+cones versus all remaining cell types) or in a pair-wise manner (cone against rod).
+Unless stated otherwise, we will be discussing the results of the "against all" contrast for the remainder of the paper.
+
+Across the cell types of the eye, we find `r diff_tab %>% nrow()` gene - celltype combinations that are significantly differentially expressed.
+We define significant as a log2 fold change greater than two as well as a p-value (corrected) less than $1 x 10^-5#.
+We find a mean of
+`r diff_tab_sup$"Homo sapiens" %>% mean() %>% round()`, 
+`r diff_tab_sup$"Mus musculus" %>% mean() %>% round()`, and 
+`r diff_tab_sup$"Macaca fascicularis" %>% mean() %>% round()` differentially expressed genes in human, macaque, and mouse respectively (`r supTab_cap(name = 'diff_tab_sup_tab', display = 'cite')`).
+These variable results directly reflect the differing number of independent studies - mouse and human have substantially more than macaque (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
+The significant filtered differential table is made available as supplementary file "differential_expression_table.csv.gz" and the full differential tables are available for download at plae.nei.nih.gov -> Data.
+
+## Well supported ocular cell types identify high confidence cell type markers
+
+Cell type marker genes can be proposed on the basis of existing knowledge about their function and on high expression differences when comparing to other cell types.
+We can further narrow down the list of markers by using the high diversity of studies and organisms in scEiaD to propose a set of community supported cell type markers.
+We first identify a set of well supported cell types which we define as cell types which have two or more independent studies across both human and mouse (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
+These cell types were then assessed to identify differentially expressed genes which met the following criteria across human, mouse, and macaque: 1.
+padj less than $1 x 10^-4$
+in two or more species, 2.
+mean log2 fold change greater than 2, and 3.
+mean padj less than $1 x 10^-5$.
+A final filter was applied to remove genes as candidates if they were differentially expressed (using the criteria above) in more than three different cell types.
+This left us with `r top_genes %>% nrow()` gene - cell type markers, a substantial reduction from the `r diff_tab %>% nrow()` we had initially.
+We provide a table of these genes as supplementary file "consistently_differentially_expressed_genes.csv.gz."
+
+To visualize these top markers we select up to eight genes (ordered by the mean log2 fold change across organisms) per well supported cell type for each and plot the log2 fold change of the gene expression relative to all other cell types (`r fig_cap(name = 'consist_genes_heatmap', display = 'cite')`).
+We see that human, mouse, and macaque have similar expression patterning of the proposed markers across the well supported cell types.
+The cell types are arranged by the columns by how related the expression patterns are.
+We see how the differentiated cell types of the retina are grouped together with the exception of the Müller glia, which are next to the RPC from which they derive from.
+The photoreceptors are also next to each other
+
+Of course other possibilities for ranking genes are possible.
+We show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`b and c how you can directly use our powerful web app to select candidate bipolar cell differentially expressed genes from the differential testing table and then quickly display the expression pattern of those genes in a compact heatmap visualization.
+The wealth of visualization options in plae allow for quick prototyping and hypothesis testing of gene expression patterning across a large number of cells and studies.
+
+```{r echo = F, message = F, }
+# tic()
+# ws_oct <- ws_table %>% filter(`Well Supported` == 'Yes') %>% pull(`CellType (Predict)`)
+# 
+# consist_diff <- scEiaD_2020_v01 %>%
+#   tbl("diff_testing") %>%
+#   filter(Base %in% ws_oct) %>%
+#   filter(Group == 'CellType (Predict)',
+#          Against == 'All',
+#          Organism %in% c('Homo sapiens','Mus musculus')) %>%
+#   group_by(Base, Gene) %>%
+#   summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-5), `mean padj` = mean(padj)) %>%
+#   filter(`mean log2FC` > 2.463, `sig count` > 1, `mean padj` < 1e-5) %>%
+#   collect() %>%
+#   mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>%
+#   left_join(gene_tab, by = 'GENEID') %>%
+#   relocate(SYMBOL, GENEID) %>%
+#   dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>%
+#   arrange(-`mean log2FC`)
+
+# # id diff genes only seen once across the cell types
+# uniq_g <- consist_diff %>% group_by(Symbol) %>% count() %>% filter(n == 1)
+#consist_diff %>% group_by(Base) %>% summarise(Count = n())
+#consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n())
+
+sup_tab_consist_diff <- top_genes %>% select(Symbol, Base) %>% left_join(consist_diff, by = c('Symbol','Base')) %>% group_by(Base) %>% summarise(Count = n()) %>% ggplot(aes(x=`Base`, y = `Count` )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip() + xlab("CellType (Predict)")
+#toc()
+```
+
+```{r consist_genes_heatmap, fig.height=22, fig.width=10, echo = F, message = F,  fig.cap=consist_genes_heatmap_cap}
+consist_genes_heatmap_cap <- fig_cap(name='consist_genes_heatmap', caption = 'Rows are genes, columns are cell types and organisms.
+Rows are split by what cell type the gene is a marker for.
+Columns are split by the cell types.
+More red is a higher log2 fold change of the gene - cell type relative to all other cell types and blue is more negative.')
+
+# tic()
+# g <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% slice_max(n=8, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
+# 
+# consist_diff_long <- scEiaD_2020_v01 %>%
+#   tbl("diff_testing") %>%
+#   filter(Gene %in% g) %>%
+#   collect() %>%
+#   filter(Base %in% ws_oct,
+#          Group == 'CellType (Predict)',
+#          Against == 'All') %>%
+#   select(Gene, Organism, log2FoldChange, Base)
+# 
+# consist_diff_wide <- consist_diff_long %>% 
+#   mutate(GO = paste0(Base, '_', Organism)) %>% 
+#   select(-Organism, -Base) %>% 
+#   pivot_wider(values_from=log2FoldChange, names_from = GO)
+# consist_diff_celltype <- colnames(consist_diff_wide[,-1])
+# consist_diff_wide <- consist_diff_wide %>% data.frame()
+# row.names(consist_diff_wide) <- consist_diff_wide$Gene
+# 
+# consist_diff_wide <- consist_diff_wide[,-1]
+# 
+# 
+# 
+# consist_diff_organism <- str_extract(consist_diff_celltype, '_\\w+ \\w+') %>% gsub('_','',.)
+# consist_diff_cts  <- str_extract(consist_diff_celltype, '.*_') %>% gsub('_','',.)
+# consist_diff_genes <- row.names(consist_diff_wide) %>% gsub(' \\(.*','',.)
+# 
+# consist_diff_base <- consist_diff_genes %>% as_tibble() %>% rename(Symbol = value) %>% left_join(consist_diff %>% select(Symbol, Base) %>% unique()) %>% pull(Base)
+
+consist_diff_hm <- ComplexHeatmap::Heatmap(consist_diff_wide, column_labels = consist_diff_organism, 
+                                           column_split = consist_diff_cts, 
+                                           row_split = consist_diff_base, show_row_dend = FALSE, 
+                                           row_labels = consist_diff_genes, 
+                                           column_title_rot = 90, 
+                                           name = 'log2FC', use_raster = TRUE,
+                                           row_title_rot = 0)
+ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(0, 0, 1.5, 0), "in"), row_title = "Marker Genes")
+#toc()
+```
+
+## Tease apart the similar neurogenic and RPC cell types
+
+```{r neuro, fig.width=30, fig.height=8, echo = F, message = F, fig.cap=neuro_cap}
+neuro_cap <- fig_cap(name='neuro', caption = 'A.
+Genes proposed by Lu et al.
+as being significantly differentially expressed during the RPC to neurogenic differentiation in human.
+We plot the differential expression (log2(fold change) directly between RPC and neurogenic cells.
+The standard error (calculated by DESeq2) is plotted.
+Genes with a padj < 0.1 and padj < 0.01 are labelled with "*" and "**", respectively.
+B.
+Genes from Lu et al.
+that we replicated with the scEiaD data.
+C.
+Novel genes we propose as being differentially expressed between RPC and neurogenic cells')
+# tic()
+# pairwise_neurogenic_test<-  scEiaD_2020_v01 %>% 
+#   tbl("diff_testing") %>% 
+#   filter(Against  == "RPC",   
+#          Group == 'CellType (Predict)', 
+#          Base == "Neurogenic Cell") %>% 
+#   group_by(Base, Against, Gene) %>% 
+#   summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
+#   arrange(-`mean log2FC`) %>%   
+#   collect() %>% 
+#   filter(!is.na(`mean log2FC`))
+# 
+# 
+# # mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
+# # left_join(gene_tab, by = 'GENEID') %>% 
+# # relocate(SYMBOL, GENEID) %>% 
+# # dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
+# # arrange(-`mean log2FC`)
+#
+
+lu_et_al <- c("PTTG1", "SCG3", "SHD", "SOX11", "ONECUT2", "VSX1", "ASCL1", "TCF4", "SOX4", "RORB", "GADD45G", "MYBL1", "MFAP4", "HES6", "NEUROD1", "C8orf46", "OTX2", "OLIG2", "INSM1", "PCBP4", "NEUROD4", "FAM131C", "HES4", "HES5", "HMGA1", "YBX1", "GADD45A", "ATOH7", "DLX2", "DLX1", "SFRP1", "MYB", "NFIB", "SOX9", "ZFP36L1", "FOS", "JUN", "ZFP36L2", "SOX2", "HES1", "EGR1", "JUNB", "NR4A1", "ETV5", "FOXP1", "ID1", "ID2", "ID3", "ID4", "FOXP2")
+
+g_lu <- scEiaD_2020_v01 %>% tbl('genes') %>% collect() %>% filter(grepl(paste(paste0('^',lu_et_al), collapse = ' |'), Gene)) %>% pull(Gene)
+
+down <- pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` < 0, `mean padj` < 0.05) %>% mutate(d = `mean log2FC` - lfcSE) %>% arrange(-d) %>% pull(Gene)
+up <-  pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` > 2, `mean padj` < 0.05) %>% mutate(d = `mean log2FC` - lfcSE) %>% arrange(-d) %>% pull(Gene)
+
+source('src/make_exp_plot.R')
+
+input <- list()
+input$exp_plot_groups <- c('study_accession')
+input$exp_plot_facet <- c('CellType_predict', 'organism')
+input$exp_plot_genes <- c(down, up[!up %in% g_lu])
+input$exp_filter_cat <- 'CellType_predict'
+#input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
+input$exp_filter_on <- c("Neurogenic Cell", "RPC" )
+input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
+#input$exp_plot_ylab <- '% of Cells Detected'
+#input$exp_plot_groups <- c('organism')
+input$exp_plot_col_num <- 16
+input$exp_filter_min_cell_number <- 25
+input$flip_facet <- FALSE
+input$figure <- TRUE
+
+exp_plot_new <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+
+input$exp_plot_genes <- up[up %in% g_lu]
+exp_plot_existing <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+a_legend <- get_legend(exp_plot_new)
+neuro_plot <- plot_grid(plot_grid(exp_plot_existing + theme(legend.position="none") + xlab(''),
+                                  exp_plot_new + theme(legend.position="none") + ylab('') + xlab(''), 
+                                  labels = c('b','c'),
+                                  rel_widths = c(0.6,1)),
+                        a_legend, ncol = 1, rel_heights = c(1,0.2))
+
+lu_long <-  scEiaD_2020_v01 %>% 
+  tbl("diff_testing") %>% 
+  filter(Gene %in% g_lu,
+         Against  == "RPC",   
+         Group == 'CellType (Predict)', 
+         Base == "Neurogenic Cell") %>% 
+  #group_by(Base, Against, Gene) %>% 
+  #summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
+  #arrange(-`mean log2FC`) %>%   
+  collect() %>% 
+  mutate(Gene = str_extract(Gene, '\\w+ ') %>% 
+           gsub(' ', '', .)) %>% 
+  mutate(Gene = factor(Gene, levels = rev(lu_et_al)))
+
+#order <- lu_et_al %>% enframe() %>% mutate(Gene = value) %>% left_join(g_lu %>% enframe() %>% mutate(full = value, Gene = str_extract(value, '\\w+ ') %>% gsub(' ', '', .)) %>% select(-value), by = 'Gene') %>% data.frame() %>% filter(!is.na(full)) %>% pull(full)
+
+lu_rep_plot <- lu_long %>% 
+  mutate(Significance = case_when(padj < 0.01 ~ '**', padj < 0.1 ~ '*', TRUE ~ '')) %>% 
+  ggplot(aes(x=Gene,y=log2FoldChange,  label = Significance)) + 
+  geom_bar(stat='identity', position = position_dodge2(), fill = 'gray') + 
+  geom_errorbar(aes(ymin = log2FoldChange - lfcSE,
+                    ymax = log2FoldChange + lfcSE), width = 0.5,
+                position = position_dodge2()) +
+  ylab('Neurogenic / RPC Gene Expression (log2 Fold Change)') +
+  coord_flip() + 
+  cowplot::theme_cowplot()  + geom_text(aes(y=7), position = position_dodge2(width = 1)) + facet_wrap(~Organism)
+#g_lu[!g_lu %in% lu_long$Gene]
+
+cowplot::plot_grid(lu_rep_plot, neuro_plot, rel_widths = c(0.6,1), labels = c('a'))
+```
+
+While the well supported markers we propose clearly define most of the cell types, we see that the neurogenic cells and RPCs have very similar expression patterns across the gene sets.
+These cell types are closely related as the neurogenic cells are derived from the multi-potent RPC to later specify neural cell types.
+We use the set of genes identified by Lu et al.
+[@luSingleCellAnalysisHuman2020b] in their study of human retinal differentiation to determine whether our scEiaD database retains the same trends (`r fig_cap(name = 'neuro', display = 'cite')`A).
+As expected, we see several patterns.
+First, we see most of the gene expression changes between RPC and neurogenic are matched between human.
+We further show that these genes also show the same trends in mouse datasets.
+Next we see that the direction of the changes largely matches Lu et al (`r supFig_cap(name='lu_et_al', display = 'cite')`).
+Finally, we see a small number of genes are confirmed in our database as consistently differentially expressed with a padj < 0.1.
+
+To determine whether we can propose any new genes that distinguish RPC and neurogenic states we use our DESeq2 differential expression contrast that directly compares these two cell types.
+We find `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Homo sapiens') %>% nrow()` human genes and `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Mus musculus') %>% nrow()` mouse genes that have an absolute log2 fold change greater than two and a adjusted p-value below 0.05 (`r supTab_cap(name = 'neuro_table', display = 'cite')`).
+To identify candidate genes that are consistently differentially expressed at this transitional state in mouse and human we apply similar logic that we used in identifying well supported marker genes across the retina cell types.
+We filter to genes with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two or less than negative one (`r fig_cap(name = 'neuro', display = 'cite')`B and C).
+We find one gene, ZBTB38, that meets this criteria and drops in expression when comparing RPC to neurogenic.
+We find `r up %>% length()` genes that increase in expression from RPC to neurogenic, `r up[!up %in% g_lu] %>% length()` of which are not previously identified by Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`C).
+We replicate `r up[up %in% g_lu] %>% str_extract(., '\\w+') %>% paste(., collapse = ', ') %>% str_to_title()` from Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`B) as differentially expressed between neurogenic and RPC.
+Futhermore we show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`a how this figure can be replicated directly in our web app.
+
+## Our pan-study macula vs peripheral cone test replicates a published macaque finding and proposes six new markers
+
+```{r region_fig, fig.width=18, fig.height=3,  echo = F, message = F, fig.cap = region_fig_cap}
+region_fig_cap <- fig_cap(name='regino_fig', caption = 'Plot of markers distinguishing human macula or peripheral punch derived cones.')
+
+source('src/make_exp_plot.R')
+#load('data/pseudoBulk_cone_region_files.Rdata')
+region_table <- read_csv('supplemental_files/pseudoBulk_cone_region_table.csv.gz')
+input <- list()
+input$exp_plot_groups <- c('study_accession')
+input$exp_plot_facet <- c('CellType_predict', 'organism', 'retina_region')
+input$exp_plot_genes <- c('PRPH2 (ENSG00000112619)')
+input$exp_filter_cat <- c('CellType_predict', 'organism')
+#input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
+input$exp_filter_on <- c("Cone", "Homo sapiens" )
+input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
+#input$exp_plot_ylab <- '% of Cells Detected'
+#input$exp_plot_groups <- c('organism')
+input$exp_plot_col_num <- 16
+input$exp_filter_min_cell_number <- 10
+input$flip_facet <- FALSE
+input$figure <- FALSE
+input$figure_fovea <- TRUE
+
+prph2 <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') +  theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() ) 
+input$exp_plot_genes <-  'RS1 (ENSG00000102104)';rs1 <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') +  theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+
+input$exp_plot_genes <-  'GNGT1 (ENSG00000127928)';gngt1 <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') +  theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+#input$exp_plot_genes <-  'RS1 (ENSG00000102104)';rs1 <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+
+diff_region_cone_genes <- region_table %>% filter(padj < 0.01) %>% 
+  arrange(log2FoldChange) %>% 
+  mutate(id = paste0(Gene, ' (', Ensembl, ')')) %>% 
+  pull(id)
+input$exp_filter_on <- c("Cone", "Homo sapiens" )
+input$exp_plot_genes <- diff_region_cone_genes[1]; a <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') +  theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[2]; b <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[3]; c <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[4]; d <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + ylab('') + xlab('') + theme(axis.ticks.x=element_blank(), axis.text.x=element_blank() ) 
+input$exp_plot_genes <- diff_region_cone_genes[5]; e <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[6]; f <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+input$exp_plot_genes <- diff_region_cone_genes[7]; g <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
+
+region_legend <- get_legend(d)
+
+plot_grid(plot_grid(a,b,c,d + theme(legend.position="none"),e,f,g, nrow = 1),
+          region_legend, 
+          rel_heights = c(1,0.2), 
+          nrow = 2)
+```
+
+While we have provided data in many forms via the web app at plae.nei.nih.gov, there are many more dissections of the data that can be performed.
+To facilitate custom testing with the scEiaD database we provide, as an example, a human-specific macula versus peripheral cone differential testing analysis document as supplementary file "pseudobulk_cone_region.freeze01.html".
+In our analysis document we use the Seurat object containing the full scEiaD dataset we provide at plae.nei.nih.gov to build a custom pseudobulk counts matrix containing only cones from human studies that were taken in a region specific manner.
+This left us with 26678 cells across six studies.
+Yan et al.
+previously proposed Prph2 and Rs1 as markers to distinguish macula and peripheral located human cones[@yanCellAtlasHuman2020a] and Peng et al.
+proposed Calb1 and Gngt1 as cone region markers in macaque [@pengMolecularClassificationComparative2019a].
+
+Our differential testing found that while Prph2, Rs1, and Gngt1 had expression differences matching that the findings Yan et al.
+and Peng et al.
+published, they failed to replicate in our test with pvalue and padj being above 0.1 for all three.
+In contrast Calb1 did replicate strongly with the sixth lowest padj (0.00721) across our entire test and a log2 fold change of 2.87.
+We find six more genes with a padj < 0.01: Pde6g, Akap12, Hspb1, Gadd45b, Cc2d2a, and RP1l1 (`r fig_cap(name = 'region_fig', display = 'cite')` and `r supTab_cap(name = 'region_table', display = 'cite')`).
+None of these candidate cone region marker genes display regionally different expression in rods (`r supFig_cap(name = 'rod_region_fig', display = 'cite')`).
+
+# Conclusions
+
+## Limitations of scEiaD
+
+The longer term goal is to continue adding datasets to the scEiaD at plae.nei.nih.gov until we have three or more independent datasets per species and nearly all cell types.
+Earlier we proposed well supported cell types.
+Right now, most of these well supported cell types are in the retina.
+The RPE behind the retina, the cornea and lens in front, and the outflow tract and supporting musculature still could use more independent studies.
+Many common human diseases underlie the RPE (e.g.
+AMD) and the cornea (e.g.
+Keratoconus) and these tissues will remain a focus for the next update of our database.
+While we have curated a large number of cell types across the eye, many of these cell types have "sub" cell types.
+For examplethe human has three types of cones that are optimized for short, medium, and long wavelengths.
+More dramatically, Yan et al.
+have proposed over sixty types of mouse amacrine cell types [@yanMouseRetinalCell2020a].
+While we have found it fairly straightforward to use machine learning to transfer the broad (e.g.
+rod, cone, amacrine) cell type labels across all datasets, we have found it very challenging to transfer the "sub" cell type labels.
+This is largely because while we have high diversity of studies for the broad cell types, this is not true for most of the sub celltypes.
+We believe more datasets that carefully dissect the sub-celltypes may be necessary to confidently propose the sub-celltypes of the retina.
+Finally, while our scVI based model can theoretically be used by an outside group to transfer their internal data into the scEiaD latent space, it is very challenging to implement as the the software landscape is still rapidly changing and it is very difficult to architect consistent outputs over time with regularly changing software versions.
+We are investigating simpler methods to enable transfer of knowledge from scEiaD to outside data.
+
+## High feature web app provides quick access to over one million transcriptomes across `r meta_filter$study_accession %>% unique %>% length` datasets
+
+We have improved on the scEiaD dataset we published in Swamy et al.
+[@swamyBuildingMegaSinglecell2021] by adding another species (chicken), new ocular tissues, a human non-ocular reference set; in total `r meta_filter$Citation %>% unique() %>% length() - meta_filter_v0$Citation %>% unique() %>% length()` new single cell datasets.
+We re-structured the data into a custom database to enable our custom built high performance web app to quickly access the data.
+This democratizes access to over a million cells across `r meta_filter$study_accession %>% unique() %>% length()` studies to anyone with a web browser.
+We recognize that while we have built a powerful visualization platform we cannot anticipate all uses and for those researchers who wish to make custom queries of the data, we provide the data underlying plae in a wide variety of formats, including anndata and Seurat objects for each study, count tables, cell level metadata, and the DESeq2 based differential testing across three species.
+We provide a short example of how a custom query could work by running a custom differential test between macula and peripheral derived human cones and demonstrate how this can both validate existing proposed genes and suggest new ones.
+
+## High diversity of studies enables reproduction and extension of cell type gene expression findings
+
+A great many groups have proposed cell atlases.
+These range from efforts focusing on a subset of cells from a single layer of a tissue (for example [@tranSingleCellProfilesRetinal2019a] ) to a huge consortium encompassing multiple organs [@consortiumTabulaSapiensSingle2021].
+We propose a complementary approach wherein we use the research data from the community at large to create a meta-atlas.
+This approach enables a small group to produce an atlas which rivals the size from massive multi-group consortiums.
+Furthermore the inherent variability between the data from independent groups can be used as a feature to get insight into whether a gene cell type signal is stable across an entire community's datasets.
+We briefly demonstrate how the high study diversity in scEiaD can be used to validate proposed RPC to neurogenic gene expression changes.
+
+# Supplementary Figures and Tables
+
+```{r decontX, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=decontX_cap}
+decontX_cap <- supFig_cap(name='decontX', caption = 'DecontX *in silico* ambient RNA contamination tool substantially removes Rhodopsin expression in non-rod cells while retaining high expression rods.
+First shown (a) is Rhodopsin expression across the scEiaD v0 dataset, without DecontX optimization.
+Note Rhodopsin expression is noticeable in many cell types beyond the rods.
+Next (b) is the scEiaD v1 dataset with DecontX optimization.
+Rhodopsin is nearly exclusively expressed in labelled rod cells.
+Despite the computational ambient RNA removal, Rhodopsin expression remains high in the Rods.')
+source('src/make_exp_plot.R')
+
+input <- list()
+input$exp_plot_groups <- c('organism')
+input$exp_plot_facet <- c('CellType_predict')
+input$exp_plot_genes <- c('RHO (ENSG00000163914)')
+input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
+input$exp_plot_col_num <- 16
+input$exp_filter_min_cell_number <- 50
+input$flip_facet <- FALSE
+input$figure <- FALSE
+input$figure_fovea <- FALSE
+
+v0_rho <- make_exp_plot(input, scEiaD_2020_v00, meta_filter_v0)
+v1_rho <-  make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+input$exp_filter_min_cell_number <- 1000
+v1.1_rho <- make_exp_plot(input, scEiaD_2020_v01, meta_filter %>% filter(CellType_predict %in% unique(meta_filter_v0$CellType_predict)))
+plot_grid(v0_rho, v1_rho, ncol = 1, labels = c('a','b'), rel_heights = c(1,1.5))
 ```
@@
 
-## PLAE resource contains several crucial features relative to other single cell resources
-
-Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources', display = 'cite')`).
-The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are not ocular focused, but do contain some ocular data.
-The PanglaoDB and UCSC are resources which have ingressed a large number of published single cell datasets The UCSC Cell Browser has around sixty thousand ocular single cell transcriptomes while the PangloaDB has about 47 thousand.
-However, these resources provide no curation of the metadata and process the data for each study individually.
-Thus each study is effectively silo-ed off from the rest and cross comparisons are impractical.
-The Tabula Sapiens project is an effort to sequence all cells in the mouse; while it only contains around around 10 thousand ocular cells it can be more readily cross compared to non-ocular tissues / cells.
-The Protein Atlas resource has recently used one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g.
-one can see whether a gene of interest is enriched in Cones).
-The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has collated a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app.
-PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence, and, crucially, full availability of the underlying data for outside usage and integrating the independent studies to deliver cross study comparisons.
-
-```{r web_resources, fig.height = 15, fig.width = 20, echo = F, message = F, results = 'asis', fig.cap=web_resources_cap}
-#tic()()
-i = 16.5 # width of the side borders in the word_document output (in centimeters)
-w = i*0.3937 # width of the side borders in the word_document output (in inches)
-
-web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
-
-cat("<div id=\"tbl:web_resources\">")
-
-feature_table <- rbind(
-  c('PLAE','Yes','Yes', 'Yes', '1136 ', 'https://plae.nei.nih.gov'),
-  c('Spectacle', 'Yes' ,'No', 'No', '1500', 'https://singlecell-eye.org'),
-  c('UCSC Cell Browser', 'Yes', 'No', 'Yes', '62', 'https://cells.ucsc.edu' ),
-  c('Tabula Sapiens', 'No', 'N/A', 'Yes', '10', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
-  c('PanglaoDB', 'Yes', 'No', 'Yes', '47', 'https://panglaodb.se'),
-  c('Protein Atlas', 'Yes*','No', 'Yes', '5',  'https://www.proteinatlas.org/humanproteome/celltype')) %>% 
-  as_tibble()
-colnames(feature_table) <- c('Resource Name', 'Independent\nDatasets', 'Harmonization', 'Ocular and\nnon-Ocular\n Data', 'Ocular Cell\nCount\n(thousands)', 'URL')
-feature_table %>% flextable() %>%  fontsize(size = 10, part = 'all') %>% autofit() %>% fit_to_width(7)#width(., width = c(1,0.75,0.75,0.75,0.75,3))
-# as of 2021 10 25
-#toc()()
-
-#Output Caption (* * for italics)
-cat(paste0("\n*",web_resources_cap,"*\n"))
-# Output </div>
-cat("</div>")
-```
-
-While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured differently[@voigtSpectacleInteractiveResource2020].
-PLAE is built around the scEiaD database, which use a multi-stage pipeline to build and identify a high performing scVI-based single cell RNA eye model that is used to integrate all datasets together [@swamyBuildingMegaSinglecell2021].
-This allows for gene queries to be constructed and analyzed across studies.
-In contrast, Spectacle, as of March 2022, is a compilation of study-level datasets and thus you cannot run queries across studies.
-Furthermore, Spectacle contains no non-ocular dataset and the cell types labels have not been curated or harmonized between studies.
-Finally, the Spectacle web app takes many seconds to load an individual dataset and many more to display the data.
-
-## Extensive optimization makes the PLAE web app highly responsive despite containing huge amounts of data
-
-There are two interwoven reasons why the PLAE app is quick, despite the huge amount of data it contains.
-First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables.
-A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage.
-The major downside is that this approach requires a large amount of disk storage, which makes local (on personal computer) usage onerous; the scEiaD v2022-03-22 sqlite database is about 368 GB.
-Second, the visualization code in PLAE (e.g.
-the UMAP view) pull from the custom database directly and use a highly optimized plotting system (scattermore) to reduce drawing time.
-Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study (tested on 2022-03-23).
-In the same amount of time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies.
-
-## Rich web-available visualizations
-
-PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values.
-So thus a user can see Crx expression across over a million cells and separately see cell type labels for those same million cells.
-The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells.
-Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`a).
-
-For users interested in seeing how expression of a gene relates to study or other covariates, the Expression Plot view displays expression by cell type, predicted cell type, or cluster.
-The values can then be split and colored by multiple covariates like study or organism.
-This allows a user to see whether a gene pattern is consistent across studies or species.
-We also offer a Dot Plot view which can display a large number of genes in a space efficient manner.
-This allows a users to see whether a set of genes has related (or not) expression patterns across multiple cell types (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`b and c).
-
-While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina.
-The structure of the cartoon retina was based on the human retina.
-This unique visualization displays the major cell types of the retina and the RPE / choroid layer behind it.
-The cell types are colored by the relative amounts of expression in the user selected gene.
-Like all visualizations in PLAE, the user can use powerful filtering options to only display data from certain metadata characteristics, like species or publication (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`d).
-
-We have also added a Heatmap visualization which summaries the relative expression differences between user selected cell types or clusters across human, macaque, and mouse.
-This plot differs from the Dot Plot view as the expression values are relative instead of absolute.
-That is to say that while the Dot Plot view is proportionate to counts, the Heatmap view is showing the difference in expression between a given cell type (or cluster) and all other cells.
-This view is potentially useful if a user is especially interested in gene expression patterns across a variety of cell types.
-
-```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
-PLAE_screenshots_cap <- fig_cap(name='PLAE_screenshots', caption = 'Screenshots from PLAE app.
-A.
-UMAP view showing CRX expression with cell type labels, B.
-Information table showing expression of Crx across celltypes and organism.
-C.
-Box plot of selected genes in a user selected cell types across different studies.
-D.
-In silico in-situ of Rho.
-E.
-Dotplot demonstating efficient display of gene expression across many cell types.')
-knitr::include_graphics('fig/fig1.png')
-```
-
-## Leveraging the high diversity of studies to identify thousands of consistently differentially expressed genes across ocular cell types
-
-We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping (see Methods).
-This created a counts matrix which has statistical properties approximating "bulk" RNA seq experiment matrices.
-This allowed the use of the more mature bulk RNA-seq tooling.
-This approach has also been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments[@squairConfrontingFalseDiscoveries2021].
-Once the pseudo-bulk matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
-We tested each species (Human, Macaque, Mouse) separately.
-Chicken was not included as we had an insufficient number of independent studies to conduct differential testing.
-The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential gene expression.
-After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g.
-cones versus all remaining cell types) or in a pair-wise manner (cone against rod).
-Unless stated otherwise, we will be discussing the results of the "against all" contrast for the remainder of the paper.
-
-Across the cell types of the eye, we find `r diff_tab %>% nrow()` gene - celltype combinations that are significantly differentially expressed.
-We define significant as a log2 fold change greater than two as well as a p-value (corrected) less than $1 x 10^-5#.
-We find a mean of
-`r diff_tab_sup$"Homo sapiens" %>% mean() %>% round()`, 
-`r diff_tab_sup$"Mus musculus" %>% mean() %>% round()`, and 
-`r diff_tab_sup$"Macaca fascicularis" %>% mean() %>% round()` differentially expressed genes in human, macaque, and mouse respectively (`r supTab_cap(name = 'diff_tab_sup_tab', display = 'cite')`).
-These variable results directly reflect the differing number of independent studies - mouse and human have substantially more than macaque (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
-The significant filtered differential table is made available as supplementary file "differential_expression_table.csv.gz" and the full differential tables are available for download at plae.nei.nih.gov -> Data.
-
-## Well supported ocular cell types identify high confidence cell type markers
-
-Cell type marker genes can be proposed on the basis of existing knowledge about their function and on high expression differences when comparing to other cell types.
-We can further narrow down the list of markers by using the high diversity of studies and organisms in scEiaD to propose a set of community supported cell type markers.
-We first identify a set of well supported cell types which we define as cell types which have two or more independent studies across both human and mouse (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
-These cell types were then assessed to identify differentially expressed genes which met the following criteria across human, mouse, and macaque: 1.
-padj less than $1 x 10^-4$
-in two or more species, 2.
-mean log2 fold change greater than 2, and 3.
-mean padj less than $1 x 10^-5$.
-A final filter was applied to remove genes as candidates if they were differentially expressed (using the criteria above) in more than three different cell types.
-This left us with `r top_genes %>% nrow()` gene - cell type markers, a substantial reduction from the `r diff_tab %>% nrow()` we had initially.
-We provide a table of these genes as supplementary file "consistently_differentially_expressed_genes.csv.gz."
-
-To visualize these top markers we select up to eight genes (ordered by the mean log2 fold change across organisms) per well supported cell type for each and plot the log2 fold change of the gene expression relative to all other cell types (`r fig_cap(name = 'consist_genes_heatmap', display = 'cite')`).
-We see that human, mouse, and macaque have similar expression patterning of the proposed markers across the well supported cell types.
-The cell types are arranged by the columns by how related the expression patterns are.
-We see how the differentiated cell types of the retina are grouped together with the exception of the Müller glia, which are next to the RPC from which they derive from.
-The photoreceptors are also next to each other
-
-Of course other possibilities for ranking genes are possible.
-We show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`b and c how you can directly use our powerful web app to select candidate bipolar cell differentially expressed genes from the differential testing table and then quickly display the expression pattern of those genes in a compact heatmap visualization.
-The wealth of visualization options in plae allow for quick prototyping and hypothesis testing of gene expression patterning across a large number of cells and studies.
-
-```{r echo = F, message = F, }
-# tic()
-# ws_oct <- ws_table %>% filter(`Well Supported` == 'Yes') %>% pull(`CellType (Predict)`)
-# 
-# consist_diff <- scEiaD_2020_v01 %>%
-#   tbl("diff_testing") %>%
-#   filter(Base %in% ws_oct) %>%
-#   filter(Group == 'CellType (Predict)',
-#          Against == 'All',
-#          Organism %in% c('Homo sapiens','Mus musculus')) %>%
-#   group_by(Base, Gene) %>%
-#   summarise(`mean log2FC` = mean(log2FoldChange), `sig count` = sum(padj < 1e-5), `mean padj` = mean(padj)) %>%
-#   filter(`mean log2FC` > 2.463, `sig count` > 1, `mean padj` < 1e-5) %>%
-#   collect() %>%
-#   mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>%
-#   left_join(gene_tab, by = 'GENEID') %>%
-#   relocate(SYMBOL, GENEID) %>%
-#   dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>%
-#   arrange(-`mean log2FC`)
-
-# # id diff genes only seen once across the cell types
-# uniq_g <- consist_diff %>% group_by(Symbol) %>% count() %>% filter(n == 1)
-#consist_diff %>% group_by(Base) %>% summarise(Count = n())
-#consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% summarise(Count = n())
-
-sup_tab_consist_diff <- top_genes %>% select(Symbol, Base) %>% left_join(consist_diff, by = c('Symbol','Base')) %>% group_by(Base) %>% summarise(Count = n()) %>% ggplot(aes(x=`Base`, y = `Count` )) + geom_bar(stat='identity', position = 'dodge') + cowplot::theme_cowplot() + coord_flip() + xlab("CellType (Predict)")
-#toc()
-```
-
-```{r consist_genes_heatmap, fig.height=22, fig.width=10, echo = F, message = F,  fig.cap=consist_genes_heatmap_cap}
-consist_genes_heatmap_cap <- fig_cap(name='consist_genes_heatmap', caption = 'Rows are genes, columns are cell types and organisms.
-Rows are split by what cell type the gene is a marker for.
-Columns are split by the cell types.
-More red is a higher log2 fold change of the gene - cell type relative to all other cell types and blue is more negative.')
-
-# tic()
-# g <- consist_diff %>% filter(Symbol %in% uniq_g$Symbol) %>% group_by(Base) %>% slice_max(n=8, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
-# 
-# consist_diff_long <- scEiaD_2020_v01 %>%
-#   tbl("diff_testing") %>%
-#   filter(Gene %in% g) %>%
-#   collect() %>%
-#   filter(Base %in% ws_oct,
-#          Group == 'CellType (Predict)',
-#          Against == 'All') %>%
-#   select(Gene, Organism, log2FoldChange, Base)
-# 
-# consist_diff_wide <- consist_diff_long %>% 
-#   mutate(GO = paste0(Base, '_', Organism)) %>% 
-#   select(-Organism, -Base) %>% 
-#   pivot_wider(values_from=log2FoldChange, names_from = GO)
-# consist_diff_celltype <- colnames(consist_diff_wide[,-1])
-# consist_diff_wide <- consist_diff_wide %>% data.frame()
-# row.names(consist_diff_wide) <- consist_diff_wide$Gene
-# 
-# consist_diff_wide <- consist_diff_wide[,-1]
-# 
-# 
-# 
-# consist_diff_organism <- str_extract(consist_diff_celltype, '_\\w+ \\w+') %>% gsub('_','',.)
-# consist_diff_cts  <- str_extract(consist_diff_celltype, '.*_') %>% gsub('_','',.)
-# consist_diff_genes <- row.names(consist_diff_wide) %>% gsub(' \\(.*','',.)
-# 
-# consist_diff_base <- consist_diff_genes %>% as_tibble() %>% rename(Symbol = value) %>% left_join(consist_diff %>% select(Symbol, Base) %>% unique()) %>% pull(Base)
-
-consist_diff_hm <- ComplexHeatmap::Heatmap(consist_diff_wide, column_labels = consist_diff_organism, 
-                                           column_split = consist_diff_cts, 
-                                           row_split = consist_diff_base, show_row_dend = FALSE, 
-                                           row_labels = consist_diff_genes, 
-                                           column_title_rot = 90, 
-                                           name = 'log2FC', use_raster = TRUE,
-                                           row_title_rot = 0)
-ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(0, 0, 1.5, 0), "in"), row_title = "Marker Genes")
-#toc()
-```
-
-## Tease apart the similar neurogenic and RPC cell types
-
-```{r neuro, fig.width=30, fig.height=8, echo = F, message = F, fig.cap=neuro_cap}
-neuro_cap <- fig_cap(name='neuro', caption = 'A.
-Genes proposed by Lu et al.
-as being significantly differentially expressed during the RPC to neurogenic differentiation in human.
-We plot the differential expression (log2(fold change) directly between RPC and neurogenic cells.
-The standard error (calculated by DESeq2) is plotted.
-Genes with a padj < 0.1 and padj < 0.01 are labelled with "*" and "**", respectively.
-B.
-Genes from Lu et al.
-that we replicated with the scEiaD data.
-C.
-Novel genes we propose as being differentially expressed between RPC and neurogenic cells')
-# tic()
-# pairwise_neurogenic_test<-  scEiaD_2020_v01 %>% 
-#   tbl("diff_testing") %>% 
-#   filter(Against  == "RPC",   
-#          Group == 'CellType (Predict)', 
-#          Base == "Neurogenic Cell") %>% 
-#   group_by(Base, Against, Gene) %>% 
-#   summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
-#   arrange(-`mean log2FC`) %>%   
-#   collect() %>% 
-#   filter(!is.na(`mean log2FC`))
-# 
-# 
-# # mutate(GENEID = str_extract(Gene, 'ENS\\w+')) %>% 
-# # left_join(gene_tab, by = 'GENEID') %>% 
-# # relocate(SYMBOL, GENEID) %>% 
-# # dplyr::rename(Symbol = SYMBOL, ID = GENEID) %>% 
-# # arrange(-`mean log2FC`)
-#
-
-lu_et_al <- c("PTTG1", "SCG3", "SHD", "SOX11", "ONECUT2", "VSX1", "ASCL1", "TCF4", "SOX4", "RORB", "GADD45G", "MYBL1", "MFAP4", "HES6", "NEUROD1", "C8orf46", "OTX2", "OLIG2", "INSM1", "PCBP4", "NEUROD4", "FAM131C", "HES4", "HES5", "HMGA1", "YBX1", "GADD45A", "ATOH7", "DLX2", "DLX1", "SFRP1", "MYB", "NFIB", "SOX9", "ZFP36L1", "FOS", "JUN", "ZFP36L2", "SOX2", "HES1", "EGR1", "JUNB", "NR4A1", "ETV5", "FOXP1", "ID1", "ID2", "ID3", "ID4", "FOXP2")
-
-g_lu <- scEiaD_2020_v01 %>% tbl('genes') %>% collect() %>% filter(grepl(paste(paste0('^',lu_et_al), collapse = ' |'), Gene)) %>% pull(Gene)
-
-down <- pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` < 0, `mean padj` < 0.05) %>% mutate(d = `mean log2FC` - lfcSE) %>% arrange(-d) %>% pull(Gene)
-up <-  pairwise_neurogenic_test %>% filter(`sig count` > 1) %>% arrange(`mean padj`) %>% filter(`mean log2FC` > 2, `mean padj` < 0.05) %>% mutate(d = `mean log2FC` - lfcSE) %>% arrange(-d) %>% pull(Gene)
-
-source('src/make_exp_plot.R')
-
-input <- list()
-input$exp_plot_groups <- c('study_accession')
-input$exp_plot_facet <- c('CellType_predict', 'organism')
-input$exp_plot_genes <- c(down, up[!up %in% g_lu])
-input$exp_filter_cat <- 'CellType_predict'
-#input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
-input$exp_filter_on <- c("Neurogenic Cell", "RPC" )
-input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
-#input$exp_plot_ylab <- '% of Cells Detected'
-#input$exp_plot_groups <- c('organism')
-input$exp_plot_col_num <- 16
-input$exp_filter_min_cell_number <- 25
-input$flip_facet <- FALSE
-input$figure <- TRUE
-
-exp_plot_new <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-
-input$exp_plot_genes <- up[up %in% g_lu]
-exp_plot_existing <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-a_legend <- get_legend(exp_plot_new)
-neuro_plot <- plot_grid(plot_grid(exp_plot_existing + theme(legend.position="none") + xlab(''),
-                                  exp_plot_new + theme(legend.position="none") + ylab('') + xlab(''), 
-                                  labels = c('b','c'),
-                                  rel_widths = c(0.6,1)),
-                        a_legend, ncol = 1, rel_heights = c(1,0.2))
-
-lu_long <-  scEiaD_2020_v01 %>% 
-  tbl("diff_testing") %>% 
-  filter(Gene %in% g_lu,
-         Against  == "RPC",   
-         Group == 'CellType (Predict)', 
-         Base == "Neurogenic Cell") %>% 
-  #group_by(Base, Against, Gene) %>% 
-  #summarise(`mean log2FC` = mean(log2FoldChange), `mean padj` = mean(padj), `sig count` = sum(pvalue < 1e-2), `na count` = sum(is.na(log2FoldChange))) %>% 
-  #arrange(-`mean log2FC`) %>%   
-  collect() %>% 
-  mutate(Gene = str_extract(Gene, '\\w+ ') %>% 
-           gsub(' ', '', .)) %>% 
-  mutate(Gene = factor(Gene, levels = rev(lu_et_al)))
-
-#order <- lu_et_al %>% enframe() %>% mutate(Gene = value) %>% left_join(g_lu %>% enframe() %>% mutate(full = value, Gene = str_extract(value, '\\w+ ') %>% gsub(' ', '', .)) %>% select(-value), by = 'Gene') %>% data.frame() %>% filter(!is.na(full)) %>% pull(full)
-
-lu_rep_plot <- lu_long %>% 
-  mutate(Significance = case_when(padj < 0.01 ~ '**', padj < 0.1 ~ '*', TRUE ~ '')) %>% 
-  ggplot(aes(x=Gene,y=log2FoldChange,  label = Significance)) + 
-  geom_bar(stat='identity', position = position_dodge2(), fill = 'gray') + 
-  geom_errorbar(aes(ymin = log2FoldChange - lfcSE,
-                    ymax = log2FoldChange + lfcSE), width = 0.5,
-                position = position_dodge2()) +
-  ylab('Neurogenic / RPC Gene Expression (log2 Fold Change)') +
-  coord_flip() + 
-  cowplot::theme_cowplot()  + geom_text(aes(y=7), position = position_dodge2(width = 1)) + facet_wrap(~Organism)
-#g_lu[!g_lu %in% lu_long$Gene]
-
-cowplot::plot_grid(lu_rep_plot, neuro_plot, rel_widths = c(0.6,1), labels = c('a'))
-```
-
-While the well supported markers we propose clearly define most of the cell types, we see that the neurogenic cells and RPCs have very similar expression patterns across the gene sets.
-These cell types are closely related as the neurogenic cells are derived from the multi-potent RPC to later specify neural cell types.
-We use the set of genes identified by Lu et al.
-[@luSingleCellAnalysisHuman2020b] in their study of human retinal differentiation to determine whether our scEiaD database retains the same trends (`r fig_cap(name = 'neuro', display = 'cite')`A).
-As expected, we see several patterns.
-First, we see most of the gene expression changes between RPC and neurogenic are matched between human.
-We further show that these genes also show the same trends in mouse datasets.
-Next we see that the direction of the changes largely matches Lu et al (`r supFig_cap(name='lu_et_al', display = 'cite')`).
-Finally, we see a small number of genes are confirmed in our database as consistently differentially expressed with a padj < 0.1.
-
-To determine whether we can propose any new genes that distinguish RPC and neurogenic states we use our DESeq2 differential expression contrast that directly compares these two cell types.
-We find `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Homo sapiens') %>% nrow()` human genes and `r pairwise_neurogenic_test_ungrouped %>% filter(abs(log2FoldChange) > 2, padj < 0.05, Organism == 'Mus musculus') %>% nrow()` mouse genes that have an absolute log2 fold change greater than two and a adjusted p-value below 0.05 (`r supTab_cap(name = 'neuro_table', display = 'cite')`).
-To identify candidate genes that are consistently differentially expressed at this transitional state in mouse and human we apply similar logic that we used in identifying well supported marker genes across the retina cell types.
-We filter to genes with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two or less than negative one (`r fig_cap(name = 'neuro', display = 'cite')`B and C).
-We find one gene, ZBTB38, that meets this criteria and drops in expression when comparing RPC to neurogenic.
-We find `r up %>% length()` genes that increase in expression from RPC to neurogenic, `r up[!up %in% g_lu] %>% length()` of which are not previously identified by Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`C).
-We replicate `r up[up %in% g_lu] %>% str_extract(., '\\w+') %>% paste(., collapse = ', ') %>% str_to_title()` from Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`B) as differentially expressed between neurogenic and RPC.
-Futhermore we show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`a how this figure can be replicated directly in our web app.
-
-## Our pan-study macula vs peripheral cone test replicates a published macaque finding and proposes six new markers
-
-```{r region_fig, fig.width=18, fig.height=3,  echo = F, message = F, fig.cap = region_fig_cap}
-region_fig_cap <- fig_cap(name='regino_fig', caption = 'Plot of markers distinguishing human macula or peripheral punch derived cones.')
-
-source('src/make_exp_plot.R')
-#load('data/pseudoBulk_cone_region_files.Rdata')
-region_table <- read_csv('supplemental_files/pseudoBulk_cone_region_table.csv.gz')
-input <- list()
-input$exp_plot_groups <- c('study_accession')
-input$exp_plot_facet <- c('CellType_predict', 'organism', 'retina_region')
-input$exp_plot_genes <- c('PRPH2 (ENSG00000112619)')
-input$exp_filter_cat <- c('CellType_predict', 'organism')
-#input$exp_filter_on <- c('AC/HC Precursor', 'Amacrine Cell', 'Astrocyte', 'Bipolar Cell', 'Cone', 'Early RPC',' Horizontal Cell','Late RPC','Microglia','Muller Glia', 'Neurogenic Cell', 'Photoreceptor Precursor','Retinal Ganglion Cell','Rod','Rod Bipolar Cell','RPC','RPE' )
-input$exp_filter_on <- c("Cone", "Homo sapiens" )
-input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
-#input$exp_plot_ylab <- '% of Cells Detected'
-#input$exp_plot_groups <- c('organism')
-input$exp_plot_col_num <- 16
-input$exp_filter_min_cell_number <- 10
-input$flip_facet <- FALSE
-input$figure <- FALSE
-input$figure_fovea <- TRUE
-
-prph2 <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') +  theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() ) 
-input$exp_plot_genes <-  'RS1 (ENSG00000102104)';rs1 <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') +  theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
-
-input$exp_plot_genes <-  'GNGT1 (ENSG00000127928)';gngt1 <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') +  theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
-#input$exp_plot_genes <-  'RS1 (ENSG00000102104)';rs1 <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-
-diff_region_cone_genes <- region_table %>% filter(padj < 0.01) %>% 
-  arrange(log2FoldChange) %>% 
-  mutate(id = paste0(Gene, ' (', Ensembl, ')')) %>% 
-  pull(id)
-input$exp_filter_on <- c("Cone", "Homo sapiens" )
-input$exp_plot_genes <- diff_region_cone_genes[1]; a <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') +  theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
-input$exp_plot_genes <- diff_region_cone_genes[2]; b <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
-input$exp_plot_genes <- diff_region_cone_genes[3]; c <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
-input$exp_plot_genes <- diff_region_cone_genes[4]; d <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + ylab('') + xlab('') + theme(axis.ticks.x=element_blank(), axis.text.x=element_blank() ) 
-input$exp_plot_genes <- diff_region_cone_genes[5]; e <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
-input$exp_plot_genes <- diff_region_cone_genes[6]; f <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
-input$exp_plot_genes <- diff_region_cone_genes[7]; g <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + xlab('') + ylab('') + theme(legend.position="none",axis.ticks.x=element_blank(), axis.text.x=element_blank() )
-
-region_legend <- get_legend(d)
-
-plot_grid(plot_grid(a,b,c,d + theme(legend.position="none"),e,f,g, nrow = 1),
-          region_legend, 
-          rel_heights = c(1,0.2), 
-          nrow = 2)
-```
-
-While we have provided data in many forms via the web app at plae.nei.nih.gov, there are many more dissections of the data that can be performed.
-To facilitate custom testing with the scEiaD database we provide, as an example, a human-specific macula versus peripheral cone differential testing analysis document as supplementary file "pseudobulk_cone_region.freeze01.html".
-In our analysis document we use the Seurat object containing the full scEiaD dataset we provide at plae.nei.nih.gov to build a custom pseudobulk counts matrix containing only cones from human studies that were taken in a region specific manner.
-This left us with 26678 cells across six studies.
-Yan et al.
-previously proposed Prph2 and Rs1 as markers to distinguish macula and peripheral located human cones[@yanCellAtlasHuman2020a] and Peng et al.
-proposed Calb1 and Gngt1 as cone region markers in macaque [@pengMolecularClassificationComparative2019a].
-
-Our differential testing found that while Prph2, Rs1, and Gngt1 had expression differences matching that the findings Yan et al.
-and Peng et al.
-published, they failed to replicate in our test with pvalue and padj being above 0.1 for all three.
-In contrast Calb1 did replicate strongly with the sixth lowest padj (0.00721) across our entire test and a log2 fold change of 2.87.
-We find six more genes with a padj < 0.01: Pde6g, Akap12, Hspb1, Gadd45b, Cc2d2a, and RP1l1 (`r fig_cap(name = 'region_fig', display = 'cite')` and `r supTab_cap(name = 'region_table', display = 'cite')`).
-None of these candidate cone region marker genes display regionally different expression in rods (`r supFig_cap(name = 'rod_region_fig', display = 'cite')`).
-
-# Conclusions
-
-## Limitations of scEiaD
-
-The longer term goal is to continue adding datasets to the scEiaD at plae.nei.nih.gov until we have three or more independent datasets per species and nearly all cell types.
-Earlier we proposed well supported cell types.
-Right now, most of these well supported cell types are in the retina.
-The RPE behind the retina, the cornea and lens in front, and the outflow tract and supporting musculature still could use more independent studies.
-Many common human diseases underlie the RPE (e.g.
-AMD) and the cornea (e.g.
-Keratoconus) and these tissues will remain a focus for the next update of our database.
-While we have curated a large number of cell types across the eye, many of these cell types have "sub" cell types.
-For examplethe human has three types of cones that are optimized for short, medium, and long wavelengths.
-More dramatically, Yan et al.
-have proposed over sixty types of mouse amacrine cell types [@yanMouseRetinalCell2020a].
-While we have found it fairly straightforward to use machine learning to transfer the broad (e.g.
-rod, cone, amacrine) cell type labels across all datasets, we have found it very challenging to transfer the "sub" cell type labels.
-This is largely because while we have high diversity of studies for the broad cell types, this is not true for most of the sub celltypes.
-We believe more datasets that carefully dissect the sub-celltypes may be necessary to confidently propose the sub-celltypes of the retina.
-Finally, while our scVI based model can theoretically be used by an outside group to transfer their internal data into the scEiaD latent space, it is very challenging to implement as the the software landscape is still rapidly changing and it is very difficult to architect consistent outputs over time with regularly changing software versions.
-We are investigating simpler methods to enable transfer of knowledge from scEiaD to outside data.
-
-## High feature web app provides quick access to over one million transcriptomes across `r meta_filter$study_accession %>% unique %>% length` datasets
-
-We have improved on the scEiaD dataset we published in Swamy et al.
-[@swamyBuildingMegaSinglecell2021] by adding another species (chicken), new ocular tissues, a human non-ocular reference set; in total `r meta_filter$Citation %>% unique() %>% length() - meta_filter_v0$Citation %>% unique() %>% length()` new single cell datasets.
-We re-structured the data into a custom database to enable our custom built high performance web app to quickly access the data.
-This democratizes access to over a million cells across `r meta_filter$study_accession %>% unique() %>% length()` studies to anyone with a web browser.
-We recognize that while we have built a powerful visualization platform we cannot anticipate all uses and for those researchers who wish to make custom queries of the data, we provide the data underlying plae in a wide variety of formats, including anndata and Seurat objects for each study, count tables, cell level metadata, and the DESeq2 based differential testing across three species.
-We provide a short example of how a custom query could work by running a custom differential test between macula and peripheral derived human cones and demonstrate how this can both validate existing proposed genes and suggest new ones.
-
-## High diversity of studies enables reproduction and extension of cell type gene expression findings
-
-A great many groups have proposed cell atlases.
-These range from efforts focusing on a subset of cells from a single layer of a tissue (for example [@tranSingleCellProfilesRetinal2019a] ) to a huge consortium encompassing multiple organs [@consortiumTabulaSapiensSingle2021].
-We propose a complementary approach wherein we use the research data from the community at large to create a meta-atlas.
-This approach enables a small group to produce an atlas which rivals the size from massive multi-group consortiums.
-Furthermore the inherent variability between the data from independent groups can be used as a feature to get insight into whether a gene cell type signal is stable across an entire community's datasets.
-We briefly demonstrate how the high study diversity in scEiaD can be used to validate proposed RPC to neurogenic gene expression changes.
-
-# Supplementary Figures and Tables
-
-```{r decontX, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=decontX_cap}
-decontX_cap <- supFig_cap(name='decontX', caption = 'DecontX *in silico* ambient RNA contamination tool substantially removes Rhodopsin expression in non-rod cells while retaining high expression rods.
-First shown (a) is Rhodopsin expression across the scEiaD v0 dataset, without DecontX optimization.
-Note Rhodopsin expression is noticeable in many cell types beyond the rods.
-Next (b) is the scEiaD v1 dataset with DecontX optimization.
-Rhodopsin is nearly exclusively expressed in labelled rod cells.
-Despite the computational ambient RNA removal, Rhodopsin expression remains high in the Rods.')
-source('src/make_exp_plot.R')
-
-input <- list()
-input$exp_plot_groups <- c('organism')
-input$exp_plot_facet <- c('CellType_predict')
-input$exp_plot_genes <- c('RHO (ENSG00000163914)')
-input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
-input$exp_plot_col_num <- 16
-input$exp_filter_min_cell_number <- 50
-input$flip_facet <- FALSE
-input$figure <- FALSE
-input$figure_fovea <- FALSE
-
-v0_rho <- make_exp_plot(input, scEiaD_2020_v00, meta_filter_v0)
-v1_rho <-  make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-input$exp_filter_min_cell_number <- 1000
-v1.1_rho <- make_exp_plot(input, scEiaD_2020_v01, meta_filter %>% filter(CellType_predict %in% unique(meta_filter_v0$CellType_predict)))
-plot_grid(v0_rho, v1_rho, ncol = 1, labels = c('a','b'), rel_heights = c(1,1.5))
-```
-
 ```{r PLAE_screenshots_supplemental, echo = F, message = F, fig.cap = PLAE_screenshots_supplemental_cap}

### 2022-08-29 src/pubmed_query.R
test pubmed query src

@@ -1,0 +1,94 @@
+# return PMID for query
+library(easyPubMed)
+library(tidyverse)
+#load('~/data/massive_integrated_eye_scRNA/top_markers.Rdata')
+library(pool)
+library(RSQLite)
+
+data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vPLAE}/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20__pointRelease01.sqlite'))
+
+pubmed_counter <- function(query, sleep_time = 0.2, api_key = '8d33dae565ca3a523495197da36cc085d308'){
+  Sys.sleep(sleep_time)
+  pub_query <- query
+  entrez_id <- get_pubmed_ids(pub_query, api_key = api_key)
+  if (entrez_id$Count == 0){
+    out <- NA
+  } else {
+    abstracts_txt <- fetch_pubmed_data(entrez_id, format = "abstract") 
+    out <- grep('PMID', abstracts_txt, value = TRUE) %>% str_extract(., '\\d\\d\\d+') %>% unique()}
+  out
+}
+
+ctr <- meta_filter %>% filter(Tissue == 'Retina') %>% pull(CellType)
+
+queries <- consist_diff %>% 
+  filter(Base %in% ctr) %>% 
+  mutate(Base2 = gsub('Cell','',Base),
+         Base2 = case_when(Base2 == 'AC/HC Precursor' ~ 'Amacrine Horizontal',
+                          grepl('RPC', Base2) ~ 'Progenitor',
+                          TRUE ~ Base2)) %>% 
+  mutate(Symbol = case_when(Symbol == 'TF' ~ 'Transferrin',
+                            Symbol == 'CP' ~ 'ceruloplasmin',
+                            Symbol == 'F3' ~ 'Coagulation Factor III',
+                            Symbol == 'HR' ~ 'hairless',
+                            TRUE ~ Symbol)) %>% 
+  mutate(
+    pm_query0 = paste0(Symbol),
+    pm_query1 = paste0(Symbol, ' AND ', Base2),
+    pm_query2 = paste0(Symbol, ' AND Retina'))
+
+pmid0 <- list()
[…]

### 2022-08-29 manuscript_00.Rmd
text update

@@
 #toc()
+
+num1 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(Tissue) %>% unique() %>% length()
+num2 <- meta_filter %>% filter(Organ != 'Eye') %>% pull(Tissue) %>% unique() %>% length()
+num3 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique() %>% length()
 ```
@@
 This allows for complicated queries to complete in seconds rather than minutes.
-To compare loading times for useful tasks we queried https://plae.nei.nih.gov, http://singlecell-eye.org, and http://tabula-sapiens-portal.ds.czbiohub.org, on 2022-08-24.
+To compare web app loading times for useful tasks we queried https://plae.nei.nih.gov, http://singlecell-eye.org, and http://tabula-sapiens-portal.ds.czbiohub.org, on 2022-08-24.
 For Spectacle, we used the "Single-Cell RNA-Seq Analysis of Retinal Development..." 10X dataset from Clark et al.
@@
 Timings are accurate to the second.
-If the functionality tested was not available for a certain web app, it was represented by a blank timing.
+If the functionality tested was not available for a certain web app, it was represented by a blank space.
 
@@
 ```{r, fig.height=14, fig.width=12, fig.cap = scEiaD_data_overview_cap}
-num1 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(Tissue) %>% unique() %>% length()
-num2 <- meta_filter %>% filter(Organ != 'Eye') %>% pull(Tissue) %>% unique() %>% length()
-num3 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique() %>% length()
 
@@
 B.
-Number of studies for each ocular cell type (over 10 cells), colored by species.
+Number of studies for each ocular predicted cell type (over 10 cells), colored by species.
 C.
 Tabulation of the number of studies present across {num1} ocular and {num2} non-ocular tissues, D.
-Web app performance (lower is better, missing functionality is shown as blank) for important functions compared between PLAE, Spectacle, and Tabula Sapiens (see methods)."))
+Web app timings (lower is better) for important functions compared between PLAE, Spectacle, and Tabula Sapiens (see methods for more details)."))
 
@@
   ungroup() %>% 
+  mutate(Organ = factor(Organ, levels = c('Eye','Body'))) %>% 
   ggplot(aes(x=Tissue,y=Count, fill = organism, label = Count, group = organism)) +
@@
                         'Abca4 Expression UMAP',
-                        'Boxplot of Abca4 Expression By CellType Across All Studies',
-                        'Boxplot of Abca4 Expression By CellType Across One Study',
+                        'Boxplot of Abca4 Expression By Cell Type Across All Studies',
+                        'Boxplot of Abca4 Expression By Cell Type Across One Study',
                         'Dotplot of Abca4, Pax6, Nrl, and Atoh7',
@@
   mutate(Test = factor(Test, levels = row.names(timings))) %>% 
-  ggplot(aes(x=name,y=value)) + 
+  mutate(Text = case_when(is.na(value) ~ 'Not possible',
+                          TRUE ~ '')) %>% 
+  ggplot(aes(x=name,y=value, label = Text)) + 
   geom_bar(stat= 'identity') + 
+  geom_text(aes(y=0), hjust = 0, angle = 90) +
   facet_wrap(~Test, scales = 'free_y', ncol = 2) + 
@@
 
-ab <- cowplot::plot_grid(cell_counts + theme(legend.position = "none"), 
-                         study_counts_by_cell, 
+ab <- cowplot::plot_grid(cell_counts + theme(legend.position = "none", axis.ticks.y = element_blank()), 
+                         study_counts_by_cell + theme(axis.ticks.y = element_blank()), 
                          nrow = 1, 
@@
 
-cd <- cowplot::plot_grid(study_count_by_tissue, 
+cd <- cowplot::plot_grid(study_count_by_tissue + theme(axis.ticks.y = element_blank()), 
                          timing_plot, 
@@
 The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are not ocular focused, but do contain some ocular data.
-The PanglaoDB and UCSC are resources which have ingressed a large number of published single cell datasets The UCSC Cell Browser has around sixty thousand ocular single cell transcriptomes while the PangloaDB has about 47 thousand.
+The PanglaoDB and UCSC are resources which have ingressed a large number of published single cell datasets The UCSC Cell Browser has around sixty thousand ocular single cell transcriptomes while the PangloaDB has about forty-seven thousand.
 However, these resources provide no curation of the metadata and process the data for each study individually.
 Thus each study is effectively silo-ed off from the rest and cross comparisons are impractical.
-The Tabula Sapiens project is an effort to sequence all cells in the mouse; while it only contains around around 10 thousand ocular cells it can be more readily cross compared to non-ocular tissues / cells.
-The Protein Atlas resource has recently used one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g.
+The Tabula Sapiens project is an effort to sequence all cell type transcriptomes in humans; while it only contains around around ten thousand ocular cells it can be more readily cross compared to non-ocular tissues / cells.
+A more substantial downside is that the cell type labels can be insufficient - for example rods and cones are not independently identified but rather labelled as photoreceptors.
+The Protein Atlas resource has recently used only one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g.
 one can see whether a gene of interest is enriched in Cones).
+This is problematic as it is unclear whether signal found is specific to the one dataset they use or is common across independent species and datasets.
+
 The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has collated a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app.
-PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence, and, crucially, full availability of the underlying data for outside usage and integrating the independent studies to deliver cross study comparisons.
+PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence in a consistent manner, and, crucially, full availability of the underlying data for outside usage and integrating the independent studies to deliver cross study comparisons.
 
@@
 
-web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.')
+web_resources_cap <- tab_cap(name = 'web_resources', caption = 'Comparison of PLAE with other web resources that contain single cell ocular datasets, as of 2021-03-15.
+The Protein Atlas is asterisked as while it contains independent datasets, it only has one ocular study.
+')
 
@@
 In contrast, Spectacle, as of March 2022, is a compilation of study-level datasets and thus you cannot run queries across studies.
-Furthermore, Spectacle contains no non-ocular dataset and the cell types labels have not been curated or harmonized between studies.
-Finally, the Spectacle web app takes many seconds to load an individual dataset and many more to display the data.
+Spectacle contains no non-ocular dataset and the cell types labels have not been curated or harmonized between studies (for example different studies can use MG, Mueller, Müller Glia, or other variations to refer to the same cell type).
+Finally, in contrast to PLAE, Spectacle does not make their datasets available for download in any form.
 
@@
 In contrast, starting from the loading page of Spectacle, it takes approximately eleven seconds to plot the cell type metadata for a several thousand cells across across a single study (tested on 2022-08-26).
-In the same amount of time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies (`r fig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')d`).
-The speed of the PLAE app is also comparable in several ways to the CELLXGENE browser while providing additional bar and dot plot functionality.
+Within the same time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies (`r fig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')d`).
+The speed of the PLAE app is also comparable to the CELLxGENE browser serving the Tabula Sapiens resource while providing additional functionality.
 
@@
 Across the cell types of the eye, we find `r diff_tab %>% nrow()` gene - celltype combinations that are significantly differentially expressed.
-We define significant as a log2 fold change greater than two as well as a p-value (corrected) less than $1 x 10^-5#.
+We define significant as a log2 fold change greater than two as well as a p-value (corrected) less than $1 x 10^-5$.
 We find a mean of

### 2022-08-29 src/pubmed_query.R
query update

@@ -84,11 +84,12 @@ for (i in pm_query3){
 #   pmid2[[i]] <- pubmed_counter(i)
 # }
 
-save(pmid0, pmid1, pmid2, file = 'data/top_marker_pmid.Rdata')
+
 queries <- queries %>% left_join(., pmid0 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% rename(query0=value), by = c("Symbol" = "name")) 
 queries <- queries %>% left_join(., pmid1 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% rename(query1=value), by = c("pm_query1" = "name")) 
 queries <- queries %>% left_join(., pmid2 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% rename(query2=value), by = c("pm_query2" = "name"))
+save(pmid0, pmid1, pmid2, pmid3, queries, file = 'data/top_marker_pmid.Rdata')
 
-save(top_markers, marker_info, exp_stats, file = '~/data/scEiaD/top_markers.Rdata')
+save(queries, file = 'data/top_markers.Rdata')
 #pmid %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe()
 #pmid %>% map(function(x) paste %>% unlist() %>% enframe()

### 2022-08-30 manuscript_00.Rmd
halfway to updated heatmap

@@
 # consist_diff_base <- consist_diff_genes %>% as_tibble() %>% rename(Symbol = value) %>% left_join(consist_diff %>% select(Symbol, Base) %>% unique()) %>% pull(Base)
-
-consist_diff_hm <- ComplexHeatmap::Heatmap(consist_diff_wide, column_labels = consist_diff_organism, 
+cbind(consist_diff_genes, consist_diff_base) %>% as_tibble() %>% left_join(queries, by = c('consist_diff_genes' = 'Symbol', 'consist_diff_base' = 'Base'))
+
+rand_nums <- floor(runif(nrow(consist_diff_wide), min=0, max=3))
+rand_col <-  circlize::colorRamp2(c(-0, 1, 2), c("white", "gray", "black"))
+
+consist_diff_hm <- ComplexHeatmap::Heatmap(consist_diff_wide, 
+                                           column_labels = consist_diff_organism, 
                                            column_split = consist_diff_cts, 
-                                           row_split = consist_diff_base, show_row_dend = FALSE, 
-                                           row_labels = consist_diff_genes, 
+                                           row_split = consist_diff_base, 
+                                           show_row_dend = FALSE, 
                                            column_title_rot = 90, 
                                            name = 'log2FC', use_raster = TRUE,
-                                           row_title_rot = 0)
+                                           row_title_rot = 0) + 
+  ComplexHeatmap::Heatmap(rand_nums,
+                          row_split = consist_diff_base,
+                          col = rand_col,
+                          row_labels = consist_diff_genes,
+                          column_labels = "Pubmed Gene Citations",
+                          column_names_side = 'top')
 ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(0, 0, 1.5, 0), "in"), row_title = "Marker Genes")

### 2022-08-30 src/pubmed_query.R
update filtering

@@ -13,6 +13,14 @@ pubmed_counter <- function(query, sleep_time = 0.2, api_key = '8d33dae565ca3a523
   pub_query <- query
   entrez_id <- get_pubmed_ids(pub_query, api_key = api_key)
   if (entrez_id$Count == 0){
+    # run one more time, with a 0.5 second delay
+    Sys.sleep(3)
+    print("RERUN")
+    entrez_id <- get_pubmed_ids(pub_query, api_key = api_key)
+  }
+  if (entrez_id$Count == 0){
+    # then give up and return NA
+    print("NA")
     out <- NA
   } else {
     abstracts_txt <- fetch_pubmed_data(entrez_id, format = "abstract") 
@@ -20,23 +28,30 @@ pubmed_counter <- function(query, sleep_time = 0.2, api_key = '8d33dae565ca3a523
   out
 }
 
-ctr <- meta_filter %>% filter(Tissue == 'Retina') %>% pull(CellType)
+#ctr <- meta_filter %>% filter(Tissue == 'Retina') %>% pull(CellType)
 
 queries <- consist_diff %>% 
-  filter(Base %in% ctr) %>% 
+  filter(Symbol %in% consist_diff_genes) %>% 
+  #filter(Base %in% ctr) %>% 
   mutate(Base2 = gsub('Cell','',Base),
          Base2 = case_when(Base2 == 'AC/HC Precursor' ~ 'Amacrine Horizontal',
                           grepl('RPC', Base2) ~ 'Progenitor',
                           TRUE ~ Base2)) %>% 
-  mutate(Symbol = case_when(Symbol == 'TF' ~ 'Transferrin',
+  mutate(Symbol2 = case_when(Symbol == 'TF' ~ 'Transferrin',
                             Symbol == 'CP' ~ 'ceruloplasmin',
                             Symbol == 'F3' ~ 'Coagulation Factor III',
                             Symbol == 'HR' ~ 'hairless',
                             TRUE ~ Symbol)) %>% 
   mutate(
-    pm_query0 = paste0(Symbol),
-    pm_query1 = paste0(Symbol, ' AND ', Base2),
-    pm_query2 = paste0(Symbol, ' AND Retina'))
+    pm_query0 = paste0(Symbol2),
+    pm_query1 = paste0(Symbol2, ' AND ', Base2),
[…]

### 2022-08-30 src/pubmed_query.R
update

@@ -111,5 +111,5 @@ for (i in pm_query4){
 queries <- queries %>% left_join(., pmid0 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% rename(query0=value), by = c("Symbol" = "name")) 
 queries <- queries %>% left_join(., pmid1 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% rename(query1=value), by = c("pm_query1" = "name")) 
 queries <- queries %>% left_join(., pmid2 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% rename(query2=value), by = c("pm_query2" = "name"))
-save(pmid0, pmid1, pmid2, pmid3, queries, file = 'data/top_marker_pmid.Rdata')
+save(pmid0, pmid1, pmid2, pmid3, pmid4, queries, file = 'data/top_marker_pmid.Rdata')
 

### 2022-08-31 manuscript_00.Rmd
update

--- 2022-08-29 manuscript_00.Rmd
@@
 
-data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
-data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
+#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+
 scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vGiga}/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite'))
@@
 # consist_diff_base <- consist_diff_genes %>% as_tibble() %>% rename(Symbol = value) %>% left_join(consist_diff %>% select(Symbol, Base) %>% unique()) %>% pull(Base)
-
-consist_diff_hm <- ComplexHeatmap::Heatmap(consist_diff_wide, column_labels = consist_diff_organism, 
+load('data/top_marker_pmid.Rdata')
+
+# pmid0 and pmid4 is just GENE
+query0 <- cbind(consist_diff_genes, consist_diff_base) %>% as_tibble() %>% 
+  left_join(queries, by = c('consist_diff_genes' = 'Symbol', 'consist_diff_base' = 'Base')) %>% 
+  mutate(citation = case_when(
+    query0 > mean(pmid4 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Many",
+    query0 > median(pmid4 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Some",
+    TRUE ~ "Few")) %>% 
+  pull(citation)
+
+# pmid2 and pmid3 is "AND RETINA"
+query2 <- cbind(consist_diff_genes, consist_diff_base) %>% 
+  as_tibble() %>% 
+  left_join(queries, by = c('consist_diff_genes' = 'Symbol', 'consist_diff_base' = 'Base')) %>% 
+   mutate(citation = case_when(
+     query2 > mean(pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Many",
+     query2 > median(pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Some",
+     TRUE ~ "Few")) %>% 
+  pull(citation)
+#rand_col <-  circlize::colorRamp2(c('0', '1', '2'), c("white", "gray", "black"))
+citation_col <-c('white','gray','black') 
+names(citation_col) <-  c('Few','Some','Many')
+consist_diff_hm <- ComplexHeatmap::Heatmap(consist_diff_wide, 
+                                           column_labels = consist_diff_organism, 
                                            column_split = consist_diff_cts, 
-                                           row_split = consist_diff_base, show_row_dend = FALSE, 
-                                           row_labels = consist_diff_genes, 
+                                           row_split = consist_diff_base, 
+                                           show_row_dend = FALSE, 
                                            column_title_rot = 90, 
                                            name = 'log2FC', use_raster = TRUE,
-                                           row_title_rot = 0)
+                                           row_title_rot = 0) +
+
+  ComplexHeatmap::Heatmap(as.character(query0),
+                          row_split = consist_diff_base,
+                          col = citation_col,
+                          row_labels = consist_diff_genes,
+                          show_heatmap_legend = FALSE,
+                          column_labels = "Pubmed Gene Citations",
+                          column_names_side = 'top') +
+  ComplexHeatmap::Heatmap(as.character(query2),
+                          row_split = consist_diff_base,
+                          col = citation_col,
+                          row_labels = consist_diff_genes,
+                          name = 'Citation Count',
+                          column_labels = "Pubmed Gene + \"Retina\" Citations",
+                          column_names_side = 'top')
 ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(0, 0, 1.5, 0), "in"), row_title = "Marker Genes")

### 2022-08-31 src/pubmed_query.R

@@ -5,8 +5,8 @@ library(tidyverse)
 library(pool)
 library(RSQLite)
 
-data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vPLAE}/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20__pointRelease01.sqlite'))
+#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+#scEiaD_2020_v01 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vPLAE}/MOARTABLES__anthology_limmaFALSE___4000-counts-universe-study_accession-scANVIprojection-15-5-0.1-50-20__pointRelease01.sqlite'))
 
 pubmed_counter <- function(query, sleep_time = 0.2, api_key = '8d33dae565ca3a523495197da36cc085d308'){
   Sys.sleep(sleep_time)
@@ -14,7 +14,7 @@ pubmed_counter <- function(query, sleep_time = 0.2, api_key = '8d33dae565ca3a523
   entrez_id <- get_pubmed_ids(pub_query, api_key = api_key)
   if (entrez_id$Count == 0){
     # run one more time, with a 0.5 second delay
-    Sys.sleep(3)
+    Sys.sleep(1)
     print("RERUN")
     entrez_id <- get_pubmed_ids(pub_query, api_key = api_key)
   }
@@ -31,7 +31,7 @@ pubmed_counter <- function(query, sleep_time = 0.2, api_key = '8d33dae565ca3a523
 #ctr <- meta_filter %>% filter(Tissue == 'Retina') %>% pull(CellType)
 
 queries <- consist_diff %>% 
-  filter(Symbol %in% consist_diff_genes) %>% 
+  #filter(Symbol %in% consist_diff_genes) %>% 
   #filter(Base %in% ctr) %>% 
   mutate(Base2 = gsub('Cell','',Base),
          Base2 = case_when(Base2 == 'AC/HC Precursor' ~ 'Amacrine Horizontal',
@@ -82,8 +82,8 @@ all_diff_symbol <- diff_tab %>% filter(padj < 0.01) %>% pull(Symbol) %>% unique(
[…]

### 2022-08-31 manuscript_00.Rmd
tighten up heatmap

--- 2022-08-30 manuscript_00.Rmd
@@
 
-## Leveraging the high diversity of studies to identify thousands of consistently differentially expressed genes across ocular cell types
+<!-- ## Leveraging the high diversity of studies to identify thousands of consistently differentially expressed genes across ocular cell types -->
+
+<!-- Across the cell types of the eye, we find `r diff_tab %>% nrow()` gene - celltype combinations that are significantly differentially expressed.
+We define significant as a log2 fold change greater than two as well as a p-value (corrected) less than $1 x 10^-5$.
+We find a mean of  -->
+<!-- `r diff_tab_sup$"Homo sapiens" %>% mean() %>% round()`,  -->
+<!-- `r diff_tab_sup$"Mus musculus" %>% mean() %>% round()`, and  -->
+<!-- `r diff_tab_sup$"Macaca fascicularis" %>% mean() %>% round()` differentially expressed genes in human, macaque, and mouse respectively (`r supTab_cap(name = 'diff_tab_sup_tab', display = 'cite')`).
+These variable results directly reflect the differing number of independent studies - mouse and human have substantially more than macaque (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
+The significant filtered differential table is made available as supplementary file "differential_expression_table.csv.gz" and the full differential tables are available for download at plae.nei.nih.gov -> Data.
+-->
+
+## Well supported ocular cell types identify high confidence cell type markers
 
@@
 
-Across the cell types of the eye, we find `r diff_tab %>% nrow()` gene - celltype combinations that are significantly differentially expressed.
-We define significant as a log2 fold change greater than two as well as a p-value (corrected) less than $1 x 10^-5$.
-We find a mean of
-`r diff_tab_sup$"Homo sapiens" %>% mean() %>% round()`, 
-`r diff_tab_sup$"Mus musculus" %>% mean() %>% round()`, and 
-`r diff_tab_sup$"Macaca fascicularis" %>% mean() %>% round()` differentially expressed genes in human, macaque, and mouse respectively (`r supTab_cap(name = 'diff_tab_sup_tab', display = 'cite')`).
-These variable results directly reflect the differing number of independent studies - mouse and human have substantially more than macaque (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
-The significant filtered differential table is made available as supplementary file "differential_expression_table.csv.gz" and the full differential tables are available for download at plae.nei.nih.gov -> Data.
-
-## Well supported ocular cell types identify high confidence cell type markers
-
 Cell type marker genes can be proposed on the basis of existing knowledge about their function and on high expression differences when comparing to other cell types.
@@
 We see how the differentiated cell types of the retina are grouped together with the exception of the Müller glia, which are next to the RPC from which they derive from.
-The photoreceptors are also next to each other
+The photoreceptors are also next to each other.
+
+Aasdfasdfasdfadsf
 

### 2022-08-31 src/diff_data_pull.R

@@ -121,6 +121,7 @@ ws_oct <- ws_table %>% filter(`Well Supported` == 'Yes') %>% pull(`CellType (Pre
 consist_diff <- scEiaD_2020_v01 %>%
   tbl("diff_testing") %>%
   filter(Base %in% ws_oct) %>%
+  #filter(Base %in% ct_retain) %>% 
   filter(Group == 'CellType (Predict)',
          Against == 'All',
          Organism %in% c('Homo sapiens','Mus musculus')) %>%
@@ -145,14 +146,14 @@ toc()
 #######################################
 
 tic()
-g <- top_genes %>% select(Symbol, Base) %>% filter(!Base %in% c('Early RPC','Late RPC')) %>% left_join(consist_diff, by = c("Symbol", "Base"))  %>% group_by(Base) %>% slice_max(n=8, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
+g <- top_genes %>% select(Symbol, Base) %>% filter(!Base %in% c('Early RPC','Late RPC', "Fibroblast","Endothelial","Oligodendrocyte", "T/NK-Cell")) %>% left_join(consist_diff, by = c("Symbol", "Base"))  %>% group_by(Base) %>% slice_max(n=8, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
 
 consist_diff_long <- scEiaD_2020_v01 %>%
   tbl("diff_testing") %>%
   filter(Gene %in% g) %>%
   collect() %>%
   filter(Base %in% top_genes$Base,
-         !Base %in% c('Early RPC','Late RPC'), 
+         !Base %in% c('Early RPC','Late RPC',"Fibroblast","Endothelial","Oligodendrocyte", "T/NK-Cell"), 
          Group == 'CellType (Predict)',
          Against == 'All') %>%
   select(Gene, Organism, log2FoldChange, Base)

### 2022-08-31 manuscript_00.Rmd
update

@@
 The cell types are arranged by the columns by how related the expression patterns are.
-We see how the differentiated cell types of the retina are grouped together with the exception of the Müller glia, which are next to the RPC from which they derive from.
+We see how the differentiated cell types of the retina are grouped together with the exception of the Müller glia, which are near the RPC from which they derive from.
 The photoreceptors are also next to each other.
 
-Aasdfasdfasdfadsf
+To assess whether this marker selection was identifying known or potentially novel marker genes we ran several sets of queries against PubMed.
+We searched for the gene name or the gene name plus " AND Retina" across all well supported genes against the PubMed database, counting how many citations were returned.
+In contrast we also selected 1000 random genes that were not considered differentially expressed and ran the same PubMed searches.
+We find a substantial enrichment of citations for our well supported differentially expressed genes (t test, p < 6.7x10-98 and p < 2.4x10-12) for the gene and gene " and retina" PubMed searches when compared against the 1000 random genes, respectively.
+To display the results of this search for each individual gene we show visually whether the number of citations returned is less or equal than the median citation count for the random gene set ("few"), more than the median ("some"), or more than the mean ("many").
+We see that nearly all genes in (`r fig_cap(name = 'consist_genes_heatmap', display = 'cite')`) reflect existing knowledge.
+
+Genes which have little existing knowledge can be considered
 
@@
 ```{r echo = F, message = F, }
+
 # tic()
@@
 Columns are split by the cell types.
-More red is a higher log2 fold change of the gene - cell type relative to all other cell types and blue is more negative.')
+More red is a higher log2 fold change of the gene - cell type relative to all other cell types and blue is more negative.
+Relative number of citations are shown for the gene or gene " and retina" when searching against PubMed compared to 1000 randomly chosen genes.')
+
+load('data/top_marker_pmid.Rdata')
 
@@
 # consist_diff_base <- consist_diff_genes %>% as_tibble() %>% rename(Symbol = value) %>% left_join(consist_diff %>% select(Symbol, Base) %>% unique()) %>% pull(Base)
-cbind(consist_diff_genes, consist_diff_base) %>% as_tibble() %>% left_join(queries, by = c('consist_diff_genes' = 'Symbol', 'consist_diff_base' = 'Base'))
-
-rand_nums <- floor(runif(nrow(consist_diff_wide), min=0, max=3))
-rand_col <-  circlize::colorRamp2(c(-0, 1, 2), c("white", "gray", "black"))
+# pmid0 and pmid4 is just GENE
+# (t.test(pmid0 %>% map(function(x) sum(!is.na(x))) %>% unlist(), pmid4 %>% map(function(x) sum(!is.na(x))) %>% unlist()))$p.value
+query0 <- cbind(consist_diff_genes, consist_diff_base) %>% as_tibble() %>%
+  left_join(queries, by = c('consist_diff_genes' = 'Symbol', 'consist_diff_base' = 'Base')) %>%
+  mutate(citation = case_when(
+    query0 > mean(pmid4 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Many",
+    query0 > median(pmid4 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Some",
+    TRUE ~ "Few")) %>%
+  pull(citation)
+
+# pmid2 and pmid3 is "AND RETINA"
+# (t.test(pmid2 %>% map(function(x) sum(!is.na(x))) %>% unlist(), pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist()))$p.value
+query2 <- cbind(consist_diff_genes, consist_diff_base) %>%
+  as_tibble() %>%
+  left_join(queries, by = c('consist_diff_genes' = 'Symbol', 'consist_diff_base' = 'Base')) %>%
+  mutate(citation = case_when(
+    query2 > mean(pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Many",
+    query2 > median(pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Some",
+    TRUE ~ "Few")) %>%
+  pull(citation)
+#rand_col <-  circlize::colorRamp2(c('0', '1', '2'), c("white", "gray", "black"))
+citation_col <-c('white','gray','black')
+names(citation_col) <-  c('Few','Some','Many')
 
@@
                                            row_title_rot = 0) +
-  ComplexHeatmap::Heatmap(rand_nums,
+  ComplexHeatmap::Heatmap(as.character(query0),
                           row_split = consist_diff_base,
-                          col = rand_col,
+                          col = citation_col,
                           row_labels = consist_diff_genes,
+                          show_heatmap_legend = FALSE,
                           column_labels = "Pubmed Gene Citations",
+                          column_names_side = 'top') +
+  ComplexHeatmap::Heatmap(as.character(query2),
+                          row_split = consist_diff_base,
+                          col = citation_col,
+                          row_labels = consist_diff_genes,
+                          name = 'Citation Count',
+                          column_labels = "Pubmed Gene + \"Retina\" Citations",
                           column_names_side = 'top')
 ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(0, 0, 1.5, 0), "in"), row_title = "Marker Genes")
+
 #toc()
@@
 
-# Conclusions
-
 ## Limitations of scEiaD
 
-The longer term goal is to continue adding datasets to the scEiaD at plae.nei.nih.gov until we have three or more independent datasets per species and nearly all cell types.
+The longer term goal is to continue adding datasets to the scEiaD at plae.nei.nih.gov until we have three or more independent datasets per species and most ocular cell types.
 Earlier we proposed well supported cell types.
@@
 
+# Conclusions
+
 ## High feature web app provides quick access to over one million transcriptomes across `r meta_filter$study_accession %>% unique %>% length` datasets
@@
 
-## High diversity of studies enables reproduction and extension of cell type gene expression findings
+## Diversity of studies enables reproduction and extension of cell type gene expression findings
 
@@
 
+```{r undersupported_genes, fig.height=48, fig.width=8, echo = FALSE}
+cone_undersupported <- queries %>% filter(query2 <= 1) %>% arrange(-`mean log2FC`) %>% filter(!grepl("Endo|Macro|Micro|Endo", Base)) %>% group_by(Symbol) %>% summarise(Count = n(), Base = paste0(Base, collapse = ', '), l2FC = mean(`mean log2FC`)) %>% filter(Count == 1) %>% filter(Base == 'Rod') %>% arrange(-l2FC)
+
+input <- list()
+input$exp_plot_groups <- c('study_accession')
+input$exp_plot_facet <- c('CellType_predict')
+input$exp_plot_genes <- scEiaD_2020_v01 %>% tbl('Genes') %>% collect() %>% mutate(Symbol = str_extract(Gene, '.* \\(') %>% gsub(' \\(','',.)) %>% filter(Symbol %in% cone_undersupported$Symbol) %>% pull(Gene)
+input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
+input$exp_plot_col_num <- 16
+input$exp_filter_min_cell_number <- 50
+input$flip_facet <- FALSE
+input$figure <- FALSE
+input$figure_bloop <- TRUE
+input$figure_fovea <- FALSE
+
+make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+
+```
 ```{r PLAE_screenshots_supplemental, echo = F, message = F, fig.cap = PLAE_screenshots_supplemental_cap}

### 2022-08-31 src/diff_data_pull.R

@@ -121,7 +121,6 @@ ws_oct <- ws_table %>% filter(`Well Supported` == 'Yes') %>% pull(`CellType (Pre
 consist_diff <- scEiaD_2020_v01 %>%
   tbl("diff_testing") %>%
   filter(Base %in% ws_oct) %>%
-  #filter(Base %in% ct_retain) %>% 
   filter(Group == 'CellType (Predict)',
          Against == 'All',
          Organism %in% c('Homo sapiens','Mus musculus')) %>%
@@ -146,14 +145,14 @@ toc()
 #######################################
 
 tic()
-g <- top_genes %>% select(Symbol, Base) %>% filter(!Base %in% c('Early RPC','Late RPC', "Fibroblast","Endothelial","Oligodendrocyte", "T/NK-Cell")) %>% left_join(consist_diff, by = c("Symbol", "Base"))  %>% group_by(Base) %>% slice_max(n=8, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
+g <- top_genes %>% select(Symbol, Base) %>% filter(!Base %in% c('Early RPC','Late RPC')) %>% left_join(consist_diff, by = c("Symbol", "Base"))  %>% group_by(Base) %>% slice_max(n=8, order_by = `mean log2FC`) %>% mutate(g = paste0(Symbol, ' (', ID, ')')) %>% pull(g)
 
 consist_diff_long <- scEiaD_2020_v01 %>%
   tbl("diff_testing") %>%
   filter(Gene %in% g) %>%
   collect() %>%
   filter(Base %in% top_genes$Base,
-         !Base %in% c('Early RPC','Late RPC',"Fibroblast","Endothelial","Oligodendrocyte", "T/NK-Cell"), 
+         !Base %in% c('Early RPC','Late RPC'), 
          Group == 'CellType (Predict)',
          Against == 'All') %>%
   select(Gene, Organism, log2FoldChange, Base)

### 2022-08-31 src/make_exp_plot.R

@@ -14,6 +14,9 @@ make_exp_plot <- function(input, db, meta_filter){
   
   grouping_features <- input$exp_plot_groups
   
+  if (!'figure_bloop' %in% names(input)){
+    input$figure_bloop <- FALSE
+  } 
   
   meta_filter_EXP <- meta_filter
   if (!is.null(input$exp_filter_cat)){
@@ -135,6 +138,19 @@ make_exp_plot <- function(input, db, meta_filter){
       scale_colour_manual(values = rep(c(pals::alphabet() %>% unname()), 20)) +
       theme(legend.position="bottom") +
       facet_wrap(ncol = as.numeric(input$exp_plot_col_num), ~Gene + retina_region) 
+  }    else if (input$figure_bloop){
+    box_data  %>% 
+      mutate(Gene = str_extract(Gene, '^\\w+ ') %>% gsub(" $", '', .)) %>%
+      ungroup() %>% 
+      ggplot(aes(x=!!as.symbol(input$exp_plot_facet), y = !!as.symbol(input$exp_plot_ylab), color = !!as.symbol(grouping_features), group = CellType_predict)) +
+      geom_boxplot(color = 'black', outlier.shape = NA) +
+      ggbeeswarm::geom_quasirandom(size = 4, groupOnX = TRUE) +
+      cowplot::theme_cowplot() +
+      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
+      scale_radius(range=c(2, 6)) +
+      scale_colour_manual(values = rep(c(pals::alphabet() %>% unname()), 20)) +
+      theme(legend.position="bottom") +
+      facet_wrap(ncol = 1, ~Gene, scales = 'free_y') 
   }
   else {
     box_data %>%

### 2022-09-01 manuscript_00.Rmd
tighten up pubmed comp

@@
   mutate(citation = case_when(
-    query0 > mean(pmid4 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Many",
-    query0 > median(pmid4 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Some",
+    query0 > mean(pmid5 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Many",
+    query0 > median(pmid5 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Some",
     TRUE ~ "Few")) %>%
@@
   mutate(citation = case_when(
-    query2 > mean(pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Many",
-    query2 > median(pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Some",
+    query2 > mean(pmid6 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Many",
+    query2 > median(pmid6 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Some",
     TRUE ~ "Few")) %>%

### 2022-09-01 src/pubmed_query.R

@@ -84,6 +84,7 @@ rand_not_marker1 <- all_genes[!all_genes %in% all_diff_symbol]
 
 rand_not_marker0_sample <- rand_not_marker0[sample(1:length(rand_not_marker0), 1000)]
 rand_not_marker1_sample <- rand_not_marker1[sample(1:length(rand_not_marker1), 1000)]
+rand_all <- all_genes[sample(1:length(all_genes), 1000)]
 
 pm_query3 <- paste0(rand_not_marker1_sample, ' AND Retina')
 pmid3 <- list()
@@ -100,6 +101,22 @@ for (i in pm_query4){
   #Sys.sleep(1)
   pmid4[[i]] <- pubmed_counter(i)
 }
+
+pm_query5 <- rand_all
+pmid5 <- list()
+for (i in pm_query5){
+  print(i)
+  #Sys.sleep(1)
+  pmid5[[i]] <- pubmed_counter(i)
+}
+
+pm_query6 <- paste0(rand_all, ' AND Retina')
+pmid6 <- list()
+for (i in pm_query6){
+  print(i)
+  #Sys.sleep(1)
+  pmid6[[i]] <- pubmed_counter(i)
+}
 # pmid2 <- list()
 # for (i in x$pm_query2){
 #   print(i)
@@ -111,5 +128,5 @@ for (i in pm_query4){
 queries <- queries %>% left_join(., pmid0 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% rename(query0=value), by = c("Symbol" = "name")) 
 queries <- queries %>% left_join(., pmid1 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% rename(query1=value), by = c("pm_query1" = "name")) 
 queries <- queries %>% left_join(., pmid2 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% rename(query2=value), by = c("pm_query2" = "name"))
-save(pmid0, pmid1, pmid2, pmid3, pmid4, queries, file = 'data/top_marker_pmid.Rdata')
+save(pmid0, pmid1, pmid2, pmid3, pmid4, pmid5, pmid6, queries, file = 'data/top_marker_pmid.Rdata')
 

### 2022-09-01 manuscript_00.Rmd
freeze 01 release

@@
 
-data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
-data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD/'
+#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vGiga}/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite'))
@@
 <!-- Genes were considered to be significantly differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2.  -->
+
+## PubMed Citation Search
+
+The R package easyPubMed was used to search for 1000 randomly chosen genes present in scEiaD for both gene name alone and gene name plus "AND Retina" with the get_pubmed_ids function.
+We then used the returned Entrez identifier to pull the abstract with the fetch_pubmed_data function so we could extract the PubMed ID (PMID).
+This was repeated with the the genes identified as being well supported differentially expressed for ocular cell types.
+To assess whether the difference in distributions was statistically different, we used the base R t.test function.
 
@@
 
-```{r, fig.height=14, fig.width=12, fig.cap = scEiaD_data_overview_cap}
+```{r, fig.height=14, fig.width=12, fig.cap = scEiaD_data_overview_cap, echo = FALSE}
 
@@
 In contrast, starting from the loading page of Spectacle, it takes approximately eleven seconds to plot the cell type metadata for a several thousand cells across across a single study (tested on 2022-08-26).
-Within the same time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies (`r fig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')d`).
+Within the same time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies (`r fig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`d).
 The speed of the PLAE app is also comparable to the CELLxGENE browser serving the Tabula Sapiens resource while providing additional functionality.
@@
 
-Genes which have little existing knowledge can be considered
-
-Of course other possibilities for ranking genes are possible.
-We show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`b and c how you can directly use our powerful web app to select candidate bipolar cell differentially expressed genes from the differential testing table and then quickly display the expression pattern of those genes in a compact heatmap visualization.
+Genes with few citations can be considered as novel markers.
+We see, for example, Onecut3 appears to well separate the horizontal cells and has few PubMed citations.
+To assess whether there are any other good candidates with few citations, we produced another heatmap (`r supFig_cap(name = 'undersupported_genes', display = 'cite')`).
+Most of these genes are differentially expressed across more than one cell type.
+We hand selected C1ql2 and Cartpt (amacrine), Frzb and Slitrk2 (Muller glia), and Onecut3 (horizontal) and plotted these five genes in the boxplot view to confirm specificity across the full scEiaD database (`r supFig_cap(name = 'lightlyCited_candidates', display = 'cite')`).
+All five look to be specific to a cell type, except for Frzb which also is expressed in the RPE.
+
+Other possibilities for ranking genes are possible.
+We show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`b and c how one can directly use our powerful web app to select candidate bipolar cell differentially expressed genes from the differential testing table and then quickly display the expression pattern of those genes in a compact heatmap visualization.
 The wealth of visualization options in plae allow for quick prototyping and hypothesis testing of gene expression patterning across a large number of cells and studies.
@@
 ComplexHeatmap::draw(consist_diff_hm, padding = unit(c(0, 0, 1.5, 0), "in"), row_title = "Marker Genes")
-
+queries %>% select(Symbol, Base, `mean log2FC`, `mean padj`, pm_query0, query0, pm_query2, query2) %>% write_csv("supplemental_files/consistently_differentially_expressed_diff_table.csv.gz")
 #toc()
@@
 num3 <- meta_filter %>% filter(Organ == 'Eye') %>% pull(CellType) %>% unique() %>% length()
-sup_scEiaD_data_overview_cap <- supFig_cap(name = '_supscEiaD_data_overview', caption = glue::glue("Distribution of tissues and cell types in scEiaD.
+sup_scEiaD_data_overview_cap <- supFig_cap(name = 'sup_scEiaD_data_overview', caption = glue::glue("Distribution of tissues and cell types in scEiaD.
 A.
@@
 
-```{r undersupported_genes, fig.height=48, fig.width=8, echo = FALSE}
-cone_undersupported <- queries %>% filter(query2 <= 1) %>% arrange(-`mean log2FC`) %>% filter(!grepl("Endo|Macro|Micro|Endo", Base)) %>% group_by(Symbol) %>% summarise(Count = n(), Base = paste0(Base, collapse = ', '), l2FC = mean(`mean log2FC`)) %>% filter(Count == 1) %>% filter(Base == 'Rod') %>% arrange(-l2FC)
-
-input <- list()
-input$exp_plot_groups <- c('study_accession')
-input$exp_plot_facet <- c('CellType_predict')
-input$exp_plot_genes <- scEiaD_2020_v01 %>% tbl('Genes') %>% collect() %>% mutate(Symbol = str_extract(Gene, '.* \\(') %>% gsub(' \\(','',.)) %>% filter(Symbol %in% cone_undersupported$Symbol) %>% pull(Gene)
-input$exp_plot_ylab <- 'Mean log2(Counts + 1)'
-input$exp_plot_col_num <- 16
-input$exp_filter_min_cell_number <- 50
-input$flip_facet <- FALSE
-input$figure <- FALSE
-input$figure_bloop <- TRUE
-input$figure_fovea <- FALSE
-
-make_exp_plot(input, scEiaD_2020_v01, meta_filter)
-
-```
+```{r undersupported_genes, fig.height=23, fig.width=8, echo = FALSE, fig.cap = undersupported_genes_cap}
+undersupported_genes_cap <- supFig_cap(name = 'undersupported_genes', caption = 'Well supported genes with a mean log2FC > 5 and zero pubmed citation hits for [gene] AND Retina search.')
+
+g_lightlyCited <- queries %>%  filter(`mean log2FC` > 5, query2 < 1) %>% filter(Gene %in% top_genes$Gene) %>% group_by(Base) %>% top_n(12, `mean log2FC`) %>% pull(Gene)
+
+consist_diff_lightlyCited_long <- scEiaD_2020_v01 %>%
+  tbl("diff_testing") %>%
+  filter(Gene %in% g_lightlyCited) %>%
+  collect() %>%
+  filter(Base %in% top_genes$Base,
+         !Base %in% c('Early RPC','Late RPC'), 
+         Group == 'CellType (Predict)',
+         Against == 'All') %>%
+  select(Gene, Organism, log2FoldChange, Base)
+
+consist_diff_lightlyCited_wide <- consist_diff_lightlyCited_long %>% 
+  mutate(GO = paste0(Base, '_', Organism)) %>% 
+  select(-Organism, -Base) %>% 
+  pivot_wider(values_from=log2FoldChange, names_from = GO)
+consist_diff_lightlyCited_celltype <- colnames(consist_diff_lightlyCited_wide[,-1])
+consist_diff_lightlyCited_wide <- consist_diff_lightlyCited_wide %>% data.frame()
+row.names(consist_diff_lightlyCited_wide) <- consist_diff_lightlyCited_wide$Gene
+
+consist_diff_lightlyCited_wide <- consist_diff_lightlyCited_wide[,-1]
+
+consist_diff_lightlyCited_organism <- str_extract(consist_diff_lightlyCited_celltype, '_\\w+ \\w+') %>% gsub('_','',.)
+consist_diff_lightlyCited_cts  <- str_extract(consist_diff_lightlyCited_celltype, '.*_') %>% gsub('_','',.)
+consist_diff_lightlyCited_genes <- row.names(consist_diff_lightlyCited_wide) %>% gsub(' \\(.*','',.)
+
+consist_diff_lightlyCited_base <- consist_diff_lightlyCited_genes %>% as_tibble() %>% 
+  rename(Symbol = value) %>% 
+  left_join(top_genes %>% 
+              mutate(Symbol = gsub(' \\(.*','',Gene)) %>% 
+              select(Symbol, Base) %>% 
+              unique()) %>% 
+  pull(Base)
+
+query0_lightlyCited <- cbind(consist_diff_lightlyCited_genes, consist_diff_lightlyCited_base) %>% as_tibble() %>%
+  left_join(queries, by = c('consist_diff_lightlyCited_genes' = 'Symbol', 'consist_diff_lightlyCited_base' = 'Base')) %>%
+  mutate(citation = case_when(
+    query0 > mean(pmid5 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Many",
+    query0 > median(pmid5 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Some",
+    TRUE ~ "Few")) %>%
+  pull(citation)
+
+# pmid2 and pmid3 is "AND RETINA"
+# (t.test(pmid2 %>% map(function(x) sum(!is.na(x))) %>% unlist(), pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist()))$p.value
+query2_lightlyCited <- cbind(consist_diff_lightlyCited_genes, consist_diff_lightlyCited_base) %>%
+  as_tibble() %>%
+  left_join(queries, by = c('consist_diff_lightlyCited_genes' = 'Symbol', 'consist_diff_lightlyCited_base' = 'Base')) %>%
+  mutate(citation = case_when(
+    query2 > mean(pmid6 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Many",
+    query2 > median(pmid6 %>% map(function(x) sum(!is.na(x))) %>% unlist()) ~ "Some",
+    TRUE ~ "Few")) %>%
+  pull(citation)
+#rand_col <-  circlize::colorRamp2(c('0', '1', '2'), c("white", "gray", "black"))
+citation_col <-c('white','gray','black')
+names(citation_col) <-  c('Few','Some','Many')
+
+consist_diff_lightlyCited_hm <- ComplexHeatmap::Heatmap(consist_diff_lightlyCited_wide,
+                                                        column_labels = consist_diff_lightlyCited_organism,
+                                                        column_split = consist_diff_lightlyCited_cts,
+                                                        row_split = consist_diff_lightlyCited_base,
+                                                        show_row_dend = FALSE,
+                                                        column_title_rot = 90,
+                                                        name = 'log2FC', use_raster = TRUE,
+                                                        row_title_rot = 0) +
+  ComplexHeatmap::Heatmap(as.character(query0_lightlyCited),
+                          row_split = consist_diff_lightlyCited_base,
+                          col = citation_col,
+                          show_heatmap_legend = TRUE,
+                          column_labels = "Pubmed Gene Citations",
+                          column_names_side = 'top') +
+  ComplexHeatmap::Heatmap(as.character(query2_lightlyCited),
+                          row_split = consist_diff_lightlyCited_base,
+                          col = citation_col,
+                          show_heatmap_legend = FALSE,
+                          row_labels = consist_diff_lightlyCited_genes,
+                          name = 'Citation Count',
+                          column_labels = "Pubmed Gene + \"Retina\" Citations",
+                          column_names_side = 'top')
+ComplexHeatmap::draw(consist_diff_lightlyCited_hm, padding = unit(c(0, 0, 1.5, 0), "in"), row_title = "Marker Genes")
+
+```
+
+```{r lightlyCited_candidates, echo = F, message = F, fig.cap = lightlyCited_candidates_cap}
+lightlyCited_candidates_cap <- supFig_cap(name='lightlyCited_candidates', caption = 'Five hand-selected genes selected from Supplementary Figure X as having fairly unique expression to a celltype while having few pubmed citations.
+Plotted as expression/boxplots to confirm specificity at the study level.')
+knitr::include_graphics('fig/lightlyCited_exp_plot.png')
+
+```
+
 ```{r PLAE_screenshots_supplemental, echo = F, message = F, fig.cap = PLAE_screenshots_supplemental_cap}

### 2022-09-01 src/diff_data_pull.R

@@ -9,10 +9,10 @@ library(glue)
 library(tictoc)
 
 tic()
-data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
-data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '/Volumes/McGaughey_S/data/scEiaD//'
+#data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
+data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
 
 
 library(pool)

### 2022-09-01 src/make_exp_plot.R

@@ -144,7 +144,7 @@ make_exp_plot <- function(input, db, meta_filter){
       ungroup() %>% 
       ggplot(aes(x=!!as.symbol(input$exp_plot_facet), y = !!as.symbol(input$exp_plot_ylab), color = !!as.symbol(grouping_features), group = CellType_predict)) +
       geom_boxplot(color = 'black', outlier.shape = NA) +
-      ggbeeswarm::geom_quasirandom(size = 4, groupOnX = TRUE) +
+      ggbeeswarm::geom_quasirandom(size = 1, groupOnX = TRUE) +
       cowplot::theme_cowplot() +
       theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
       scale_radius(range=c(2, 6)) +

### 2022-09-08 manuscript_00.Rmd
text updates

@@
 However each data deposit is independently processed, which makes cross study queries impossible.
-We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space [@swamyBuildingMegaSinglecell2021].
+We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space, the Single Cell Eye in a Disk (scEiaD) [@swamyBuildingMegaSinglecell2021].
 
@@
 "Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function [@loveModeratedEstimationFold2014a].
+
+We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping (see Methods).
+This created a counts matrix which has statistical properties approximating "bulk" RNA seq experiment matrices.
+This allowed the use of the more mature bulk RNA-seq tooling.
+This approach has also been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments[@squairConfrontingFalseDiscoveries2021].
+Once the pseudo-bulk matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
+We tested each species (Human, Macaque, Mouse) separately.
+Chicken was not included as we had an insufficient number of independent studies to conduct differential testing.
+The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential gene expression.
+After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g.
+cones versus all remaining cell types) or in a pair-wise manner (cone against rod).
+Unless stated otherwise, we will be discussing the results of the "against all" contrast for the remainder of the paper.
+
 <!-- Genes were considered to be significantly differentially expressed if they have a padj less than $1 x 10^-5 and a abs(log2FoldChange) > 2.  -->
@@
 As before, the pipeline's scVI's integration parameters were chosen rigorously by testing inegration performance across a wide range of latent dimensions and number of highly variable genes in our scPOP framework [@swamyBuildingMegaSinglecell2021].
-Each study in the resource were hand assessed to determine whether they had any unusual properties in regards to machine learnt cell type proportions or position in the 2D UMAP space.
-The small number of outlier studies were removed from the resource.
-
-After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (`r fig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`).
+Each study in the resource was hand assessed to determine whether they had any unusual properties in regards to machine learnt cell type proportions or position in the 2D UMAP space and a small number of outlier studies were removed from the resource.
+
+After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (`r supFig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`).
 The scEiaD database contains `r num2` non-ocular human and mouse tissues and `r num1` ocular tissues across human, chicken, macaque, and mouse (`r fig_cap(name = 'scEiaD_data_overview', display = 'cite')`).
@@
   mutate(Test = factor(Test, levels = row.names(timings))) %>% 
-  mutate(Text = case_when(is.na(value) ~ 'Not possible',
+  mutate(Text = case_when(is.na(value) ~ 'Not\npossible',
                           TRUE ~ '')) %>% 
@@
 The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has collated a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app.
-PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence in a consistent manner, and, crucially, full availability of the underlying data for outside usage and integrating the independent studies to deliver cross study comparisons.
+PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence in a consistent manner, integrating the independent studies to deliver cross study comparisons, and, crucially, providing full availability of the underlying data for outside usage.
 
@@
 This allows for gene queries to be constructed and analyzed across studies.
-In contrast, Spectacle, as of March 2022, is a compilation of study-level datasets and thus you cannot run queries across studies.
-Spectacle contains no non-ocular dataset and the cell types labels have not been curated or harmonized between studies (for example different studies can use MG, Mueller, Müller Glia, or other variations to refer to the same cell type).
+In contrast, Spectacle, as of August 2022, is a compilation of study-level datasets and which makes queries across studies impossible.
+Spectacle does not contain any non-ocular datasets and the cell type labels have not been curated or harmonized between studies (for example different studies can use MG, Mueller, Müller Glia, or other variations to refer to the same cell type).
 Finally, in contrast to PLAE, Spectacle does not make their datasets available for download in any form.
@@
 A sqlite database and its indexing functionality allows large amounts of data to be stored for quick retrieval with minimal loading time and memory usage.
-The major downside is that this approach requires a large amount of disk storage, which can make local (on personal computer) usage difficult; the scEiaD v2022-03-22 sqlite database is about 368 GB.
+The major downside is that this approach requires a large amount of disk storage, which can make local (on a personal computer) usage difficult; the scEiaD v2022-03-22 sqlite database is about 368 GB.
 Second, the visualization code in PLAE (e.g.
@@
 In contrast, starting from the loading page of Spectacle, it takes approximately eleven seconds to plot the cell type metadata for a several thousand cells across across a single study (tested on 2022-08-26).
-Within the same time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies (`r fig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`d).
+Within the same time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies (`r supFig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`d).
 The speed of the PLAE app is also comparable to the CELLxGENE browser serving the Tabula Sapiens resource while providing additional functionality.
@@
 PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values.
-So thus a user can see Crx expression across over a million cells and separately see cell type labels for those same million cells.
+So thus, a user can see *Crx* expression across over a million cells and separately see cell type labels for those same million cells.
 The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells.
@@
 We also offer a Dot Plot view which can display a large number of genes in a space efficient manner.
-This allows a users to see whether a set of genes has related (or not) expression patterns across multiple cell types (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`b and c).
+This allows users to see whether a set of genes has related (or not) expression patterns across multiple cell types (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`b and c).
 
@@
 A.
-UMAP view showing CRX expression with cell type labels, B.
+UMAP view showing *Crx* expression with cell type labels, B.
 Information table showing expression of Crx across celltypes and organism.
@@
 
-We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping (see Methods).
-This created a counts matrix which has statistical properties approximating "bulk" RNA seq experiment matrices.
-This allowed the use of the more mature bulk RNA-seq tooling.
-This approach has also been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments[@squairConfrontingFalseDiscoveries2021].
-Once the pseudo-bulk matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
-We tested each species (Human, Macaque, Mouse) separately.
-Chicken was not included as we had an insufficient number of independent studies to conduct differential testing.
-The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential gene expression.
-After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g.
-cones versus all remaining cell types) or in a pair-wise manner (cone against rod).
-Unless stated otherwise, we will be discussing the results of the "against all" contrast for the remainder of the paper.
-
 Cell type marker genes can be proposed on the basis of existing knowledge about their function and on high expression differences when comparing to other cell types.
 We can further narrow down the list of markers by using the high diversity of studies and organisms in scEiaD to propose a set of community supported cell type markers.
-We first identify a set of well supported cell types which we define as cell types which have two or more independent studies across both human and mouse (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
+We first identify a set of well supported cell types which we define as cell types that are detected in two or more independent studies across both human and mouse (`r supTab_cap(name = 'well_supported_celltypes_tab', display = 'cite')`).
 These cell types were then assessed to identify differentially expressed genes which met the following criteria across human, mouse, and macaque: 1.
@@
 The cell types are arranged by the columns by how related the expression patterns are.
-We see how the differentiated cell types of the retina are grouped together with the exception of the Müller glia, which are near the RPC from which they derive from.
-The photoreceptors are also next to each other.
+We see how the differentiated cell types of the retina are grouped together with the exception of the Müller glia, which are near the RPC from which they are derived.
 
@@
 Genes with few citations can be considered as novel markers.
-We see, for example, Onecut3 appears to well separate the horizontal cells and has few PubMed citations.
+We see, for example, *Onecut3* appears to well separate the horizontal cells and has few PubMed citations.
 To assess whether there are any other good candidates with few citations, we produced another heatmap (`r supFig_cap(name = 'undersupported_genes', display = 'cite')`).
 Most of these genes are differentially expressed across more than one cell type.
-We hand selected C1ql2 and Cartpt (amacrine), Frzb and Slitrk2 (Muller glia), and Onecut3 (horizontal) and plotted these five genes in the boxplot view to confirm specificity across the full scEiaD database (`r supFig_cap(name = 'lightlyCited_candidates', display = 'cite')`).
-All five look to be specific to a cell type, except for Frzb which also is expressed in the RPE.
-
-Other possibilities for ranking genes are possible.
+We hand selected *C1ql2* and *Cartpt* (amacrine), *Frzb* and *Slitrk2* (Muller glia), and *Onecut3* (horizontal) and plotted these five genes in the boxplot view to confirm specificity across the full scEiaD database (`r supFig_cap(name = 'lightlyCited_candidates', display = 'cite')`).
+All five look to be specific to a cell type, except for *Frzb* which also is expressed in the RPE.
+
+Other strategies for ranking genes are possible.
 We show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`b and c how one can directly use our powerful web app to select candidate bipolar cell differentially expressed genes from the differential testing table and then quickly display the expression pattern of those genes in a compact heatmap visualization.
-The wealth of visualization options in plae allow for quick prototyping and hypothesis testing of gene expression patterning across a large number of cells and studies.
+The wealth of visualization options in plae allow for quick prototyping and hypothesis testing of gene expression patterning across many cells and studies.
 
@@
 Columns are split by the cell types.
-More red is a higher log2 fold change of the gene - cell type relative to all other cell types and blue is more negative.
+More intense red is a higher log2 fold change of the gene - cell type relative to all other cell types and blue is more negative.
 Relative number of citations are shown for the gene or gene " and retina" when searching against PubMed compared to 1000 randomly chosen genes.')
@@
 We filter to genes with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two or less than negative one (`r fig_cap(name = 'neuro', display = 'cite')`B and C).
-We find one gene, ZBTB38, that meets this criteria and drops in expression when comparing RPC to neurogenic.
+We find one gene, *Zbtb38*, that meets this criteria and drops in expression when comparing RPC to neurogenic.
 We find `r up %>% length()` genes that increase in expression from RPC to neurogenic, `r up[!up %in% g_lu] %>% length()` of which are not previously identified by Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`C).
-We replicate `r up[up %in% g_lu] %>% str_extract(., '\\w+') %>% paste(., collapse = ', ') %>% str_to_title()` from Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`B) as differentially expressed between neurogenic and RPC.
+We replicate *`r up[up %in% g_lu] %>% str_extract(., '\\w+') %>% paste(., collapse = ', ') %>% str_to_title()`* from Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`B) as differentially expressed between neurogenic and RPC.
 Futhermore we show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`a how this figure can be replicated directly in our web app.
@@
 Yan et al.
-previously proposed Prph2 and Rs1 as markers to distinguish macula and peripheral located human cones[@yanCellAtlasHuman2020a] and Peng et al.
-proposed Calb1 and Gngt1 as cone region markers in macaque [@pengMolecularClassificationComparative2019a].
-
-Our differential testing found that while Prph2, Rs1, and Gngt1 had expression differences matching that the findings Yan et al.
+previously proposed *Prph2* and *Rs1* as markers to distinguish macula and peripheral located human cones[@yanCellAtlasHuman2020a] and Peng et al.
+proposed *Calb1* and *Gngt1* as cone region markers in macaque [@pengMolecularClassificationComparative2019a].
+
+Our differential testing found that while *Prph2*, *Rs1*, and *Gngt1* had expression differences matching that the findings Yan et al.
 and Peng et al.
 published, they failed to replicate in our test with pvalue and padj being above 0.1 for all three.
-In contrast Calb1 did replicate strongly with the sixth lowest padj (0.00721) across our entire test and a log2 fold change of 2.87.
-We find six more genes with a padj < 0.01: Pde6g, Akap12, Hspb1, Gadd45b, Cc2d2a, and RP1l1 (`r fig_cap(name = 'region_fig', display = 'cite')` and `r supTab_cap(name = 'region_table', display = 'cite')`).
+In contrast *Calb1* did replicate strongly with the sixth lowest padj (0.00721) across our entire test and a log2 fold change of 2.87.
+We find six more genes with a padj < 0.01: *Pde6g*, *Akap12*, *Hspb1*, *Gadd45b*, *Cc2d2a*, and *RP1l1* (`r fig_cap(name = 'region_fig', display = 'cite')` and `r supTab_cap(name = 'region_table', display = 'cite')`).
 None of these candidate cone region marker genes display regionally different expression in rods (`r supFig_cap(name = 'rod_region_fig', display = 'cite')`).
@@
 More dramatically, Yan et al.
-have proposed over sixty types of mouse amacrine cell types [@yanMouseRetinalCell2020a].
+have proposed over sixty different mouse amacrine cell types [@yanMouseRetinalCell2020a].
 While we have found it fairly straightforward to use machine learning to transfer the broad (e.g.
@@
 This democratizes access to over a million cells across `r meta_filter$study_accession %>% unique() %>% length()` studies to anyone with a web browser.
-We recognize that while we have built a powerful visualization platform we cannot anticipate all uses and for those researchers who wish to make custom queries of the data, we provide the data underlying plae in a wide variety of formats, including anndata and Seurat objects for each study, count tables, cell level metadata, and the DESeq2 based differential testing across three species.
-We provide a short example of how a custom query could work by running a custom differential test between macula and peripheral derived human cones and demonstrate how this can both validate existing proposed genes and suggest new ones.
+We recognize that while we have built a powerful visualization platform we cannot anticipate all uses and for those researchers who wish to make custom queries of the data, we provide the data underlying PLAE in a wide variety of formats, including anndata and Seurat objects for each study, count tables, cell level metadata, and the DESeq2 based differential testing across three species.
+We provide a short example of how a novel query could work by running a custom differential test between macula and peripheral derived human cones and demonstrate how this can both validate existing proposed genes and suggest new ones.
 
@@
 This approach enables a small group to produce an atlas which rivals the size from massive multi-group consortiums.
-Furthermore the inherent variability between the data from independent groups can be used as a feature to get insight into whether a gene cell type signal is stable across an entire community's datasets.
+Furthermore the inherent variability between the data from independent groups can be used as a feature to get insight into whether a gene-cell type signal is stable across an entire community's datasets.
 We briefly demonstrate how the high study diversity in scEiaD can be used to validate proposed RPC to neurogenic gene expression changes.

### 2022-09-09 manuscript_00.Rmd
added params

@@
 
+After evaluating the performance of the new dataset with our scPOP package we selected the following parameters for the integration with scANVI: 15 latent dimensions, 4000 highly variable genes, and 5 epochs.
+The UMAP was built with a min_dist of 0.1 and the clustering used 50 nearest neighbors.
+
 ## Cross species gene alignment
@@
 
-The code base for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is XXXXXXX.
+The code base for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is 8f2c04b.
 There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app.
-The code base for the web app (along with local installation instructions) is available at github.com/davemcg/scEiaD/plaeApp; the commit corresponding to this manuscript is XXXXXXXXXXXX.
+The code base for the web app (along with local installation instructions) is available at github.com/davemcg/scEiaD/plaeApp.
 

### 2022-09-09 manuscript_00.Rmd
citation fix

@@
 #data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
+data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'
 # setup caption-ing
@@
 #data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
+data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'
+
 scEiaD_2020_v00 <- pool::dbPool(RSQLite::SQLite(), dbname =  glue('{data_dir_vGiga}/MOARTABLES__anthology_limmaFALSE___5000-transform-counts-universe-batch-scVIprojectionSO-8-0.1-50-5.sqlite'))
@@
 
-The retina is composed of six major cell populations: the photoreceptors, horizontal, bipolar, amacrine, ganglion, and non-neuronal cells [@masland_cell_2011].
-These six populations can be further divided into dozens of cell types [@yan_cell_2020].
+The retina is composed of six major cell populations: the photoreceptors, horizontal, bipolar, amacrine, ganglion, and non-neuronal cells [@maslandCellPopulationsRetina2011].
+These six populations can be further divided into dozens of cell types [@yanCellAtlasHuman2020a].
 Furthermore, human disease of the eye can derive from damage to specific cell types.

### 2022-09-09 manuscript_00.Rmd
latest version

@@
 #data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
-data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
-data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'
+data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
+#data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'
 # setup caption-ing
@@
 #data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
-data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
-data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'
+data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
+#data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'
 

### 2022-09-12 manuscript_00.Rmd
update

--- 2022-09-09 manuscript_00.Rmd
@@
 RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encomopassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 37 publications.
-To demonstrate the value of this unified pan-eye dataset, we replicate known neurogenic and cone macula markers as well as propose six????
-new cone human region markers.
+To demonstrate the value of this unified pan-eye dataset, we replicate known neurogenic and cone macula markers as well as propose six new cone human region markers.
 CONCLUSION: The PLAE web application provides the eye community a powerful and quick means to test hypotheses related to gene expression across a highly diverse, community-derived database."
@@
 
-## Data reproducibility
+## Data reproducibility and availability
 
@@
 The code base for the web app (along with local installation instructions) is available at github.com/davemcg/scEiaD/plaeApp.
+Data is available for download at plae.nei.nih.gov (click "Data") and the Seurat object and metadata for the resource has been deposited at Zenodo with accession 10.5281/zenodo.7071682.
 

### 2022-09-12 manuscript_00.Rmd
fix methods diff section with semi-dup text

@@
 
-The aggregateAcrossCells function from the scuttle R package was used to aggregate gene counts across species, study accession, and category where category was either Cluster, CellType, or CellType (Predict) assignments for a cell [@mccarthyScaterPreprocessingQuality2017].
+We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping using the aggregateAcrossCells function from the scuttle R package [@mccarthyScaterPreprocessingQuality2017].
+This created a counts matrix which has statistical properties approximating "bulk" RNA seq experiment matrices.
+This allowed the use of the more mature bulk RNA-seq tooling.
+This approach has also been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments[@squairConfrontingFalseDiscoveries2021].
 If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test.
@@
 "Cone") against all remaining or in a pair wise manner ("Cone" vs "Rod") with the DESeq2 "results" function [@loveModeratedEstimationFold2014a].
-
-We used a "pseudo-bulk" based approach where the counts for a category (CellType, Cluster, or CellType (Predict)) were summed for each category - organism - study grouping (see Methods).
-This created a counts matrix which has statistical properties approximating "bulk" RNA seq experiment matrices.
-This allowed the use of the more mature bulk RNA-seq tooling.
-This approach has also been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments[@squairConfrontingFalseDiscoveries2021].
-Once the pseudo-bulk matrix was constructed, we use the DESeq2 differential gene expression analysis software to identify differentially expressed genes.
 We tested each species (Human, Macaque, Mouse) separately.
-Chicken was not included as we had an insufficient number of independent studies to conduct differential testing.
-The model used for DESeq2 was "~study_accession + category" in order to regress out study specific differential gene expression.
-After running these tests we extracted differentially expressed genes with two types of contrasts: either against all remaining categories (e.g.
-cones versus all remaining cell types) or in a pair-wise manner (cone against rod).
-Unless stated otherwise, we will be discussing the results of the "against all" contrast for the remainder of the paper.
 

### 2022-09-12 manuscript_00.Rmd
text update

@@
 Very briefly, we collected the raw fastq files and quantified with the kallisto / bustools system [@melstedBarcodeUMISet2019; @melstedModularEfficientPreprocessing2019].
-The quantified counts are merged into one matrix that is used for the downstream integration with scvitools create a batch corrected lower dimensional space.
+The quantified counts are merged into one matrix that is used for the downstream integration with scvitools (version 0.13.0) create a batch corrected lower dimensional space.
 The raw counts and scANVI batch corrected lower dimensional space is sent into Seurat for quality control, clustering, and 2D UMAP visualization.
@@
 
-The code base for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is 8f2c04b.
-There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app.
+The code base for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is #a99dced There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app.
 The code base for the web app (along with local installation instructions) is available at github.com/davemcg/scEiaD/plaeApp.

### 2022-09-23 supplemental_files/pseudobulk_cone_region.Rmd
small updates

--- 2022-08-02 supplemental_files/cone_fovea_vs_peripheral_diff_testing.Rmd
@@
 ---
-title: "Test Fovea vs Peripheral Cones"
+title: "Pseudobulk Differential Testing of Fovea vs Peripheral Cones"
 output: html_notebook
@@
 
-# Load
 ```{r}
@@
 library(org.Hs.eg.db)
-#system('wget -O ~/Downloads/scEiaD_all_seurat_v3.Rdata https://hpc.nih.gov/~mcgaugheyd/scEiaD/2022_03_22/scEiaD_all_seurat_v3.Rdata)
+#system('wget -O ~/data/scEiaD_2022_02/scEiaD_all_seurat_v3.Rdata https://hpc.nih.gov/~mcgaugheyd/scEiaD/2022_03_22/scEiaD_all_seurat_v3.Rdata)
 # the -O renames the file and puts it in the ~/data/scEiaD directory
-load('~/Downloads/scEiaD_all_seurat_v3.Rdata')
+load('~/data/scEiaD_2022_02/scEiaD_all_seurat_v3.Rdata')
 ```
+
 ## Compare macula cones vs peripheral cones
@@
 ```
-
 Yes, we have several human studies with macula / peripheral cone cells.
@@
 Let's demonstrate how we can quickly run a conservative differential expression test the leverages the many studies we have.
-The "traditional" scRNA diff tests often return insane numbers of differentially expressed genes, which is  annoying as then you have to plot each result individually to confirm that it does not look wacky.
+The "traditional" scRNA diff tests often return insane numbers of highly differentially expressed genes (the test we ran aboves returns many many genes with a padj < 0.05), which is highly annoying as then you really have to plot each result individually to confirm that it does not look wacky.
 We are going to build a "pseudo bulk" experiment here where we sum all the counts by gene, cell type (in this case we are doing macula cones *versus* peripheral cones), and study.
@@
 
-scEiaD is based on the Ensembl gene id (like ENSG00000185527).
-That's not very friendly to humans so I added a bit of code to link the Ensembl ID to the gene name and plot the top differentially expressed genes (padj < 0.05).
-
+scEiaD is built on the Ensembl gene id (for examples ENSG00000185527).
+That's not very friendly to humans so I added a bit of code to link the Ensembl ID to the gene name and show the top 16 diff genes (padj < 0.05) as a table.
 ```{r}
@@
 
-deseq_res %>% as_tibble(rownames = 'name') %>%  
+region_table <- deseq_res %>% as_tibble(rownames = 'name') %>%  
   left_join(symbols %>% enframe()) %>% 
   dplyr::rename(Ensembl = name, Gene = value) %>% 
+  relocate(Gene)
+
+region_table %>% 
   filter(padj < 0.05) %>% 
@@
 
+We use the "ScaleData" function from Seurat to zero center and scale the counts for each dataset
 #### PDE6G 
@@
 
-#### HSPB1
-Now we see the opposite side - the top gene that is higher in peripheral vs macula. Again looks fairly consistent across all the studies.
+#### All genes
 ```{r}
-VlnPlot(scEiaD__subset, features = c('ENSG00000099860'))
-
-VlnPlot(scEiaD__subset, c('ENSG00000099860'), split.by = 'regionCT', group.by='study_accession', ) + aes(color=scEiaD__subset$regionCT)
+for (i in region_table %>% 
+  filter(padj < 0.05) %>% 
+  pull(Ensembl)){
+  #print(i)
+  gene_name <- region_table %>% filter(Ensembl == i) %>% pull(Gene)
+  print(VlnPlot(scEiaD__subset, i, split.by = 'retina_region', group.by='study_accession', log = TRUE) + ggtitle(gene_name))
+}
 ```
 
-#### PRPH2
-Yan et al (https://www.nature.com/articles/s41598-020-66092-9) proposed PRPH2 as a cone fovea/peripheral specific gene.
-We can see whether that effect is still shown across multiple studies in scEiaD.
-We see the padj is not impressive in our diff test.
-
+# Output and Session Info
 ```{r}
-# first grab the ensembl id
-deseq_res %>% as_tibble(rownames = 'name') %>%  
-  left_join(symbols %>% enframe()) %>% 
-  dplyr::rename(Ensembl = name, Gene = value) %>% 
-  filter(Gene == 'PRPH2')
+sessionInfo()
+save(scEiaD__subset,region_table, deseq_res, file = 'pseudoBulk_cone_region_files.Rdata' )
+write_csv(region_table, file = 'pseudoBulk_cone_region_table.csv.gz' )
 ```
-
-The plots certainly do not DISPROVE the Yan et al hypothesis - you still see that the trend is (mostly) the same across all the studies.
-However the p value is likely low for because the effect size (fold change) is modest across most studies (and in one case, the effect is opposite).
-As a exercise to the reader, you could re-do the pseudobulk test, removing SRP194595 and seeing whether padj value you get returned is substantially different.
-
-```{r}
-VlnPlot(scEiaD__subset, features = c('ENSG00000112619'))
-
-VlnPlot(scEiaD__subset, c('ENSG00000112619'), split.by = 'regionCT', group.by='study_accession', ) + aes(color=scEiaD__subset$regionCT)
-
-```
-
-# Save our file
-```{r}
-save(scEiaD__subset, file = 'scEiaD_fovea_vs_peripheral_cone.seuratV3.Rdata')
-```
-
-```{r}
-devtools::session_info()
-```

### 2022-09-23 supplemental_files/pseudobulk_cone_region.Rmd
pub update

@@
 We use the "ScaleData" function from Seurat to zero center and scale the counts for each dataset
+
 #### PDE6G 
@@
 ```{r}
-for (i in region_table %>% 
-  filter(padj < 0.05) %>% 
-  pull(Ensembl)){
+for (i in region_table %>% filter(padj < 0.05) %>% pull(Ensembl)){
   #print(i)
-  gene_name <- region_table %>% filter(Ensembl == i) %>% pull(Gene)
-  print(VlnPlot(scEiaD__subset, i, split.by = 'retina_region', group.by='study_accession', log = TRUE) + ggtitle(gene_name))
+  gene <- region_table %>% filter(Ensembl == i) %>% pull(Gene)
+  print(VlnPlot(scEiaD__subset, i, split.by = 'retina_region', group.by='study_accession', log = TRUE)  + ggtitle(gene))
 }

### 2023-02-16 manuscript_00.Rmd
text updates for resubmission

@@
   - Vinay Swamy:
-      institute: bg
+      institute: cu
   - Zachary Batz:
@@
 institute:
+  - cu: Department of Biomedical Informatics, Columbia University
   - bg: Bioinformatics Group, Ophthalmic Genetics & Visual Function Branch, National Eye Institute, National Institutes of Health
   - nnrl: Neurobiology, Neurodegeneration & Repair Laboratory, National Eye Institute, National Institutes of Health
-  - mgog: Medical Genetics and Ophthalmic Genomics Unit, National Eye Institute, National Institutes of Health
 
@@
 METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it.
-RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encomopassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 37 publications.
+RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encompassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 37 publications.
 To demonstrate the value of this unified pan-eye dataset, we replicate known neurogenic and cone macula markers as well as propose six new cone human region markers.
@@
 #data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
-#data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
-#data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
+data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'
 # setup caption-ing
@@
 #data_dir_vPLAE <- '/Volumes/McGaughey_S/data/scEiaD_2022_02/'
-data_dir_vGiga <- '~/data/scEiaD_gigascience/'
-data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
-#data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
-#data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'
+#data_dir_vGiga <- '~/data/scEiaD_gigascience/'
+#data_dir_vPLAE <- '~/data/scEiaD_2022_02/'
+data_dir_vGiga <- '/Volumes/Thunder/data/scEiaD/'
+data_dir_vPLAE <- '/Volumes/Thunder/data/scEiaD_2022_02/'
 
@@
 
-The modern era of single cell transcriptomics ha been marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment.
+The modern era of single cell transcriptomics has been marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment.
 This approach was pioneered by Macoksko et al.
-and enabled the profiling of gene expression in 44,808 single cells in the adult mouse retina.
-[@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a].
+and enabled the profiling of gene expression in 44,808 single cells in the adult mouse retina [@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a].
 The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since Macosko publication.
@@
 
-The scEiaD pipeline is largely the same as in Swamy et al.
+Our full scEiaD pipeline was published previously in Swamy et al.
 [@swamyBuildingMegaSinglecell2021].
-Very briefly, we collected the raw fastq files and quantified with the kallisto / bustools system [@melstedBarcodeUMISet2019; @melstedModularEfficientPreprocessing2019].
-The quantified counts are merged into one matrix that is used for the downstream integration with scvitools (version 0.13.0) create a batch corrected lower dimensional space.
-The raw counts and scANVI batch corrected lower dimensional space is sent into Seurat for quality control, clustering, and 2D UMAP visualization.
-We use our previously published xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamyBuildingMegaSinglecell2021].
-
-The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300.
-Second, we use the DecontX algorithm[@yangDecontaminationAmbientRNA2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination on a per-study basis.
+It is composed of three Snakefiles which perform three core workflow tasks[@molderSustainableDataAnalysis2021].
+The "SnakeQUANT" pipeline takes the raw fastq input and produces a gene by cell count matrix with kallisto and bustools [@melstedBarcodeUMISet2019; @melstedModularEfficientPreprocessing2019].
+This matrix is used in the "SnakePOP" pipeline to identify the optimal parameters for the scANVI based batch correction process (`r supFig_cap(name = 'scPOP_performance', display = 'cite')`).
+Outputs from this pipeline include the batch corrected lower dimensional space, the UMAP coordinates, cell cluster assignments, and the cell type predictions.
+The outputs from this pipeline are used in the "SnakeSCEIAD" which performs differential gene testing across multiple conditions and wraps the various data types into the scEiaD sqlite database for use in the plae app (`r supFig_cap(name = 'workflow_network', display = 'cite')`).
+
+## Quality Control
+
+To decide whether a droplet is considered a viable cell, we use the "barcodeRanks" algorithm from DropletUtils to automatically identify empty droplets.
+We then kept cells with the following criteria: less than 20% of the counts were in mitochondrial genes, a cell had more than 600 total counts, at least 300 unique genes quantified, and fewer than 100,000 total counts.
+To identify likely doublets we used Scrublet [@wolockScrubletComputationalIdentification2019] and doubletCells (from scran) [@lunStepbystepWorkflowLowlevel2016] to generate per cell doublet probability scores.
+To identify cells for removal we used the scores in two ways.
+First, if an individual cell had a Scrublet score greater than 0.8 (where 1.0 is 100% probability of doublet) and a doubletCell score greater than 1,000,000 (unitless measure where higher value is more likely to be doublet) the cell is labelled as a doublet.
+Second, if a cluster has an average Scrublet score greater than 0.2 or an average doubletCell score greater than 12 the entire cluster was marked as doublets.
+These cutoffs were selected by hand inspecting density distributions of the Scrublet and doubletCell scores.
+
+## Alterations from Swamy et al. to increase stringency of the dataset
+
+The following alterations were made to Swamy et al.: First, the cutoff to retain a cell was raised from 200 quantified unique genes to 300.
+Second, we use the DecontX algorithm[@yangDecontaminationAmbientRNA2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination on a per-study basis to produce a updated count matrix.
 Third, we altered the procedure for aligning gene names across species (see below for details).
-Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process [@xuProbabilisticHarmonizationAnnotation2021].
+Fourth, we use the scANVI (scvitools based) batch correction method, which leverages information from the known cell type labels in the batch correction process [@xuProbabilisticHarmonizationAnnotation2021].
 
@@
 
-We found that while nearly all datasets we curated could be integrated as a whole, a small number had strange behavior in the two dimensional UMAP space and/or nearly all cells from a study were machine labelled as a single cell type.
+## Optimal parameter choices for scANVI batch correction
+
+We found in Swamy et al.
+that scVI had the best batch integration performance relative to other tools like Seurat's CCA, fastMNN, etc.
+For this study we chose to use a variant of scVI, scANVI, We also found that adjusting the number of latent dimensions, highly variable genes, and epochs had substantial effects on the integration performance.
+For this update of scEiaD we ran integration with 3000, 4000, 5000, and 6000 highly variable genes, 5 and 15 epochs, and 6, 8, 10, 15, and 20 scANVI outputted latent dimensions.
+After evaluating the performance of the new scEiaD dataset with our scPOP package we selected the following parameters for the integration with scANVI: 15 latent dimensions, 4000 highly variable genes, and 5 epochs [@swamyBuildingMegaSinglecell2021].
+The UMAP was built with a min_dist of 0.1 and the clustering used 50 nearest neighbors.
+
+While nearly all datasets we curated could be integrated as a whole, a small number either did not integrate with other data (clusters were formed which were entirely composed of one study) and/or nearly all cells from a study were machine labelled as a single cell type.
 These studies were hand-removed from the resource.
-
-After evaluating the performance of the new dataset with our scPOP package we selected the following parameters for the integration with scANVI: 15 latent dimensions, 4000 highly variable genes, and 5 epochs.
-The UMAP was built with a min_dist of 0.1 and the clustering used 50 nearest neighbors.
+The metadata for these samples is made available as supplementary file "samples_hand_removed.csv."
 
@@
 As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify, we must identify shared genes.
-We found homologs/orthologs by pulling from the BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference.
+We found homologs and orthologs by downloading from the BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference.
 We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus musculus*.
@@
 Rod, Muller Glia), we wrote custom code for each study that made this table available to link their cell type assignments back to the matching barcoded cell.
-We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels.
+We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels and the later are the original cell type labels.
 We normalized names for CellType, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).
-We did retain the original published (except for the broader name normalization) labels under SubCellType.
 
@@
 This allowed the use of the more mature bulk RNA-seq tooling.
-This approach has also been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments[@squairConfrontingFalseDiscoveries2021].
+This approach has also been found to be more specific in identifying differentially expressed genes in single cell RNA-seq experiments [@squairConfrontingFalseDiscoveries2021].
 If there were fewer than 50 cells in organism - study accession - category combination, they were discarded for the differential test.
@@
 
-## New studies and improvements to the scEiaD database
+## New studies and processing improvements enhance the scEiaD database
 
 Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021].
-We updated the scEiaD database for the PLAE v0.91 web app in six important ways.
+We updated the scEiaD database for the PLAE v0.93 web app in six important ways.
 First, we added a new species, chicken (*Gallus gallus*) [@yamagataCellAtlasChick2021].
@@
 Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons [@SinglecellTranscriptomics20].
-As before, the pipeline's scVI's integration parameters were chosen rigorously by testing inegration performance across a wide range of latent dimensions and number of highly variable genes in our scPOP framework [@swamyBuildingMegaSinglecell2021].
+As before, the pipeline's scVI integration parameters were chosen rigorously by testing integration performance across a wide range of latent dimensions and number of highly variable genes in our scPOP framework [@swamyBuildingMegaSinglecell2021] (`r supFig_cap(name = 'scPOP_performance', display = 'cite')`).
 Each study in the resource was hand assessed to determine whether they had any unusual properties in regards to machine learnt cell type proportions or position in the 2D UMAP space and a small number of outlier studies were removed from the resource.
+
+Of the `r meta_filter %>% filter(Organ == 'Eye') %>% nrow()` ocular cells, only `r meta_filter %>% filter(Organ == 'Eye', !is.na(CellType)) %>% nrow()` had cell type labels we could recover from the publications or data deposits.
+To label most of the remaining cells we used our xgboost approach from Swamy et al.
+which we previousy found performed well when compared to other cell type transfer approaches.
+The overall accuracy is 96% and the confusion matrix shows the predicted cell type to published cell type recall is very high for most cell types (`r supFig_cap(name = 'sup_CT_confusion', display = 'cite')`) and that cell type labelling works well across the different species  (`r supTab_cap(name = 'sup_CT_table', display = 'cite')`).
+The cell types with lower labelled performance tended to be ciliary, limbal, and corneal cells (which were often labelled as fibroblasts).
+There are fewer unique datasets for these cell types and we expect that future runs with more data will improve this performance
 
@@
 Several single cell RNA-seq web-based resources contain ocular data (`r tab_cap(name = 'web_resources', display = 'cite')`).
-The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are not ocular focused, but do contain some ocular data.
-The PanglaoDB and UCSC are resources which have ingressed a large number of published single cell datasets The UCSC Cell Browser has around sixty thousand ocular single cell transcriptomes while the PangloaDB has about forty-seven thousand.
-However, these resources provide no curation of the metadata and process the data for each study individually.
-Thus each study is effectively silo-ed off from the rest and cross comparisons are impractical.
+The UCSC Cell Browser, Tabula Sapiens portal, PanglaoDB, and Protein Atlas are general purpose single cell repositories that contain some ocular data.
+The PanglaoDB and UCSC resources have ingressed a large number of published single cell datasets.
+Af of late 2022, the UCSC Cell Browser has around sixty thousand ocular single cell transcriptomes while the PangloaDB has about forty-seven thousand.
+However, these resources  do not curate and normalize metadata terms (e.g.
+ensuring the same naming system for cell types).
+Furthermore, each study is computationally processed separately.
+In practice each study is silo-ed off from the rest, making comparisons across studies impractical.
 The Tabula Sapiens project is an effort to sequence all cell type transcriptomes in humans; while it only contains around around ten thousand ocular cells it can be more readily cross compared to non-ocular tissues / cells.
-A more substantial downside is that the cell type labels can be insufficient - for example rods and cones are not independently identified but rather labelled as photoreceptors.
-The Protein Atlas resource has recently used only one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g.
+A downside is that the ocular cell type labels can be insufficient - for example rods and cones are together labelled as photoreceptors.
+The Protein Atlas resource has used only one retina single cell RNA-seq paper to provide cell type label focused information for gene queries (e.g.
 one can see whether a gene of interest is enriched in Cones).
-This is problematic as it is unclear whether signal found is specific to the one dataset they use or is common across independent species and datasets.
+Protein Atlas's use of only one study is problematic as it is unclear whether signal found is specific to the one dataset they use or is common across independent species and datasets.
 
@@
 The Protein Atlas is asterisked as while it contains independent datasets, it only has one ocular study.
-')
+Export-able means whether a resource allows bulk downloads of the count matrices and metadata.
+Organisms refers to the number of ocular species used.')
 
@@
 feature_table <- rbind(
-  c('PLAE','Yes','Yes', 'Yes', '1136 ', 'https://plae.nei.nih.gov'),
-  c('Spectacle', 'Yes' ,'No', 'No', '1500', 'https://singlecell-eye.org'),
-  c('UCSC Cell Browser', 'Yes', 'No', 'Yes', '62', 'https://cells.ucsc.edu' ),
-  c('Tabula Sapiens', 'No', 'N/A', 'Yes', '10', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
-  c('PanglaoDB', 'Yes', 'No', 'Yes', '47', 'https://panglaodb.se'),
-  c('Protein Atlas', 'Yes*','No', 'Yes', '5',  'https://www.proteinatlas.org/humanproteome/celltype')) %>% 
+  c('PLAE','Yes','Yes', 'Yes', '4','Yes',  '1136 ', 'https://plae.nei.nih.gov'),
+  c('Spectacle', 'Yes' ,'No', 'No', '5','No',  '1500', 'https://singlecell-eye.org'),
+  c('UCSC Cell Browser', 'Yes', 'No', 'Yes', '2','Yes',  '62', 'https://cells.ucsc.edu' ),
+  c('Tabula Sapiens', 'No', 'N/A', 'Yes', '1','Yes',  '10', 'https://tabula-sapiens-portal.ds.czbiohub.org'),
+  c('PanglaoDB', 'Yes', 'No', 'Yes',  '2','Yes', '47',  'https://panglaodb.se'),
+  c('Protein Atlas', 'Yes*','No', 'No',  '1', 'Yes', '5',  'https://www.proteinatlas.org/humanproteome/celltype')) %>% 
   as_tibble()
-colnames(feature_table) <- c('Resource Name', 'Independent\nDatasets', 'Harmonization', 'Ocular and\nnon-Ocular\n Data', 'Ocular Cell\nCount\n(thousands)', 'URL')
+colnames(feature_table) <- c('Resource Name', 'Independent\nDatasets', 'Harmonization', 'Export-able', 'Organisms', 'Ocular and\nnon-Ocular\n Data', 'Ocular Cell\nCount\n(thousands)', 'URL')
 feature_table %>% flextable() %>%  fontsize(size = 10, part = 'all') %>% autofit() %>% fit_to_width(7)#width(., width = c(1,0.75,0.75,0.75,0.75,3))
@@
 Spectacle does not contain any non-ocular datasets and the cell type labels have not been curated or harmonized between studies (for example different studies can use MG, Mueller, Müller Glia, or other variations to refer to the same cell type).
+Spectacle does contain more cells than PLAE because it also contains organoid and perturbed studies.
 Finally, in contrast to PLAE, Spectacle does not make their datasets available for download in any form.
@@
 In contrast, starting from the loading page of Spectacle, it takes approximately eleven seconds to plot the cell type metadata for a several thousand cells across across a single study (tested on 2022-08-26).
-Within the same time, PLAE can plot the cell type labels for over a million cells across `r meta_filter_v0$study_accession %>% unique() %>% length()` studies (`r supFig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`d).
+Within the same time, PLAE can plot the cell type labels for over a million cells across `r meta_filter$Citation %>% unique() %>% length()` publications (`r supFig_cap(name = 'sup_scEiaD_data_overview', display = 'cite')`d).
 The speed of the PLAE app is also comparable to the CELLxGENE browser serving the Tabula Sapiens resource while providing additional functionality.
@@
 We hand selected *C1ql2* and *Cartpt* (amacrine), *Frzb* and *Slitrk2* (Muller glia), and *Onecut3* (horizontal) and plotted these five genes in the boxplot view to confirm specificity across the full scEiaD database (`r supFig_cap(name = 'lightlyCited_candidates', display = 'cite')`).
-All five look to be specific to a cell type, except for *Frzb* which also is expressed in the RPE.
+All five look to be specific to a cell type, except for *FRZB* which also is expressed in the RPE.
 
@@
 We filter to genes with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two or less than negative one (`r fig_cap(name = 'neuro', display = 'cite')`B and C).
-We find one gene, *Zbtb38*, that meets this criteria and drops in expression when comparing RPC to neurogenic.
+We find one gene, *ZBTB38*, that meets this criteria and drops in expression when comparing RPC to neurogenic.
 We find `r up %>% length()` genes that increase in expression from RPC to neurogenic, `r up[!up %in% g_lu] %>% length()` of which are not previously identified by Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`C).
-We replicate *`r up[up %in% g_lu] %>% str_extract(., '\\w+') %>% paste(., collapse = ', ') %>% str_to_title()`* from Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`B) as differentially expressed between neurogenic and RPC.
+We replicate *`r up[up %in% g_lu] %>% str_extract(., '\\w+') %>% paste(., collapse = ', ') %>% toupper()`* from Lu et al (`r fig_cap(name = 'neuro', display = 'cite')`B) as differentially expressed between neurogenic and RPC.
 Futhermore we show in `r supFig_cap(name = 'PLAE_screenshots_supplemental', display = 'cite')`a how this figure can be replicated directly in our web app.
@@
 While we have curated a large number of cell types across the eye, many of these cell types have "sub" cell types.
-For examplethe human has three types of cones that are optimized for short, medium, and long wavelengths.
+For example the human retina has three types of cones that are optimized for short, medium, and long wavelengths.
 More dramatically, Yan et al.
@@
 A great many groups have proposed cell atlases.
-These range from efforts focusing on a subset of cells from a single layer of a tissue (for example [@tranSingleCellProfilesRetinal2019a] ) to a huge consortium encompassing multiple organs [@consortiumTabulaSapiensSingle2021].
+These range from efforts focusing on a subset of cells from a single layer of a tissue [@tranSingleCellProfilesRetinal2019a] to a huge consortium encompassing multiple organ [@consortiumTabulaSapiensSingle2021].
 We propose a complementary approach wherein we use the research data from the community at large to create a meta-atlas.
-This approach enables a small group to produce an atlas which rivals the size from massive multi-group consortiums.
-Furthermore the inherent variability between the data from independent groups can be used as a feature to get insight into whether a gene-cell type signal is stable across an entire community's datasets.
-We briefly demonstrate how the high study diversity in scEiaD can be used to validate proposed RPC to neurogenic gene expression changes.
+This strategy enables a small group to produce an atlas which rivals the size from massive multi-group consortiums.
+Furthermore, the inherent variability between the data from independent groups can be used as a feature to get insight into whether a gene-cell type signal is stable across an entire community’s datasets.
+We use this feature to demonstrate how the high study diversity in scEiaD can be used to validate proposed RPC to neurogenic gene expression changes in a R analysis document.
+Our community powered single cell ocular meta-atlas enables the entire ocular community to use the wealth of research it has generated.
+
+# Acknowledgments
+
+This research was funded by the National Institutes of Health (National Eye Institute) intramural program.
+This work utilized the computational resources of the NIH HPC Biowulf cluster (http://hpc.nih.gov).
 
 # Supplementary Figures and Tables
+```{r scPOP_performance, fig.height = 8, fig.width = 12, echo = F, message = F, fig.cap=scPOP_performance_cap}
+
+scPOP_performance_cap <- supFig_cap(name='scPOP_performance', caption = 'scPOP based performance scoring (higher is better) across different numbers of highly variable genes, epochs of scANVI used, and number of outputted latent dimensions.')
+
+source('src/assess_integration.R')
+pop_perf
+```
+
+```{r workflow_network, fig.height = 6, fig.width = 8, echo = F, message = F, fig.cap=workflow_network_cap}
+
+workflow_network_cap <- supFig_cap(name='workflow_network', caption = 'Data flow overview from published cell types, sample information, raw fastq counts to produce the sqlite database used in plae.nei.nih.gov')
+
+source('src/workflow_fig.R')
+workflow_plot  + theme(plot.margin = margin(0,0,0,0 , "cm"))
+```
 
@@
 plot_grid(v0_rho, v1_rho, ncol = 1, labels = c('a','b'), rel_heights = c(1,1.5))
+```
+
+```{r sup_CT_confusion, fig.height = 12, fig.width = 14, echo = F, message = F, fig.cap=sup_CT_confusion_cap}
+sup_CT_confusion_cap <- supFig_cap(name = 'sup_CT_confusion', caption = "Confusion matrix of cell type label consistency between predicted cell types (columns) and published cell types (rows).")
+
+source('src/ct_confusion.R')
+ct_confusion
 ```
@@
 ```{r lightlyCited_candidates, echo = F, message = F, fig.cap = lightlyCited_candidates_cap}
-lightlyCited_candidates_cap <- supFig_cap(name='lightlyCited_candidates', caption = 'Five hand-selected genes selected from Supplementary Figure X as having fairly unique expression to a celltype while having few pubmed citations.
+lightlyCited_candidates_cap <- supFig_cap(name='lightlyCited_candidates', caption = 'Five hand-selected genes selected from Supplementary Figure 3 as having fairly unique expression to a celltype while having few pubmed citations.
 Plotted as expression/boxplots to confirm specificity at the study level.')
@@
 
+```{r sup_CT_table,fig.height = 15, fig.width = 20, echo = F, message = F, results = 'asis',fig.cap=sup_CT_table_cap}
+# 0
+sup_CT_table_cap <- supTab_cap(name = 'sup_CT_table', caption = 'Count and ratio of cell type and predicted cell type labelling assignments split by organisms shows overall strong labelling performance.
+Cell type to predicted cell type comparisons with a ratio less than 0.01 were filtered out.')
+cat("<div id=\"tbl:sup_CT_table\">")
+ct_table %>% filter(Ratio > 0.01) %>% flextable() %>% autofit() %>% fit_to_width(7)
+
+#Output Caption (* * for italics)
+cat(paste0("\n*",sup_CT_table_cap,"*\n"))
+# Output </div>
+cat("</div>")
+```
+
 ```{r studies_table, fig.height = 15, fig.width = 20, echo = F, message = F, results = 'asis',fig.cap=studies_table_cap}

### 2023-02-16 src/assess_integration.R

@@ -24,7 +24,7 @@ accuracy %>%
   summarise(bad_score_sum = sum(score < 0.8), xgboost_score = mean(score)) %>% 
   arrange(-xgboost_score)
 
-perf %>% 
+pop_perf <- perf %>% 
   filter(partition %in% c('universe')) %>% 
   select(-clusterN, -clusterMedian, -subset) %>% 
   pivot_wider(names_from = c('Score','Group'), values_from = Value) %>% 

### 2023-02-16 src/ct_confusion.R

@@ -20,7 +20,7 @@ ct_mat <- ct_mat %>% as.matrix()
 row.names(ct_mat) <- rowN
 ct_mat <- ct_mat[row.names(ct_mat) %>% rev(), ]
 
-ct_confusion <- ComplexHeatmap::Heatmap(ct_mat, cluster_rows = FALSE, cluster_columns = FALSE, col=viridis::viridis(20), name =  'Recall')
+ct_confusion <- ComplexHeatmap::Heatmap(ct_mat, cluster_rows = FALSE, cluster_columns = FALSE, col=viridis::viridis(20), name =  'Recall', row_title = 'Published Cell Type', column_title = 'Predicted Cell Type')
 
 
 ct_table <- meta_filter %>% 

### 2023-02-17 src/workflow_fig.R
workflow fig code

@@ -1,0 +1,55 @@
+library(tidyverse)
+library(tidygraph)
+library(ggdag)
+library(ggraph)
+library(cowplot)
+library(patchwork)
+
+dagified <- dagify('SnakeQUANT' ~ 'fastq',
+                   'counts' ~ 'SnakeQUANT',
+                   'SnakePOP' ~ 'CellType',
+                   'SnakePOP' ~ 'counts',
+                   'LatentDims' ~ 'SnakePOP',
+                   'cluster' ~ 'LatentDims',
+                   'umap' ~ 'LatentDims' ,
+                   'ML' ~ 'CellType',
+                   'projectedCellType' ~ 'ML',
+                   'ML' ~ 'LatentDims',
+                   'SnakeSCEIAD' ~ 'counts',
+                   'SnakeSCEIAD' ~ 'projectedCellType',
+                   'SnakeSCEIAD' ~  'CellType',
+                   'SnakeSCEIAD' ~   'cluster',
+                   'SnakeSCEIAD' ~   'umap',
+                   'SnakeSCEIAD' ~ 'SampleInfo',
+                   'SnakeQUANT' ~ 'SampleInfo',
+                   'SnakePOP' ~ 'SampleInfo',
+                   'DiffTesting' ~ 'SnakeSCEIAD',
+                   'scEiaD' ~ 'SnakeSCEIAD',
+                   'scEiaD' ~ 'DiffTesting',
+                   'plae' ~ 'scEiaD')
+
+tidy_dagitty(dagified)
+set.seed(51345)
+workflow_plot <- tidy_dagitty(dagified) %>% as_tbl_graph() %>% 
+  mutate(type = case_when(name %in% c('fastq', 'CellType', 'SampleInfo') ~ 'Input',
+                          name %in% c('SnakeQUANT', 'counts') ~ 'SQ',
+                          name %in% c('umap', 'projectedCellType','cluster', 'counts', 'LatentDims', 'SnakePOP', 'ML') ~ 'SP',
+                          name %in% c('scEiaD','DiffTesting', 'SnakeSCEIAD') ~ 'SS'),
+         name = case_when(name == 'counts' ~ 'ambient cleaned counts',
+                          name == 'CellType' ~ 'Published\nCell Types',
+                          name == 'projectedCellType' ~ 'Learned\nCell Types',
+                          name == 'DiffTesting' ~ 'Diff\nTesting',
[…]

### 2023-02-17 manuscript_00.Rmd
plotting and text updates

@@
 
-The retina is composed of six major cell populations: the photoreceptors, horizontal, bipolar, amacrine, ganglion, and non-neuronal cells [@maslandCellPopulationsRetina2011].
+The outer layer of the eye consists of the cornea and sclera.
+Beneath the sclera is the uvea which is made of the iris, ciliary body, and the choroid.
+The inner most layer of the eye is the retina, which is backed by the retinal pigmented epithelium (RPE).
+The retina alone is composed of six major cell populations: the photoreceptors, horizontal, bipolar, amacrine, ganglion, and non-neuronal cells [@maslandCellPopulationsRetina2011].
 These six populations can be further divided into dozens of cell types [@yanCellAtlasHuman2020a].
@@
 
-The modern era of single cell transcriptomics has been marked by the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment.
+The modern era of single cell transcriptomics starts with the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment [@kleinDropletBarcodingSingleCell2015].
 This approach was pioneered by Macoksko et al.
-and enabled the profiling of gene expression in 44,808 single cells in the adult mouse retina [@kleinDropletBarcodingSingleCell2015; @macoskoHighlyParallelGenomewide2015a].
-The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since Macosko publication.
-
-These publications's foci includes, but are not limited to, studying developmental transcriptomic dynamics, identifying genes to distinguish core cell types, proposing finer gradations of cell types within existing categories, and comparing ocular transcriptomes between organisms.
-Independent querying of these datasets requires a laborious series of steps.
+and enabled the profiling of gene expression in 44,808 single cells in the adult mouse retina [@macoskoHighlyParallelGenomewide2015a].
+The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since the Macosko publication.
+An incompmlete list of what the ocular community has used single cell technology across dozens of studies would include studing developmental transcriptomic dynamics, identifying genes to distinguish cell types, proposing finer gradations of cell types within existing categories, comparing and contrasting ocular transcriptomes between organisms, and identify differences in spatially different cells.
 
@@
 Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging.
-In contrast to these, the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020].
+In contrast,the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020].
 However each data deposit is independently processed, which makes cross study queries impossible.
@@
 We then kept cells with the following criteria: less than 20% of the counts were in mitochondrial genes, a cell had more than 600 total counts, at least 300 unique genes quantified, and fewer than 100,000 total counts.
-To identify likely doublets we used Scrublet [@wolockScrubletComputationalIdentification2019] and doubletCells (from scran) [@lunStepbystepWorkflowLowlevel2016] to generate per cell doublet probability scores.
+To identify likely doublets we used Scrublet [@wolockScrubletComputationalIdentification2019a] and doubletCells (from scran) [@lunStepbystepWorkflowLowlevel2016a] to generate per cell doublet probability scores.
 To identify cells for removal we used the scores in two ways.
@@
 
-The code base for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is #a99dced There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app.
+The code base for scEiaD is available at github.com/davemcg/scEiaD; the commit corresponding to this manuscript is #a99dced.
+There are three relevant Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which uses the optimal parameters identified in SnakePOP to run the cell type machine learning and differential gene testing that is incorporated into the scEiaD sqlite database for the app.
 The code base for the web app (along with local installation instructions) is available at github.com/davemcg/scEiaD/plaeApp.
@@
 
-```{r, fig.height=14, fig.width=12, fig.cap = scEiaD_data_overview_cap, echo = FALSE}
+```{r, fig.height=13, fig.width=11, fig.cap = scEiaD_data_overview_cap, echo = FALSE}
 
@@
   geom_bar(stat='identity', position='stack') +
-  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5)) +
+  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4) +
   cowplot::theme_cowplot() +
   coord_flip() +
-  theme(legend.position = "none") +
   scale_y_continuous(expand = c(0, 0))  + xlab('') +
@@
   geom_bar(stat='identity', position='stack') +
-  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5)) + 
+  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4) + 
   cowplot::theme_cowplot() +
@@
 
-colnames(timings) <- c('PLAE',"Spectacle",'Tabula Sapiens\n(CELLxGENE)')
+colnames(timings) <- c('PLAE',"Spectacle",'Tabula\nSapiens\n(CELLxGENE)')
 
@@
   mutate(Test = factor(Test, levels = row.names(timings))) %>% 
-  mutate(Text = case_when(is.na(value) ~ 'Not\npossible',
+  mutate(Text = case_when(is.na(value) ~ 'Cannot',
                           TRUE ~ '')) %>% 
@@
 ab <- cowplot::plot_grid(cell_counts + theme(legend.position = "none", axis.ticks.y = element_blank()), 
-                         study_counts_by_cell + theme(axis.ticks.y = element_blank()), 
+                         study_counts_by_cell + theme(legend.position = "none", axis.ticks.y = element_blank()), 
                          nrow = 1, 
@@
 
-cd <- cowplot::plot_grid(study_count_by_tissue + theme(axis.ticks.y = element_blank()), 
+cd <- cowplot::plot_grid(study_count_by_tissue + theme(legend.position = c(0.4, 0.3)) + theme(axis.ticks.y = element_blank()), 
                          timing_plot, 
@@
 
-cowplot::plot_grid(ab, cd, ncol = 1, rel_heights = c(1,0.8))
+cowplot::plot_grid(ab, cd, ncol = 1, rel_heights = c(1,0.85))
 
@@
 The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has collated a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app.
-PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing of the data from the raw sequence in a consistent manner, integrating the independent studies to deliver cross study comparisons, and, crucially, providing full availability of the underlying data for outside usage.
+PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing the faw sequence data in a consistent framework, batch correcting the independent studies, and, crucially, providing full availability of the underlying data for outside usage.
 
@@
 
-```{r PLAE_screenshots, fig.height = 18, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
+```{r PLAE_screenshots, fig.height = 10, fig.width = 10, echo = F, message = F, fig.cap=PLAE_screenshots_cap}
 PLAE_screenshots_cap <- fig_cap(name='PLAE_screenshots', caption = 'Screenshots from PLAE app.
@@
 Dotplot demonstating efficient display of gene expression across many cell types.')
-knitr::include_graphics('fig/fig1.png')
+knitr::include_graphics('fig/figure2 copy.png')
 ```
@@
 
-```{r consist_genes_heatmap, fig.height=22, fig.width=10, echo = F, message = F,  fig.cap=consist_genes_heatmap_cap}
+```{r consist_genes_heatmap, fig.height=23, fig.width=12, echo = F, message = F,  fig.cap=consist_genes_heatmap_cap}
 consist_genes_heatmap_cap <- fig_cap(name='consist_genes_heatmap', caption = 'Rows are genes, columns are cell types and organisms.
@@
 
-```{r neuro, fig.width=30, fig.height=8, echo = F, message = F, fig.cap=neuro_cap}
+```{r neuro, fig.width=16, fig.height=8, echo = F, message = F, fig.cap=neuro_cap}
 neuro_cap <- fig_cap(name='neuro', caption = 'A.
@@
 
-exp_plot_new <- make_exp_plot(input, scEiaD_2020_v01, meta_filter)
+exp_plot_new <- make_exp_plot(input, scEiaD_2020_v01, meta_filter) + theme(legend.box="vertical", legend.justification="center")
 
@@
 a_legend <- get_legend(exp_plot_new)
-neuro_plot <- plot_grid(plot_grid(exp_plot_existing + theme(legend.position="none") + xlab(''),
-                                  exp_plot_new + theme(legend.position="none") + ylab('') + xlab(''), 
+neuro_plot <- plot_grid(plot_grid(exp_plot_existing + theme(legend.position="none") + ylab("CellType (Predict)") ,
+                                  exp_plot_new + theme(legend.position="none", axis.text.y =element_blank(), axis.ticks.y = element_blank()) + ylab('') , 
                                   labels = c('b','c'),
@@
 
-```{r undersupported_genes, fig.height=23, fig.width=8, echo = FALSE, fig.cap = undersupported_genes_cap}
+```{r undersupported_genes, fig.height=23, fig.width=12, echo = FALSE, fig.cap = undersupported_genes_cap}
 undersupported_genes_cap <- supFig_cap(name = 'undersupported_genes', caption = 'Well supported genes with a mean log2FC > 5 and zero pubmed citation hits for [gene] AND Retina search.')

### 2023-02-17 src/make_exp_plot.R
swap axes

@@ -116,7 +116,7 @@ make_exp_plot <- function(input, db, meta_filter){
   } else if (input$figure){
     box_data  %>% 
       mutate(Gene = str_extract(Gene, '^\\w+ ') %>% gsub(" $", '', .)) %>%
-      ggplot(aes(x=!!as.symbol(input$exp_plot_facet), y = !!as.symbol(input$exp_plot_ylab), color = !!as.symbol(grouping_features), group = CellType_predict, shape = organism)) +
+      ggplot(aes(y=!!as.symbol(input$exp_plot_facet), x = !!as.symbol(input$exp_plot_ylab), color = !!as.symbol(grouping_features), group = CellType_predict, shape = organism)) +
       geom_boxplot(color = 'black', outlier.shape = NA) +
       ggbeeswarm::geom_quasirandom(size = 4, groupOnX = TRUE) +
       cowplot::theme_cowplot() +
@@ -124,7 +124,7 @@ make_exp_plot <- function(input, db, meta_filter){
       scale_radius(range=c(2, 6)) +
       scale_colour_manual(values = rep(c(pals::alphabet() %>% unname()), 20)) +
       theme(legend.position="bottom") +
-      facet_wrap(ncol = as.numeric(input$exp_plot_col_num), scales = 'free', 'Gene') 
+      facet_wrap(nrow = 2, scales = 'free_x', 'Gene') 
   }  else if (input$figure_fovea){
     box_data  %>% 
       filter(!is.na(retina_region)) %>% 

### 2023-02-18 manuscript_00.Rmd
text and figs

@@
 METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it.
-RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encompassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 37 publications.
+RESULTS: The new portal provides quick visualization of over a million individual cells from vertebrate eye-and body transcriptomes encompassing four species, 60 cell types, six ocular tissues, and 23 body tissues across 35 publications.
 To demonstrate the value of this unified pan-eye dataset, we replicate known neurogenic and cone macula markers as well as propose six new cone human region markers.
@@
          Citation = case_when(study_accession == 'OGVFB_Hufnagel_iPSC_RPE' ~ 'Swamy VS, Fufa TD, Hufnagel RB, McGaughey DM Building the Mega Single Cell Transcriptome Ocular Meta-Atlas Gigascience 2021 Oct 13;10(10)',
+                              grepl('axon guidance in early-born retinal neurons.', Citation) ~ 'Lo Giudice Q, Leleu M, La Manno G, Fabre PJ.
+Single-cell transcriptional logic of cell-fate specification and axon guidance in early-born retinal neurons.
+Development 2019 Sep 9;146(17)',
+                              Citation == 'Lu Y, Shiau F, Yi W, Lu S et al.
+Single-Cell Analysis of Human Retina Identifies Evolutionarily Conserved and Species-Specific Mechanisms Controlling Development.
+Dev Cell 2020 May 18;53(4):473-491.e9.' ~ 'Lu Y, Shiau F, Yi W, Lu S et al.
+Single-Cell Analysis of Human Retina Identifies Evolutionarily Conserved and Species-Specific Mechanisms Controlling Development.
+Dev Cell 2020 May 18;53(4):473-491.e9',
                               TRUE ~ Citation),
@@
 The inner most layer of the eye is the retina, which is backed by the retinal pigmented epithelium (RPE).
+Distinct cell types are used to compose each of these layers.
 The retina alone is composed of six major cell populations: the photoreceptors, horizontal, bipolar, amacrine, ganglion, and non-neuronal cells [@maslandCellPopulationsRetina2011].
-These six populations can be further divided into dozens of cell types [@yanCellAtlasHuman2020a].
+These six populations can be subdivided into dozens of cell types [@yanCellAtlasHuman2020a].
 Furthermore, human disease of the eye can derive from damage to specific cell types.
-Glaucoma is characterized by damage to the retinal ganglion cells while cone and cone-rod dystrophies are characterized by photoreceptor degeneration.
-Efforts to untangle the transcriptomics of these diverse cell types with single cell transcriptomics date to the early 2000s when researchers [@hagstromConePigmentGene2000b; @jabsEvidenceP2XP2X2000; @liLocalizationDopamineD1receptor2004; @lindqvistSingleCellRTPCR2002a; @reidRetinoschisinPhotoreceptorsecretedProtein2003]  used RT-PCR approaches to amplify select transcripts in small numbers of cells.
+
+Damage to some of these cell types has been implicated in human eye disease.
+For example glaucoma is characterized by damage to the retinal ganglion cells while cone and cone-rod dystrophies are characterized by photoreceptor degeneration.
+Efforts to untangle the gene expression patterns of these diverse cell types with single cell transcriptomics date to the early 2000s when researchers [@hagstromConePigmentGene2000b; @jabsEvidenceP2XP2X2000; @liLocalizationDopamineD1receptor2004; @lindqvistSingleCellRTPCR2002a; @reidRetinoschisinPhotoreceptorsecretedProtein2003] used RT-PCR approaches to amplify select transcripts in small numbers of cells.
 Later, researchers used microarray platforms to analyze a broader array of transcripts[@cherryDevelopmentDiversificationRetinal2009; @gustincichGeneDiscoveryGenetically2004a; @kimIdentificationMolecularMarkers2008a; @roeschTranscriptomeRetinalMuller2008; @trimarchiIndividualRetinalProgenitor2008; @trimarchiMolecularHeterogeneityDeveloping2007b; @wahlinMethodAnalysisGene2004a]
 
-The modern era of single cell transcriptomics starts with the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment [@kleinDropletBarcodingSingleCell2015].
+The modern era of single cell transcriptomics started with the introduction of droplet-based technologies, which enabled quantitation of thousands of cells in a single experiment [@kleinDropletBarcodingSingleCell2015].
 This approach was pioneered by Macoksko et al.
 and enabled the profiling of gene expression in 44,808 single cells in the adult mouse retina [@macoskoHighlyParallelGenomewide2015a].
-The commercialization of the droplet approach has enabled broad access to this technology and the ocular community have responded by publishing several dozen ocular studies using single cell transcriptome technology since the Macosko publication.
-An incompmlete list of what the ocular community has used single cell technology across dozens of studies would include studing developmental transcriptomic dynamics, identifying genes to distinguish cell types, proposing finer gradations of cell types within existing categories, comparing and contrasting ocular transcriptomes between organisms, and identify differences in spatially different cells.
-
-Unfortunately manuscripts can only present a sliver of the underlying data and further exploration is laborious, technically challenging, and computationally expensive.
-Briefly, one would have to identify the relevant publication, find the associated data deposit, download the cell - count matrix, find the cell label table (and optionally the t-SNE or UMAP coordinates), and then load the data into R or python to run the desired queries.
-If one desired to compare across publications, then one would also have to align the cell type labels and should re-create the count quantification with consistent bioinformatic tooling.
+Commercialization of the droplet technology has enabled broad access and the ocular community has responded by using single cell RNA sequencing in several dozen ocular studies.An incomplete list of how the ocular community has used single cell technology would include studying developmental transcriptomic dynamics, identifying genes to distinguish cell types, proposing finer gradations of cell types within existing categories, comparing and contrasting ocular transcriptomes between organisms, and identifying differences in spatially distinct cells.
+
+Unfortunately manuscripts can only present a sliver of the underlying data.
+Further data exploration can be laborious, technically challenging, and computationally expensive.
+Briefly, one would have to identify the relevant publication, find the associated data deposit, download the cell by gene count matrix, find the cell label table (and optionally the t-SNE or UMAP coordinates) with the matching fields to align them, and then load the data into R or python to run the desired queries.
+If one desired to compare across publications, then one would also have to align the cell type label schema and ideally should re-create the count quantification with consistent bioinformatic tooling.
 It is desirable, perhaps even necessary, to use multiple datasets to assess gene expression to confirm whether a gene expression pattern is consistent and reproducible.
@@
 Single cell RNA-seq web resources that contain ocular tissues include the UCSC Cell Browser, Tabula Sapiens, PanglaoDB, and the Protein Atlas [@consortiumTabulaSapiensSingle2021; @franzenPanglaoDBWebServer2019; @karlssonSingleCellType2021; @speirUCSCCellBrowser2021].
-Primary limitations to these resources include incomplete cell type labels and sample metadata that make ocular focused queries challenging.
-In contrast,the Spectacle resource is an ocular specific web app which has curated over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020].
-However each data deposit is independently processed, which makes cross study queries impossible.
-We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single combined space, the Single Cell Eye in a Disk (scEiaD) [@swamyBuildingMegaSinglecell2021].
-
-To broaden the usage of this powerful meta-atlas, we restructured the data into a custom sqlite-based database and built a reactive web application to access it.
+However, ocular focused queries are challenging as these resources have incomplete and inconsistent cell type labels and sample metadata.
+In contrast to there general purpose resources, Spectacle (http://singlecell-eye.org) is an ocular specific web app which contains over a million cells from dozens of independent resources [@voigtSpectacleInteractiveResource2020].
+However each data deposit is independently and inconsistently  processed, which makes cross study queries impossible.
+
+  We recently created a unified ocular data resource which placed dozens of independent single-cell RNA-seq ocular and non-ocular datasets in a single batch corrected space, the Single Cell Eye in a Disk (scEiaD) [@swamyBuildingMegaSinglecell2021].
+To broaden the usage of this powerful meta-atlas dataset, we restructured the data into a custom sqlite-based database and built a reactive web application to access it.
 As many new studies were published since our last data build, we re-ran the scEiaD pipeline in March of 2022.
 In aggregate we have collated `r meta_filter$study_accession %>% unique() %>% length()` datasets across `r meta_filter$PMID %>% unique() %>% length()` publications, four species, and `r meta_filter$Tissue %>% unique() %>% length()` tissues.
-We hand-curated `r meta_filter$CellType_predict %>% unique() %>% length()` cell type labels, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the `r length(meta_filter$CellType_predict[!is.na(meta_filter$CellType_predict)]) - length(meta_filter$CellType[!is.na(meta_filter$CellType)])` more cells.
+We hand-curated `r meta_filter$CellType_predict %>% unique() %>% length()` published cell type labels, totaling `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` cells and used those ground-truth labels as the basis for a machine learning-based algorithm which labels the `r length(meta_filter$CellType_predict[!is.na(meta_filter$CellType_predict)]) - length(meta_filter$CellType[!is.na(meta_filter$CellType)])` previously unannotated cells.
 Quality control metrics were tuned and applied to remove lower quality cells.
@@
 
-Our full scEiaD pipeline was published previously in Swamy et al.
-[@swamyBuildingMegaSinglecell2021].
-It is composed of three Snakefiles which perform three core workflow tasks[@molderSustainableDataAnalysis2021].
-The "SnakeQUANT" pipeline takes the raw fastq input and produces a gene by cell count matrix with kallisto and bustools [@melstedBarcodeUMISet2019; @melstedModularEfficientPreprocessing2019].
+Our full scEiaD pipeline was published previously in Swamy et al [@swamyBuildingMegaSinglecell2021].
+It was composed of two Snakefiles that run the workflows [@molderSustainableDataAnalysis2021].
+The "SnakeQUANT" workflow takes the raw fastq input and produces a gene by cell count matrix with kallisto and bustools [@melstedBarcodeUMISet2019; @melstedModularEfficientPreprocessing2019].
 This matrix is used in the "SnakePOP" pipeline to identify the optimal parameters for the scANVI based batch correction process (`r supFig_cap(name = 'scPOP_performance', display = 'cite')`).
 Outputs from this pipeline include the batch corrected lower dimensional space, the UMAP coordinates, cell cluster assignments, and the cell type predictions.
-The outputs from this pipeline are used in the "SnakeSCEIAD" which performs differential gene testing across multiple conditions and wraps the various data types into the scEiaD sqlite database for use in the plae app (`r supFig_cap(name = 'workflow_network', display = 'cite')`).
+The outputs from this pipeline are used in a our new Snakefile "SnakeSCEIAD" which performs differential gene testing across multiple conditions and wraps the various data types into the scEiaD sqlite database for use in the plae app (`r supFig_cap(name = 'workflow_network', display = 'cite')`).
 
@@
 Second, if a cluster has an average Scrublet score greater than 0.2 or an average doubletCell score greater than 12 the entire cluster was marked as doublets.
-These cutoffs were selected by hand inspecting density distributions of the Scrublet and doubletCell scores.
+All of these cutoff values were selected by hand inspecting density distributions of the Scrublet and doubletCell scores.
 
@@
 
-The following alterations were made to Swamy et al.: First, the cutoff to retain a cell was raised from 200 quantified unique genes to 300.
+The following alterations were made to the Swamy et al.
+pipeline: First, the cutoff to retain a cell was raised from 200 quantified unique genes to 300.
 Second, we use the DecontX algorithm[@yangDecontaminationAmbientRNA2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination on a per-study basis to produce a updated count matrix.
@@
 A custom cutoff for the minimum required number of unique genes identified per cell were used for the SRP362101 cornea dataset as an abnormally high number of cells were returned from our default value of 300.
-We set the cutoff for this studies to be 800 as this value resulted in approximately the same number of cells being returned as the authors reported in their papers.
+We set the cutoff for this study to 800 as this value resulted in approximately the same number of cells being returned as the authors reported in their paper.
 
@@
 We found in Swamy et al.
-that scVI had the best batch integration performance relative to other tools like Seurat's CCA, fastMNN, etc.
-For this study we chose to use a variant of scVI, scANVI, We also found that adjusting the number of latent dimensions, highly variable genes, and epochs had substantial effects on the integration performance.
+that scVI had the best batch integration performance relative to other tools like Seurat's CCA, fastMNN, harmony, ComBat, scanorama, and others.
+We previously found that adjusting the number of latent dimensions, highly variable genes, and epochs had substantial effects on the integration performance.
 For this update of scEiaD we ran integration with 3000, 4000, 5000, and 6000 highly variable genes, 5 and 15 epochs, and 6, 8, 10, 15, and 20 scANVI outputted latent dimensions.
-After evaluating the performance of the new scEiaD dataset with our scPOP package we selected the following parameters for the integration with scANVI: 15 latent dimensions, 4000 highly variable genes, and 5 epochs [@swamyBuildingMegaSinglecell2021].
+After evaluating the performance of the new scEiaD dataset with our scPOP package we selected the following parameters for the integration: 15 latent dimensions, 4000 highly variable genes, and 5 epochs [@swamyBuildingMegaSinglecell2021].
 The UMAP was built with a min_dist of 0.1 and the clustering used 50 nearest neighbors.
@@
 
-Every study brought into scEiaD was hand curated to identify unique biological samples, published paper ID (where available), organ, tissue (e.g.
+Every study brought into scEiaD was hand curated to identify the unique biological samples, published paper ID (where available), organ, tissue (e.g.
 Cornea), source (iPSC based or tissue), single cell platform (e.g.
@@
 
-The R package easyPubMed was used to search for 1000 randomly chosen genes present in scEiaD for both gene name alone and gene name plus "AND Retina" with the get_pubmed_ids function.
+The R package easyPubMed was used to search for 1000 randomly chosen genes present in scEiaD for both gene name alone and gene name plus "AND Retina" with the get_pubmed_ids function .
 We then used the returned Entrez identifier to pull the abstract with the fetch_pubmed_data function so we could extract the PubMed ID (PMID).
@@
 
-Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021].
+Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` datasets, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021].
 We updated the scEiaD database for the PLAE v0.93 web app in six important ways.
@@
 Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (`r supFig_cap(name = 'decontX', display = 'cite')`).
-Fifth, we increased the cell retention cutoff for detected unique transcripts per cell from 200 to 300.
+Fifth, we increased the stringency of the cell retention cutoff for detected unique transcripts per cell from 200 to 300.
 Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris project to more easily provide non eye cell type comparisons [@SinglecellTranscriptomics20].
 As before, the pipeline's scVI integration parameters were chosen rigorously by testing integration performance across a wide range of latent dimensions and number of highly variable genes in our scPOP framework [@swamyBuildingMegaSinglecell2021] (`r supFig_cap(name = 'scPOP_performance', display = 'cite')`).
-Each study in the resource was hand assessed to determine whether they had any unusual properties in regards to machine learnt cell type proportions or position in the 2D UMAP space and a small number of outlier studies were removed from the resource.
 
@@
 which we previousy found performed well when compared to other cell type transfer approaches.
-The overall accuracy is 96% and the confusion matrix shows the predicted cell type to published cell type recall is very high for most cell types (`r supFig_cap(name = 'sup_CT_confusion', display = 'cite')`) and that cell type labelling works well across the different species  (`r supTab_cap(name = 'sup_CT_table', display = 'cite')`).
-The cell types with lower labelled performance tended to be ciliary, limbal, and corneal cells (which were often labelled as fibroblasts).
+The overall accuracy is 96% and the confusion matrix shows the predicted cell type to published cell type recall is very high for most cell types (`r supFig_cap(name = 'sup_CT_confusion', display = 'cite')`).
+We also see that the cell type labelling works well across the different species  (`r supTab_cap(name = 'sup_CT_table', display = 'cite')`).
+The cell types with lower labelling performance tended to be ciliary, limbal, and corneal cells (which were often labelled as fibroblasts).
 There are fewer unique datasets for these cell types and we expect that future runs with more data will improve this performance
@@
 The PanglaoDB and UCSC resources have ingressed a large number of published single cell datasets.
-Af of late 2022, the UCSC Cell Browser has around sixty thousand ocular single cell transcriptomes while the PangloaDB has about forty-seven thousand.
-However, these resources  do not curate and normalize metadata terms (e.g.
-ensuring the same naming system for cell types).
-Furthermore, each study is computationally processed separately.
-In practice each study is silo-ed off from the rest, making comparisons across studies impractical.
+As of late 2022, the UCSC Cell Browser had around sixty thousand ocular single cell transcriptomes while the PangloaDB has about forty-seven thousand.
+However, these resources  do not curate and normalize metadata terms.
+This means that cell types have inconsistent naming which makes cross study comparisons very challenging.
 The Tabula Sapiens project is an effort to sequence all cell type transcriptomes in humans; while it only contains around around ten thousand ocular cells it can be more readily cross compared to non-ocular tissues / cells.
@@
 The Spectacle web resource is the closest comparison to PLAE, as it is a ocular focused resource which has collated a large number of ocular single cell RNA-seq resources (both organoid and tissue based datasets total well over a million cells) and made them available on a web app.
-PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing the faw sequence data in a consistent framework, batch correcting the independent studies, and, crucially, providing full availability of the underlying data for outside usage.
+PLAE is distinguished from these resources by hand curation of the cell and sample level metadata, re-processing the raw sequence data in a consistent framework, batch correcting the independent studies, and, crucially, providing full availability of the underlying data for outside usage.
 
@@
 Export-able means whether a resource allows bulk downloads of the count matrices and metadata.
-Organisms refers to the number of ocular species used.')
+Organisms refers to the number of ocular species with ocular data')
 
@@
 Spectacle does contain more cells than PLAE because it also contains organoid and perturbed studies.
-Finally, in contrast to PLAE, Spectacle does not make their datasets available for download in any form.
+Finally, unlike the PLAE resource, Spectacle does not make their datasets available for download in any form.
 
@@
 PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values.
-So thus, a user can see *Crx* expression across over a million cells and separately see cell type labels for those same million cells.
-The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells.
+So thus, a user can see *Crx* expression across over a million cells and dozens of studies and separately see cell type labels for those same million cells.
+The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells by clicking and dragging to set a zoom in region.
 Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area (`r fig_cap(name = 'PLAE_screenshots', display = 'cite')`a).
@@
 
-While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina.
+While scEiaD does not have spatial scRNA seq datasets as of 2023, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina.
 The structure of the cartoon retina was based on the human retina.
@@
 This plot differs from the Dot Plot view as the expression values are relative instead of absolute.
-That is to say that while the Dot Plot view is proportionate to counts, the Heatmap view is showing the difference in expression between a given cell type (or cluster) and all other cells.
+That is to say that while the Dot Plot view is proportionate to counts, the Heatmap view is showing the difference in expression between a given cell type (or cluster) and all other cells as calculated by DESeq2.
 This view is potentially useful if a user is especially interested in gene expression patterns across a variety of cell types.
+One downside is that because we currently only have one *Gallus gallus* dataset, we cannot use that data in this visualization because you cannot calculate a meaningful differential expression with only one sample.
 
@@
 
-Genes with few citations can be considered as novel markers.
+Genes with relatively few citations can be considered as novel markers.
 We see, for example, *Onecut3* appears to well separate the horizontal cells and has few PubMed citations.
@@
 These cell types are closely related as the neurogenic cells are derived from the multi-potent RPC to later specify neural cell types.
-We use the set of genes identified by Lu et al.
+We first use the set of genes identified by Lu et al.
 [@luSingleCellAnalysisHuman2020b] in their study of human retinal differentiation to determine whether our scEiaD database retains the same trends (`r fig_cap(name = 'neuro', display = 'cite')`A).
@@
 To identify candidate genes that are consistently differentially expressed at this transitional state in mouse and human we apply similar logic that we used in identifying well supported marker genes across the retina cell types.
-We filter to genes with a padj less than 0.05 in both human and mouse, a mean log2 fold change greater than two or less than negative one (`r fig_cap(name = 'neuro', display = 'cite')`B and C).
+We filter to genes with a adjusted p value less than 0.05 in both human and mouse and a mean log2 fold change greater than two or less than negative one (`r fig_cap(name = 'neuro', display = 'cite')`B and C).
 We find one gene, *ZBTB38*, that meets this criteria and drops in expression when comparing RPC to neurogenic.
@@
 and Peng et al.
-published, they failed to replicate in our test with pvalue and padj being above 0.1 for all three.
+published, they failed to replicate in our test with a p value being above 0.1 for all three.
 In contrast *Calb1* did replicate strongly with the sixth lowest padj (0.00721) across our entire test and a log2 fold change of 2.87.
@@
 We have improved on the scEiaD dataset we published in Swamy et al.
-[@swamyBuildingMegaSinglecell2021] by adding another species (chicken), new ocular tissues, a human non-ocular reference set; in total `r meta_filter$Citation %>% unique() %>% length() - meta_filter_v0$Citation %>% unique() %>% length()` new single cell datasets.
+[@swamyBuildingMegaSinglecell2021] by adding another species (*Gallus gallus*), new ocular tissues, and a human non-ocular reference set; in total `r meta_filter$Citation %>% unique() %>% length() - meta_filter_v0$Citation %>% unique() %>% length()` new single cell datasets.
 We re-structured the data into a custom database to enable our custom built high performance web app to quickly access the data.

### 2023-02-21 manuscript_00.Rmd
small text changes

@@
     reference_docx: word-styles-reference-01.docx
+    keep_md: true
     fig_caption: yes
@@
 csl: investigative-ophthalmology-and-visual-science.csl
-abstract: "PURPOSE: To create a high performance reactive web application to query single cell gene expression data across cell type, species, study, and other factors.
+abstract: "PURPOSE: To create a high-performance reactive web application to query single cell gene expression data across cell type, species, study, and other factors.
 METHODS: We updated the content and structure of the underlying data (single cell Eye in a Disk, scEiaD) and wrote the web application PLAE (https://plae.nei.nih.gov) to visualize and explore it.
@@
 ---
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(dev = 'svg') # set output device to svg
+```
 
@@
 Efforts to untangle the gene expression patterns of these diverse cell types with single cell transcriptomics date to the early 2000s when researchers [@hagstromConePigmentGene2000b; @jabsEvidenceP2XP2X2000; @liLocalizationDopamineD1receptor2004; @lindqvistSingleCellRTPCR2002a; @reidRetinoschisinPhotoreceptorsecretedProtein2003] used RT-PCR approaches to amplify select transcripts in small numbers of cells.
-Later, researchers used microarray platforms to analyze a broader array of transcripts[@cherryDevelopmentDiversificationRetinal2009; @gustincichGeneDiscoveryGenetically2004a; @kimIdentificationMolecularMarkers2008a; @roeschTranscriptomeRetinalMuller2008; @trimarchiIndividualRetinalProgenitor2008; @trimarchiMolecularHeterogeneityDeveloping2007b; @wahlinMethodAnalysisGene2004a]
+Later, researchers used microarray platforms to analyze a broader array of transcripts[@cherryDevelopmentDiversificationRetinal2009; @gustincichGeneDiscoveryGenetically2004a; @kimIdentificationMolecularMarkers2008a; @roeschTranscriptomeRetinalMuller2008; @trimarchiIndividualRetinalProgenitor2008; @trimarchiMolecularHeterogeneityDeveloping2007b; @wahlinMethodAnalysisGene2004a].
 
@@
 
-To decide whether a droplet is considered a viable cell, we use the "barcodeRanks" algorithm from DropletUtils to automatically identify empty droplets.
+To decide whether a droplet is considered a viable cell, we use the "barcodeRanks" algorithm from DropletUtils to automatically identify empty droplets [@lunEmptyDropsDistinguishingCells2019a].
 We then kept cells with the following criteria: less than 20% of the counts were in mitochondrial genes, a cell had more than 600 total counts, at least 300 unique genes quantified, and fewer than 100,000 total counts.
@@
 Second, if a cluster has an average Scrublet score greater than 0.2 or an average doubletCell score greater than 12 the entire cluster was marked as doublets.
-All of these cutoff values were selected by hand inspecting density distributions of the Scrublet and doubletCell scores.
+All cutoff values were selected by hand inspecting density distributions of the Scrublet and doubletCell scores.
 
@@
 Every study brought into scEiaD was hand curated to identify the unique biological samples, published paper ID (where available), organ, tissue (e.g.
-Cornea), source (iPSC based or tissue), single cell platform (e.g.
+Cornea), source (tissue), single cell platform (e.g.
 10Xv2, DropSeq, etc.).
@@
 We curated two types of cell type labels: "CellType" and "SubCellType" where the former is our normalized cell types taken from the published labels and the later are the original cell type labels.
-We normalized names for CellType, for example changing "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).
+We normalized names for CellType, for example changing "MG" to "Muller Glia" and dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level).
 
@@
 The cell types are arranged by the columns by how related the expression patterns are.
-We see how the differentiated cell types of the retina are grouped together with the exception of the Müller glia, which are near the RPC from which they are derived.
+We see how the differentiated cell types of the retina are grouped together apart from the Müller glia, which are near the RPC from which they are derived.
 
@@
 To facilitate custom testing with the scEiaD database we provide, as an example, a human-specific macula versus peripheral cone differential testing analysis document as supplementary file "pseudobulk_cone_region.freeze01.html".
-In our analysis document we use the Seurat object containing the full scEiaD dataset we provide at plae.nei.nih.gov to build a custom pseudobulk counts matrix containing only cones from human studies that were taken in a region specific manner.
+In our analysis document we use the Seurat object containing the full scEiaD dataset we provide at plae.nei.nih.gov to build a custom pseudobulk counts matrix containing only cones from human studies that were taken in a region-specific manner.
 This left us with 26678 cells across six studies.
 Yan et al.
-previously proposed *Prph2* and *Rs1* as markers to distinguish macula and peripheral located human cones[@yanCellAtlasHuman2020a] and Peng et al.
+previously proposed *Prph2* and *Rs1* as markers to distinguish macula and peripheral located human cones [@yanCellAtlasHuman2020a] and Peng et al.
 proposed *Calb1* and *Gngt1* as cone region markers in macaque [@pengMolecularClassificationComparative2019a].

### 2023-02-21 src/workflow_fig.R
url change

@@ -42,6 +42,7 @@ workflow_plot <- tidy_dagitty(dagified) %>% as_tbl_graph() %>%
                           name == 'scVIrefquery' ~ 'Optimal\nscVI Model',
                           name == 'LatentDims' ~ 'scVI Batch Corrected\nLatent Dims',
                           name == 'SampleInfo' ~ 'Sample\nInfo',
+                          name == 'plae' ~ 'plae.nei.nih.gov',
                           TRUE ~ name)) %>% 
   ggraph(layout = 'auto') + 
   geom_edge_bend2(arrow = arrow(length = unit(3, 'mm')), 

### 2023-08-15 manuscript_00.Rmd
resubmission changes

@@
     fig_caption: yes
-    pandoc_args:
-      - '--lua-filter=scholarly-metadata.lua'
-      - '--lua-filter=author-info-blocks.lua'
+
 bibliography: zotero_plaeApp.bib
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(dev = 'svg') # set output device to svg
+knitr::opts_chunk$set(dev = 'pdf', dpi = 1200) # set output device to pdf
 ```
@@
 To decide whether a droplet is considered a viable cell, we use the "barcodeRanks" algorithm from DropletUtils to automatically identify empty droplets [@lunEmptyDropsDistinguishingCells2019a].
-We then kept cells with the following criteria: less than 20% of the counts were in mitochondrial genes, a cell had more than 600 total counts, at least 300 unique genes quantified, and fewer than 100,000 total counts.
+We then kept cells with the following criteria: less than 10% of the counts were in mitochondrial genes (20% for 10x v3 as this platform has overall higher mitochondrial counts), a cell had more than 600 total counts, at least 300 unique genes quantified, and fewer than 100,000 total counts.
 To identify likely doublets we used Scrublet [@wolockScrubletComputationalIdentification2019a] and doubletCells (from scran) [@lunStepbystepWorkflowLowlevel2016a] to generate per cell doublet probability scores.
@@
 All cutoff values were selected by hand inspecting density distributions of the Scrublet and doubletCell scores.
+Study level distributions of percent mitochondrial genes, total gene counts, and unique genes identified are shown in `r supFig_cap(name = 'qc_metrics', display = 'cite')`.
 
@@
 Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` datasets, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types [@swamyBuildingMegaSinglecell2021].
-We updated the scEiaD database for the PLAE v0.93 web app in six important ways.
+We updated the scEiaD database for the PLAE v0.94 web app in six important ways.
 First, we added a new species, chicken (*Gallus gallus*) [@yamagataCellAtlasChick2021].
@@
 
-```{r, fig.height=13, fig.width=11, fig.cap = scEiaD_data_overview_cap, echo = FALSE}
+```{r scEiaD_data_overview, fig.height=13, fig.width=11, fig.cap = scEiaD_data_overview_cap, echo = FALSE}
 
@@
   geom_bar(stat='identity', position='stack') +
-  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4) +
+  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4, bg.color = 'white', color = 'black') +
   cowplot::theme_cowplot() +
@@
   geom_bar(stat='identity', position='stack') +
-  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4) + 
+  shadowtext::geom_shadowtext(position = position_stack(vjust = 0.5), size = 4, bg.color = 'white', color = 'black') + 
   cowplot::theme_cowplot() +
@@
 UMAP view showing *Crx* expression with cell type labels, B.
-Information table showing expression of Crx across celltypes and organism.
-C.
-Box plot of selected genes in a user selected cell types across different studies.
+In silico in-situ of Cltb, C.
+Dotplot demonstating efficient display of gene expression across many cell types.
 D.
-In silico in-situ of Rho.
-E.
-Dotplot demonstating efficient display of gene expression across many cell types.')
-knitr::include_graphics('fig/figure2 copy.png')
+Box plot of eight genes, with each overlaid point representing an individual study.')
+knitr::include_graphics('fig/fig2v02.png')
 ```
@@
 
-```{r echo = F, message = F, }
+```{r sup_tab_consist_diff, echo = F, message = F }
 
@@
                                            show_row_dend = FALSE,
+                                           raster_quality = 10,
                                            column_title_rot = 90,
@@
                           show_heatmap_legend = FALSE,
+                          raster_quality = 10,
                           column_labels = "Pubmed Gene Citations",
@@
                           col = citation_col,
+                          raster_quality = 10,
                           row_labels = consist_diff_genes,
@@
 workflow_plot  + theme(plot.margin = margin(0,0,0,0 , "cm"))
+```
+
+```{r quality_control,  fig.height = 10, fig.width = 14, echo = F, message = F, fig.cap=qc_metrics_cap}
+
+qc_metrics_cap <- supFig_cap(name='qc_metrics', caption = 'Study level box plot showing unique genes identified, total gene counts, and percent mitochondrial reads.')
+
+sample_info <- data.table::fread('~/git/scEiaD/data/sample_run_layout_organism_tech_biosample_organ_2022_03_04.tsv')
+a <- meta_filter  %>% 
+  left_join(sample_info %>% select(sample_accession, Platform) %>% unique()) %>% 
+  select(Platform, study_accession, nFeature_RNA) %>% 
+  ggplot(aes(y=study_accession, x = nFeature_RNA)) + 
+  geom_boxplot() + 
+  ggforce::facet_col(~Platform, scales = 'free', space = 'free') +
+  cowplot::theme_cowplot()
+
+b <- meta_filter  %>% 
+  left_join(sample_info %>% select(sample_accession, Platform) %>% unique()) %>% 
+  select(Platform, study_accession, nCount_RNA) %>% 
+  ggplot(aes(y=study_accession, x = nCount_RNA)) + 
+  geom_boxplot() + 
+  ggforce::facet_col(~Platform, scales = 'free', space = 'free') +
+  ylab('') +
+  cowplot::theme_cowplot() +
+  theme(axis.text.y=element_blank())
+
+c <- meta_filter  %>% 
+  left_join(sample_info %>% select(sample_accession, Platform) %>% unique()) %>% 
+  select(Platform, study_accession, percent_mt) %>% 
+  ggplot(aes(y=study_accession, x = percent_mt)) + 
+  geom_boxplot() + 
+  ggforce::facet_col(~Platform, scales = 'free', space = 'free') +
+  ylab('') +
+  cowplot::theme_cowplot() +
+  theme(axis.text.y=element_blank())
+
+plot_grid(a,b,c, ncol = 3)
 ```
@@
 cat("<div id=\"tbl:sup_CT_table\">")
-ct_table %>% filter(Ratio > 0.01) %>% flextable() %>% autofit() %>% fit_to_width(7)
+ct_table %>% filter(Ratio > 0.01) %>% flextable() %>% autofit() %>% fit_to_width(12)
 
